### 2021-07-18 manuscript/manuscript.Rmd
added manuscript

---
title: Dynamic shifts in predator diel activity due to habitat type and fear
author:
  - name: Matthew W. Rees
    email: matt.wayne.rees@gmail.com
    affiliation: UOM
    footnote: 1
  - name: Jack H. Pascoe
    email: 
    affiliation: CEC
  - name: Mark Le Pla
    email: 
    affiliation: CEC
  - name: Emma K. Birnbaum
    email: 
    affiliation: CEC
    footnote: 
  - name: Alan Robley
    email: 
    affiliation: ARI
  - name: Nick Golding
    email: 
    affiliation: CU,TKI
  - name: Brendan A. Wintle
    email: 
    affiliation: UOM
  - name: Bronwyn A. Hradsky
    email: 
    affiliation: UOM
    footnote: 
address:
  - code: UOM
    address: Quantitative & Applied Ecology Group, School of Ecosystem and Forest Science, The University of Melbourne, Parkville, VIC, Australia
  - code: CEC 
    address: Conservation Ecology Centre, Otway Lighthouse Rd, Cape Otway, VIC, Australia
  - code: ARI 
    address: Department of Environment, Land, Water and Planning, Arthur Rylah Institute for Environmental Research, Heidelberg, Australia
  - code: CU 
    address: Curtin University, Bentley, WA, Australia
  - code: TKI 
    address: Telethon Kids Institute, Perth Children's Hospital, Nedlands, WA, Australia
footnote:
  - code: 1
    text: "Corresponding Author"
abstract: |
  1. Background.
  2. Methods.
  3. Results.
  4. *Synthesis and applications.*

keywords: "daily activity patterns; camera trap; *Felis catus*; invasive predator; interspecific competition; mesopredator release; diel activity; hierarchical generalised additive models; species interactions; *Vulpes vulpes*."
journal: "Ecography"
date: "`r Sys.Date()`"
bibliography: mybibfile.bib
header-includes:
    - \usepackage{setspace}\doublespacing
    - \usepackage{float}
    - \floatplacement{figure}{H}
linenumbers: true
#numbersections: true
csl: ecography.csl
geometry: margin=1.1in
fontsize: 12pt
output:
  bookdown::pdf_book:
    base_format: rticles::elsevier_article
editor_options: 
  chunk_output_type: console
---
\parskip=12pt

\newpage

```{r setup, include = FALSE}

knitr::opts_chunk$set(echo = FALSE, cache = FALSE, warning = FALSE, message = FALSE)

```

INTRODUCTION
==========================

The activity of an animal varies across space and time. 
Spatial activity is major research focus, however, temporal activity is much less studied. 
How temporal activity varies across space is even less understood.

The physical environment and presence of other species shapes animal diel activity. 
Directly and indirectly.

Species interactions - coexistence or avoidance - are important to understand for management. 
Species can avoid other species through spatial and/or temporal avoidance.

Classic models of diel activity overlap- flawed. 
No covariates - how does this change across space. 
Haphazard methods to faciltiate this e.g. (xx).

GAMs increasingly being used for diel activity (but not necessarily species overlap). 
Particularly because you can model covariates. 
However GAMs have  features which lend themselves to this analysis, but have not being used - flexible joint spatial and temporal heirarchical modelling.

Interactions between invasive predators are important to understand.

In this study, we...

\newpage

MATERIALS AND METHODS
==========================

## Study area

We collated camera-trap data from two distinct regions in south-west Victoria, Australia (Fig. 1). The terrestrial mammalian predator guild is depauperate here. Dingoes *Canis familiaris* are now absent throughout, as is the the tiger quoll *Dasyurus maculatus*, more recently absent in the Otway Ranges (last sighted in 2014 despite extensive camera-trapping). Introduced red foxes and feral cat are therefore the only medium-large functional mammalian terrestrial predators across both regions. Across broad sections of each region, government land managers conduct ongoing targeted fox control for biodiversity conservation - poison baits containing 3 mg of sodium mono-fluroacetate (1080) are buried at a depth of 10 cm at 1-km intervals along accessible forest tracks and roads. The purpose of each  collated camera-trap study was to to experimentally survey changes in the mammalian community due to fox control, and so, our study covers a wide range contexts with and without fox-baiting.

In the Glenelg region, Gunditjmara country, natural vegetation has been highly fragmented. Here, camera-survey sites were situated in six distinct forest blocks; two of which separated by a river, others separated by mostly pastoral farming and residential property). Foxes in three forests blocks have been subject to lethal control since October 2005, with 1080 poison-baits replaced fortnightly [@ROBLEY2014262].

The Otway Ranges, Gadubanud and XX country, is a largely continuous patch of natural vegetation with a strong east-west rainfall and elevation gradient [@swan2021species]. Foxes are controlled through 1080 poison-baits with monthly replacement across most of the Otway Ranges (a large section to the north-west remains unbaited). The majority of this baiting began in 2016 - 2017, although fox-baiting commenced in some small sections in 2008.

Alongside fragmentation, the extremely rugged, wet forest and rainforest matrix in the western Otway Ranges is the main difference between the two regions. Prescribed fire, primarily for the purpose of asset protection, is another key management action undertaken throughout both regions (except in very wet sections of Otway Ranges).

## Camera-trapping surveys

We compiled camera-trap data from four distinct survey across these regions. After removing camera-traps which were operational for less than 14 days, this totalled 4276 deployments of camera-traps at 1527 camera-trap sites. Survey durations ranged between 14 and 93 days (mean 43), totalling 183,678 trapnights. The number of sites, deployments and trapnights were nearly equal across both regions. Camera-trap were deployed in the Glenelg region between 2013 – 2019, whereas in the Otway Ranges surveys were conducted in 2016 – 2019. Camera-trap spacing was highly variable; with minimum distance between sites ranging between 40 - 4150 m (mean: 597 m).

All camera-trap deployments consisted of a Reconyx (Holmen, Wisconsin) brand camera-trap (both white and infrared flashes), attached to a tree or a metal picket, facing a lure. Camera-traps were set-up in two ways depending on the key aim of the project, either targeted toward predators or prey species. One study across both regions positioned camera-traps lower on a tree (around 15 - 30 cm above the ground – angled only slightly downwards) facing a tuna oil lure approximately 2 - 2.5 m away [detailed in @REES2019108287]. The remaining camera-traps were positioned higher on a tree or a metal picket (at least 40 cm above ground) and angled downwards more strongly - facing a lure approximately 1 - 1.5 m away. These lures consisted of peanut butter, golden syrup and rolled oats were mixed into a small ball, placed within a tea strainer or PVC pipe container and secured either to the ground, or 20 - 60 cm above ground on a wooden stake. All set-ups were effective in detecting both predator and prey species.

## Data preparation
We first created a table of species detections and deployment information (coordinates, dates) for every camera-trap deployment in R version 3.6.3 [@R]. We discarded camera-traps which were active for less than 14 days. We fit lorelograms using methods provided by @https://doi.org/10.1111/2041-210X.13308 to identify that removing repeat site detections of a species within 30 minutes was sufficient to reduce temporal autocorrelation. To account for different sunrise and sunset times across space and at different times of the year, we converted detection times to solar time; with sunrise and sunset times standardised at 06:00 and 18:00 respectively. We did so using the average double anchoring approach described by @https://doi.org/10.1111/2041-210X.13290 in the ‘activity’ r-package [@rowcliffe_activity].

We attached Ecological Vegetation Class (hereafter 'EVC') groups [@delwp2020] to each camera-trap location. Rainforests are interspersed (primarily in low lying gullies) throughout wet and damp forests at a finescale (relevant to predator movements), and so, we merged these groupings (hereafter 'wet forests'). For each deployment, we calculated the number of poison fox-baits within a 2.3 km radius around each camera-trap - the average maximum distance foxes in these region travel from their home range centre [@hradsky2017human].

Using the 'reshape' r-package [@reshape], we manipulated the detection table into a dataframe with a row for each hour of the day, for every camera-trap deployment, recording the total number of 'independent' fox and feral cat detections within each hour for the entire camera-trap survey duration.

## Generalised additive models of diel activity 
We modelled predator counts (response variable) with generalised additive mixed-effects models (hereafter 'GAMMs') implemented in the 'mgcv' r-package [@wood2017]. We first fit these models using the poisson family, but switched to the negative binomial family as overdispersion, but not zero-inflation, was detected using the 'DHARMa' r-package [@DHARMa]. We specified a model offset to account for differences in camera-trap survey durations and a random intercept for each site to account for repeat sampling. We used a duchon spline static spatial smooth to directly model spatial autocorrelation and differences in predator activity across space (i.e. the effect of unmodelled environmental covariates). For the fox model, we additionally included a thin plate regression spline for poison-bait density, with a factor smoother for each region (due to differences in bait replacement schedules). This formed the base model specification for each model we fitted. Models (n = 4) differed in their specification of the cyclic hour smooth in order to explore (1) variations in predator diel activity across EVC groups, as well as how (2a) feral cat diel activity differed with the presence-absence of foxes, and (2b) across the gradient of observed fox counts (detailed in the sections below). We plotted models using 'ggplot2' and 'gratia' R-packages [@ggplot2; @gratia].

We expected predator activity to be lowest during the day, but potentially having small shifts in relative activity between sunset and sunrise across the different EVC groups (1). And so, fitting separate responses for each EVC group would throw away this shared information, which is particularly costly for the least surveyed groups -- resulting in more noisy estimates. We therefore specified a global smoother for hour (i.e. average response), along with group-level smoothers with shared wiggliness for each EVC group [i.e. model GS in @pedersen2019hierarchical]. This model structure penalises functions which deviate strongly from the average response--allowing confidence that they are true deviations opposed to the result of sampling noise.

Because feral cats are likely subordinate to foxes, we investigated changes in their diel activity in response to both the (2a) presence-absence and (2b) counts of foxes. We added extra columns to our dataframe for whether foxes were detected on each camera-trap or not, as well as the number of fox detections divided by the log of the survey days to account for differences in camera-trap survey durations. Because we saw different fox diel activity patterns in wet forests (more active during the day) relative to the other EVC groups (nocturnal), we expected feral cats to temporally avoid foxes in opposing ways between these wet and dry vegetation groups. We therefore modelled fox-induced changes in feral cat diel activity separately for wet and dry vegetation groups. Because of our large sample size, we split dry vegetation groups by region to for spatial replication. We refer to this variable as 'habitat type'; having three levels:  wet forests of the Otway Ranges ('wet_otways') dry EVC groups of the Otway Ranges ('dry_otways') and all of the Glenelg region data ('dry_glenelg').

To fit a model which estimate feral cat diel activity where foxes were also detected, and camera-trap sites where foxes not detected (2a), we specified an cyclic hourly by variable smooth (for the presence-absence of foxes), with a factor smooth basis for habitat type. To fit a model which allowing feral cat diel activity to change across the range of adjusted fox counts (2b), we used a tensor product of hour and adjusted fox counts, also employing a by variable smooth to estimate this separately across each habitat type. We used a tensor product rather than a linear by variable for fox counts because predator interactions are unlikely to be linear [@https://doi.org/10.1111/j.1365-2664.2009.01650.x].

\newpage

```{r map, out.width="100%", fig.cap = "Locations of study sites in south-west Victoria, Australia. The grids of camera-traps are denoted by white dots. The Glenelg region is to the west and Otway region to the east. Native vegetation is indicated by dark green, with hill shading. Map tiles by Stamen Design, under CC BY 3.0, map data by OpenStreetMap, under CC BY SA."}
knitr::include_graphics("../figs/fig1_map.png")
```

\newpage

RESULTS
============

## Differences in diel activity across vegetation types

## Fox-induced changes in feral cat diel activity

\newpage

```{r tab1}
library(pander)
library(tidyverse)

# load records
records <- read_csv("../derived_data/counts_hour_cleaned.csv")
# drop hour
records <- distinct(records, station_year, .keep_all = TRUE)
# get unique sites dataframe
camdata <- distinct(records, station, .keep_all = TRUE)

table1 <- records %>%
  group_by(region, vegetation_group) %>%
  mutate(sites = length(unique(station)),
         deployments = length(unique(station_year)),
         fox_count_vr = sum(fox_count),
         cat_count_vr = sum(cat_count),
         fox_present = sum(fox_pa == "present"),
         cat_present = sum(cat_pa == "present"),
         fox_naive = (fox_present / deployments),
         cat_naive = (cat_present / deployments)
         )

table1 <- table1 %>%
  select(region, vegetation_group, sites, deployments, fox_count_vr, cat_count_vr, fox_naive, cat_naive)  %>% 
  distinct() %>%
  arrange(vegetation_group, region)

# rename region 
table1$region <- factor(table1$region)
levels(table1$region) <- c("Glenelg", "Otway")

# make table
pander(table1, style = "simple", split.table = Inf, caption = "Summary of camera-trap deployments in unique sites within Ecological Vegetation Class groups, as well as independent counts of predator detections and naive occupancy rates, across two broad regions in south-west Victoria, Australia.", round = 2,
              col.names = c("Region", "Vegetation", "Sites", "Deployments",  "Fox counts", "Cat counts", "Fox present", "Cat present"),
)

```

\newpage

```{r veg, out.width="100%", fig.cap = "Fox (A) and feral cat (B) diel activity patterns across Ecological Vegetation Class groups in south-west Victoria, Australia. Dotted, vertical lines respresent sunrise and sunset times. Shaded areas indicate 95\\% confidence intervals."}
knitr::include_graphics("../figs/predator_veg.png")
```

\newpage

```{r count, out.width="100%", fig.cap="Predicted feral cat diel activity (A) and uncertainty (B) across each habitat type in response to foxes (log counts adjusted for survey effort). Grey vertical lines respresent sunrise and sunset times. Shaded areas indicate 95\\% confidence intervals."}
knitr::include_graphics("../figs/cat_fox_count.png")
```

\newpage

DISCUSSION
===================

\newpage

ACKNOWLEDGEMENTS
===================
We acknowledge and pay respect to the Gadubanud and Gunditjmara people on whose traditional lands this study took place. This experiment was a collaborative effort between the Glenelg Ark (Department of Environment, Land Water and Planning) and Otway Ark (Parks Victoria) working groups. We are extremely thankful to those who led and assisted with these camera-trap deployments and image processing. Univeristy of Melbounre led surveys were conducted under University of Melbourne Animal Ethics Committee approval 1714119 and Victorian Government Department of Environment, Land Water and Planning Research Permit 10008273. Our study was generously supported by the Conservation Ecology Centre, the Victorian Government Department of Environment, Land Water and Planning, Arthur Rylah Institute for Environmental Research, Parks Victoria and the Australian Government’s National Environmental Science Program through the Threatened Species Recovery Hub, and ARC Linkage Project LP170101134. MR also receives support from an Australian Government Research Training Program Scholarship.

OPEN RESEARCH
===================

Raw data and code are on Github link xx.  
Data will be deposited on the Dryad Digital Repository after acceptance.

\newpage

REFERENCES {#references .unnumbered}
==========

### 2021-07-18 manuscript/manuscript.tex
added manuscript

\documentclass[]{elsarticle} %review=doublespace preprint=single 5p=2 column
%%% Begin My package additions %%%%%%%%%%%%%%%%%%%
\usepackage[hyphens]{url}

  \journal{Ecography} % Sets Journal name

\usepackage{lineno} % add
  \linenumbers % turns line numbering on
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

\usepackage{graphicx}
\usepackage{booktabs} % book-quality tables
%%%%%%%%%%%%%%%% end my additions to header

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{fontspec}
  \ifxetex
    \usepackage{xltxtra,xunicode}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\usepackage[margin=1.1in]{geometry}
\bibliographystyle{elsarticle-harv}
\usepackage{longtable}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={Dynamic shifts in predator diel activity due to habitat type and fear},
            colorlinks=false,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls

\setcounter{secnumdepth}{5}
% Pandoc toggle for numbering sections (defaults to be off)

% Pandoc citation processing

% Pandoc header
\usepackage{setspace}\doublespacing
\usepackage{float}
\floatplacement{figure}{H}

\begin{document}
\begin{frontmatter}

  \title{Dynamic shifts in predator diel activity due to habitat type and fear}
    \author[UOM]{Matthew W. Rees\corref{1}}
   \ead{matt.wayne.rees@gmail.com} 
    \author[CEC]{Jack H. Pascoe}

    \author[CEC]{Mark Le Pla}

    \author[CEC]{Emma K. Birnbaum}

    \author[ARI]{Alan Robley}

    \author[CU,TKI]{Nick Golding}

    \author[UOM]{Brendan A. Wintle}

    \author[UOM]{Bronwyn A. Hradsky}

      \address[UOM]{Quantitative \& Applied Ecology Group, School of Ecosystem and Forest Science, The University of Melbourne, Parkville, VIC, Australia}
    \address[CEC]{Conservation Ecology Centre, Otway Lighthouse Rd, Cape Otway, VIC, Australia}
    \address[ARI]{Department of Environment, Land, Water and Planning, Arthur Rylah Institute for Environmental Research, Heidelberg, Australia}
    \address[CU]{Curtin University, Bentley, WA, Australia}
    \address[TKI]{Telethon Kids Institute, Perth Children's Hospital, Nedlands, WA, Australia}
      \cortext[1]{Corresponding Author}

  \begin{abstract}
  \begin{enumerate}
  \def\labelenumi{\arabic{enumi}.}
  \tightlist
  \item
    Background.
  \item
    Methods.
  \item
    Results.
  \item
    \emph{Synthesis and applications.}
  \end{enumerate}
  \end{abstract}
   \begin{keyword} daily activity patterns; camera trap; \emph{Felis catus}; invasive predator; interspecific competition; mesopredator release; diel activity; hierarchical generalised additive models; species interactions; \emph{Vulpes vulpes}.\end{keyword}
 \end{frontmatter}

\parskip=12pt

\newpage

\hypertarget{introduction}{%
\section{INTRODUCTION}\label{introduction}}

The activity of an animal varies across space and time.
Spatial activity is major research focus, however, temporal activity is much less studied.
How temporal activity varies across space is even less understood.

The physical environment and presence of other species shapes animal diel activity.
Directly and indirectly.

Species interactions - coexistence or avoidance - are important to understand for management.
Species can avoid other species through spatial and/or temporal avoidance.

Classic models of diel activity overlap- flawed.
No covariates - how does this change across space.
Haphazard methods to faciltiate this e.g.~(xx).

GAMs increasingly being used for diel activity (but not necessarily species overlap).
Particularly because you can model covariates.
However GAMs have features which lend themselves to this analysis, but have not being used - flexible joint spatial and temporal heirarchical modelling.

Interactions between invasive predators are important to understand.

In this study, we\ldots{}

\newpage

\hypertarget{materials-and-methods}{%
\section{MATERIALS AND METHODS}\label{materials-and-methods}}

\hypertarget{study-area}{%
\subsection{Study area}\label{study-area}}

We collated camera-trap data from two distinct regions in south-west Victoria, Australia (Fig. 1). The terrestrial mammalian predator guild is depauperate here. Dingoes \emph{Canis familiaris} are now absent throughout, as is the the tiger quoll \emph{Dasyurus maculatus}, more recently absent in the Otway Ranges (last sighted in 2014 despite extensive camera-trapping). Introduced red foxes and feral cat are therefore the only medium-large functional mammalian terrestrial predators across both regions. Across broad sections of each region, government land managers conduct ongoing targeted fox control for biodiversity conservation - poison baits containing 3 mg of sodium mono-fluroacetate (1080) are buried at a depth of 10 cm at 1-km intervals along accessible forest tracks and roads. The purpose of each collated camera-trap study was to to experimentally survey changes in the mammalian community due to fox control, and so, our study covers a wide range contexts with and without fox-baiting.

In the Glenelg region, Gunditjmara country, natural vegetation has been highly fragmented. Here, camera-survey sites were situated in six distinct forest blocks; two of which separated by a river, others separated by mostly pastoral farming and residential property). Foxes in three forests blocks have been subject to lethal control since October 2005, with 1080 poison-baits replaced fortnightly (Robley et al. 2014).

The Otway Ranges, Gadubanud and XX country, is a largely continuous patch of natural vegetation with a strong east-west rainfall and elevation gradient (Swan et al. 2021). Foxes are controlled through 1080 poison-baits with monthly replacement across most of the Otway Ranges (a large section to the north-west remains unbaited). The majority of this baiting began in 2016 - 2017, although fox-baiting commenced in some small sections in 2008.

Alongside fragmentation, the extremely rugged, wet forest and rainforest matrix in the western Otway Ranges is the main difference between the two regions. Prescribed fire, primarily for the purpose of asset protection, is another key management action undertaken throughout both regions (except in very wet sections of Otway Ranges).

\hypertarget{camera-trapping-surveys}{%
\subsection{Camera-trapping surveys}\label{camera-trapping-surveys}}

We compiled camera-trap data from four distinct survey across these regions. After removing camera-traps which were operational for less than 14 days, this totalled 4276 deployments of camera-traps at 1527 camera-trap sites. Survey durations ranged between 14 and 93 days (mean 43), totalling 183,678 trapnights. The number of sites, deployments and trapnights were nearly equal across both regions. Camera-trap were deployed in the Glenelg region between 2013 -- 2019, whereas in the Otway Ranges surveys were conducted in 2016 -- 2019. Camera-trap spacing was highly variable; with minimum distance between sites ranging between 40 - 4150 m (mean: 597 m).

All camera-trap deployments consisted of a Reconyx (Holmen, Wisconsin) brand camera-trap (both white and infrared flashes), attached to a tree or a metal picket, facing a lure. Camera-traps were set-up in two ways depending on the key aim of the project, either targeted toward predators or prey species. One study across both regions positioned camera-traps lower on a tree (around 15 - 30 cm above the ground -- angled only slightly downwards) facing a tuna oil lure approximately 2 - 2.5 m away (detailed in Rees et al. 2019). The remaining camera-traps were positioned higher on a tree or a metal picket (at least 40 cm above ground) and angled downwards more strongly - facing a lure approximately 1 - 1.5 m away. These lures consisted of peanut butter, golden syrup and rolled oats were mixed into a small ball, placed within a tea strainer or PVC pipe container and secured either to the ground, or 20 - 60 cm above ground on a wooden stake. All set-ups were effective in detecting both predator and prey species.

\hypertarget{data-preparation}{%
\subsection{Data preparation}\label{data-preparation}}

We first created a table of species detections and deployment information (coordinates, dates) for every camera-trap deployment in R version 3.6.3 (R Core Team 2020). We discarded camera-traps which were active for less than 14 days. We fit lorelograms using methods provided by Iannarilli et al. (2019) to identify that removing repeat site detections of a species within 30 minutes was sufficient to reduce temporal autocorrelation. To account for different sunrise and sunset times across space and at different times of the year, we converted detection times to solar time; with sunrise and sunset times standardised at 06:00 and 18:00 respectively. We did so using the average double anchoring approach described by Vazquez et al. (2019) in the `activity' r-package (Rowcliffe 2019).

We attached Ecological Vegetation Class (hereafter `EVC') groups (DELWP 2020) to each camera-trap location. Rainforests are interspersed (primarily in low lying gullies) throughout wet and damp forests at a finescale (relevant to predator movements), and so, we merged these groupings (hereafter `wet forests'). For each deployment, we calculated the number of poison fox-baits within a 2.3 km radius around each camera-trap - the average maximum distance foxes in these region travel from their home range centre (Hradsky et al. 2017).

Using the `reshape' r-package (Wickham 2007), we manipulated the detection table into a dataframe with a row for each hour of the day, for every camera-trap deployment, recording the total number of `independent' fox and feral cat detections within each hour for the entire camera-trap survey duration.

\hypertarget{generalised-additive-models-of-diel-activity}{%
\subsection{Generalised additive models of diel activity}\label{generalised-additive-models-of-diel-activity}}

We modelled predator counts (response variable) with generalised additive mixed-effects models (hereafter `GAMMs') implemented in the `mgcv' r-package (Wood 2017). We first fit these models using the poisson family, but switched to the negative binomial family as overdispersion, but not zero-inflation, was detected using the `DHARMa' r-package (Hartig 2020). We specified a model offset to account for differences in camera-trap survey durations and a random intercept for each site to account for repeat sampling. We used a duchon spline static spatial smooth to directly model spatial autocorrelation and differences in predator activity across space (i.e.~the effect of unmodelled environmental covariates). For the fox model, we additionally included a thin plate regression spline for poison-bait density, with a factor smoother for each region (due to differences in bait replacement schedules). This formed the base model specification for each model we fitted. Models (n = 4) differed in their specification of the cyclic hour smooth in order to explore (1) variations in predator diel activity across EVC groups, as well as how (2a) feral cat diel activity differed with the presence-absence of foxes, and (2b) across the gradient of observed fox counts (detailed in the sections below). We plotted models using `ggplot2' and `gratia' R-packages (Wickham 2016, Simpson 2021).

We expected predator activity to be lowest during the day, but potentially having small shifts in relative activity between sunset and sunrise across the different EVC groups (1). And so, fitting separate responses for each EVC group would throw away this shared information, which is particularly costly for the least surveyed groups -- resulting in more noisy estimates. We therefore specified a global smoother for hour (i.e.~average response), along with group-level smoothers with shared wiggliness for each EVC group (i.e.~model GS in Pedersen et al. 2019). This model structure penalises functions which deviate strongly from the average response--allowing confidence that they are true deviations opposed to the result of sampling noise.

Because feral cats are likely subordinate to foxes, we investigated changes in their diel activity in response to both the (2a) presence-absence and (2b) counts of foxes. We added extra columns to our dataframe for whether foxes were detected on each camera-trap or not, as well as the number of fox detections divided by the log of the survey days to account for differences in camera-trap survey durations. Because we saw different fox diel activity patterns in wet forests (more active during the day) relative to the other EVC groups (nocturnal), we expected feral cats to temporally avoid foxes in opposing ways between these wet and dry vegetation groups. We therefore modelled fox-induced changes in feral cat diel activity separately for wet and dry vegetation groups. Because of our large sample size, we split dry vegetation groups by region to for spatial replication. We refer to this variable as `habitat type'; having three levels: wet forests of the Otway Ranges (`wet\_otways') dry EVC groups of the Otway Ranges (`dry\_otways') and all of the Glenelg region data (`dry\_glenelg').

To fit a model which estimate feral cat diel activity where foxes were also detected, and camera-trap sites where foxes not detected (2a), we specified an cyclic hourly by variable smooth (for the presence-absence of foxes), with a factor smooth basis for habitat type. To fit a model which allowing feral cat diel activity to change across the range of adjusted fox counts (2b), we used a tensor product of hour and adjusted fox counts, also employing a by variable smooth to estimate this separately across each habitat type. We used a tensor product rather than a linear by variable for fox counts because predator interactions are unlikely to be linear (Johnson and VanDerWal 2009).

\newpage

\begin{figure}
\includegraphics[width=1\linewidth]{../figs/fig1_map} \caption{Locations of study sites in south-west Victoria, Australia. The grids of camera-traps are denoted by white dots. The Glenelg region is to the west and Otway region to the east. Native vegetation is indicated by dark green, with hill shading. Map tiles by Stamen Design, under CC BY 3.0, map data by OpenStreetMap, under CC BY SA.}\label{fig:map}
\end{figure}

\newpage

\hypertarget{results}{%
\section{RESULTS}\label{results}}

\hypertarget{differences-in-diel-activity-across-vegetation-types}{%
\subsection{Differences in diel activity across vegetation types}\label{differences-in-diel-activity-across-vegetation-types}}

\hypertarget{fox-induced-changes-in-feral-cat-diel-activity}{%
\subsection{Fox-induced changes in feral cat diel activity}\label{fox-induced-changes-in-feral-cat-diel-activity}}

\newpage

\begin{longtable}[]{@{}cccccccc@{}}
\caption{Summary of camera-trap deployments in unique sites within Ecological Vegetation Class groups, as well as independent counts of predator detections and naive occupancy rates, across two broad regions in south-west Victoria, Australia.}\tabularnewline
\toprule
Region & Vegetation & Sites & Deployments & Fox counts & Cat counts & Fox present & Cat present\tabularnewline
\midrule
\endfirsthead
\toprule
Region & Vegetation & Sites & Deployments & Fox counts & Cat counts & Fox present & Cat present\tabularnewline
\midrule
\endhead
Glenelg & Dry Forest & 25 & 68 & 347 & 9 & 0.66 & 0.1\tabularnewline
Otway & Dry Forest & 111 & 310 & 341 & 158 & 0.42 & 0.28\tabularnewline
Glenelg & Heathland & 40 & 119 & 265 & 59 & 0.43 & 0.34\tabularnewline
Otway & Heathland & 3 & 9 & 8 & 6 & 0.22 & 0.44\tabularnewline
Glenelg & Heathy Woodland & 154 & 424 & 574 & 96 & 0.31 & 0.14\tabularnewline
Otway & Heathy Woodland & 82 & 256 & 160 & 66 & 0.29 & 0.14\tabularnewline
Glenelg & Herb-rich Woodland & 59 & 372 & 860 & 197 & 0.5 & 0.27\tabularnewline
Otway & Herb-rich Woodland & 2 & 6 & 3 & 2 & 0.33 & 0.17\tabularnewline
Glenelg & Lowland Forest & 383 & 1046 & 1900 & 290 & 0.46 & 0.18\tabularnewline
Otway & Lowland Forest & 52 & 162 & 190 & 35 & 0.42 & 0.14\tabularnewline
Glenelg & Swampy Scrub & 4 & 10 & 19 & 8 & 0.6 & 0.5\tabularnewline
Otway & Swampy Scrub & 36 & 97 & 64 & 88 & 0.32 & 0.33\tabularnewline
Otway & Wet Forest & 281 & 780 & 715 & 1187 & 0.31 & 0.54\tabularnewline
\bottomrule
\end{longtable}

\newpage

\begin{figure}
\includegraphics[width=1\linewidth]{../figs/predator_veg} \caption{Fox (A) and feral cat (B) diel activity patterns across Ecological Vegetation Class groups in south-west Victoria, Australia. Dotted, vertical lines respresent sunrise and sunset times. Shaded areas indicate 95\% confidence intervals.}\label{fig:veg}
\end{figure}

\newpage

\begin{figure}
\includegraphics[width=1\linewidth]{../figs/cat_fox_count} \caption{Predicted feral cat diel activity (A) and uncertainty (B) across each habitat type in response to foxes (log counts adjusted for survey effort). Grey vertical lines respresent sunrise and sunset times. Shaded areas indicate 95\% confidence intervals.}\label{fig:count}
\end{figure}

\newpage

\hypertarget{discussion}{%
\section{DISCUSSION}\label{discussion}}

\newpage

\hypertarget{acknowledgements}{%
\section{ACKNOWLEDGEMENTS}\label{acknowledgements}}

We acknowledge and pay respect to the Gadubanud and Gunditjmara people on whose traditional lands this study took place. This experiment was a collaborative effort between the Glenelg Ark (Department of Environment, Land Water and Planning) and Otway Ark (Parks Victoria) working groups. We are extremely thankful to those who led and assisted with these camera-trap deployments and image processing. Univeristy of Melbounre led surveys were conducted under University of Melbourne Animal Ethics Committee approval 1714119 and Victorian Government Department of Environment, Land Water and Planning Research Permit 10008273. Our study was generously supported by the Conservation Ecology Centre, the Victorian Government Department of Environment, Land Water and Planning, Arthur Rylah Institute for Environmental Research, Parks Victoria and the Australian Government's National Environmental Science Program through the Threatened Species Recovery Hub, and ARC Linkage Project LP170101134. MR also receives support from an Australian Government Research Training Program Scholarship.

\hypertarget{open-research}{%
\section{OPEN RESEARCH}\label{open-research}}

Raw data and code are on Github link xx.\\
Data will be deposited on the Dryad Digital Repository after acceptance.

\newpage

\hypertarget{references}{%
\section*{REFERENCES}\label{references}}
\addcontentsline{toc}{section}{REFERENCES}

\hypertarget{refs}{}
\leavevmode\hypertarget{ref-delwp2020}{}%
DELWP 2020. Bioregions and evc benchmarks.

\leavevmode\hypertarget{ref-DHARMa}{}%
Hartig, F. 2020. DHARMa: Residual diagnostics for hierarchical (multi-level / mixed) regression models.

\leavevmode\hypertarget{ref-hradsky2017human}{}%
Hradsky, B. A. et al. 2017. Human-modified habitats facilitate forest-dwelling populations of an invasive predator, vulpes vulpes. - Scientific Reports 7: 1--12.

\leavevmode\hypertarget{ref-https:ux2fux2fdoi.orgux2f10.1111ux2f2041-210X.13308}{}%
Iannarilli, F. et al. 2019. Using lorelograms to measure and model correlation in binary data: Applications to ecological studies. - Methods in Ecology and Evolution 10: 2153--2162.

\leavevmode\hypertarget{ref-https:ux2fux2fdoi.orgux2f10.1111ux2fj.1365-2664.2009.01650.x}{}%
Johnson, C. N. and VanDerWal, J. 2009. Evidence that dingoes limit abundance of a mesopredator in eastern australian forests. - Journal of Applied Ecology 46: 641--646.

\leavevmode\hypertarget{ref-pedersen2019hierarchical}{}%
Pedersen, E. J. et al. 2019. Hierarchical generalized additive models in ecology: An introduction with mgcv. - PeerJ 7: e6876.

\leavevmode\hypertarget{ref-R}{}%
R Core Team 2020. R: A language and environment for statistical computing. - R Foundation for Statistical Computing.

\leavevmode\hypertarget{ref-REES2019108287}{}%
Rees, M. W. et al. 2019. Unexpectedly high densities of feral cats in a rugged temperate forest. - Biological Conservation 239: 108287.

\leavevmode\hypertarget{ref-ROBLEY2014262}{}%
Robley, A. et al. 2014. Long-term and large-scale control of the introduced red fox increases native mammal occupancy in australian forests. - Biological Conservation 180: 262--269.

\leavevmode\hypertarget{ref-rowcliffe_activity}{}%
Rowcliffe, M. 2019. Activity: Animal activity statistics.

\leavevmode\hypertarget{ref-gratia}{}%
Simpson, G. L. 2021. Gratia: Graceful 'ggplot'-based graphics and other functions for gams fitted using 'mgcv'.

\leavevmode\hypertarget{ref-swan2021species}{}%
Swan, M. et al. 2021. Species distribution models for conservation planning in fire-prone landscapes. - Biodiversity and Conservation 30: 1119--1136.

\leavevmode\hypertarget{ref-https:ux2fux2fdoi.orgux2f10.1111ux2f2041-210X.13290}{}%
Vazquez, C. et al. 2019. Comparing diel activity patterns of wildlife across latitudes and seasons: Time transformations using day length. - Methods in Ecology and Evolution 10: 2057--2066.

\leavevmode\hypertarget{ref-reshape}{}%
Wickham, H. 2007. Reshaping data with the reshape package. - Journal of Statistical Software 21: 1--20.

\leavevmode\hypertarget{ref-ggplot2}{}%
Wickham, H. 2016. Ggplot2: Elegant graphics for data analysis. - Springer-Verlag New York.

\leavevmode\hypertarget{ref-wood2017}{}%
Wood, S. N. 2017. Generalized additive models: An introduction with r. - CRC press.

\end{document}

### 2021-07-18 r_scripts/5_gams_diel_activity.R

@@ -44,14 +44,16 @@ table(records$XGROUPNAME)
 
 
 # ADD FOX COUNTS PER CAMERA -----------------------------------------------
-## Single fox count for each cam-trap
+## Single fox count for each cam-trap (and do the same for cats for good measure)
 records <- records %>%
   group_by(station_year)  %>%
-  mutate(fox_count = sum(fox))
+  mutate(fox_count = sum(fox),
+         cat_count = sum(cat))
 
 # fox presence / absence
 records$fox_pa <- if_else(records$fox_count > 0, "present", "absent")
-table(records$fox_pa)
+records$cat_pa <- if_else(records$cat_count > 0, "present", "absent")
+xtabs(~records$fox_pa + records$cat_pa)
 
 # adjust (positive only - we don't want negative counts) fox counts for survey effort 
 records$fox_count_adj <- ifelse(records$fox_count > 0, records$fox_count/log(records$survey_duration), records$fox_count)
@@ -82,6 +84,9 @@ records <- transform(records,
                      survey_duration = as.integer(survey_duration)
 )
 summary(records)
+
+# save modified dataframe
+write.csv(records, "derived_data/counts_hour_cleaned.csv")
 
 
 # GAMS --------------------------------------------------------------------
@@ -237,7 +242,7 @@ plot_range_se <- ggplot(aes(hour, fox_count_adj, fill = se.fit), data = df_int) 
   xlab("Hour") 
 
 # plot / save
-png("figs/cat_fox_count.png", width = 12, height = 8.5, res = 600, units = "in")
+png("figs/cat_fox_count.png", width = 9, height = 6.5, res = 600, units = "in")
 plot_range / plot_range_se + plot_annotation(tag_levels = "A")
 dev.off()
 

### 2021-07-28 manuscript/manuscript.Rmd
select = TRUE

@@
 
-The activity of an animal varies across space and time. 
-Spatial activity is major research focus, however, temporal activity is much less studied. 
-How temporal activity varies across space is even less understood.
+Fear of another species is a powerful driver of animal behaviour, but difficult to measure.
+Both threats and animal responses to perceived threats are dynamic - varying across space and time.
 
-The physical environment and presence of other species shapes animal diel activity. 
-Directly and indirectly.
+Avoidance behaviour depends on need (how much of a threat is this?) and ability (how much is avoidance going to cost me?)
+For example, avoidance behaviours are less tenable in resource poor environments. Some environment more easily facilitate avoidance (e.g. shelter availability). 
+Landscape context is therefore key.
 
-Species interactions - coexistence or avoidance - are important to understand for management. 
-Species can avoid other species through spatial and/or temporal avoidance.
+Subordinate animals may avoid dominant ones by altering what time of day they are active, which places they are active at, or which which places they are active at during certain times of the day. 
+Space use well studied, time not so much. Space and time - extremely rarely.
 
-Classic models of diel activity overlap- flawed. 
-No covariates - how does this change across space. 
-Haphazard methods to faciltiate this e.g. (xx).
+Methods to monitor spatial interactions. 
+Simple correlations to JSDMs.
 
-GAMs increasingly being used for diel activity (but not necessarily species overlap). 
-Particularly because you can model covariates. 
-However GAMs have  features which lend themselves to this analysis, but have not being used - flexible joint spatial and temporal heirarchical modelling.
+Methods to monitor temporal interactions. 
+Coefficient of overlap -- arbitrary, don't allow covariates, highly susceptible to noise from small sample sizes.
 
-Interactions between invasive predators are important to understand.
+(Lack of) methods to monitor spatio-temporal behaviour. 
+GAMs increasingly used for both spatial and diel activity, but not for both (despite easily facilitating this).
 
-In this study, we...
+Species interactions are important to understand for management. 
+Particularly invasive predators. 
+Cat-fox context.
+
+In this study, we used GAMs to (1) assess how predators change diel activity across vegetation types, and (2) changes in both diel and spatial activity of a mesopredator in response to an apex predator.
+Used a simple predator system, as well as manipulated apex predator activity through lethal control.
 
@@
 
-We compiled camera-trap data from four distinct survey across these regions.
-After removing camera-traps which were operational for less than 14 days, this totalled 4276 deployments of camera-traps at 1527 camera-trap sites.
-Survey durations ranged between 14 and 93 days (mean 43), totalling 183,678 trapnights.
-The number of sites, deployments and trapnights were nearly equal across both regions.
+We compiled camera-trap data from three distinct studies across the two regions.
+After removing camera-traps which were operational for less than 14 days, this totalled 3659 deployments of camera-traps at 1232 camera-trap sites.
+Survey durations ranged between 14 and 93 days (mean 47), totalling 171,966 trapnights.
+Camera-trap spacing was variable; the average minimum distance between simultaneously deployed camera-traps was 445, 853 and 1266 metres for each study.
 Camera-trap were deployed in the Glenelg region between 2013 – 2019, whereas in the Otway Ranges surveys were conducted in 2016 – 2019.
-Camera-trap spacing was highly variable; with minimum distance between sites ranging between 40 - 4150 m (mean: 597 m).
 
@@
 We first created a table of species detections and deployment information (coordinates, dates) for every camera-trap deployment in R version 3.6.3 [@R].
-We discarded camera-traps which were active for less than 14 days.
-We fit lorelograms using methods provided by @https://doi.org/10.1111/2041-210X.13308 to identify that removing repeat site detections of a species within 30 minutes was sufficient to reduce temporal autocorrelation.
-To account for different sunrise and sunset times across space and at different times of the year, we converted detection times to solar time; with sunrise and sunset times standardised at 06:00 and 18:00 respectively.
-We did so using the average double anchoring approach described by @https://doi.org/10.1111/2041-210X.13290 in the ‘activity’ r-package [@rowcliffe_activity].
+We used lorelograms to identify the minimum interval to approximate independence [@https://doi.org/10.1111/2041-210X.13308]; discarding repeat detections of a species within 30 minutes to reduce temporal autocorrelation.
+To account for day length variation across space and time, we extracted sunrise and sunset times for each camera-trap deployment using the 'maptools' R-package [@maptools] and adjusted detection times to be relative to sunrise and sunset using the average double anchoring approach described by @https://doi.org/10.1111/2041-210X.13290.
 
@@
 
-Using the 'reshape' r-package [@reshape], we manipulated the detection table into a dataframe with a row for each hour of the day, for every camera-trap deployment, recording the total number of 'independent' fox and feral cat detections within each hour for the entire camera-trap survey duration.
+Using the 'reshape' R-package [@reshape], we manipulated the detection table into a dataframe with a row for each hour of the day, for every camera-trap deployment, recording the total number of 'independent' fox and feral cat detections within each hour for the entire camera-trap survey duration.
 
 ## Generalised additive models of diel activity 
-We modelled predator counts (response variable) with generalised additive mixed-effects models (hereafter 'GAMMs') implemented in the 'mgcv' r-package [@wood2017].
-We first fit these models using the poisson family, but switched to the negative binomial family as overdispersion, but not zero-inflation, was detected using the 'DHARMa' r-package [@DHARMa].
+We modelled predator counts (response variable) with generalised additive mixed-effects models (hereafter 'GAMMs') implemented in the 'mgcv' R-package [@wood2017].
+We first fit these models using the poisson family, but switched to the negative binomial family as overdispersion, but not zero-inflation, was detected using the 'DHARMa' R-package [@DHARMa].
 We specified a model offset to account for differences in camera-trap survey durations and a random intercept for each site to account for repeat sampling.
-We used a duchon spline static spatial smooth to directly model spatial autocorrelation and differences in predator activity across space (i.e.
+We used a duchon spline static spatial smooth to model both spatial autocorrelation and differences in predator activity across space (i.e.
 the effect of unmodelled environmental covariates).
-For the fox model, we additionally included a thin plate regression spline for poison-bait density, with a factor smoother for each region (due to differences in bait replacement schedules).
+For the fox model, we additionally included a thin plate regression spline for poison-bait density, with a factor smoother for each region, as bait replacement schedules differed between the regions.
 This formed the base model specification for each model we fitted.
-Models (n = 4) differed in their specification of the cyclic hour smooth in order to explore (1) variations in predator diel activity across EVC groups, as well as how (2a) feral cat diel activity differed with the presence-absence of foxes, and (2b) across the gradient of observed fox counts (detailed in the sections below).
+Models (n = 3) differed in their specification of the cyclical hour smooth in order to explore (1) variations in predator diel activity across EVC groups, as well as (2) whether feral cat diel activity changed across the gradient of observed fox counts (detailed in the sections below).
 We plotted models using 'ggplot2' and 'gratia' R-packages [@ggplot2; @gratia].
@@
 model GS in @pedersen2019hierarchical].
-This model structure penalises functions which deviate strongly from the average response--allowing confidence that they are true deviations opposed to the result of sampling noise.
-
-Because feral cats are likely subordinate to foxes, we investigated changes in their diel activity in response to both the (2a) presence-absence and (2b) counts of foxes.
-We added extra columns to our dataframe for whether foxes were detected on each camera-trap or not, as well as the number of fox detections divided by the log of the survey days to account for differences in camera-trap survey durations.
-Because we saw different fox diel activity patterns in wet forests (more active during the day) relative to the other EVC groups (nocturnal), we expected feral cats to temporally avoid foxes in opposing ways between these wet and dry vegetation groups.
+This model structure penalises functions which deviate strongly from the average response--increasing confidence that they are true deviations opposed to the result of sampling noise.
+Because feral cats are likely subordinate to foxes, we investigated changes in their diel activity in response to counts of foxes.
+We added an extra column to our dataframe with the number of fox detections divided by the log of the survey days to account for differences in camera-trap survey durations (hereafter 'adjusted fox counts').
+Because we saw different fox diel activity patterns in wet forests relative to the other EVC groups, we expected feral cats to temporally avoid foxes in opposing ways between these wet and dry vegetation groups.
 We therefore modelled fox-induced changes in feral cat diel activity separately for wet and dry vegetation groups.
-Because of our large sample size, we split dry vegetation groups by region to for spatial replication.
+Because of our large sample size, we further split dry vegetation groups by region for spatial replication.
 We refer to this variable as 'habitat type'; having three levels: wet forests of the Otway Ranges ('wet_otways') dry EVC groups of the Otway Ranges ('dry_otways') and all of the Glenelg region data ('dry_glenelg').
-
-To fit a model which estimate feral cat diel activity where foxes were also detected, and camera-trap sites where foxes not detected (2a), we specified an cyclic hourly by variable smooth (for the presence-absence of foxes), with a factor smooth basis for habitat type.
-To fit a model which allowing feral cat diel activity to change across the range of adjusted fox counts (2b), we used a tensor product of hour and adjusted fox counts, also employing a by variable smooth to estimate this separately across each habitat type.
-We used a tensor product rather than a linear by variable for fox counts because predator interactions are unlikely to be linear [@https://doi.org/10.1111/j.1365-2664.2009.01650.x].
+We used a tensor product of hour (cyclic cubic regression spline) and log adjusted fox counts (thin plate regression spline) to model feral cat diel activity across the range of observed fox activity, separately for each habitat type.
 
@@
 
-## Differences in diel activity across vegetation types
+## Predator diel activity across vegetation types
+
+On average, both predator species occupied similar times of the day. Lowest during the day, peaking during sunset and sunrise, high at night. 
+Foxes had a stronger diel activity patterns than cats, much less likely to be out during the day. 
+Both predators had slight variations in diel activity across EVC groups, except for wet forests, were more substantial changes were seen. 
+Foxes barely had any diel activity pattern here, cats more nocturnal.
+Deviance explained and Rsq values for the two models were XX.
 
 ## Fox-induced changes in feral cat diel activity
+
+We saw evidence of changes in cat diel activity due to fox activity in all three habitat types.
+Similar responses in dry habitats of both regions -- cat spatial activity increased with fox spatial activity, but became more diurnal.
+In Otway wet forests, no real change in spatial activity of cats due to foxes, but cats became more strongly nocturnal.  
+Deviance explained and Rsq value was XX.
+Slightly higher uncertainty where fox spatial activity is high, this is because we had a largest sample size for fox absences, and fewest at the extreme of fox activity.
 
@@
 ```{r veg, out.width="100%", fig.cap = "Fox (A) and feral cat (B) diel activity patterns across Ecological Vegetation Class groups in south-west Victoria, Australia.
-Dotted, vertical lines respresent sunrise and sunset times.
+Dotted, vertical lines respresent average sunrise and sunset times.
 Shaded areas indicate 95\\% confidence intervals."}
@@
 ```{r count, out.width="100%", fig.cap="Predicted feral cat diel activity (A) and uncertainty (B) across each habitat type in response to foxes (log counts adjusted for survey effort).
-Grey vertical lines respresent sunrise and sunset times.
+Grey vertical lines respresent average sunrise and sunset times.
 Shaded areas indicate 95\\% confidence intervals."}
@@
 
+Foxes likely impacted the spatiotemporal behaviour of feral cats. One of the few studies to actually demonstrate and replicate this across ecosystems. 
+Happens in different ways - context key.
+Spatial only, as well as temporal only models would have had opposite inference - high overlap.
+Joint spatiotemporal analysis is key to understand predator interactions.
+
+Our approach is simple and flexible. 
+Small sample sizes are usually a problem. We demonstrate hierarchical model specification to share information - ability to pool all available data. 
+Our approach is more aligned with null hypothesis testing - small sample sizes shrink to a flat line (or to the average), in contrast to circular overlap. 
+Could be easily adapted to other contexts. E.g. instead of vegetation type, could do individual variation in diel activity.
+
+Combined spatiotemporal modelling - differences to other approaches. 
+Cunningham paper most similar to our approach - different in diel activity only, and linear responses.
+We used a tensor product for fox counts because predator interactions may not be linear [@https://doi.org/10.1111/j.1365-2664.2009.01650.x].
+
+Limitations of our approach. 
+Detectability ignored. 
+Individual heterogeneity ignored.
+Correlation does not = causation. 
+Does not (clearly) seperate numerical responses from behavioural (Borchers SCR paper)
+
+Invasive predators are useful species to test this on, as they are extremely adaptable to different conditions.
+Implications for prey.
+
+Conclusions.
+
 \newpage
@@
 We are extremely thankful to those who led and assisted with these camera-trap deployments and image processing.
-Univeristy of Melbounre led surveys were conducted under University of Melbourne Animal Ethics Committee approval 1714119 and Victorian Government Department of Environment, Land Water and Planning Research Permit 10008273.
+University of Melbourne led surveys were conducted under University of Melbourne Animal Ethics Committee approval 1714119 and Victorian Government Department of Environment, Land Water and Planning Research Permit 10008273.
 Our study was generously supported by the Conservation Ecology Centre, the Victorian Government Department of Environment, Land Water and Planning, Arthur Rylah Institute for Environmental Research, Parks Victoria and the Australian Government’s National Environmental Science Program through the Threatened Species Recovery Hub, and ARC Linkage Project LP170101134.

### 2021-07-28 manuscript/manuscript.tex
select = TRUE

@@
 
-The activity of an animal varies across space and time.
-Spatial activity is major research focus, however, temporal activity is much less studied.
-How temporal activity varies across space is even less understood.
+Fear of another species is a powerful driver of animal behaviour, but difficult to measure.
+Both threats and animal responses to perceived threats are dynamic - varying across space and time.
 
-The physical environment and presence of other species shapes animal diel activity.
-Directly and indirectly.
+Avoidance behaviour depends on need (how much of a threat is this?) and ability (how much is avoidance going to cost me?)
+For example, avoidance behaviours are less tenable in resource poor environments. Some environment more easily facilitate avoidance (e.g.~shelter availability).
+Landscape context is therefore key.
 
-Species interactions - coexistence or avoidance - are important to understand for management.
-Species can avoid other species through spatial and/or temporal avoidance.
+Subordinate animals may avoid dominant ones by altering what time of day they are active, which places they are active at, or which which places they are active at during certain times of the day.
+Space use well studied, time not so much. Space and time - extremely rarely.
 
-Classic models of diel activity overlap- flawed.
-No covariates - how does this change across space.
-Haphazard methods to faciltiate this e.g.~(xx).
+Methods to monitor spatial interactions.
+Simple correlations to JSDMs.
 
-GAMs increasingly being used for diel activity (but not necessarily species overlap).
-Particularly because you can model covariates.
-However GAMs have features which lend themselves to this analysis, but have not being used - flexible joint spatial and temporal heirarchical modelling.
+Methods to monitor temporal interactions.
+Coefficient of overlap -- arbitrary, don't allow covariates, highly susceptible to noise from small sample sizes.
 
-Interactions between invasive predators are important to understand.
+(Lack of) methods to monitor spatio-temporal behaviour.
+GAMs increasingly used for both spatial and diel activity, but not for both (despite easily facilitating this).
 
-In this study, we\ldots{}
+Species interactions are important to understand for management.
+Particularly invasive predators.
+Cat-fox context.
+
+In this study, we used GAMs to (1) assess how predators change diel activity across vegetation types, and (2) changes in both diel and spatial activity of a mesopredator in response to an apex predator.
+Used a simple predator system, as well as manipulated apex predator activity through lethal control.
 
@@
 
-We compiled camera-trap data from four distinct survey across these regions.
-After removing camera-traps which were operational for less than 14 days, this totalled 4276 deployments of camera-traps at 1527 camera-trap sites.
-Survey durations ranged between 14 and 93 days (mean 43), totalling 183,678 trapnights.
-The number of sites, deployments and trapnights were nearly equal across both regions.
+We compiled camera-trap data from three distinct studies across the two regions.
+After removing camera-traps which were operational for less than 14 days, this totalled 3659 deployments of camera-traps at 1232 camera-trap sites.
+Survey durations ranged between 14 and 93 days (mean 47), totalling 171,966 trapnights.
+Camera-trap spacing was variable; the average minimum distance between simultaneously deployed camera-traps was 445, 853 and 1266 metres for each study.
 Camera-trap were deployed in the Glenelg region between 2013 -- 2019, whereas in the Otway Ranges surveys were conducted in 2016 -- 2019.
-Camera-trap spacing was highly variable; with minimum distance between sites ranging between 40 - 4150 m (mean: 597 m).
 
@@
 We first created a table of species detections and deployment information (coordinates, dates) for every camera-trap deployment in R version 3.6.3 (R Core Team 2020).
-We discarded camera-traps which were active for less than 14 days.
-We fit lorelograms using methods provided by Iannarilli et al.
-(2019) to identify that removing repeat site detections of a species within 30 minutes was sufficient to reduce temporal autocorrelation.
-To account for different sunrise and sunset times across space and at different times of the year, we converted detection times to solar time; with sunrise and sunset times standardised at 06:00 and 18:00 respectively.
-We did so using the average double anchoring approach described by Vazquez et al.
-(2019) in the `activity' r-package (Rowcliffe 2019).
+We used lorelograms to identify the minimum interval to approximate independence (Iannarilli et al.
+2019); discarding repeat detections of a species within 30 minutes to reduce temporal autocorrelation.
+To account for day length variation across space and time, we extracted sunrise and sunset times for each camera-trap deployment using the `maptools' R-package (Bivand and Lewin-Koh 2021) and adjusted detection times to be relative to sunrise and sunset using the average double anchoring approach described by Vazquez et al.
+(2019).
 
@@
 
-Using the `reshape' r-package (Wickham 2007), we manipulated the detection table into a dataframe with a row for each hour of the day, for every camera-trap deployment, recording the total number of `independent' fox and feral cat detections within each hour for the entire camera-trap survey duration.
+Using the `reshape' R-package (Wickham 2007), we manipulated the detection table into a dataframe with a row for each hour of the day, for every camera-trap deployment, recording the total number of `independent' fox and feral cat detections within each hour for the entire camera-trap survey duration.
 
@@
 
-We modelled predator counts (response variable) with generalised additive mixed-effects models (hereafter `GAMMs') implemented in the `mgcv' r-package (Wood 2017).
-We first fit these models using the poisson family, but switched to the negative binomial family as overdispersion, but not zero-inflation, was detected using the `DHARMa' r-package (Hartig 2020).
+We modelled predator counts (response variable) with generalised additive mixed-effects models (hereafter `GAMMs') implemented in the `mgcv' R-package (Wood 2017).
+We first fit these models using the poisson family, but switched to the negative binomial family as overdispersion, but not zero-inflation, was detected using the `DHARMa' R-package (Hartig 2020).
 We specified a model offset to account for differences in camera-trap survey durations and a random intercept for each site to account for repeat sampling.
-We used a duchon spline static spatial smooth to directly model spatial autocorrelation and differences in predator activity across space (i.e.~the effect of unmodelled environmental covariates).
-For the fox model, we additionally included a thin plate regression spline for poison-bait density, with a factor smoother for each region (due to differences in bait replacement schedules).
+We used a duchon spline static spatial smooth to model both spatial autocorrelation and differences in predator activity across space (i.e.~the effect of unmodelled environmental covariates).
+For the fox model, we additionally included a thin plate regression spline for poison-bait density, with a factor smoother for each region, as bait replacement schedules differed between the regions.
 This formed the base model specification for each model we fitted.
-Models (n = 4) differed in their specification of the cyclic hour smooth in order to explore (1) variations in predator diel activity across EVC groups, as well as how (2a) feral cat diel activity differed with the presence-absence of foxes, and (2b) across the gradient of observed fox counts (detailed in the sections below).
+Models (n = 3) differed in their specification of the cyclical hour smooth in order to explore (1) variations in predator diel activity across EVC groups, as well as (2) whether feral cat diel activity changed across the gradient of observed fox counts (detailed in the sections below).
 We plotted models using `ggplot2' and `gratia' R-packages (Wickham 2016, Simpson 2021).
@@
 2019).
-This model structure penalises functions which deviate strongly from the average response--allowing confidence that they are true deviations opposed to the result of sampling noise.
-
-Because feral cats are likely subordinate to foxes, we investigated changes in their diel activity in response to both the (2a) presence-absence and (2b) counts of foxes.
-We added extra columns to our dataframe for whether foxes were detected on each camera-trap or not, as well as the number of fox detections divided by the log of the survey days to account for differences in camera-trap survey durations.
-Because we saw different fox diel activity patterns in wet forests (more active during the day) relative to the other EVC groups (nocturnal), we expected feral cats to temporally avoid foxes in opposing ways between these wet and dry vegetation groups.
+This model structure penalises functions which deviate strongly from the average response--increasing confidence that they are true deviations opposed to the result of sampling noise.
+Because feral cats are likely subordinate to foxes, we investigated changes in their diel activity in response to counts of foxes.
+We added an extra column to our dataframe with the number of fox detections divided by the log of the survey days to account for differences in camera-trap survey durations (hereafter `adjusted fox counts').
+Because we saw different fox diel activity patterns in wet forests relative to the other EVC groups, we expected feral cats to temporally avoid foxes in opposing ways between these wet and dry vegetation groups.
 We therefore modelled fox-induced changes in feral cat diel activity separately for wet and dry vegetation groups.
-Because of our large sample size, we split dry vegetation groups by region to for spatial replication.
+Because of our large sample size, we further split dry vegetation groups by region for spatial replication.
 We refer to this variable as `habitat type'; having three levels: wet forests of the Otway Ranges (`wet\_otways') dry EVC groups of the Otway Ranges (`dry\_otways') and all of the Glenelg region data (`dry\_glenelg').
-
-To fit a model which estimate feral cat diel activity where foxes were also detected, and camera-trap sites where foxes not detected (2a), we specified an cyclic hourly by variable smooth (for the presence-absence of foxes), with a factor smooth basis for habitat type.
-To fit a model which allowing feral cat diel activity to change across the range of adjusted fox counts (2b), we used a tensor product of hour and adjusted fox counts, also employing a by variable smooth to estimate this separately across each habitat type.
-We used a tensor product rather than a linear by variable for fox counts because predator interactions are unlikely to be linear (Johnson and VanDerWal 2009).
+We used a tensor product of hour (cyclic cubic regression spline) and log adjusted fox counts (thin plate regression spline) to model feral cat diel activity across the range of observed fox activity, separately for each habitat type.
 
@@
 
-\hypertarget{differences-in-diel-activity-across-vegetation-types}{%
-\subsection{Differences in diel activity across vegetation types}\label{differences-in-diel-activity-across-vegetation-types}}
+\hypertarget{predator-diel-activity-across-vegetation-types}{%
+\subsection{Predator diel activity across vegetation types}\label{predator-diel-activity-across-vegetation-types}}
+
+On average, both predator species occupied similar times of the day. Lowest during the day, peaking during sunset and sunrise, high at night.
+Foxes had a stronger diel activity patterns than cats, much less likely to be out during the day.
+Both predators had slight variations in diel activity across EVC groups, except for wet forests, were more substantial changes were seen.
+Foxes barely had any diel activity pattern here, cats more nocturnal.
+Deviance explained and Rsq values for the two models were XX.
 
@@
 \subsection{Fox-induced changes in feral cat diel activity}\label{fox-induced-changes-in-feral-cat-diel-activity}}
+
+We saw evidence of changes in cat diel activity due to fox activity in all three habitat types.
+Similar responses in dry habitats of both regions -- cat spatial activity increased with fox spatial activity, but became more diurnal.
+In Otway wet forests, no real change in spatial activity of cats due to foxes, but cats became more strongly nocturnal.\\
+Deviance explained and Rsq value was XX.
+Slightly higher uncertainty where fox spatial activity is high, this is because we had a largest sample size for fox absences, and fewest at the extreme of fox activity.
 
@@
 \includegraphics[width=1\linewidth]{../figs/predator_veg} \caption{Fox (A) and feral cat (B) diel activity patterns across Ecological Vegetation Class groups in south-west Victoria, Australia.
-Dotted, vertical lines respresent sunrise and sunset times.
+Dotted, vertical lines respresent average sunrise and sunset times.
 Shaded areas indicate 95\% confidence intervals.}\label{fig:veg}
@@
 \includegraphics[width=1\linewidth]{../figs/cat_fox_count} \caption{Predicted feral cat diel activity (A) and uncertainty (B) across each habitat type in response to foxes (log counts adjusted for survey effort).
-Grey vertical lines respresent sunrise and sunset times.
+Grey vertical lines respresent average sunrise and sunset times.
 Shaded areas indicate 95\% confidence intervals.}\label{fig:count}
@@
 \section{DISCUSSION}\label{discussion}}
+
+Foxes likely impacted the spatiotemporal behaviour of feral cats. One of the few studies to actually demonstrate and replicate this across ecosystems.
+Happens in different ways - context key.
+Spatial only, as well as temporal only models would have had opposite inference - high overlap.
+Joint spatiotemporal analysis is key to understand predator interactions.
+
+Our approach is simple and flexible.
+Small sample sizes are usually a problem. We demonstrate hierarchical model specification to share information - ability to pool all available data.
+Our approach is more aligned with null hypothesis testing - small sample sizes shrink to a flat line (or to the average), in contrast to circular overlap.
+Could be easily adapted to other contexts. E.g. instead of vegetation type, could do individual variation in diel activity.
+
+Combined spatiotemporal modelling - differences to other approaches.
+Cunningham paper most similar to our approach - different in diel activity only, and linear responses.
+We used a tensor product for fox counts because predator interactions may not be linear (Johnson and VanDerWal 2009).
+
+Limitations of our approach.
+Detectability ignored.
+Individual heterogeneity ignored.
+Correlation does not = causation.
+Does not (clearly) seperate numerical responses from behavioural (Borchers SCR paper)
+
+Invasive predators are useful species to test this on, as they are extremely adaptable to different conditions.
+Implications for prey.
+
+Conclusions.
 
@@
 We are extremely thankful to those who led and assisted with these camera-trap deployments and image processing.
-Univeristy of Melbounre led surveys were conducted under University of Melbourne Animal Ethics Committee approval 1714119 and Victorian Government Department of Environment, Land Water and Planning Research Permit 10008273.
+University of Melbourne led surveys were conducted under University of Melbourne Animal Ethics Committee approval 1714119 and Victorian Government Department of Environment, Land Water and Planning Research Permit 10008273.
 Our study was generously supported by the Conservation Ecology Centre, the Victorian Government Department of Environment, Land Water and Planning, Arthur Rylah Institute for Environmental Research, Parks Victoria and the Australian Government's National Environmental Science Program through the Threatened Species Recovery Hub, and ARC Linkage Project LP170101134.
@@
 \hypertarget{refs}{}
+\leavevmode\hypertarget{ref-maptools}{}%
+Bivand, R. and Lewin-Koh, N. 2021. Maptools: Tools for handling spatial objects.
+
 \leavevmode\hypertarget{ref-delwp2020}{}%
@@
 
-\leavevmode\hypertarget{ref-rowcliffe_activity}{}%
-Rowcliffe, M. 2019. Activity: Animal activity statistics.
-
 \leavevmode\hypertarget{ref-gratia}{}%

### 2021-07-28 r_scripts/5_gams_diel_activity.R

@@ -98,7 +98,7 @@ gam_fox <- bam(fox ~ s(hour, bs = "cc", k = 8) +
                      s(longitude, latitude, bs = "ds",  m = c(1, 0.5), k = 200) +
                      s(station, bs = "re") +  
                      offset(log(survey_duration)), 
-                    data = records, family = nb, knots = list(hour = c(0, 23)), nthreads = 3, discrete = TRUE)
+                    data = records, family = nb, knots = list(hour = c(0, 23)), nthreads = 3, discrete = TRUE, select = TRUE)
 summary(gam_fox)
 plot(gam_fox, pages = 1, scheme = 2, seWithMean = TRUE, shade = TRUE, scale = 0)
 
@@ -109,7 +109,7 @@ gam_cat_1 <- bam(cat ~ s(hour, bs = "cc", k = 8) +
                        s(longitude, latitude, bs = "ds",  m = c(1, 0.5), k = 200) +
                        s(station, bs = "re") +  
                        offset(log(survey_duration)), 
-                 data = records, family = nb, knots = list(hour = c(0, 23)), nthreads = 3, discrete = TRUE)
+                 data = records, family = nb, knots = list(hour = c(0, 23)), nthreads = 3, discrete = TRUE, select = TRUE)
 summary(gam_cat_1)
 plot(gam_cat_1, pages = 1, scheme = 2, seWithMean = TRUE, shade = TRUE, scale = 0)
 
@@ -120,7 +120,7 @@ gam_cat_2a <- bam(cat ~ fox_pa + s(hour, habitat_type, by = fox_pa, bs = "fs", x
                         s(longitude, latitude, bs = "ds",  m = c(1, 0.5), k = 200) +
                         s(station, bs = "re") +  
                         offset(log(survey_duration)), 
-                  data = records, family = nb, knots = list(hour = c(0, 23)), nthreads = 3, discrete = TRUE)
+                  data = records, family = nb, knots = list(hour = c(0, 23)), nthreads = 3, discrete = TRUE, select = TRUE)
 summary(gam_cat_2a)
 plot(gam_cat_2a, pages = 1, scheme = 2, seWithMean = TRUE, shade = TRUE, scale = 0)
 
@@ -129,10 +129,9 @@ gam_cat_2b <- bam(cat ~ habitat_type + t2(hour, fox_count_adj, by = habitat_type
[…]

### 2021-07-28 manuscript/manuscript.Rmd
removed naive occ from table

@@
          fox_present = sum(fox_pa == "present"),
-         cat_present = sum(cat_pa == "present"),
-         fox_naive = (fox_present / deployments),
-         cat_naive = (cat_present / deployments)
+         cat_present = sum(cat_pa == "present")
          )
@@
 table1 <- table1 %>%
-  select(region, vegetation_group, sites, deployments, fox_count_vr, cat_count_vr, fox_naive, cat_naive)  %>% 
+  select(region, vegetation_group, sites, deployments, fox_count_vr, cat_count_vr)  %>% 
   distinct() %>%
@@
 pander(table1, style = "simple", split.table = Inf, caption = "Summary of camera-trap deployments in unique sites within Ecological Vegetation Class groups, as well as independent counts of predator detections and naive occupancy rates, across two broad regions in south-west Victoria, Australia.", round = 2,
-              col.names = c("Region", "Vegetation", "Sites", "Deployments",  "Fox counts", "Cat counts", "Fox present", "Cat present"),
+              col.names = c("Region", "Vegetation", "Sites", "Deployments",  "Fox counts", "Cat counts"),
 )

### 2021-07-28 manuscript/manuscript.tex
removed naive occ from table

@@
 
-\begin{longtable}[]{@{}cccccccc@{}}
+\begin{longtable}[]{@{}cccccc@{}}
 \caption{Summary of camera-trap deployments in unique sites within Ecological Vegetation Class groups, as well as independent counts of predator detections and naive occupancy rates, across two broad regions in south-west Victoria, Australia.}\tabularnewline
 \toprule
-Region & Vegetation & Sites & Deployments & Fox counts & Cat counts & Fox present & Cat present\tabularnewline
+Region & Vegetation & Sites & Deployments & Fox counts & Cat counts\tabularnewline
 \midrule
@@
 \toprule
-Region & Vegetation & Sites & Deployments & Fox counts & Cat counts & Fox present & Cat present\tabularnewline
+Region & Vegetation & Sites & Deployments & Fox counts & Cat counts\tabularnewline
 \midrule
 \endhead
-Glenelg & Dry Forest & 25 & 68 & 347 & 9 & 0.66 & 0.1\tabularnewline
-Otway & Dry Forest & 111 & 310 & 341 & 158 & 0.42 & 0.28\tabularnewline
-Glenelg & Heathland & 40 & 119 & 265 & 59 & 0.43 & 0.34\tabularnewline
-Otway & Heathland & 3 & 9 & 8 & 6 & 0.22 & 0.44\tabularnewline
-Glenelg & Heathy Woodland & 154 & 424 & 574 & 96 & 0.31 & 0.14\tabularnewline
-Otway & Heathy Woodland & 82 & 256 & 160 & 66 & 0.29 & 0.14\tabularnewline
-Glenelg & Herb-rich Woodland & 59 & 372 & 860 & 197 & 0.5 & 0.27\tabularnewline
-Otway & Herb-rich Woodland & 2 & 6 & 3 & 2 & 0.33 & 0.17\tabularnewline
-Glenelg & Lowland Forest & 383 & 1046 & 1900 & 290 & 0.46 & 0.18\tabularnewline
-Otway & Lowland Forest & 52 & 162 & 190 & 35 & 0.42 & 0.14\tabularnewline
-Glenelg & Swampy Scrub & 4 & 10 & 19 & 8 & 0.6 & 0.5\tabularnewline
-Otway & Swampy Scrub & 36 & 97 & 64 & 88 & 0.32 & 0.33\tabularnewline
-Otway & Wet Forest & 281 & 780 & 715 & 1187 & 0.31 & 0.54\tabularnewline
+Glenelg & Dry Forest & 25 & 68 & 347 & 9\tabularnewline
+Otway & Dry Forest & 111 & 310 & 341 & 158\tabularnewline
+Glenelg & Heathland & 40 & 119 & 265 & 59\tabularnewline
+Otway & Heathland & 3 & 9 & 8 & 6\tabularnewline
+Glenelg & Heathy Woodland & 154 & 424 & 574 & 96\tabularnewline
+Otway & Heathy Woodland & 82 & 256 & 160 & 66\tabularnewline
+Glenelg & Herb-rich Woodland & 59 & 372 & 860 & 197\tabularnewline
+Otway & Herb-rich Woodland & 2 & 6 & 3 & 2\tabularnewline
+Glenelg & Lowland Forest & 383 & 1046 & 1900 & 290\tabularnewline
+Otway & Lowland Forest & 52 & 162 & 190 & 35\tabularnewline
+Glenelg & Swampy Scrub & 4 & 10 & 19 & 8\tabularnewline
+Otway & Swampy Scrub & 36 & 97 & 64 & 88\tabularnewline
+Otway & Wet Forest & 281 & 780 & 715 & 1187\tabularnewline
 \bottomrule

### 2021-08-03 manuscript/manuscript.Rmd
ms update

@@
 ---
-title: Dynamic shifts in predator diel activity due to habitat type and fear
+title: Joint spatiotemporal models reveal dynamic shifts in mesopredator spatial and diel activity for apex predator avoidance
 author:
@@
 
-Fear of another species is a powerful driver of animal behaviour, but difficult to measure.
-Both threats and animal responses to perceived threats are dynamic - varying across space and time.
+Effective wildlife management relies on understanding species population and behavioural dynamics to accurately predict intervention outcomes.
+These dynamics depend on both environmental conditions and those of other species, but few statistical approaches incorporate both factors, or allow responses to be dynamic across space and time.
+This is key because species interactions and niche preferences are multidimensional.
 
-Avoidance behaviour depends on need (how much of a threat is this?) and ability (how much is avoidance going to cost me?)
-For example, avoidance behaviours are less tenable in resource poor environments. Some environment more easily facilitate avoidance (e.g. shelter availability). 
-Landscape context is therefore key.
+Fear is a near universal driver of animal behaviour.
+Fear may be from other individuals of their species, a dominant species (including humans) or an event (e.g.
+fire).
+Animals may be naive to threats (e.g.
+alien predators), unable to respond to threats (e.g.
+scarce prey), or the threat may be too minimal to warrant a response.
+Some environments more easily facilitate avoidance (e.g.
+high food and shelter accessibility), and so accounting for landscape context is key.
+A threat response may be routine, or only employed or enhanced when the threat is heightened.
+This likely depends on how constant and predictable the threat is.
 
-Subordinate animals may avoid dominant ones by altering what time of day they are active, which places they are active at, or which which places they are active at during certain times of the day. 
-Space use well studied, time not so much. Space and time - extremely rarely.
+Animals may respond to threats by altering their activity patterns (i.e.
+avoidance).
+An animal may avoid locations where the threat is likely to occur (hereafter ‘spatial avoidance’), avoid times of the day when a threat is likely to occur (hereafter ‘temporal avoidance’), or avoid both locations and times a threat is heightened (hereafter ‘spatiotemporal avoidance’).
+Such avoidance likely has trade-offs, which may or may not incur costs to individual and population fitness.
+Modelling approaches which do not allow spatiotemporal and dynamic fear reponses can only make limited inference about species interactions.
 
-Methods to monitor spatial interactions. 
-Simple correlations to JSDMs.
+Species interactions are usually considered in degrees of overlap between activity patterns.
+Spatiotemporal avoidance is most often considered, using joint species distribution models (including MSOMs).
+However, temporal avoidance is increasingly being used (perhaps just allowed due to widely available GPS and camera technology) with kernel density estimation methods.
+Although, these methods don’t allow environmental covariates, and are particularly susceptible to noisy estimates.
+Importantly, some are beginning to combine these methods to allow spatiotemporal responses (Azzou et al 2021; Farris et al 2020).
+However, degrees of overlap in activity patterns only get at routine avoidance, we are generally interested in whether changes in threat levels lead to altered behaviour.
+This is more indicative of species interactions because static levels of overlap may just reflect different niche preferences rather than avoidance.
 
-Methods to monitor temporal interactions. 
-Coefficient of overlap -- arbitrary, don't allow covariates, highly susceptible to noise from small sample sizes.
+Dynamic behavioural responses are also important from a management perspective because they want to know whether an action will cause a change in animal behaviour /  range?).
+For example, after the reintroduction of wolves in Yellowstone, it was widely thought that this had limited elk populations.
+Using GPS data and habitat selection methods,  Kohl et al.
+2018 found that elk used riskier parts of landscape at times when wolves were least active -- suggesting coexistence without fitness costs.
+However, there are many other large carnivores in this ecosystem which are likely to impact elk behaviour.
+And so, will elk respond to wolves in the same way in different environmental contexts?
+There are little-no examples of replicated spatiotemporal interactions studies.
 
-(Lack of) methods to monitor spatio-temporal behaviour. 
-GAMs increasingly used for both spatial and diel activity, but not for both (despite easily facilitating this).
+There is contention around the interactions between two invasive predators - feral cats and foxes.
+Foxes are routinely culled, and there is concern that this could lead to a mesopredator release of feral cats.
+Mesopredator release may manifest in both behavioural and numerical changes, which my previous chapter identified - but that behavioral responses (ranging and detectability) were not universal.
+Here, we aim to dig into the behavioral response further - in particular diel activity times.
 
-Species interactions are important to understand for management. 
-Particularly invasive predators. 
-Cat-fox context.
-
-In this study, we used GAMs to (1) assess how predators change diel activity across vegetation types, and (2) changes in both diel and spatial activity of a mesopredator in response to an apex predator.
-Used a simple predator system, as well as manipulated apex predator activity through lethal control.
+In this study, we used generalised additive models to assess (1) how predator (fox and cat) diel activity varies between vegetation types, and (2) whether an introduced mesopredator (cat) changes its diel and/or spatial activity in response to the activity of an introduced apex predator (fox).
+Because foxes and cats fill a similar function niche, we expected that, on average, they would have similar diel activity patterns across vegetation types (hypothesis 1).
+However, where fox detections were high, we hypothesised that the subordinate cat to be less active (hypothesis 2) and shift diel activity to times when foxes are least active (hypothesis 3).
+We tested this on a simple predator system where only cats and foxes occur, and fox activity is manipulated through lethal control.
+This allowed focus on the sole interactions between these two predators, and ensured differences in correlations wasn’t just due to differences in niche preferences.
 
@@
 
-We collated camera-trap data from two distinct regions in south-west Victoria, Australia (Fig.
+We collated camera-trap data on these two predator species from three distinct studies across two regions of south-west Victoria, Australia (Fig.
 1).
-The terrestrial mammalian predator guild is depauperate here.
+The purpose of each study was to experimentally survey changes in the mammalian community due to fox control, and so our dataset covered a wide range contexts with and without fox-baiting, allowing us to investigate fox-cat interactions across a gradient of apex predator (fox) activity.
+
+The terrestrial mammalian predator guild is depauperate throughout both regions.
 Dingoes *Canis familiaris* are now absent throughout, as is the the tiger quoll *Dasyurus maculatus*, more recently absent in the Otway Ranges (last sighted in 2014 despite extensive camera-trapping).
-Introduced red foxes and feral cat are therefore the only medium-large functional mammalian terrestrial predators across both regions.
-Across broad sections of each region, government land managers conduct ongoing targeted fox control for biodiversity conservation - poison baits containing 3 mg of sodium mono-fluroacetate (1080) are buried at a depth of 10 cm at 1-km intervals along accessible forest tracks and roads.
-The purpose of each  collated camera-trap study was to to experimentally survey changes in the mammalian community due to fox control, and so, our study covers a wide range contexts with and without fox-baiting.
+Introduced red foxes and feral cats are therefore the only medium-large functional mammalian terrestrial predators across both regions.
 
-In the Glenelg region, Gunditjmara country, natural vegetation has been highly fragmented.
-Here, camera-survey sites were situated in six distinct forest blocks; two of which separated by a river, others separated by mostly pastoral farming and residential property).
-Foxes in three forests blocks have been subject to lethal control since October 2005, with 1080 poison-baits replaced fortnightly [@ROBLEY2014262].
+In the baited parts of the regions, government land managers conduct ongoing targeted fox control for biodiversity conservation.
+Poison baits containing 3 mg of sodium mono-fluroacetate (1080) are buried at a depth of 10 cm at 1-km intervals along accessible forest tracks and roads.
 
-The Otway Ranges, Gadubanud and XX country, is a largely continuous patch of natural vegetation with a strong east-west rainfall and elevation gradient [@swan2021species].
-Foxes are controlled through 1080 poison-baits with monthly replacement across most of the Otway Ranges (a large section to the north-west remains unbaited).
+In the Glenelg region, Gunditjmara country, natural vegetation is highly fragmented.
+Here, camera-survey sites were situated in six distinct forest blocks: two are separated by a river, the others are separated by at least *XX* km of pastoral farming and residential property.
+Foxes in three of the forest blocks have been subject to lethal control since October 2005, with 1080 poison-baits replaced fortnightly [@ROBLEY2014262].
+
+The Otway Ranges, Gadubanud country, is a largely continuous patch of natural vegetation with a strong east-west rainfall and elevation gradient [@swan2021species].
+Foxes are controlled through 1080 poison-baits with monthly replacement across most of the Otway Ranges, but a large section to the north-west remains unbaited.
 The majority of this baiting began in 2016 - 2017, although fox-baiting commenced in some small sections in 2008.
@@
 
-## Camera-trapping surveys
+## Camera-trap surveys
 
@@
 The remaining camera-traps were positioned higher on a tree or a metal picket (at least 40 cm above ground) and angled downwards more strongly - facing a lure approximately 1 - 1.5 m away.
-These lures consisted of peanut butter, golden syrup and rolled oats were mixed into a small ball, placed within a tea strainer or PVC pipe container and secured either to the ground, or 20 - 60 cm above ground on a wooden stake.
-All set-ups were effective in detecting both predator and prey species.
+These lures consisted of peanut butter, golden syrup and rolled oats mixed into a small ball, placed within a tea strainer or PVC pipe container and secured either to the ground, or 20 - 60 cm above ground on a wooden stake.
+All set-ups were effective in detecting both predator species.
 
 ## Data preparation
-We first created a table of species detections and deployment information (coordinates, dates) for every camera-trap deployment in R version 3.6.3 [@R].
-We used lorelograms to identify the minimum interval to approximate independence [@https://doi.org/10.1111/2041-210X.13308]; discarding repeat detections of a species within 30 minutes to reduce temporal autocorrelation.
+We first created a table of species detections and deployment information (coordinates, dates) for every camera-trap deployment.
+All data analysis was conducted in R version 3.6.3 [@R].
+We used lorelograms to identify the minimum interval to approximate independence [@https://doi.org/10.1111/2041-210X.13308]; this indicated 30 minutes discarding repeat detections of a species within 30 minutes was sufficient in reducing temporal autocorrelation.
 To account for day length variation across space and time, we extracted sunrise and sunset times for each camera-trap deployment using the 'maptools' R-package [@maptools] and adjusted detection times to be relative to sunrise and sunset using the average double anchoring approach described by @https://doi.org/10.1111/2041-210X.13290.
-
-We attached Ecological Vegetation Class (hereafter 'EVC') groups [@delwp2020] to each camera-trap location.
-Rainforests are interspersed (primarily in low lying gullies) throughout wet and damp forests at a finescale (relevant to predator movements), and so, we merged these groupings (hereafter 'wet forests').
-For each deployment, we calculated the number of poison fox-baits within a 2.3 km radius around each camera-trap - the average maximum distance foxes in these region travel from their home range centre [@hradsky2017human].
 
@@
 
-## Generalised additive models of diel activity 
+We identified the broad vegetation type (Ecological Vegetation Class (hereafter 'EVC') group [@delwp2020]) at each camera-trap location.
+Six vegetation types were surveyed in both regions, although to varying degrees (Table 1).
+In the Otway region, rainforests are interspersed (primarily in low lying gullies) throughout wet and damp forests at a finescale, and we merged these groups (hereafter 'wet forests').
+
+For each deployment, we calculated the number of poison fox-baits within a 2.3 km radius around each camera-trap - the average maximum distance foxes in these regions travel from their home range centre [@hradsky2017human].
+
+## Generalised additive models 
 We modelled predator counts (response variable) with generalised additive mixed-effects models (hereafter 'GAMMs') implemented in the 'mgcv' R-package [@wood2017].
 We first fit these models using the poisson family, but switched to the negative binomial family as overdispersion, but not zero-inflation, was detected using the 'DHARMa' R-package [@DHARMa].
-We specified a model offset to account for differences in camera-trap survey durations and a random intercept for each site to account for repeat sampling.
+We specified a model offset to account for differences in camera-trap survey duration, and a random intercept for each site to account for repeat sampling.
 We used a duchon spline static spatial smooth to model both spatial autocorrelation and differences in predator activity across space (i.e.
@@
 This formed the base model specification for each model we fitted.
-Models (n = 3) differed in their specification of the cyclical hour smooth in order to explore (1) variations in predator diel activity across EVC groups, as well as (2) whether feral cat diel activity changed across the gradient of observed fox counts (detailed in the sections below).
+Models (n = 2) differed in their specification of the cyclical hour smooth in order to explore (1) variations in predator diel activity across EVC groups, as well as (2) whether feral cat diel activity changed across the gradient of observed fox counts (detailed in the sections below).
+We used the double penalty model selection approach, which penalises the null space in addition to the range space (i.e.
+shrinking wiggly terms to linear terms) of the spline basis, meaning covariate effects can be entirely removed from the model [@MARRA20112372].
 We plotted models using 'ggplot2' and 'gratia' R-packages [@ggplot2; @gratia].
 
-We expected predator activity to be lowest during the day, but potentially having small shifts in relative activity between sunset and sunrise across the different EVC groups (1).
-And so, fitting separate responses for each EVC group would throw away this shared information, which is particularly costly for the least surveyed groups -- resulting in more noisy estimates.
+We expected the activity of each predator species to follow a broadly consistent pattern (e.g.
+lower during the day), but potentially have small shifts in relative activity between sunset and sunrise across the different EVC groups (1).
+Fitting separate responses for each EVC group would throw away shared information, which is particularly costly for the least surveyed groups, and would result in more noisy estimates.
 We therefore specified a global smoother for hour (i.e.
@@
 model GS in @pedersen2019hierarchical].
-This model structure penalises functions which deviate strongly from the average response--increasing confidence that they are true deviations opposed to the result of sampling noise.
-Because feral cats are likely subordinate to foxes, we investigated changes in their diel activity in response to counts of foxes.
-We added an extra column to our dataframe with the number of fox detections divided by the log of the survey days to account for differences in camera-trap survey durations (hereafter 'adjusted fox counts').
-Because we saw different fox diel activity patterns in wet forests relative to the other EVC groups, we expected feral cats to temporally avoid foxes in opposing ways between these wet and dry vegetation groups.
+This model structure penalises functions which deviate strongly from the average response, increasing confidence that they are true deviations opposed to the result of sampling noise.
+
+To investigate changes in feral cat  diel activity in response to counts of foxes.
+We added an extra column to our dataframe with the number of fox detections divided by the log of the number of survey days to account for differences in camera-trap survey durations (hereafter 'adjusted fox counts').
+
+Analysis of stage 1 showed that fox diel activity patterns in wet forests differed relative to all other EVC groups.
+We therefore expected feral cats might to temporally avoid foxes in a different way in wet forests than the dry vegetation groups, but that responses would be similar in all the drier vegetation groups.
 We therefore modelled fox-induced changes in feral cat diel activity separately for wet and dry vegetation groups.
-Because of our large sample size, we further split dry vegetation groups by region for spatial replication.
-We refer to this variable as 'habitat type'; having three levels: wet forests of the Otway Ranges ('wet_otways') dry EVC groups of the Otway Ranges ('dry_otways') and all of the Glenelg region data ('dry_glenelg').
-We used a tensor product of hour (cyclic cubic regression spline) and log adjusted fox counts (thin plate regression spline) to model feral cat diel activity across the range of observed fox activity, separately for each habitat type.
+We further split dry vegetation groups by region, to examine whether patterns remained consistent across space?
+for spatial replication.
+We refer to the resulting variable as 'habitat type', which had three levels: wet forests of the Otway Ranges ('wet_otways') dry EVC groups of the Otway Ranges ('dry_otways') and all of the Glenelg region data ('dry_glenelg').
+
+We used a tensor product of hour (using cyclic cubic regression spline) and log adjusted fox count (using thin plate regression spline) smooths to model feral cat diel activity across the range of observed fox activity, separately for each habitat type.
+By also having double penalty shrinkage turned on, this meant that feral cat activity could vary nonlinearly and be penalised consistently across both dimensions.
+This model specification allows five different scenarios: that there was (1) no effect of hour or foxes on feral cats (2) a static hourly effect only, (3) a spatial response to foxes only, (4) a spatial response to foxes and an unrelated static hourly effect, (5) a spatial response to foxes and a hourly effect which changes across the range of fox counts.
 
@@
 
-On average, both predator species occupied similar times of the day. Lowest during the day, peaking during sunset and sunrise, high at night. 
+On average, both predator species occupied similar times of the day--mostly nocturnal with activity peaks around sunrise and sunset.
+The main difference was that fox activity peaked just after sunset rather than on sunset.
+For foxes, this diel pattern was largely exhibited across all vegetation types except wet forest.
+In wet forests and rainforests, the hour of the day had little effect on foxes (especially considering diel activity curve here is being drawn towards a nocturnal pattern).
+
+lowest activity during the middle of the day, peaking at sunrise and sunset times, while remaining relatively high at night and dawn. 
 Foxes had a stronger diel activity patterns than cats, much less likely to be out during the day.
-Both predators had slight variations in diel activity across EVC groups, except for wet forests, were more substantial changes were seen. 
-Foxes barely had any diel activity pattern here, cats more nocturnal.
+
+Diel activity patterns were similar across the different vegetation types for each predator,, except for wet forests, where more substantial changes were seen.
+In the wet forests, ox activity remained fairly constant throughout day and night barely had any diel activity pattern here, cats more nocturnal.
 Deviance explained and Rsq values for the two models were XX.
@@
 # make table
-pander(table1, style = "simple", split.table = Inf, caption = "Summary of camera-trap deployments in unique sites within Ecological Vegetation Class groups, as well as independent counts of predator detections and naive occupancy rates, across two broad regions in south-west Victoria, Australia.", round = 2,
+pander(table1, style = "simple", split.table = Inf, caption = "Summary of camera-trap deployments in unique sites within Ecological Vegetation Class groups and independent counts of predator detections across two broad regions in south-west Victoria, Australia.", round = 2,
               col.names = c("Region", "Vegetation", "Sites", "Deployments",  "Fox counts", "Cat counts"),
@@
 ```{r veg, out.width="100%", fig.cap = "Fox (A) and feral cat (B) diel activity patterns across Ecological Vegetation Class groups in south-west Victoria, Australia.
-Dotted, vertical lines respresent average sunrise and sunset times.
+Dotted, vertical lines represent average sunrise and sunset times.
 Shaded areas indicate 95\\% confidence intervals."}

### 2021-08-03 manuscript/manuscript.tex
ms update

@@
             pdfauthor={},
-            pdftitle={Dynamic shifts in predator diel activity due to habitat type and fear},
+            pdftitle={Joint spatiotemporal models reveal dynamic shifts in mesopredator spatial and diel activity for apex predator avoidance},
             colorlinks=false,
@@
 
-  \title{Dynamic shifts in predator diel activity due to habitat type and fear}
+  \title{Joint spatiotemporal models reveal dynamic shifts in mesopredator spatial and diel activity for apex predator avoidance}
     \author[UOM]{Matthew W. Rees\corref{1}}
@@
 
-Fear of another species is a powerful driver of animal behaviour, but difficult to measure.
-Both threats and animal responses to perceived threats are dynamic - varying across space and time.
+Effective wildlife management relies on understanding species population and behavioural dynamics to accurately predict intervention outcomes.
+These dynamics depend on both environmental conditions and those of other species, but few statistical approaches incorporate both factors, or allow responses to be dynamic across space and time.
+This is key because species interactions and niche preferences are multidimensional.
 
-Avoidance behaviour depends on need (how much of a threat is this?) and ability (how much is avoidance going to cost me?)
-For example, avoidance behaviours are less tenable in resource poor environments. Some environment more easily facilitate avoidance (e.g.~shelter availability).
-Landscape context is therefore key.
+Fear is a near universal driver of animal behaviour.
+Fear may be from other individuals of their species, a dominant species (including humans) or an event (e.g.~fire).
+Animals may be naive to threats (e.g.~alien predators), unable to respond to threats (e.g.~scarce prey), or the threat may be too minimal to warrant a response.
+Some environments more easily facilitate avoidance (e.g.~high food and shelter accessibility), and so accounting for landscape context is key.
+A threat response may be routine, or only employed or enhanced when the threat is heightened.
+This likely depends on how constant and predictable the threat is.
 
-Subordinate animals may avoid dominant ones by altering what time of day they are active, which places they are active at, or which which places they are active at during certain times of the day.
-Space use well studied, time not so much. Space and time - extremely rarely.
+Animals may respond to threats by altering their activity patterns (i.e.~avoidance).
+An animal may avoid locations where the threat is likely to occur (hereafter `spatial avoidance'), avoid times of the day when a threat is likely to occur (hereafter `temporal avoidance'), or avoid both locations and times a threat is heightened (hereafter `spatiotemporal avoidance').
+Such avoidance likely has trade-offs, which may or may not incur costs to individual and population fitness.
+Modelling approaches which do not allow spatiotemporal and dynamic fear reponses can only make limited inference about species interactions.
 
-Methods to monitor spatial interactions.
-Simple correlations to JSDMs.
+Species interactions are usually considered in degrees of overlap between activity patterns.
+Spatiotemporal avoidance is most often considered, using joint species distribution models (including MSOMs).
+However, temporal avoidance is increasingly being used (perhaps just allowed due to widely available GPS and camera technology) with kernel density estimation methods.
+Although, these methods don't allow environmental covariates, and are particularly susceptible to noisy estimates.
+Importantly, some are beginning to combine these methods to allow spatiotemporal responses (Azzou et al 2021; Farris et al 2020).
+However, degrees of overlap in activity patterns only get at routine avoidance, we are generally interested in whether changes in threat levels lead to altered behaviour.
+This is more indicative of species interactions because static levels of overlap may just reflect different niche preferences rather than avoidance.
 
-Methods to monitor temporal interactions.
-Coefficient of overlap -- arbitrary, don't allow covariates, highly susceptible to noise from small sample sizes.
+Dynamic behavioural responses are also important from a management perspective because they want to know whether an action will cause a change in animal behaviour / range?).
+For example, after the reintroduction of wolves in Yellowstone, it was widely thought that this had limited elk populations.
+Using GPS data and habitat selection methods, Kohl et al.~2018 found that elk used riskier parts of landscape at times when wolves were least active -- suggesting coexistence without fitness costs.
+However, there are many other large carnivores in this ecosystem which are likely to impact elk behaviour.
+And so, will elk respond to wolves in the same way in different environmental contexts?
+There are little-no examples of replicated spatiotemporal interactions studies.
 
-(Lack of) methods to monitor spatio-temporal behaviour.
-GAMs increasingly used for both spatial and diel activity, but not for both (despite easily facilitating this).
+There is contention around the interactions between two invasive predators - feral cats and foxes.
+Foxes are routinely culled, and there is concern that this could lead to a mesopredator release of feral cats.
+Mesopredator release may manifest in both behavioural and numerical changes, which my previous chapter identified - but that behavioral responses (ranging and detectability) were not universal.
+Here, we aim to dig into the behavioral response further - in particular diel activity times.
 
-Species interactions are important to understand for management.
-Particularly invasive predators.
-Cat-fox context.
-
-In this study, we used GAMs to (1) assess how predators change diel activity across vegetation types, and (2) changes in both diel and spatial activity of a mesopredator in response to an apex predator.
-Used a simple predator system, as well as manipulated apex predator activity through lethal control.
+In this study, we used generalised additive models to assess (1) how predator (fox and cat) diel activity varies between vegetation types, and (2) whether an introduced mesopredator (cat) changes its diel and/or spatial activity in response to the activity of an introduced apex predator (fox).
+Because foxes and cats fill a similar function niche, we expected that, on average, they would have similar diel activity patterns across vegetation types (hypothesis 1).
+However, where fox detections were high, we hypothesised that the subordinate cat to be less active (hypothesis 2) and shift diel activity to times when foxes are least active (hypothesis 3).
+We tested this on a simple predator system where only cats and foxes occur, and fox activity is manipulated through lethal control.
+This allowed focus on the sole interactions between these two predators, and ensured differences in correlations wasn't just due to differences in niche preferences.
 
@@
 
-We collated camera-trap data from two distinct regions in south-west Victoria, Australia (Fig.
+We collated camera-trap data on these two predator species from three distinct studies across two regions of south-west Victoria, Australia (Fig.
 1).
-The terrestrial mammalian predator guild is depauperate here.
+The purpose of each study was to experimentally survey changes in the mammalian community due to fox control, and so our dataset covered a wide range contexts with and without fox-baiting, allowing us to investigate fox-cat interactions across a gradient of apex predator (fox) activity.
+
+The terrestrial mammalian predator guild is depauperate throughout both regions.
 Dingoes \emph{Canis familiaris} are now absent throughout, as is the the tiger quoll \emph{Dasyurus maculatus}, more recently absent in the Otway Ranges (last sighted in 2014 despite extensive camera-trapping).
-Introduced red foxes and feral cat are therefore the only medium-large functional mammalian terrestrial predators across both regions.
-Across broad sections of each region, government land managers conduct ongoing targeted fox control for biodiversity conservation - poison baits containing 3 mg of sodium mono-fluroacetate (1080) are buried at a depth of 10 cm at 1-km intervals along accessible forest tracks and roads.
-The purpose of each collated camera-trap study was to to experimentally survey changes in the mammalian community due to fox control, and so, our study covers a wide range contexts with and without fox-baiting.
+Introduced red foxes and feral cats are therefore the only medium-large functional mammalian terrestrial predators across both regions.
 
-In the Glenelg region, Gunditjmara country, natural vegetation has been highly fragmented.
-Here, camera-survey sites were situated in six distinct forest blocks; two of which separated by a river, others separated by mostly pastoral farming and residential property).
-Foxes in three forests blocks have been subject to lethal control since October 2005, with 1080 poison-baits replaced fortnightly (Robley et al.
+In the baited parts of the regions, government land managers conduct ongoing targeted fox control for biodiversity conservation.
+Poison baits containing 3 mg of sodium mono-fluroacetate (1080) are buried at a depth of 10 cm at 1-km intervals along accessible forest tracks and roads.
+
+In the Glenelg region, Gunditjmara country, natural vegetation is highly fragmented.
+Here, camera-survey sites were situated in six distinct forest blocks: two are separated by a river, the others are separated by at least \emph{XX} km of pastoral farming and residential property.
+Foxes in three of the forest blocks have been subject to lethal control since October 2005, with 1080 poison-baits replaced fortnightly (Robley et al.
 2014).
 
-The Otway Ranges, Gadubanud and XX country, is a largely continuous patch of natural vegetation with a strong east-west rainfall and elevation gradient (Swan et al.
+The Otway Ranges, Gadubanud country, is a largely continuous patch of natural vegetation with a strong east-west rainfall and elevation gradient (Swan et al.
 2021).
-Foxes are controlled through 1080 poison-baits with monthly replacement across most of the Otway Ranges (a large section to the north-west remains unbaited).
+Foxes are controlled through 1080 poison-baits with monthly replacement across most of the Otway Ranges, but a large section to the north-west remains unbaited.
 The majority of this baiting began in 2016 - 2017, although fox-baiting commenced in some small sections in 2008.
@@
 
-\hypertarget{camera-trapping-surveys}{%
-\subsection{Camera-trapping surveys}\label{camera-trapping-surveys}}
+\hypertarget{camera-trap-surveys}{%
+\subsection{Camera-trap surveys}\label{camera-trap-surveys}}
 
@@
 The remaining camera-traps were positioned higher on a tree or a metal picket (at least 40 cm above ground) and angled downwards more strongly - facing a lure approximately 1 - 1.5 m away.
-These lures consisted of peanut butter, golden syrup and rolled oats were mixed into a small ball, placed within a tea strainer or PVC pipe container and secured either to the ground, or 20 - 60 cm above ground on a wooden stake.
-All set-ups were effective in detecting both predator and prey species.
+These lures consisted of peanut butter, golden syrup and rolled oats mixed into a small ball, placed within a tea strainer or PVC pipe container and secured either to the ground, or 20 - 60 cm above ground on a wooden stake.
+All set-ups were effective in detecting both predator species.
 
@@
 
-We first created a table of species detections and deployment information (coordinates, dates) for every camera-trap deployment in R version 3.6.3 (R Core Team 2020).
+We first created a table of species detections and deployment information (coordinates, dates) for every camera-trap deployment.
+All data analysis was conducted in R version 3.6.3 (R Core Team 2020).
 We used lorelograms to identify the minimum interval to approximate independence (Iannarilli et al.
-2019); discarding repeat detections of a species within 30 minutes to reduce temporal autocorrelation.
+2019); this indicated 30 minutes discarding repeat detections of a species within 30 minutes was sufficient in reducing temporal autocorrelation.
 To account for day length variation across space and time, we extracted sunrise and sunset times for each camera-trap deployment using the `maptools' R-package (Bivand and Lewin-Koh 2021) and adjusted detection times to be relative to sunrise and sunset using the average double anchoring approach described by Vazquez et al.
@@
 
-We attached Ecological Vegetation Class (hereafter `EVC') groups (DELWP 2020) to each camera-trap location.
-Rainforests are interspersed (primarily in low lying gullies) throughout wet and damp forests at a finescale (relevant to predator movements), and so, we merged these groupings (hereafter `wet forests').
-For each deployment, we calculated the number of poison fox-baits within a 2.3 km radius around each camera-trap - the average maximum distance foxes in these region travel from their home range centre (Hradsky et al.
+Using the `reshape' R-package (Wickham 2007), we manipulated the detection table into a dataframe with a row for each hour of the day, for every camera-trap deployment, recording the total number of `independent' fox and feral cat detections within each hour for the entire camera-trap survey duration.
+
+We identified the broad vegetation type (Ecological Vegetation Class (hereafter `EVC') group (DELWP 2020)) at each camera-trap location.
+Six vegetation types were surveyed in both regions, although to varying degrees (Table 1).
+In the Otway region, rainforests are interspersed (primarily in low lying gullies) throughout wet and damp forests at a finescale, and we merged these groups (hereafter `wet forests').
+
+For each deployment, we calculated the number of poison fox-baits within a 2.3 km radius around each camera-trap - the average maximum distance foxes in these regions travel from their home range centre (Hradsky et al.
 2017).
 
-Using the `reshape' R-package (Wickham 2007), we manipulated the detection table into a dataframe with a row for each hour of the day, for every camera-trap deployment, recording the total number of `independent' fox and feral cat detections within each hour for the entire camera-trap survey duration.
-
-\hypertarget{generalised-additive-models-of-diel-activity}{%
-\subsection{Generalised additive models of diel activity}\label{generalised-additive-models-of-diel-activity}}
+\hypertarget{generalised-additive-models}{%
+\subsection{Generalised additive models}\label{generalised-additive-models}}
 
@@
 We first fit these models using the poisson family, but switched to the negative binomial family as overdispersion, but not zero-inflation, was detected using the `DHARMa' R-package (Hartig 2020).
-We specified a model offset to account for differences in camera-trap survey durations and a random intercept for each site to account for repeat sampling.
+We specified a model offset to account for differences in camera-trap survey duration, and a random intercept for each site to account for repeat sampling.
 We used a duchon spline static spatial smooth to model both spatial autocorrelation and differences in predator activity across space (i.e.~the effect of unmodelled environmental covariates).
@@
 This formed the base model specification for each model we fitted.
-Models (n = 3) differed in their specification of the cyclical hour smooth in order to explore (1) variations in predator diel activity across EVC groups, as well as (2) whether feral cat diel activity changed across the gradient of observed fox counts (detailed in the sections below).
+Models (n = 2) differed in their specification of the cyclical hour smooth in order to explore (1) variations in predator diel activity across EVC groups, as well as (2) whether feral cat diel activity changed across the gradient of observed fox counts (detailed in the sections below).
+We used the double penalty model selection approach, which penalises the null space in addition to the range space (i.e.~shrinking wiggly terms to linear terms) of the spline basis, meaning covariate effects can be entirely removed from the model (Marra and Wood 2011).
 We plotted models using `ggplot2' and `gratia' R-packages (Wickham 2016, Simpson 2021).
 
-We expected predator activity to be lowest during the day, but potentially having small shifts in relative activity between sunset and sunrise across the different EVC groups (1).
-And so, fitting separate responses for each EVC group would throw away this shared information, which is particularly costly for the least surveyed groups -- resulting in more noisy estimates.
+We expected the activity of each predator species to follow a broadly consistent pattern (e.g.~lower during the day), but potentially have small shifts in relative activity between sunset and sunrise across the different EVC groups (1).
+Fitting separate responses for each EVC group would throw away shared information, which is particularly costly for the least surveyed groups, and would result in more noisy estimates.
 We therefore specified a global smoother for hour (i.e.~average response), along with group-level smoothers with shared wiggliness for each EVC group (i.e.~model GS in Pedersen et al.
 2019).
-This model structure penalises functions which deviate strongly from the average response--increasing confidence that they are true deviations opposed to the result of sampling noise.
-Because feral cats are likely subordinate to foxes, we investigated changes in their diel activity in response to counts of foxes.
-We added an extra column to our dataframe with the number of fox detections divided by the log of the survey days to account for differences in camera-trap survey durations (hereafter `adjusted fox counts').
-Because we saw different fox diel activity patterns in wet forests relative to the other EVC groups, we expected feral cats to temporally avoid foxes in opposing ways between these wet and dry vegetation groups.
+This model structure penalises functions which deviate strongly from the average response, increasing confidence that they are true deviations opposed to the result of sampling noise.
+
+To investigate changes in feral cat diel activity in response to counts of foxes.
+We added an extra column to our dataframe with the number of fox detections divided by the log of the number of survey days to account for differences in camera-trap survey durations (hereafter `adjusted fox counts').
+
+Analysis of stage 1 showed that fox diel activity patterns in wet forests differed relative to all other EVC groups.
+We therefore expected feral cats might to temporally avoid foxes in a different way in wet forests than the dry vegetation groups, but that responses would be similar in all the drier vegetation groups.
 We therefore modelled fox-induced changes in feral cat diel activity separately for wet and dry vegetation groups.
-Because of our large sample size, we further split dry vegetation groups by region for spatial replication.
-We refer to this variable as `habitat type'; having three levels: wet forests of the Otway Ranges (`wet\_otways') dry EVC groups of the Otway Ranges (`dry\_otways') and all of the Glenelg region data (`dry\_glenelg').
-We used a tensor product of hour (cyclic cubic regression spline) and log adjusted fox counts (thin plate regression spline) to model feral cat diel activity across the range of observed fox activity, separately for each habitat type.
+We further split dry vegetation groups by region, to examine whether patterns remained consistent across space?
+for spatial replication.
+We refer to the resulting variable as `habitat type', which had three levels: wet forests of the Otway Ranges (`wet\_otways') dry EVC groups of the Otway Ranges (`dry\_otways') and all of the Glenelg region data (`dry\_glenelg').
+
+We used a tensor product of hour (using cyclic cubic regression spline) and log adjusted fox count (using thin plate regression spline) smooths to model feral cat diel activity across the range of observed fox activity, separately for each habitat type.
+By also having double penalty shrinkage turned on, this meant that feral cat activity could vary nonlinearly and be penalised consistently across both dimensions.
+This model specification allows five different scenarios: that there was (1) no effect of hour or foxes on feral cats (2) a static hourly effect only, (3) a spatial response to foxes only, (4) a spatial response to foxes and an unrelated static hourly effect, (5) a spatial response to foxes and a hourly effect which changes across the range of fox counts.
 
@@
 
-On average, both predator species occupied similar times of the day. Lowest during the day, peaking during sunset and sunrise, high at night.
+On average, both predator species occupied similar times of the day. Lowest during the middle of the day, peaking during sunset and remaining relatively high at night and dawn.
 Foxes had a stronger diel activity patterns than cats, much less likely to be out during the day.
-Both predators had slight variations in diel activity across EVC groups, except for wet forests, were more substantial changes were seen.
-Foxes barely had any diel activity pattern here, cats more nocturnal.
+
+Diel activity patterns were similar across the different vegetation types for each predator,, except for wet forests, where more substantial changes were seen.
+In the wet forests, ox activity remained fairly constant throughout day and night barely had any diel activity pattern here, cats more nocturnal.
 Deviance explained and Rsq values for the two models were XX.
@@
 \begin{longtable}[]{@{}cccccc@{}}
-\caption{Summary of camera-trap deployments in unique sites within Ecological Vegetation Class groups, as well as independent counts of predator detections and naive occupancy rates, across two broad regions in south-west Victoria, Australia.}\tabularnewline
+\caption{Summary of camera-trap deployments in unique sites within Ecological Vegetation Class groups and independent counts of predator detections across two broad regions in south-west Victoria, Australia.}\tabularnewline
 \toprule
@@
 \includegraphics[width=1\linewidth]{../figs/predator_veg} \caption{Fox (A) and feral cat (B) diel activity patterns across Ecological Vegetation Class groups in south-west Victoria, Australia.
-Dotted, vertical lines respresent average sunrise and sunset times.
+Dotted, vertical lines represent average sunrise and sunset times.
 Shaded areas indicate 95\% confidence intervals.}\label{fig:veg}
@@
 
+\leavevmode\hypertarget{ref-MARRA20112372}{}%
+Marra, G. and Wood, S. N. 2011. Practical variable selection for generalized additive models. - Computational Statistics \& Data Analysis 55: 2372--2387.
+
 \leavevmode\hypertarget{ref-pedersen2019hierarchical}{}%

### 2021-08-04 r_scripts/6_individual_cat_variation.R
added individual cat variations

@@ -1,0 +1,320 @@
+## REFORMAT CAT IDENTIFIED RECORDS FOR DIEL ACTIVITY GAMS
+
+rm(list = ls())
+library(stringr)
+library(maptools)        
+library(dplyr)
+library(activity) 
+library(lubridate)
+library(reshape2)
+library(mgcv)
+library(gratia)
+library(ggplot2)
+library(patchwork)
+
+# load function to change clock time to radial time for overlap package
+source("r_scripts/function_clock_time_to_radian.R")
+
+# Set the default theme for ggplot objects 
+theme_set(theme_bw())
+theme_update(panel.grid = element_blank())
+
+# load cat records
+records <- read.csv("raw_data/cat_id_detections.csv")
+records$X <- NULL
+
+# load camdata - but just matts deployments
+camdata <- read.csv("raw_data/camdata.csv")
+camdata <- filter(camdata, data_source == "matt")
+camdata$X <- NULL
+camdata$X.1 <- NULL
+
+
+
+# ADD SOLAR TIMES ---------------------------------------------------------
+# take a matrix of the coordinates
+coords <- matrix(c(records$longitude, records$latitude), nrow = length(records$latitude))
+# add a coordinate reference system (can't be UTM)
+coords <- sp::SpatialPoints(coords, proj4string=sp::CRS("+proj=longlat +datum=WGS84")) 
+
+# specify detection time as a date class, as well as timezone (use Brisbane because it doesn't have daylight savings)
+records$date_time <- ymd_hms(records$date_time, tz = "Australia/Brisbane")
+
+# calculate the time of sunrise given the location and timing of each species detection 
+sunriset_df <- as.data.frame(sunriset(coords, records$date_time, direction = "sunrise", POSIXct.out=TRUE))
+# convert to radian time 
+sunrise_radian <- ClocktimeToTimeRad(sunriset_df$time) 
+# calculate the time of sunset given the location and timing of each species detection 
+sunsett_df <- as.data.frame(sunriset(coords, records$date_time, direction = "sunset", POSIXct.out=TRUE))
+# convert to radian 
+sunset_radian <- ClocktimeToTimeRad(sunsett_df$time) 
+
+# save these as a separate matrix for the transtime function below 
+anchors <- matrix(c(sunrise_radian, sunset_radian), nrow = length(sunset_radian))
+
+# we also need to convert the detection time to radian time
+det_radian <- ClocktimeToTimeRad(records$date_time)
+
[…]

### 2021-08-05 manuscript/manuscript.Rmd
removed model 2a

@@
 ---
-title: Joint spatiotemporal models reveal dynamic shifts in mesopredator spatial and diel activity for apex predator avoidance
+title: Joint spatiotemporal models reveal dramatic shifts in mesopredator diel activity across individuals, habitat types and apex predator activity levels
 author:
@@
 
+```{r ind, out.width="100%", fig.cap = "Individual variations in feral cat diel activity patterns across the Glenelg region (A) and wet forests of the western Otway Ranges (B) in south-west Victoria, Australia.
+Dotted, vertical lines represent average sunrise and sunset times.
+Shaded areas indicate 95\\% confidence intervals."}
+knitr::include_graphics("../figs/cat_ind.png")
+```
+
+\newpage
+
 DISCUSSION

### 2021-08-05 manuscript/manuscript.tex
removed model 2a

@@
             pdfauthor={},
-            pdftitle={Joint spatiotemporal models reveal dynamic shifts in mesopredator spatial and diel activity for apex predator avoidance},
+            pdftitle={Joint spatiotemporal models reveal dramatic shifts in mesopredator diel activity across individuals, habitat types and apex predator activity levels},
             colorlinks=false,
@@
 
-  \title{Joint spatiotemporal models reveal dynamic shifts in mesopredator spatial and diel activity for apex predator avoidance}
+  \title{Joint spatiotemporal models reveal dramatic shifts in mesopredator diel activity across individuals, habitat types and apex predator activity levels}
     \author[UOM]{Matthew W. Rees\corref{1}}
@@
 
-On average, both predator species occupied similar times of the day. Lowest during the middle of the day, peaking during sunset and remaining relatively high at night and dawn.
+On average, both predator species occupied similar times of the day--mostly nocturnal with activity peaks around sunrise and sunset.
+The main difference was that fox activity peaked just after sunset rather than on sunset.
+For foxes, this diel pattern was largely exhibited across all vegetation types except wet forest.
+In wet forests and rainforests, the hour of the day had little effect on foxes (especially considering diel activity curve here is being drawn towards a nocturnal pattern).
+
+lowest activity during the middle of the day, peaking at sunrise and sunset times, while remaining relatively high at night and dawn.
 Foxes had a stronger diel activity patterns than cats, much less likely to be out during the day.
@@
 Shaded areas indicate 95\% confidence intervals.}\label{fig:count}
+\end{figure}
+
+\newpage
+
+\begin{figure}
+\includegraphics[width=1\linewidth]{../figs/cat_ind} \caption{Individual variations in feral cat diel activity patterns across the Glenelg region (A) and wet forests of the western Otway Ranges (B) in south-west Victoria, Australia.
+Dotted, vertical lines represent average sunrise and sunset times.
+Shaded areas indicate 95\% confidence intervals.}\label{fig:ind}
 \end{figure}

### 2021-08-05 r_scripts/5_gams_diel_activity.R

@@ -98,7 +98,7 @@ gam_fox <- bam(fox ~ s(hour, bs = "cc", k = 8) +
                      s(longitude, latitude, bs = "ds",  m = c(1, 0.5), k = 200) +
                      s(station, bs = "re") +  
                      offset(log(survey_duration)), 
-                    data = records, family = nb, knots = list(hour = c(0, 23)), nthreads = 3, discrete = TRUE, select = TRUE)
+                    data = records, family = nb, knots = list(hour = c(0, 23)), nthreads = 3, discrete = TRUE)
 summary(gam_fox)
 plot(gam_fox, pages = 1, scheme = 2, seWithMean = TRUE, shade = TRUE, scale = 0)
 
@@ -109,20 +109,10 @@ gam_cat_1 <- bam(cat ~ s(hour, bs = "cc", k = 8) +
                        s(longitude, latitude, bs = "ds",  m = c(1, 0.5), k = 200) +
                        s(station, bs = "re") +  
                        offset(log(survey_duration)), 
-                 data = records, family = nb, knots = list(hour = c(0, 23)), nthreads = 3, discrete = TRUE, select = TRUE)
+                 data = records, family = nb, knots = list(hour = c(0, 23)), nthreads = 3, discrete = TRUE)
 summary(gam_cat_1)
 plot(gam_cat_1, pages = 1, scheme = 2, seWithMean = TRUE, shade = TRUE, scale = 0)
 
-
-# 2) DO CATS CHANGE DIEL ACTIVITY BASED ON FOX ACTIVITY?
-# 2a) fox presence / absence
-gam_cat_2a <- bam(cat ~ fox_pa + s(hour, habitat_type, by = fox_pa, bs = "fs", xt = list(bs = "cc"), k = 8) + 
-                        s(longitude, latitude, bs = "ds",  m = c(1, 0.5), k = 200) +
-                        s(station, bs = "re") +  
-                        offset(log(survey_duration)), 
-                  data = records, family = nb, knots = list(hour = c(0, 23)), nthreads = 3, discrete = TRUE, select = TRUE)
-summary(gam_cat_2a)
-plot(gam_cat_2a, pages = 1, scheme = 2, seWithMean = TRUE, shade = TRUE, scale = 0)
 
 # 2b) fox counts
[…]

### 2021-08-05 r_scripts/6_individual_cat_variation.R

@@ -267,7 +267,7 @@ x = sapply(gam_cat_ind_g$smooth, "[[",  "label")
 df <- cbind(df, predict.gam(gam_cat_ind_g, newdata = df, se.fit = TRUE, type = "link", newdata.guaranteed = TRUE, exclude = c(x[3:6], "s(survey_duration)")))
 # shorten cat names
 df$individual <- as.character(df$individual)
-df$individual <- recode(df$individual, annya_swirl_lola = "faith", c_tabby_quasimodo = "brisket", m_swirl_socksy = "devon", annya_swirl_ben = "sheila", annya_swirl_motley = "motley", hotspur_tabby_1 = "sprout", m_tabby_murray = "murray")
+df$individual <- recode(df$individual, annya_swirl_lola = "faith", c_tabby_quasimodo = "brisket", m_swirl_socksy = "boots", annya_swirl_ben = "dolly", annya_swirl_motley = "motley", hotspur_tabby_1 = "sprout", m_tabby_murray = "murray")
 # plot top 7 individuals
 plot_ind <- ggplot(data=df, aes(x=hour, y=fit, group=individual)) +
   geom_line(aes(y=fit, x=hour), lwd = 0.7) +
@@ -277,7 +277,7 @@ plot_ind <- ggplot(data=df, aes(x=hour, y=fit, group=individual)) +
   labs(subtitle = "iii.   Predicted activity", x = "Hour", y = "log(count)") 
 
 # plot / save
-png("figs/cat_ind_g.png", width = 9, height = 6, res = 600, units = "in")
+png("figs/cat_ind_g.png", width = 9, height = 5, res = 600, units = "in")
 (plot_global | plot_difference) / plot_ind  + plot_annotation(title = "A.   Glenelg region")
 dev.off()
 
@@ -310,8 +310,8 @@ plot_ind <- ggplot(data=df, aes(x=hour, y=fit, group=individual)) +
   labs(subtitle = "iii.   Predicted activity", x = "Hour", y = "log(count)") 
 
 # plot / save
-png("figs/cat_ind_o.png", width = 9, height = 6, res = 600, units = "in")
-(plot_global | plot_difference) / plot_ind  + plot_annotation(title = "B.   Western Otways")
+png("figs/cat_ind_o.png", width = 9, height = 5, res = 600, units = "in")
+(plot_global | plot_difference) / plot_ind  + plot_annotation(title = "B.   Western Otway Ranges")
 dev.off()
 
 # to combine figures, type in the terminal (using imagemagick): 

### 2021-08-06 manuscript/manuscript.Rmd
update

@@
   1. Background.
+
   2. Methods.
+
   3. Results.
+
   4. *Synthesis and applications.*
@@
 
-Effective wildlife management relies on understanding species population and behavioural dynamics to accurately predict intervention outcomes.
-These dynamics depend on both environmental conditions and those of other species, but few statistical approaches incorporate both factors, or allow responses to be dynamic across space and time.
-This is key because species interactions and niche preferences are multidimensional.
-
-Fear is a near universal driver of animal behaviour.
-Fear may be from other individuals of their species, a dominant species (including humans) or an event (e.g.
-fire).
-Animals may be naive to threats (e.g.
-alien predators), unable to respond to threats (e.g.
-scarce prey), or the threat may be too minimal to warrant a response.
-Some environments more easily facilitate avoidance (e.g.
-high food and shelter accessibility), and so accounting for landscape context is key.
-A threat response may be routine, or only employed or enhanced when the threat is heightened.
-This likely depends on how constant and predictable the threat is.
+Diel activity patterns - the distribution of activity throughout the daily cycle - are an often overlooked behavioural aspect in ecology.
+A species must allocate particular times of the day to rest, foraging, socialisation and reproduction; which are traded-off against threat avoidance and deterrence against resource loss.
+These trade-offs may or may not incur costs to individual and population fitness.Diel activity may be bound by physiological and environmental constraints.
+For example, animal must be nocturnal as to not overheat or because that is when prey is most accessible.
+Other animals have varied diel activity, which may alter their ecosystem impact.
+Rigid or flexible animal diel activity may be the difference between adapting or succumbing to a threat.
 
@@
 avoidance).
-An animal may avoid locations where the threat is likely to occur (hereafter ‘spatial avoidance’), avoid times of the day when a threat is likely to occur (hereafter ‘temporal avoidance’), or avoid both locations and times a threat is heightened (hereafter ‘spatiotemporal avoidance’).
-Such avoidance likely has trade-offs, which may or may not incur costs to individual and population fitness.
-Modelling approaches which do not allow spatiotemporal and dynamic fear reponses can only make limited inference about species interactions.
+An animal may avoid times of the day when a threat is likely to occur (hereafter ‘temporal avoidance’), or locations where the threat is likely to occur (hereafter ‘spatial avoidance’), avoid or avoid both locations and times a threat is heightened (hereafter ‘spatiotemporal avoidance’).
+Modelling approaches which do not consider both spatial and temporal niches in a combined framework miss important nuance in animal behaviour.
+This is particularly the case when threats vary spatiotemporally, such as other species.
 
@@
 
-Dynamic behavioural responses are also important from a management perspective because they want to know whether an action will cause a change in animal behaviour /  range?).
+Dynamic behavioural responses are important from a management perspective because they want to know whether an action will cause a change in animal behaviour /  range?).
 For example, after the reintroduction of wolves in Yellowstone, it was widely thought that this had limited elk populations.
@@
 
+Animals may be naive to threats (e.g.
+alien predators), unable to respond to threats (e.g.
+scarce prey), or the threat may be too minimal to warrant a response.
+Some environments more easily facilitate avoidance (e.g.
+high food and shelter accessibility), and so accounting for landscape context is key.
+A threat response may be routine, or only employed or enhanced when the threat is heightened.
+This likely depends on how constant and predictable the threat is.
+
+Invasive species are highly adaptable to new conditions, which may facilitated by flexible spatiotemporal behaviour. 
 There is contention around the interactions between two invasive predators - feral cats and foxes.
 Foxes are routinely culled, and there is concern that this could lead to a mesopredator release of feral cats.
-Mesopredator release may manifest in both behavioural and numerical changes, which my previous chapter identified - but that behavioral responses (ranging and detectability) were not universal.
-Here, we aim to dig into the behavioral response further - in particular diel activity times.
+Mesopredator release may manifest in both behavioural and numerical changes, which my previous chapter identified - but that behavioural responses (ranging and detectability) were not universal.
+Here, we aim to dig into the behavioural response further - in particular diel activity times.
 
@@
 ## Camera-trap surveys
-
 We compiled camera-trap data from three distinct studies across the two regions.
@@
 ## Generalised additive models 
-We modelled predator counts (response variable) with generalised additive mixed-effects models (hereafter 'GAMMs') implemented in the 'mgcv' R-package [@wood2017].
-We first fit these models using the poisson family, but switched to the negative binomial family as overdispersion, but not zero-inflation, was detected using the 'DHARMa' R-package [@DHARMa].
+We modelled predator counts (response variable) with generalised additive models (hereafter 'GAMs') implemented in the 'mgcv' R-package [@wood2017].
+GAMs learn from the data to specify nonlinear relationships between response and explanatory variables.
+A maximum wiggliness is specified for each explanatory variable, with complexity automatically reduced (all the way to completely linear function) through a smoothing penalty to avoid overfitting.
+We used REML estimation and the negative binomial family (as overdispersion, but not zero-inflation, was initially detected with a Poisson distribution using the 'DHARMa' R-package [@DHARMa]).
+
 We specified a model offset to account for differences in camera-trap survey duration, and a random intercept for each site to account for repeat sampling.
 We used a duchon spline static spatial smooth to model both spatial autocorrelation and differences in predator activity across space (i.e.
-the effect of unmodelled environmental covariates).
+the effect of unaccounted for environmental covariates).
 For the fox model, we additionally included a thin plate regression spline for poison-bait density, with a factor smoother for each region, as bait replacement schedules differed between the regions.
 This formed the base model specification for each model we fitted.
-Models (n = 2) differed in their specification of the cyclical hour smooth in order to explore (1) variations in predator diel activity across EVC groups, as well as (2) whether feral cat diel activity changed across the gradient of observed fox counts (detailed in the sections below).
-We used the double penalty model selection approach, which penalises the null space in addition to the range space (i.e.
-shrinking wiggly terms to linear terms) of the spline basis, meaning covariate effects can be entirely removed from the model [@MARRA20112372].
+Models (n = 3) differed in their specification of the cyclical hour smooth in order to explore variations in (1) predator diel activity across EVC groups, (2) feral cat diel activity across the gradient of observed fox counts, as well as (3) individual feral cat diel activity, which are detailed in the sections below.
 We plotted models using 'ggplot2' and 'gratia' R-packages [@ggplot2; @gratia].
 
+### Variation in predator diel activity across vegetation types
 We expected the activity of each predator species to follow a broadly consistent pattern (e.g.
@@
 
+### Variation in feral cat diel activity across apex predator activity
 To investigate changes in feral cat  diel activity in response to counts of foxes.
@@
 We used a tensor product of hour (using cyclic cubic regression spline) and log adjusted fox count (using thin plate regression spline) smooths to model feral cat diel activity across the range of observed fox activity, separately for each habitat type.
-By also having double penalty shrinkage turned on, this meant that feral cat activity could vary nonlinearly and be penalised consistently across both dimensions.
+We used the double penalty model selection approach, which penalises the null space in addition to the range space (i.e.
+shrinking wiggly terms to linear terms) of the spline basis, meaning covariate effects can be entirely removed from the model [@MARRA20112372].
+This meant that feral cat activity could vary nonlinearly and be penalised consistently across both dimensions.
 This model specification allows five different scenarios: that there was (1) no effect of hour or foxes on feral cats (2) a static hourly effect only, (3) a spatial response to foxes only, (4) a spatial response to foxes and an unrelated static hourly effect, (5) a spatial response to foxes and a hourly effect which changes across the range of fox counts.
+
+### Individual variation in feral cat diel activity
 
@@
 ==========
+
+GRAVEYARD
+===================
+
+Diel activity patterns are a product of physiological constraints, species adaptiveness and landscape context.
+These dynamics depend on both environmental conditions and those of other species, but few statistical approaches incorporate both factors, or allow responses to be dynamic across space and time.
+This is key because species interactions and niche preferences are multidimensional.
+
+Effective wildlife management relies on understanding species behavioural and population dynamics to accurately predict intervention outcomes.
+Behavioural dynamics can vary across individuals, environmental conditions and with the activity of different species, and this may or may not lead to population-level impacts.
+
+Fear is a near universal driver of animal behaviour. Fear may be from other individuals of their species, a dominant species (including humans) or an event (e.g. fire).

### 2021-08-06 manuscript/manuscript.tex
update

@@
   \def\labelenumi{\arabic{enumi}.}
-  \tightlist
   \item
@@
 
-Effective wildlife management relies on understanding species population and behavioural dynamics to accurately predict intervention outcomes.
-These dynamics depend on both environmental conditions and those of other species, but few statistical approaches incorporate both factors, or allow responses to be dynamic across space and time.
-This is key because species interactions and niche preferences are multidimensional.
-
-Fear is a near universal driver of animal behaviour.
-Fear may be from other individuals of their species, a dominant species (including humans) or an event (e.g.~fire).
-Animals may be naive to threats (e.g.~alien predators), unable to respond to threats (e.g.~scarce prey), or the threat may be too minimal to warrant a response.
-Some environments more easily facilitate avoidance (e.g.~high food and shelter accessibility), and so accounting for landscape context is key.
-A threat response may be routine, or only employed or enhanced when the threat is heightened.
-This likely depends on how constant and predictable the threat is.
+Diel activity patterns - the distribution of activity throughout the daily cycle - are an often overlooked behavioural aspect in ecology.
+A species must allocate particular times of the day to rest, foraging, socialisation and reproduction; which are traded-off against threat avoidance and deterrence against resource loss.
+These trade-offs may or may not incur costs to individual and population fitness.Diel activity may be bound by physiological and environmental constraints.
+For example, animal must be nocturnal as to not overheat or because that is when prey is most accessible.
+Other animals have varied diel activity, which may alter their ecosystem impact.
+Rigid or flexible animal diel activity may be the difference between adapting or succumbing to a threat.
 
 Animals may respond to threats by altering their activity patterns (i.e.~avoidance).
-An animal may avoid locations where the threat is likely to occur (hereafter `spatial avoidance'), avoid times of the day when a threat is likely to occur (hereafter `temporal avoidance'), or avoid both locations and times a threat is heightened (hereafter `spatiotemporal avoidance').
-Such avoidance likely has trade-offs, which may or may not incur costs to individual and population fitness.
-Modelling approaches which do not allow spatiotemporal and dynamic fear reponses can only make limited inference about species interactions.
+An animal may avoid times of the day when a threat is likely to occur (hereafter `temporal avoidance'), or locations where the threat is likely to occur (hereafter `spatial avoidance'), avoid or avoid both locations and times a threat is heightened (hereafter `spatiotemporal avoidance').
+Modelling approaches which do not consider both spatial and temporal niches in a combined framework miss important nuance in animal behaviour.
+This is particularly the case when threats vary spatiotemporally, such as other species.
 
@@
 
-Dynamic behavioural responses are also important from a management perspective because they want to know whether an action will cause a change in animal behaviour / range?).
+Dynamic behavioural responses are important from a management perspective because they want to know whether an action will cause a change in animal behaviour / range?).
 For example, after the reintroduction of wolves in Yellowstone, it was widely thought that this had limited elk populations.
@@
 
+Animals may be naive to threats (e.g.~alien predators), unable to respond to threats (e.g.~scarce prey), or the threat may be too minimal to warrant a response.
+Some environments more easily facilitate avoidance (e.g.~high food and shelter accessibility), and so accounting for landscape context is key.
+A threat response may be routine, or only employed or enhanced when the threat is heightened.
+This likely depends on how constant and predictable the threat is.
+
+Invasive species are highly adaptable to new conditions, which may facilitated by flexible spatiotemporal behaviour.
 There is contention around the interactions between two invasive predators - feral cats and foxes.
 Foxes are routinely culled, and there is concern that this could lead to a mesopredator release of feral cats.
-Mesopredator release may manifest in both behavioural and numerical changes, which my previous chapter identified - but that behavioral responses (ranging and detectability) were not universal.
-Here, we aim to dig into the behavioral response further - in particular diel activity times.
+Mesopredator release may manifest in both behavioural and numerical changes, which my previous chapter identified - but that behavioural responses (ranging and detectability) were not universal.
+Here, we aim to dig into the behavioural response further - in particular diel activity times.
 
@@
 
-We modelled predator counts (response variable) with generalised additive mixed-effects models (hereafter `GAMMs') implemented in the `mgcv' R-package (Wood 2017).
-We first fit these models using the poisson family, but switched to the negative binomial family as overdispersion, but not zero-inflation, was detected using the `DHARMa' R-package (Hartig 2020).
+We modelled predator counts (response variable) with generalised additive models (hereafter `GAMs') implemented in the `mgcv' R-package (Wood 2017).
+GAMs learn from the data to specify nonlinear relationships between response and explanatory variables.
+A maximum wiggliness is specified for each explanatory variable, with complexity automatically reduced (all the way to completely linear function) through a smoothing penalty to avoid overfitting.
+We used REML estimation and the negative binomial family (as overdispersion, but not zero-inflation, was initially detected with a Poisson distribution using the `DHARMa' R-package (Hartig 2020)).
+
 We specified a model offset to account for differences in camera-trap survey duration, and a random intercept for each site to account for repeat sampling.
-We used a duchon spline static spatial smooth to model both spatial autocorrelation and differences in predator activity across space (i.e.~the effect of unmodelled environmental covariates).
+We used a duchon spline static spatial smooth to model both spatial autocorrelation and differences in predator activity across space (i.e.~the effect of unaccounted for environmental covariates).
 For the fox model, we additionally included a thin plate regression spline for poison-bait density, with a factor smoother for each region, as bait replacement schedules differed between the regions.
 This formed the base model specification for each model we fitted.
-Models (n = 2) differed in their specification of the cyclical hour smooth in order to explore (1) variations in predator diel activity across EVC groups, as well as (2) whether feral cat diel activity changed across the gradient of observed fox counts (detailed in the sections below).
-We used the double penalty model selection approach, which penalises the null space in addition to the range space (i.e.~shrinking wiggly terms to linear terms) of the spline basis, meaning covariate effects can be entirely removed from the model (Marra and Wood 2011).
+Models (n = 3) differed in their specification of the cyclical hour smooth in order to explore variations in (1) predator diel activity across EVC groups, (2) feral cat diel activity across the gradient of observed fox counts, as well as (3) individual feral cat diel activity, which are detailed in the sections below.
 We plotted models using `ggplot2' and `gratia' R-packages (Wickham 2016, Simpson 2021).
+
+\hypertarget{variation-in-predator-diel-activity-across-vegetation-types}{%
+\subsubsection{Variation in predator diel activity across vegetation types}\label{variation-in-predator-diel-activity-across-vegetation-types}}
 
@@
 This model structure penalises functions which deviate strongly from the average response, increasing confidence that they are true deviations opposed to the result of sampling noise.
+
+\hypertarget{variation-in-feral-cat-diel-activity-across-apex-predator-activity}{%
+\subsubsection{Variation in feral cat diel activity across apex predator activity}\label{variation-in-feral-cat-diel-activity-across-apex-predator-activity}}
 
@@
 We used a tensor product of hour (using cyclic cubic regression spline) and log adjusted fox count (using thin plate regression spline) smooths to model feral cat diel activity across the range of observed fox activity, separately for each habitat type.
-By also having double penalty shrinkage turned on, this meant that feral cat activity could vary nonlinearly and be penalised consistently across both dimensions.
+We used the double penalty model selection approach, which penalises the null space in addition to the range space (i.e.~shrinking wiggly terms to linear terms) of the spline basis, meaning covariate effects can be entirely removed from the model (Marra and Wood 2011).
+This meant that feral cat activity could vary nonlinearly and be penalised consistently across both dimensions.
 This model specification allows five different scenarios: that there was (1) no effect of hour or foxes on feral cats (2) a static hourly effect only, (3) a spatial response to foxes only, (4) a spatial response to foxes and an unrelated static hourly effect, (5) a spatial response to foxes and a hourly effect which changes across the range of fox counts.
+
+\hypertarget{individual-variation-in-feral-cat-diel-activity}{%
+\subsubsection{Individual variation in feral cat diel activity}\label{individual-variation-in-feral-cat-diel-activity}}
 
@@
 
+\hypertarget{graveyard}{%
+\section{GRAVEYARD}\label{graveyard}}
+
+Diel activity patterns are a product of physiological constraints, species adaptiveness and landscape context.
+These dynamics depend on both environmental conditions and those of other species, but few statistical approaches incorporate both factors, or allow responses to be dynamic across space and time.
+This is key because species interactions and niche preferences are multidimensional.
+
+Effective wildlife management relies on understanding species behavioural and population dynamics to accurately predict intervention outcomes.
+Behavioural dynamics can vary across individuals, environmental conditions and with the activity of different species, and this may or may not lead to population-level impacts.
+
+Fear is a near universal driver of animal behaviour. Fear may be from other individuals of their species, a dominant species (including humans) or an event (e.g.~fire).
+
 \hypertarget{refs}{}

### 2021-08-06 r_scripts/7_coat_cat_variation.R

@@ -110,6 +110,19 @@ records <- filter(records, is.na(individual) | individual != "annya_swirl_ben_&_
 # now add the new ones in 
 records <- rbind(records, records_double)
 
+unique(records$coat_type)
+
+# new simple coat type
+records$coat <- if_else(records$coat_type == "tortoise", "tabby", "temp")
+records$coat <- if_else(records$coat_type == "tabby", "tabby", records$coat)
+records$coat <- if_else(records$coat_type == "ginger", "tabby", records$coat)
+records$coat <- if_else(records$coat_type == "blackwhite", "black", records$coat)
+records$coat <- if_else(records$coat_type == "kittens_&_black", "black", records$coat)
+records$coat <- if_else(records$coat_type == "kittens_&_tortoise_&_tabby", "tabby", records$coat)
+records$coat <- if_else(records$coat_type == "tortoise_&_tabby", "tabby", records$coat)
+records <- filter(records, coat != "temp")
+unique(records$coat)
+
 
 # RESHAPE BY HOUR ---------------------------------------------------------
 # DATAFRAME 1:  HOUR ----------------------------------------------------
@@ -132,12 +145,12 @@ gam_data_g <- melt(records_g, id.var = colnames(records_g), measure.var = "occ")
 gam_data_o <- melt(records_o, id.var = colnames(records_o), measure.var = "occ")
 
 # cast to derive count per year per station_year long format
-gam_data_g = dcast(gam_data_g,  individual + station_year + hour ~ species, fill = 0, fun = sum)
-gam_data_o = dcast(gam_data_o,  individual + station_year + hour ~ species, fill = 0, fun = sum)
+gam_data_g = dcast(gam_data_g,  coat + station_year + hour ~ species, fill = 0, fun = sum)
+gam_data_o = dcast(gam_data_o,  coat + station_year + hour ~ species, fill = 0, fun = sum)
 
 # make a table with all unique combinations - (use camdata so we don't just take the sites with detections)
-df_g <- expand.grid(station_year = unique(camdata_g$station_year), individual = unique(records_g$individual), hour = 0:23)
-df_o <- expand.grid(station_year = unique(camdata_o$station_year), individual = unique(records_o$individual), hour = 0:23)
+df_g <- expand.grid(station_year = unique(camdata_g$station_year), coat = c("tabby", "black"), hour = 0:23)
[…]

### 2021-09-05 diel_activity_analysis.Rmd
fox ~ ts, region ~ re

---
title: "Diel activity GAMs"
author: "Matthew Rees"
date: "12/08/2021"
output:
  html_document:
    depth: 1
    number_sections: no
    theme: sandstone
    toc: yes
    toc_float: yes
    always_allow_html: yes
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, warnings = FALSE, message = FALSE)
options(scipen = 999)

library(dplyr)
library(mgcv)
library(ggplot2)
library(gganimate)
library(patchwork)
library(ggpubr)
library(viridis)
library(gratia)
library(chron)
library(sp)
library(sf)
library(reshape2)
library(lubridate)
library(stringr)
library(maptools)        
library(activity) 
library(lubridate)
library(pander)

# load function to change clock time to radial time for overlap package
source("r_scripts/function_clock_time_to_radian.R")

# Set the default theme for ggplot objects 
theme_set(theme_bw())
theme_update(panel.grid = element_blank())

```

```{r data prep, echo=FALSE}

# PREPARE DATA ------------------------------------------------------------
## load gam data - 23 rows for each hour of every deployment
records <- read.csv("derived_data/counts_hour.csv")

## EXCLUDE DATA
# drop zoi's data 
records <- filter(records, data_source != "zoi")
# exclude stations left out for less than 2 weeks 
records <- filter(records, survey_duration >= 10)
#hist(records$survey_duration / 23, breaks = 50)
#summary(records$survey_duration)

## MODIFY EVC GROUPS
# as we only had three "Riverine Grassy Woodlands or Forests" unique survey sites --> drop em
records <- filter(records, XGROUPNAME != "Riverine Grassy Woodlands or Forests")
# as we only had 20 "Rainforests" unique survey sites, all of which interspersed in "Wet Forests" in the Otways, reclassify as "Wet Forests". 
records$XGROUPNAME <- if_else(records$XGROUPNAME == "Rainforests", "Wet or Damp Forests", as.character(records$XGROUPNAME))
# abbreviate EVC group names for plotting
records$XGROUPNAME <- if_else(records$XGROUPNAME == "Wet or Damp Forests", "Wet Forests", records$XGROUPNAME)
records$XGROUPNAME <- if_else(records$XGROUPNAME == "Riparian Scrubs or Swampy Scrubs and Woodlands", "Swampy Scrubs", records$XGROUPNAME)
records$XGROUPNAME <- substr(records$XGROUPNAME, 1, nchar(records$XGROUPNAME) - 1)
#table(records$XGROUPNAME)

# add xy coordinates
records_sf <- st_as_sf(records, coords = c("longitude", "latitude"), crs = 4326) %>% 
  st_transform(crs = 32754)
records$x <- st_coordinates(records_sf)[,1]
records$y <- st_coordinates(records_sf)[,2]
#head(records)

# ADD FOX COUNTS PER CAMERA -----------------------------------------------
## Single fox count for each cam-trap (and do the same for cats for good measure)
records <- records %>%
  group_by(station_year)  %>%
  mutate(fox_count = sum(fox),
         cat_count = sum(cat))

# fox presence / absence
records$fox_pa <- if_else(records$fox_count > 0, "present", "absent")
records$cat_pa <- if_else(records$cat_count > 0, "present", "absent")
#xtabs(~records$fox_pa + records$cat_pa)

# adjust (positive only - we don't want negative counts) fox counts for survey effort 
records$fox_count_adj <- ifelse(records$fox_count > 0, records$fox_count/log(records$survey_duration), records$fox_count)

# and take the log (add 1 because you cant log 0)
records$fox_count_adj <- log(records$fox_count_adj + 1)
#hist(records$fox_count_adj)

## Add habitat type cov (for models 2/3)
records$habitat_type <- if_else(records$XGROUPNAME == "Wet Forest", "wet_otway", "dry_otway")
records$habitat_type <- if_else(records$region == "glenelg", "dry_glenelg", records$habitat_type)
records$habitat_type <- if_else(records$region == "glenelg", "dry_glenelg", records$habitat_type)
records$habitat_type <- if_else(is.na(records$habitat_type), "dry_otway", records$habitat_type)
#unique(records$habitat_type)

## Transform variable class  for GAMs
records <- transform(records,
                     hour = as.integer(hour),
                     cat = as.integer(cat), 
                     fox = as.integer(fox), 
                     foxbaits = as.integer(foxbaits), 
                     fox_pa = factor(fox_pa, ordered = FALSE), 
                     fox_count_adj = as.numeric(fox_count_adj), 
                     region = factor(region, ordered = FALSE), 
                     station = factor(station, ordered = FALSE), 
                     vegetation_group = factor(XGROUPNAME, ordered = FALSE), 
                     habitat_type = factor(habitat_type, ordered = FALSE), 
                     survey_duration = as.integer(survey_duration)
)
#summary(records)

# save modified dataframe
write.csv(records, "derived_data/counts_hour_cleaned.csv")

# split by region
records_g <- filter(records, region == "glenelg")
records_o <- filter(records, region == "otways")

```

## Changes spatial activity across the daily cycle

```{r space fox, cache = TRUE}

gam_fox_sp <- bam(fox ~ t2(x, y, hour, d = c(2, 1), bs = c("ds", "cc"), k = c(80, 8), m = c(1, 0.5), full = TRUE) +
                         s(foxbaits, bs = "tp", k = 4) + 
                         s(station, bs = "re") +  
                         offset(log(survey_duration)), 
                 data = records, family = nb, knots = list(hour = c(0, 23)), nthreads = 3, discrete = TRUE)
summary(gam_fox_sp)

gam_cat_sp <- bam(cat ~ t2(x, y, hour, d = c(2, 1), bs = c("ds", "cc"), k = c(80, 8), m = c(1, 0.5), full = TRUE) +
                         s(station, bs = "re") +  
                         offset(log(survey_duration)), 
                 data = records, family = nb, knots = list(hour = c(0, 23)), nthreads = 3, discrete = TRUE)
summary(gam_cat_sp)

```

### Spatial variation in the strength of diel patterns

```{r space plot creation, echo=FALSE}
#### FOX
### GLENELG
## STEP 1) Make buffer zone around cameras to restrict plotting - optional 
#take just the cams
records_g_cams <- distinct(records_g, station, .keep_all = TRUE)
# change it to sf class
records_g_cams <- st_as_sf(records_g_cams, coords = c("x", "y"), crs = 32754) 
# make a 4km buffer around each camera
glenelg_cams_buffer = st_buffer(records_g_cams, 4000)
# dissolve the buffer
glenelg_cams_buffer = st_union(glenelg_cams_buffer)
# convert back to dataframe
glenelg_buffer_df <- fortify(as_Spatial(glenelg_cams_buffer))%>%
  transmute(x = long, y = lat, order = order, group = group)
# split by each group (seperate polygon for each grid)
buffer_df1 <- glenelg_buffer_df[which(glenelg_buffer_df$group == "ID1.1"),]
buffer_df2 <- glenelg_buffer_df[which(glenelg_buffer_df$group == "ID1.2"),]
buffer_df3 <- glenelg_buffer_df[which(glenelg_buffer_df$group == "ID1.3"),]
buffer_df4 <- glenelg_buffer_df[which(glenelg_buffer_df$group == "ID1.4"),]

## STEP 2) MAKE A GRID TO PREDICT DATA INTO
data_g_plot = expand.grid(
  x = seq(min(glenelg_buffer_df$x), 
          max(glenelg_buffer_df$x),
          length=50),
  y = seq(min(glenelg_buffer_df$y),
          max(glenelg_buffer_df$y),
          length=50),
  survey_duration = 60,
  foxbaits = 0, 
  hour = c(0:23),
  station = "A001")

# subset to just locations within  of each camera 
data_g_plot1 = data_g_plot[with(data_g_plot, inSide(buffer_df1, x, y)),]
data_g_plot2 = data_g_plot[with(data_g_plot, inSide(buffer_df2, x, y)),]
data_g_plot3 = data_g_plot[with(data_g_plot, inSide(buffer_df3, x, y)),]
data_g_plot4 = data_g_plot[with(data_g_plot, inSide(buffer_df4, x, y)),]
data_g_plot <- rbind(data_g_plot1, data_g_plot2, data_g_plot3, data_g_plot4)

# predict model results into dataframe
data_g_plot <- cbind(data_g_plot, predict.gam(gam_fox_sp, newdata = data_g_plot, se.fit = TRUE, type = "link", exclude = c("s(station)", "s(survey_duration)")))
data_g_plot <- rename(data_g_plot, fox_predicted = fit,  fox_predicted_se = se.fit) # rename
# give each row / location a unique name to group by
data_g_plot$location_id <- paste0(data_g_plot$x, "_", data_g_plot$y)
# get difference in min max diel activity for each row
data_g_plot <- data_g_plot %>% 
  group_by(location_id) %>% 
  mutate(diel_diff = max(fox_predicted) - min(fox_predicted))

# fox plot: spatial activity per hour
g_fox_plot_hr <- ggplot(aes(x, y, fill = fox_predicted),
                     data = data_g_plot) +
  geom_tile()+
  scale_fill_viridis("log(activity)", option = "viridis", limits = c(-5.27, -1.13)) +
  geom_point(data = records_g, fill = NA, col = "white", size = 0.7, alpha = 0.1, shape = 3) +
  theme_bw(10) + 
  theme(axis.title = element_blank()) + 
    labs(title = 'Glenelg region; red fox; Hour: {frame_time}:00') +
  transition_time(hour) +
  ease_aes('linear')

# fox plot: spatial difference
g_fox_plot_diff <- ggplot(aes(x, y, fill = diel_diff),
                     data = data_g_plot) +
  geom_tile()+
  scale_fill_viridis("Relative diel pattern strength", option = "viridis", limits = c(0.5, 1.55), breaks=c(0.5, 1.03, 1.55),labels=c("low", "medium", "high")) +
  geom_point(data = records_g, fill = NA, col = "white", size = 0.7, alpha = 0.015, shape = 3) +
  theme_bw(10) + 
  ggtitle("a.  Glenelg region; red fox") +
  theme(axis.title = element_blank())

### OTWAYS
## STEP 1) Make buffer zone around cameras to restrict plotting - optional 
#take just the cams
records_o_cams <- distinct(records_o, station, .keep_all = TRUE)
# change it to sf class
records_o_cams <- st_as_sf(records_o_cams, coords = c("x", "y"), crs = 32754) 
# make a 4km buffer around each camera
otways_cams_buffer = st_buffer(records_o_cams, 6500)
# dissolve the buffer
otways_cams_buffer = st_union(otways_cams_buffer)
# convert back to dataframe
otways_buffer_df <- fortify(as_Spatial(otways_cams_buffer))%>%
  transmute(x = long, y = lat, order = order, group = group)

## STEP 2) MAKE A GRID TO PREDICT DATA INTO
data_o_plot = expand.grid(
  x = seq(min(otways_buffer_df$x), 
          max(otways_buffer_df$x),
          length=50),
  y = seq(min(otways_buffer_df$y),
          max(otways_buffer_df$y),
          length=50),
  station = "T053", 
  year = 2018,
  survey_duration = 60,
  hour = 0:23,
  foxbaits = 0)

# subset to just locations within buffer zone
data_o_plot = data_o_plot[with(data_o_plot, inSide(otways_buffer_df, x, y)),]
data_o_plot <- cbind(data_o_plot, predict.gam(gam_fox_sp, newdata = data_o_plot, se.fit = TRUE, type = "link", exclude = c("s(station)", "s(survey_duration)")))
data_o_plot <- rename(data_o_plot, fox_predicted = fit,  fox_predicted_se = se.fit) # rename
# give each row / location a unique name to group by
data_o_plot$location_id <- paste0(data_o_plot$x, "_", data_o_plot$y)
# get difference in min max diel activity for each row
data_o_plot <- data_o_plot %>% 
  group_by(location_id) %>% 
  mutate(diel_diff = max(fox_predicted) - min(fox_predicted))

# fox plot: spatial activity per hour
o_fox_plot_hr <- ggplot(aes(x, y, fill = fox_predicted),
                     data = data_o_plot) +
  geom_tile()+
  scale_fill_viridis("log(activity)", option = "viridis", limits = c(-5.27, -1.13)) +
  geom_point(data = records_o, fill = NA, col = "white", size = 0.7, alpha = 0.1, shape = 3) +
  theme_bw(10) + 
  theme(axis.title = element_blank()) + 
    labs(title = 'Otway Ranges; red fox; Hour: {frame_time}:00') +
  transition_time(hour) +
  ease_aes('linear')

# fox plot: spatial difference
o_fox_plot_diff <- ggplot(aes(x, y, fill = diel_diff),
                     data = data_o_plot) +
  geom_tile()+
  scale_fill_viridis("Relative diel pattern strength", option = "viridis", limits = c(0.5, 1.55), breaks=c(0.5, 1.03, 1.55),labels=c("low", "medium", "high")) +
  geom_point(data = records_o, fill = NA, col = "white", size = 0.7, alpha = 0.015, shape = 3) +
  theme_bw(10) + 
  ggtitle("c.  Otway Ranges; red fox") +
  theme(axis.title = element_blank())

#### CAT 
### GLENELG
# predict model results into dataframe
data_g_plot <- cbind(data_g_plot, predict.gam(gam_cat_sp, newdata = data_g_plot, se.fit = TRUE, type = "link", exclude = c("s(station)", "s(survey_duration)")))
data_g_plot <- rename(data_g_plot, cat_predicted = fit,  cat_predicted_se = se.fit) # rename
# give each row / location a unique name to group by
data_g_plot$location_id <- paste0(data_g_plot$x, "_", data_g_plot$y)
# get difference in min max diel activity for each row
data_g_plot <- data_g_plot %>% 
  group_by(location_id) %>% 
  mutate(diel_diff = max(cat_predicted) - min(cat_predicted))

# cat plot: spatial activity per hour
g_cat_plot_hr <- ggplot(aes(x, y, fill = cat_predicted),
                     data = data_g_plot) +
  geom_tile()+
  scale_fill_viridis("log(activity)", option = "viridis", limits = c(-5.27, -1.13)) +
  geom_point(data = records_g, fill = NA, col = "white", size = 0.7, alpha = 0.1, shape = 3) +
  theme_bw(10) + 
  theme(axis.title = element_blank()) + 
    labs(title = 'Glenelg region; feral cat; Hour: {frame_time}:00') +
  transition_time(hour) +
  ease_aes('linear')

# cat plot: spatial difference
g_cat_plot_diff <- ggplot(aes(x, y, fill = diel_diff),
                     data = data_g_plot) +
  geom_tile()+
  scale_fill_viridis("Relative diel pattern strength", option = "viridis", limits = c(0.5, 1.55), breaks=c(0.5, 1.03, 1.55),labels=c("low", "medium", "high")) +
  geom_point(data = records_g, fill = NA, col = "white", size = 0.7, alpha = 0.015, shape = 3) +
  theme_bw(10) + 
  ggtitle("b.  Glenelg region; feral cat") +
   theme(axis.title = element_blank())

### OTWAYS
# predict model results into dataframe
data_o_plot <- cbind(data_o_plot, predict.gam(gam_cat_sp, newdata = data_o_plot, se.fit = TRUE, type = "link", exclude = c("s(station)", "s(survey_duration)")))
data_o_plot <- rename(data_o_plot, cat_predicted = fit,  cat_predicted_se = se.fit) # rename
# give each row / location a unique name to group by
data_o_plot$location_id <- paste0(data_o_plot$x, "_", data_o_plot$y)
# get difference in min max diel activity for each row
data_o_plot <- data_o_plot %>% 
  group_by(location_id) %>% 
  mutate(diel_diff = max(cat_predicted) - min(cat_predicted))

# cat plot: spatial activity per hour
o_cat_plot_hr <- ggplot(aes(x, y, fill = cat_predicted),
                     data = data_o_plot) +
  geom_tile()+
  scale_fill_viridis("log(activity)", option = "viridis", limits = c(-5.27, -1.13)) +
  geom_point(data = records_o, fill = NA, col = "white", size = 0.7, alpha = 0.1, shape = 3) +
  theme_bw(10) + 
  theme(axis.title = element_blank()) + 
    labs(title = 'Otway Ranges; feral cat; Hour: {frame_time}:00') +
  transition_time(hour) +
  ease_aes('linear')

# cat plot: spatial difference
o_cat_plot_diff <- ggplot(aes(x, y, fill = diel_diff),
                     data = data_o_plot) +
  geom_tile()+
  scale_fill_viridis("Relative diel pattern strength", option = "viridis", limits = c(0.5, 1.55), breaks=c(0.5, 1.03, 1.55),labels=c("low", "medium", "high")) +
  geom_point(data = records_o, fill = NA, col = "white", size = 0.7, alpha = 0.015, shape = 3) +
  theme_bw(10) + 
  ggtitle("d.  Otway Ranges; feral cat") +
  theme(axis.title = element_blank())

```

```{r x, echo=FALSE, out.width = '80%'}
g_fox_plot_hr 
```

```{r , echo=FALSE, out.width = '80%'}
g_cat_plot_hr 
```

```{r , echo=FALSE, out.width = '80%'}
o_fox_plot_hr 
```

```{r , echo=FALSE, out.width = '80%'}
o_cat_plot_hr
```

```{r space plot diff, echo=FALSE}
# save 
png("figs/diel_strength_600dpi.png", width = 10, height = 9, res = 600, units = "in")
ggarrange(g_fox_plot_diff, g_cat_plot_diff, o_fox_plot_diff, o_cat_plot_diff, ncol=2, nrow=2, common.legend = TRUE, legend="bottom", heights=c(4.25, 3.25))
dev.off()

```

## Spatiotemporal activity across vegetation types

```{r , cache=TRUE}
gam_fox_veg <- bam(fox ~ s(hour, bs = "cc", k = 8) +   
                         s(hour, vegetation_group, bs = "fs", xt = list(bs = "cc"), k = 8) + 
                         s(foxbaits, region, bs = "fs", xt = list(bs = "tp"), k = 4) + 
                         s(region, bs = "re") +  
                         s(station, bs = "re") +  
                         offset(log(survey_duration)), 
                   data = records, family = nb, knots = list(hour = c(0, 23)), nthreads = 3, discrete = TRUE)

summary(gam_fox_veg)

gam_cat_veg <- bam(cat ~ s(hour, bs = "cc", k = 8) +   
                         s(hour, vegetation_group, bs = "fs", xt = list(bs = "cc"), k = 8) +  
                         s(region, bs = "re") +  
                         s(station, bs = "re") +  
                         offset(log(survey_duration)), 
                   data = records, family = nb, knots = list(hour = c(0, 23)), nthreads = 3, discrete = TRUE)

summary(gam_cat_veg)
```

```{r , echo=FALSE}
# PLOT MODEL 1 ------------------------------------------------------------------
# get average sunrise / sunset times for plots
#mean(times(records$sunrise))
#mean(times(records$sunset))

# function to run and plot GAMs for each species
gam_plot <- function(gam_model, data){
  ## Plot global smooth (using the gratia package)
  plot_global <- draw(evaluate_smooth(gam_model, overall_uncertainty = TRUE, "s(hour)"), xlab = "Hour",  title = "", subtitle = "i.   Average") + 
    geom_vline(xintercept = c(6.16,18.34), colour = "black", size = 0.6, linetype="dotted") 
  ## Plot the difference smooths using the gratia package
  plot_difference <- draw(evaluate_smooth(gam_model, "s(hour,vegetation_group)"), xlab = "Hour",  title = "", subtitle = "ii.   Vegetation-specific deviations from the average") + 
    geom_vline(xintercept = c(6.16,18.34), colour = "black", size = 0.6, linetype="dotted") 
  ## Plot the actual hourly activity for each vegetation type   
  # make a new dataframe to predict into
  df <- expand.grid(hour = 0:23,
                    vegetation_group = levels(data$vegetation_group),
                    foxbaits = 0, 
                    region = c("glenelg", "otways"),
                    survey_duration = mean(data$survey_duration))
  # get smooths to exlude
  x = sapply(gam_model$smooth, "[[",  "label")
  # predict model estimates (and uncertainty) into this dataframe, excluding the impact of variables other than hour:
  df <- cbind(df, predict.gam(gam_model, newdata = df, se.fit = TRUE, type = "link", newdata.guaranteed = TRUE, exclude = c(x[3:7], "s(survey_duration)")))
  plot_veg <- ggplot(data=df, aes(x=hour, y=fit, group=vegetation_group)) +
    # ylim(-13, -2) +
    geom_line(aes(y=fit, x=hour), lwd = 0.7) +
    facet_wrap(~vegetation_group, nrow =1) +
    geom_ribbon(aes(ymin=(fit-2*se.fit), ymax=(fit+2*se.fit)), alpha=0.2) +
    geom_vline(xintercept = c(6.16,18.34), colour = "black", size = 0.6, linetype="dotted") + 
    labs(subtitle = "iii.   Predicted activity", x = "Hour", y = "log(count)") 
  ## Assemble plots
  patch <- (plot_global | plot_difference) / plot_veg 
  patch
}

# save
png("figs/fox_veg.png", width = 9, height = 5, res = 600, units = "in")
gam_plot(gam_fox_veg, records) + plot_annotation(title = "a.   Fox")
dev.off()

png("figs/cat_veg.png", width = 9, height = 5, res = 600, units = "in")
gam_plot(gam_cat_veg, records) + plot_annotation(title = "b.   Feral cat")
dev.off()

# to combine figures, type in the terminal (using imagemagick): 
# convert figs/fox_veg.png figs/cat_veg.png  -append figs/predator_veg.png

```

## Spatiotemporal activity in response to apex predator threat

```{r , cache = TRUE}
gam_cat_fox <- bam(cat ~ habitat_type + t2(hour, fox_count_adj, by = habitat_type, bs = c("cc", "ts"), k = c(8, 5), full = TRUE) +  
                    s(longitude, latitude, bs = "ds",  m = c(1, 0.5), k = 200) +
                    s(station, bs = "re") +  
                    offset(log(survey_duration)), 
                  data = records, family = nb, knots = list(hour = c(0, 23)), nthreads = 3, discrete = TRUE)

summary(gam_cat_fox)
```

```{r , echo=FALSE}

# PLOT MODEL 2B ------------------------------------------------------------------
df = expand.grid(hour = 0:23,
                 habitat_type = levels(records$habitat_type),
                 longitude = mean(records$longitude),
                 latitude = mean(records$latitude),
                 fox_count_adj = seq(min(records$fox_count_adj), max(records$fox_count_adj), length=150),
                 station = "T052",
                 survey_duration = mean(records$survey_duration))
# predict model results into dataframe
x = sapply(gam_cat_fox$smooth, "[[",  "label")
df_int <- cbind(df, predict(gam_cat_fox, newdata = df, se.fit = TRUE, type = "link", exclude = c(x[4:5], "s(survey_duration)")))
# plot
plot_range <- ggplot(aes(hour, fox_count_adj, fill = fit), data = df_int) +
  geom_tile() +
  facet_wrap(~habitat_type, nrow =1) +
  scale_fill_viridis("log(count)", option = "viridis") + 
  ggtitle("", subtitle = "Feral cat activity") +
  geom_vline(xintercept = c(6.16,18.34), colour = "grey75", size = 1) + 
  ylab("log(adjusted fox count)") + 
  xlab("Hour")

# plot standard error 
plot_range_se <- ggplot(aes(hour, fox_count_adj, fill = se.fit), data = df_int) +
  geom_tile() +
  facet_wrap(~habitat_type, nrow =1) +
  scale_fill_viridis("Standard error", option = "viridis") + 
  ggtitle("", subtitle = "Uncertainty") +
  geom_vline(xintercept = c(6.16,18.34), colour = "grey75", size = 1) + 
  ylab("log(adjusted fox count)") + 
  xlab("Hour")

#  save
png("figs/cat_fox_count.png", width = 9, height = 6.5, res = 600, units = "in")
plot_range / plot_range_se + plot_annotation(tag_levels = "a")
dev.off()

```

## Individual heterogeneity in diel activity

```{r , echo=FALSE}
## LOAD PREPARE INDIVIDUAL CAT DATA FROM MATTS PHD

# load cat records
records <- read.csv("raw_data/cat_id_detections.csv")
records$X <- NULL

# load camdata - but just matts deployments
camdata <- read.csv("raw_data/camdata.csv")
camdata <- filter(camdata, data_source == "matt")
camdata$X <- NULL
camdata$X.1 <- NULL

# ADD SOLAR TIMES ---------------------------------------------------------
# take a matrix of the coordinates
coords <- matrix(c(records$longitude, records$latitude), nrow = length(records$latitude))
# add a coordinate reference system (can't be UTM)
coords <- sp::SpatialPoints(coords, proj4string=sp::CRS("+proj=longlat +datum=WGS84"))

# specify detection time as a date class, as well as timezone (use Brisbane because it doesn't have daylight savings)
records$date_time <- ymd_hms(records$date_time, tz = "Australia/Brisbane")

# calculate the time of sunrise given the location and timing of each species detection 
sunriset_df <- as.data.frame(sunriset(coords, records$date_time, direction = "sunrise", POSIXct.out=TRUE))
# convert to radian time 
sunrise_radian <- ClocktimeToTimeRad(sunriset_df$time) 
# calculate the time of sunset given the location and timing of each species detection 
sunsett_df <- as.data.frame(sunriset(coords, records$date_time, direction = "sunset", POSIXct.out=TRUE))
# convert to radian 
sunset_radian <- ClocktimeToTimeRad(sunsett_df$time)

# save these as a separate matrix for the transtime function below 
anchors <- matrix(c(sunrise_radian, sunset_radian), nrow = length(sunset_radian))

# we also need to convert the detection time to radian time
det_radian <- ClocktimeToTimeRad(records$date_time)

# we can now calculate the average anchored time (time relative to both sunrise and sunset) for every detection
det_anchored <- transtime(det_radian, anchors)

# add to dataframe
records$aa_radian <- det_anchored
# as well as in clock time (new cols)
records$aa_clock <- records$aa_radian * 3.81971

# REMOVE REPEAT DETECTIONS ------------------------------------------------

## remove records of the same species at a particular cam-trap within 30 minutes
time_to_independence <- 1800 # number of seconds in 30 minutes

# get time (in seconds) since last visit
records <- records %>%
  group_by(station_year, individual) %>%
  arrange(date_time) %>%
  mutate(seconds_diff = date_time - lag(date_time))

# change NA's to a value greater than the time to independence so they get kept when we filter out 
records$seconds_diff[is.na(records$seconds_diff)] <- time_to_independence + 1

# filter out records less than 60 minutes
records <- filter(records, seconds_diff > time_to_independence)

# remove seconds_diff col
records$seconds_diff <- NULL

# SPLIT COMBINED CATS ---------------------------------------------------------
# duplicate row and split individual name
records_double1 <- filter(records, individual == "otway_swirl_venus_&_otway_swirl_peter" | individual == "annya_swirl_ben_&_annya_swirl_motley")
records_double2 <- filter(records, individual == "otway_swirl_venus_&_otway_swirl_peter" | individual == "annya_swirl_ben_&_annya_swirl_motley")
records_double1$individual <- str_split_fixed(records_double1$individual, "_&_", 2)[,1]
records_double2$individual <- str_split_fixed(records_double2$individual, "_&_", 2)[,2]
records_double1$coat_type <- str_split_fixed(records_double1$coat_type, "_&_", 2)[,1]
records_double2$coat_type <- if_else(records_double2$coat_type == "tortoise_&_tabby", "tabby", as.character(records_double2$coat_type))
records_double <- rbind(records_double1, records_double2)

# remove these rows from records
records <- filter(records, is.na(individual) | individual != "otway_swirl_venus_&_otway_swirl_peter")
records <- filter(records, is.na(individual) | individual != "annya_swirl_ben_&_annya_swirl_motley")

# now add the new ones in 
records <- rbind(records, records_double)

# RESHAPE BY HOUR ---------------------------------------------------------
# DATAFRAME 1:  HOUR ----------------------------------------------------
# need to do this for summing
records$occ <- 1

# make solar hour an integer, of character class
records$hour <- as.integer(records$aa_clock)
records$species <- "cat"

# split by region
records_g <- filter(records, region == "glenelg")
records_o <- filter(records, region == "otways")
# do same for camdata
camdata_g <- filter(camdata, region == "glenelg")
camdata_o <- filter(camdata, region == "otways")

# melt the dataframe
gam_data_g <- melt(records_g, id.var = colnames(records_g), measure.var = "occ")
gam_data_o <- melt(records_o, id.var = colnames(records_o), measure.var = "occ")

# cast to derive count per year per station_year long format
gam_data_g = dcast(gam_data_g,  individual + station_year + hour ~ species, fill = 0, fun = sum)
gam_data_o = dcast(gam_data_o,  individual + station_year + hour ~ species, fill = 0, fun = sum)

# make a table with all unique combinations - (use camdata so we don't just take the sites with detections)
df_g <- expand.grid(station_year = unique(camdata_g$station_year), individual = unique(records_g$individual), hour = 0:23)
df_o <- expand.grid(station_year = unique(camdata_o$station_year), individual = unique(records_o$individual), hour = 0:23)

# merge into gam_data so every hour for every station is provided (even with no predator detections) - fills missing values with NA
gam_data_g <- left_join(df_g, gam_data_g)
gam_data_o <- left_join(df_o, gam_data_o)

# change NA's to 0's
gam_data_g$cat <- ifelse(is.na(gam_data_g$cat), 0, gam_data_g$cat)
gam_data_o$cat <- ifelse(is.na(gam_data_o$cat), 0, gam_data_o$cat)

# sort by station, hour
gam_data_g <- arrange(gam_data_g, individual, station_year, hour)
gam_data_o <- arrange(gam_data_o, individual, station_year, hour)

## merge in site covariates
gam_data_g <- left_join(gam_data_g, camdata_g, by = "station_year")
gam_data_o <- left_join(gam_data_o, camdata_o, by = "station_year")

# ADD SUNSET SUNRISE TIMES FOR STATIONS ------------------------------------------
## glenelg 
# make date class
#gam_data_g$date_start <- ymd(gam_data_g$date_start, tz = "Australia/Brisbane")
#gam_data_g$date_end <- ymd(gam_data_g$date_end, tz = "Australia/Brisbane")
## take a matrix of the coordinates
#coords <- matrix(c(gam_data_g$longitude, gam_data_g$latitude), nrow = length(gam_data_g$latitude))
## add a coordinate reference system (can't be UTM)
#coords <- sp::SpatialPoints(coords, proj4string=sp::CRS("+proj=longlat +datum=WGS84")) 
## calculate the time of sunrise given the location and timing of each species detection - midway point for survey
#sunriset_df <- as.data.frame(sunriset(coords, gam_data_g$date_start + days(as.integer(gam_data_g$survey_duration / 2)), direction = #"sunrise", POSIXct.out=TRUE))
## calculate the time of sunset given the location and timing of each species detection - midway point for survey
#sunsett_df <- as.data.frame(sunriset(coords, gam_data_g$date_start + days(as.integer(gam_data_g$survey_duration / 2)), direction = #"sunset", POSIXct.out=TRUE))
## load hms r package here so it does't conflict with lubridate^
#library(hms)
## add time only to dataframe using hms package
#gam_data_g$sunrise <- hms::as_hms(lubridate::ymd_hms(sunriset_df$time, tz = "Australia/Brisbane"))
#gam_data_g$sunset <- hms::as_hms(lubridate::ymd_hms(sunsett_df$time, tz = "Australia/Brisbane"))
#
## otways
## make date class
#gam_data_o$date_start <- ymd(gam_data_o$date_start, tz = "Australia/Brisbane")
#gam_data_o$date_end <- ymd(gam_data_o$date_end, tz = "Australia/Brisbane")
## take a matrix of the coordinates
#coords <- matrix(c(gam_data_o$longitude, gam_data_o$latitude), nrow = length(gam_data_o$latitude))
## add a coordinate reference system (can't be UTM)
#coords <- sp::SpatialPoints(coords, proj4string=sp::CRS("+proj=longlat +datum=WGS84")) 
## calculate the time of sunrise given the location and timing of each species detection - midway point for survey
#sunriset_df <- as.data.frame(sunriset(coords, gam_data_o$date_start + days(as.integer(gam_data_o$survey_duration / 2)), direction = #"sunrise", POSIXct.out=TRUE))
## calculate the time of sunset given the location and timing of each species detection - midway point for survey
#sunsett_df <- as.data.frame(sunriset(coords, gam_data_o$date_start + days(as.integer(gam_data_o$survey_duration / 2)), direction = #"sunset", POSIXct.out=TRUE))
## add time only to dataframe using hms package
#gam_data_o$sunrise <- hms::as_hms(lubridate::ymd_hms(sunriset_df$time, tz = "Australia/Brisbane"))
#gam_data_o$sunset <- hms::as_hms(lubridate::ymd_hms(sunsett_df$time, tz = "Australia/Brisbane"))
#
## get averages (use in plot)
#library(chron)
#mean(times(gam_data_g$sunrise))
#mean(times(gam_data_g$sunset))
#mean(times(gam_data_o$sunrise))
#mean(times(gam_data_o$sunset))

## Transform variable class  for GAMs
gam_data_o <- transform(gam_data_o,
                       hour = as.integer(hour),
                       individual = factor(individual, ordered = FALSE), 
                       station = factor(station, ordered = FALSE), 
                       survey_duration = as.integer(survey_duration)
)

gam_data_g <- transform(gam_data_g,
                        hour = as.integer(hour),
                        individual = factor(individual, ordered = FALSE), 
                        station = factor(station, ordered = FALSE), 
                        survey_duration = as.integer(survey_duration)
)

#levels(gam_data_g$individual)
#levels(gam_data_o$individual)

```

```{r , cache=TRUE}

gam_cat_ind_o <- bam(cat ~ s(hour, bs = "cc", k = 8) +   
                           s(hour, individual, bs = "fs", xt = list(bs = "cc"), k = 8) +  
                           s(longitude, latitude, bs = "ds",  m = c(1, 0.5), k = 150) +
                           s(station, bs = "re") +  
                           offset(log(survey_duration)), 
                     data = gam_data_o, family = nb, knots = list(hour = c(0, 23)), nthreads = 3, discrete = TRUE)

summary(gam_cat_ind_o)

gam_cat_ind_g <- bam(cat ~ s(hour, bs = "cc", k = 8) +   
                           s(hour, individual, bs = "fs", xt = list(bs = "cc"), k = 8) +  
                           s(longitude, latitude, bs = "ds",  m = c(1, 0.5), k = 150) +
                           offset(log(survey_duration)), 
                     data = gam_data_g, family = nb, knots = list(hour = c(0, 23)), nthreads = 3, discrete = TRUE)

summary(gam_cat_ind_g)
```

```{r , echo=FALSE}
# plot individuals
## 1st get top 7 cats with most detections
top7_g <- slice_max(as.data.frame(table(records_g$individual)), order_by = Freq, n = 7, with_ties = FALSE)
top7_o <- slice_max(as.data.frame(table(records_o$individual)), order_by = Freq, n = 7, with_ties = FALSE)

## glenelg
## Plot global smooth (using the gratia package)
plot_global <- draw(evaluate_smooth(gam_cat_ind_g, overall_uncertainty = TRUE, "s(hour)"), xlab = "Hour",  title = "", subtitle = "i.   Average") + 
  geom_vline(xintercept = c(7.01,18.12), colour = "black", size = 0.6, linetype="dotted") 
## Plot the difference smooths using the gratia package
plot_difference <- draw(evaluate_smooth(gam_cat_ind_g, "s(hour,individual)"), xlab = "Hour",  title = "", subtitle = "ii.   Individual deviations from the average") + 
  geom_vline(xintercept = c(7.01,18.12), colour = "black", size = 0.6, linetype="dotted") 
## Plot the actual hourly activity for each vegetation type   
# make a new dataframe to predict into
df <- expand.grid(hour = 0:23,
                  individual = unique(top7_g$Var1),
                  survey_duration = mean(gam_data_g$survey_duration))
# get smooths to exlude
x = sapply(gam_cat_ind_g$smooth, "[[",  "label")
# predict model estimates (and uncertainty) into this dataframe, excluding the impact of variables other than hour:
df <- cbind(df, predict.gam(gam_cat_ind_g, newdata = df, se.fit = TRUE, type = "link", newdata.guaranteed = TRUE, exclude = c(x[3:6], "s(survey_duration)")))
# shorten cat names
df$individual <- as.character(df$individual)
df$individual <- recode(df$individual, annya_swirl_lola = "faith", c_tabby_quasimodo = "brisket", m_swirl_socksy = "boots", annya_swirl_ben = "dolly", annya_swirl_motley = "motley", hotspur_tabby_1 = "sprout", m_tabby_murray = "murray")
# plot top 7 individuals
plot_ind <- ggplot(data=df, aes(x=hour, y=fit, group=individual)) +
  geom_line(aes(y=fit, x=hour), lwd = 0.7) +
  facet_wrap(~individual, nrow =1) +
  geom_ribbon(aes(ymin=(fit-2*se.fit), ymax=(fit+2*se.fit)), alpha=0.2) +
  geom_vline(xintercept = c(7.01,18.12), colour = "black", size = 0.6, linetype="dotted") + 
  labs(subtitle = "iii.   Predicted activity", x = "Hour", y = "log(count)")

# plot / save
png("figs/cat_ind_g.png", width = 9, height = 5, res = 600, units = "in")
(plot_global | plot_difference) / plot_ind  + plot_annotation(title = "a.   Glenelg region")
dev.off()

## otways
## Plot global smooth (using the gratia package)
plot_global <- draw(evaluate_smooth(gam_cat_ind_o, overall_uncertainty = TRUE, "s(hour)"), xlab = "Hour",  title = "", subtitle = "i.   Average") + 
  geom_vline(xintercept = c(7.27,17.38), colour = "black", size = 0.6, linetype="dotted") 
## Plot the difference smooths using the gratia package
plot_difference <- draw(evaluate_smooth(gam_cat_ind_o, "s(hour,individual)"), xlab = "Hour",  title = "", subtitle = "ii.   Individual deviations from the average") + 
  geom_vline(xintercept = c(7.27,17.38), colour = "black", size = 0.6, linetype="dotted") 
## Plot the actual hourly activity for each vegetation type   
# make a new dataframe to predict into
df <- expand.grid(hour = 0:23,
                  individual = unique(top7_o$Var1),
                  survey_duration = mean(gam_data_o$survey_duration))
# get smooths to exlude
x = sapply(gam_cat_ind_o$smooth, "[[",  "label")
# predict model estimates (and uncertainty) into this dataframe, excluding the impact of variables other than hour:
df <- cbind(df, predict.gam(gam_cat_ind_o, newdata = df, se.fit = TRUE, type = "link", newdata.guaranteed = TRUE, exclude = c(x[3:6], "s(survey_duration)")))
# shorten cat names
df$individual <- as.character(df$individual)
df$individual <- recode(df$individual, otway_tabby_javier = "pav", otway_swirl_skip = "skip", otway_tabby_crass = "crass", otway_tabby_meg = "pauline", otway_swirl_bluey = "bluey", otway_swirl_chowder = "chowder", otway_tabby_catsup = "percy")
# plot top 7 individuals
plot_ind <- ggplot(data=df, aes(x=hour, y=fit, group=individual)) +
  geom_line(aes(y=fit, x=hour), lwd = 0.7) +
  facet_wrap(~individual, nrow =1) +
  geom_ribbon(aes(ymin=(fit-2*se.fit), ymax=(fit+2*se.fit)), alpha=0.2) +
  geom_vline(xintercept = c(7.27,17.38), colour = "black", size = 0.6, linetype="dotted") + 
  labs(subtitle = "iii.   Predicted activity", x = "Hour", y = "log(count)")

# plot / save
png("figs/cat_ind_o.png", width = 9, height = 5, res = 600, units = "in")
(plot_global | plot_difference) / plot_ind  + plot_annotation(title = "b.   Western Otway Ranges")
dev.off()

# to combine figures, type in the terminal (using imagemagick): 
# convert figs/cat_ind_g.png figs/cat_ind_o.png  -append figs/cat_ind.png

```

## Session information

```{r , echo = FALSE}

sessionInfo() %>% pander()

```

### 2021-09-05 manuscript/manuscript.Rmd
fox ~ ts, region ~ re

--- 2021-08-03 manuscript/manuscript.Rmd
@@
 ---
-title: Joint spatiotemporal models reveal dynamic shifts in mesopredator spatial and diel activity for apex predator avoidance
+title: "Quantifying spatiotemporal shifts in predator behaviour across habitats, individuals and threats." 
 author:
@@
     - \floatplacement{figure}{H}
-linenumbers: true
+linenumbers: false
 #numbersections: true
@@
 
-Effective wildlife management relies on understanding species population and behavioural dynamics to accurately predict intervention outcomes.
-These dynamics depend on both environmental conditions and those of other species, but few statistical approaches incorporate both factors, or allow responses to be dynamic across space and time.
-This is key because species interactions and niche preferences are multidimensional.
+Diel activity patterns - the distribution of activity throughout the daily cycle - are an important but overlooked aspect of behavioural ecology.
+Animals must allocate particular times of the day to rest, forage, socialise and reproduce, which are traded-off against threat avoidance (e.g.
+predation) and deterrence against resource loss (i.e.
+inter and intraspecific competition).
+Diel activity can therefore vary spatially across different environmental conditions (e.g.
+food, shelter availability) and species assemblages.
+This is important to understand because constrained diel behaviour can incur fitness costs to individuals and populations (Lamb 2020 PNAS).
 
-Fear is a near universal driver of animal behaviour.
-Fear may be from other individuals of their species, a dominant species (including humans) or an event (e.g.
-fire).
-Animals may be naive to threats (e.g.
-alien predators), unable to respond to threats (e.g.
-scarce prey), or the threat may be too minimal to warrant a response.
-Some environments more easily facilitate avoidance (e.g.
-high food and shelter accessibility), and so accounting for landscape context is key.
-A threat response may be routine, or only employed or enhanced when the threat is heightened.
-This likely depends on how constant and predictable the threat is.
+Flexible spatiotemporal behaviour can allow animals to adapt to changing conditions.
+For example, elk in Yellowstone National Park use risky, open parts of the landscape when the activity of reintroduced wolves is low, suggesting coexistence without major fitness costs (Kohl et al.
+2019).
+Diel activity decisions are clearly linked to habitat type, and there is growing recognition of individual-level variation in these decisions (Basille et al 2015; time-limited seabirds; urban bears).
+This is important to consider because populations with high individual heterogeneity are likely more adaptable to threats (Wolf and Weissing 2012; Dingemanse and Wolf 2013).
+Most studies examine spatiotemporal behavioural responses by repeating spatial analyses at different time periods (generally night and day), but few consider continuous shifts in diel activity across spatial contexts.
 
-Animals may respond to threats by altering their activity patterns (i.e.
-avoidance).
-An animal may avoid locations where the threat is likely to occur (hereafter ‘spatial avoidance’), avoid times of the day when a threat is likely to occur (hereafter ‘temporal avoidance’), or avoid both locations and times a threat is heightened (hereafter ‘spatiotemporal avoidance’).
-Such avoidance likely has trade-offs, which may or may not incur costs to individual and population fitness.
-Modelling approaches which do not allow spatiotemporal and dynamic fear reponses can only make limited inference about species interactions.
+Diel activity patterns are increasingly being considered as technology like telemetry devices and camera-traps have become more widely available.
+Telemetry devices like GPS collars and accelerometers provide high resolution data at the individual-level.
+However, resource constraints generally limit these studies to few individuals (Harmsen 2009), and target similar personalities which are susceptible to physical capture.
+Camera-traps are particularly useful because they concurrently survey multiple species--including the elusive individuals--providing inference on ecological differences and species interactions.
+However, individual-level inference would be limited to species with unique markings.
+Distiller et al 2020 recently used camera-traps to individually identify jaguars and estimate sex difference in diel activity patters.
+Despite having information on individual diel activity patterns, studies predominantly estimate at the population-level due to small sample sizes.
 
-Species interactions are usually considered in degrees of overlap between activity patterns.
-Spatiotemporal avoidance is most often considered, using joint species distribution models (including MSOMs).
-However, temporal avoidance is increasingly being used (perhaps just allowed due to widely available GPS and camera technology) with kernel density estimation methods.
-Although, these methods don’t allow environmental covariates, and are particularly susceptible to noisy estimates.
-Importantly, some are beginning to combine these methods to allow spatiotemporal responses (Azzou et al 2021; Farris et al 2020).
-However, degrees of overlap in activity patterns only get at routine avoidance, we are generally interested in whether changes in threat levels lead to altered behaviour.
-This is more indicative of species interactions because static levels of overlap may just reflect different niche preferences rather than avoidance.
+A variety of statistical approaches are used to examine how diel activity patterns vary among habitat types, individuals and co-occurring species.
+Kernel density estimation is most often used to estimate hourly activity curves (Linkie & Ridout, 2011), however this methods is susceptible to noise from sample sizes (Iannaerlli) and does allow interactions with covariates (Frey 2017).
+They remain popular because largely because they estimate a coefficient of overlap between two diel activity curves - either between two species, or one species in two different contexts.
+However, rarely are there only two interacting species or categorical contexts.
+Joint approaches with spatial overlap coefficients are being developed (e.g.
+Azzou et al 2021; Farris et al 2020) to account for the spatiotemporal nature of species avoidance strategies.
+However, activity overlap coefficients may just reflect different niche preferences.
+A threat response may not be routine, but only employed or enhanced when the threat is heightened.
+And so, whether an animal *changes* its activity patterns with an increasing threat level (e.g.
+higher apex predator activity) is more interesting and relevant to management than arbitrary overlap.
 
-Dynamic behavioural responses are also important from a management perspective because they want to know whether an action will cause a change in animal behaviour /  range?).
-For example, after the reintroduction of wolves in Yellowstone, it was widely thought that this had limited elk populations.
-Using GPS data and habitat selection methods,  Kohl et al.
-2018 found that elk used riskier parts of landscape at times when wolves were least active -- suggesting coexistence without fitness costs.
-However, there are many other large carnivores in this ecosystem which are likely to impact elk behaviour.
-And so, will elk respond to wolves in the same way in different environmental contexts?
-There are little-no examples of replicated spatiotemporal interactions studies.
+Straightforward, integrated methods for examining changes in spatiotemporal animal behaviour are needed.
+Generalised additive models (hereafter 'GAMs') are increasingly used to estimate diel activity patterns (e.g xx).
+GAMs learn from data to specify nonlinear relationships between response and explanatory variables.
+Unlike kernel density estimation (Linke and Ridout), GAMs include smoothing penalties to avoid overfitting (making them less susceptible to noisy estimates due to small sample sizes) and allow diel activity to vary with explanatory variables.
+GAMs are also being used to test spatial associations between species (e.g.
+my chapter 2) as a dominant species may only suppress subordinate species at high densities (Johnson).
+Despite this potential, diel activity curves are rarely modelled with categorical or continuous covariates, particularly using another species as that covariate (although see Cunningham).
+Further, GAMs can share information across conditions using heirarchical smooths, which estimate an average response, as well as group-level estimates [@pedersen2019hierarchical].
+Group-level estimates are more penalised as they deviate from the average, allowing confidence that these are true deviations across groups (e.g individuals, categorical environmental conditions).
 
-There is contention around the interactions between two invasive predators - feral cats and foxes.
-Foxes are routinely culled, and there is concern that this could lead to a mesopredator release of feral cats.
-Mesopredator release may manifest in both behavioural and numerical changes, which my previous chapter identified - but that behavioral responses (ranging and detectability) were not universal.
-Here, we aim to dig into the behavioral response further - in particular diel activity times.
-
-In this study, we used generalised additive models to assess (1) how predator (fox and cat) diel activity varies between vegetation types, and (2) whether an introduced mesopredator (cat) changes its diel and/or spatial activity in response to the activity of an introduced apex predator (fox).
-Because foxes and cats fill a similar function niche, we expected that, on average, they would have similar diel activity patterns across vegetation types (hypothesis 1).
-However, where fox detections were high, we hypothesised that the subordinate cat to be less active (hypothesis 2) and shift diel activity to times when foxes are least active (hypothesis 3).
-We tested this on a simple predator system where only cats and foxes occur, and fox activity is manipulated through lethal control.
-This allowed focus on the sole interactions between these two predators, and ensured differences in correlations wasn’t just due to differences in niche preferences.
+Here we provide a case-study to illustrate how generalised additive models can be used to assess spatiotemporal behaviour of two predator species.
+The Red fox *Vulpes vulpes* (hereafter 'fox') and feral cats *Felis catus* (hereafter 'cat') have some of the widest distributions of any carnivores, and co-occur across most of their global range.
+Because of their devastating impacts on native prey in their introduced range, understanding differences in their behavioural ecology across space, and how foxes may impact feral cat behaviour is key to effective management.
+Numerous studies have tested spatial and temporal associations between these invasive predators and environmental covariates, but not in a spatially explicit or joint modelling framework.
+We therefore investigated how invasive predator activity varies across (1) space and (2) vegetation types, as well as how feral cat activity varies across fox activity levels (3) and individuals in two distinct study regions.
+We tested this in a simple predator system where only cats and foxes occur, and fox activity is manipulated through lethal control.
+This allowed focus on the sole interactions between these two predators, and ensuring correlations doesn't simply reflect different niche preferences.
 
@@
 ## Study area
+We collated camera-trap data on these two predator species from three distinct studies across two regions of south-west Victoria, Australia (Fig.
+\@ref(fig:diel-map)).
+The terrestrial mammalian predator guild is depauperate throughout both regions.
+Dingoes *Canis familiaris* are now absent throughout, as is the tiger quoll *Dasyurus maculatus*, more recently absent in the Otway Ranges (last sighted in 2014 despite extensive camera-trapping).
+Introduced foxes and cats are therefore the only medium-large functional mammalian terrestrial predators across both regions.
 
-We collated camera-trap data on these two predator species from three distinct studies across two regions of south-west Victoria, Australia (Fig.
-1).
-The purpose of each study was to experimentally survey changes in the mammalian community due to fox control, and so our dataset covered a wide range contexts with and without fox-baiting, allowing us to investigate fox-cat interactions across a gradient of apex predator (fox) activity.
-
-The terrestrial mammalian predator guild is depauperate throughout both regions.
-Dingoes *Canis familiaris* are now absent throughout, as is the the tiger quoll *Dasyurus maculatus*, more recently absent in the Otway Ranges (last sighted in 2014 despite extensive camera-trapping).
-Introduced red foxes and feral cats are therefore the only medium-large functional mammalian terrestrial predators across both regions.
-
-In the baited parts of the regions, government land managers conduct ongoing targeted fox control for biodiversity conservation.
+The purpose of each collated study was to experimentally survey changes in the mammalian community due to fox control.
+In broad sections of each region, government land managers conduct ongoing targeted fox control for biodiversity conservation.
 Poison baits containing 3 mg of sodium mono-fluroacetate (1080) are buried at a depth of 10 cm at 1-km intervals along accessible forest tracks and roads.
+This allowed us to investigate fox-cat interactions across a gradient of apex predator (fox) activity and other landscape contexts.
 
 In the Glenelg region, Gunditjmara country, natural vegetation is highly fragmented.
-Here, camera-survey sites were situated in six distinct forest blocks: two are separated by a river, the others are separated by at least *XX* km of pastoral farming and residential property.
+Here, camera-survey sites were situated in six distinct forest blocks: two are separated by a river, the others are separated by at least **XX** km of pastoral farming and residential property (Fig.
+\@ref(fig:map)).
 Foxes in three of the forest blocks have been subject to lethal control since October 2005, with 1080 poison-baits replaced fortnightly [@ROBLEY2014262].
-
 The Otway Ranges, Gadubanud country, is a largely continuous patch of natural vegetation with a strong east-west rainfall and elevation gradient [@swan2021species].
@@
 The majority of this baiting began in 2016 - 2017, although fox-baiting commenced in some small sections in 2008.
-
 Alongside fragmentation, the extremely rugged, wet forest and rainforest matrix in the western Otway Ranges is the main difference between the two regions.
-Prescribed fire, primarily for the purpose of asset protection, is another key management action undertaken throughout both regions (except in very wet sections of Otway Ranges).
+Prescribed fire is another key management action undertaken throughout both regions (except in very wet sections of Otway Ranges), primarily for asset protection.
 
 ## Camera-trap surveys
-
 We compiled camera-trap data from three distinct studies across the two regions.
-After removing camera-traps which were operational for less than 14 days, this totalled 3659 deployments of camera-traps at 1232 camera-trap sites.
-Survey durations ranged between 14 and 93 days (mean 47), totalling 171,966 trapnights.
+After removing camera-traps which were operational for less than 14 days, this totalled 3694 deployments of camera-traps at 1232 camera-trap sites.
+Camera-traps were active for an average of 47 days, totalling 172,167 trapnights.
 Camera-trap spacing was variable; the average minimum distance between simultaneously deployed camera-traps was 445, 853 and 1266 metres for each study.
-Camera-trap were deployed in the Glenelg region between 2013 – 2019, whereas in the Otway Ranges surveys were conducted in 2016 – 2019.
+Camera-traps were deployed in the Glenelg region between 2013 – 2019, and in the Otway Ranges between 2016 – 2019.
 
@@
 To account for day length variation across space and time, we extracted sunrise and sunset times for each camera-trap deployment using the 'maptools' R-package [@maptools] and adjusted detection times to be relative to sunrise and sunset using the average double anchoring approach described by @https://doi.org/10.1111/2041-210X.13290.
-
 Using the 'reshape' R-package [@reshape], we manipulated the detection table into a dataframe with a row for each hour of the day, for every camera-trap deployment, recording the total number of 'independent' fox and feral cat detections within each hour for the entire camera-trap survey duration.
 
-We identified the broad vegetation type (Ecological Vegetation Class (hereafter 'EVC') group [@delwp2020]) at each camera-trap location.
-Six vegetation types were surveyed in both regions, although to varying degrees (Table 1).
+We attached the Ecological Vegetation Class (hereafter 'EVC') group [@delwp2020]) at each camera-trap location.
+**Six?** vegetation types were surveyed in both regions, although to varying degrees (Table 1).
 In the Otway region, rainforests are interspersed (primarily in low lying gullies) throughout wet and damp forests at a finescale, and we merged these groups (hereafter 'wet forests').
-
 For each deployment, we calculated the number of poison fox-baits within a 2.3 km radius around each camera-trap - the average maximum distance foxes in these regions travel from their home range centre [@hradsky2017human].
-
-## Generalised additive models 
-We modelled predator counts (response variable) with generalised additive mixed-effects models (hereafter 'GAMMs') implemented in the 'mgcv' R-package [@wood2017].
-We first fit these models using the poisson family, but switched to the negative binomial family as overdispersion, but not zero-inflation, was detected using the 'DHARMa' R-package [@DHARMa].
-We specified a model offset to account for differences in camera-trap survey duration, and a random intercept for each site to account for repeat sampling.
-We used a duchon spline static spatial smooth to model both spatial autocorrelation and differences in predator activity across space (i.e.
-the effect of unmodelled environmental covariates).
-For the fox model, we additionally included a thin plate regression spline for poison-bait density, with a factor smoother for each region, as bait replacement schedules differed between the regions.
-This formed the base model specification for each model we fitted.
-Models (n = 2) differed in their specification of the cyclical hour smooth in order to explore (1) variations in predator diel activity across EVC groups, as well as (2) whether feral cat diel activity changed across the gradient of observed fox counts (detailed in the sections below).
-We used the double penalty model selection approach, which penalises the null space in addition to the range space (i.e.
-shrinking wiggly terms to linear terms) of the spline basis, meaning covariate effects can be entirely removed from the model [@MARRA20112372].
-We plotted models using 'ggplot2' and 'gratia' R-packages [@ggplot2; @gratia].
-
-We expected the activity of each predator species to follow a broadly consistent pattern (e.g.
-lower during the day), but potentially have small shifts in relative activity between sunset and sunrise across the different EVC groups (1).
-Fitting separate responses for each EVC group would throw away shared information, which is particularly costly for the least surveyed groups, and would result in more noisy estimates.
-We therefore specified a global smoother for hour (i.e.
-average response), along with group-level smoothers with shared wiggliness for each EVC group [i.e.
-model GS in @pedersen2019hierarchical].
-This model structure penalises functions which deviate strongly from the average response, increasing confidence that they are true deviations opposed to the result of sampling noise.
-
 To investigate changes in feral cat diel activity in response to counts of foxes.
@@
 
-Analysis of stage 1 showed that fox diel activity patterns in wet forests differed relative to all other EVC groups.
-We therefore expected feral cats might to temporally avoid foxes in a different way in wet forests than the dry vegetation groups, but that responses would be similar in all the drier vegetation groups.
+## Generalised additive models
+We modelled the total predator counts for each camera-trap deployment (response variable) with generalised additive mixed-effect models implemented in the 'mgcv' R-package [@wood2017].
+GAMs learn from the data to specify nonlinear relationships between response and explanatory variables.
+A maximum wiggliness is specified for each explanatory variable, with complexity automatically reduced (ultimately to a linear function) through a smoothing penalty to avoid overfitting.
+We used REML estimation and the negative binomial family (as overdispersion, but not zero-inflation, was detected using the 'DHARMa' R-package [@DHARMa]).
+We specified a model offset to account for differences in camera-trap survey duration, and a random intercept for each site to account for repeat sampling.
+For fox models (n = 2), we included a smooth effect of poison-bait density using a thin plate regression spline basis.
+This formed the base model specification for each model we fitted.
+Models differed in their specification of the cyclical hour smooth in order to provide inference on variations of predator diel activity across (1) space (i.e.
+coordinates) (2) EVC groups, as across for feral cats across the range of observed fox counts (3), and across each identified individual cat (4), detailed in the sections below.
+This model specification allowed both spatial activity to be higher in each context (i.e.
+across coordinates, vegetation types, fox activity levels, individual cats) in addition to variable diel activity patterns.
+Code used to specify each model set is presented in the Supporting Information.
+We plotted models using 'ggplot2' and 'gratia' R-packages [@ggplot2; @gratia].
+
+### Spatial variation in predator activity
+We fit a model separately for both foxes and cats which included a tensor product interaction between a spatial smooth and hourly smooth.
+This allowed predators to have different activity levels across space (static across the years surveyed), as well as a variation in diel activity across space.
+Space was modelled using camera-trap coordinates and a duchon spline basis (Miller and Wood).
+To compare relative diel activity pattern strengths--how much activity patterns vary across the daily cycle--we plotted the difference between the minimum and maximum activity estimate for every predicted location.
+
+### Variation in predator activity across vegetation types
+We estimated diel activity across EVC group using a hierarchical model specification.
+This included a global smoother for hour (i.e.
+average response), along with group-level smoothers for each EVC group with shared shared wiggliness [i.e.
+model GS in @pedersen2019hierarchical].
+This model structure penalises functions which deviate strongly from the average response.
+
+### Feral cat spatiotemporal avoidance of foxes
+Analysis of fox diel activity across EVC groups (section 4.2.2) showed strong similarity between all vegetation groups except wet forests.
+We therefore hypothesised that cats would temporally avoid foxes by becoming more diurnal in dry vegetation groups, but not in wet forests.
 We therefore modelled fox-induced changes in feral cat diel activity separately for wet and dry vegetation groups.
-We further split dry vegetation groups by region, to examine whether patterns remained consistent across space?
-for spatial replication.
-We refer to the resulting variable as 'habitat type', which had three levels: wet forests of the Otway Ranges ('wet_otways') dry EVC groups of the Otway Ranges ('dry_otways') and all of the Glenelg region data ('dry_glenelg').
+We further split dry vegetation groups by region for replication.
+We refer to the resulting variable as 'habitat type', which had three levels: (1) wet forests and rainforests in the western Otway Ranges ('wet_otways'), (2) dry EVC groups of the Otway Ranges ('dry_otways') and the Glenelg region ('dry_glenelg').
 
-We used a tensor product of hour (using cyclic cubic regression spline) and log adjusted fox count (using thin plate regression spline) smooths to model feral cat diel activity across the range of observed fox activity, separately for each habitat type.
-By also having double penalty shrinkage turned on, this meant that feral cat activity could vary nonlinearly and be penalised consistently across both dimensions.
+We used a tensor product of hour (using a cyclic cubic regression spline) and log adjusted fox count (using a thin plate regression spline) smooths to model feral cat diel activity across the range of observed fox activity.
+We specified this with a by variable factor smooth to model separate responses for each habitat type.
+We turned on the double penalty model selection, which penalises the null space in addition to the range space (i.e.
+shrinking wiggly terms to linear functions) of the spline basis, meaning covariate effects can be entirely removed from the model [@MARRA20112372].
+This meant that feral cat activity could vary nonlinearly and be penalised consistently across both dimensions.
 This model specification allows five different scenarios: that there was (1) no effect of hour or foxes on feral cats (2) a static hourly effect only, (3) a spatial response to foxes only, (4) a spatial response to foxes and an unrelated static hourly effect, (5) a spatial response to foxes and a hourly effect which changes across the range of fox counts.
+Additionally, we included a separate smooth for space (using a duchon spline) to account for the effect of unmodelled environmental covariates and spatial autocorrelation -- increasing confidence that only cat responses to foxes were being estimated.
+
+### Individual variation in feral cat diel activity
+One of the collated datasets identified identified individual cats based on unique pelage patterns across the wet forests of the western Otway four forest blocks in the glenelg region (matt chapter 2).
+We used only this dataset to estimated individual heterogeneity in cat diel activity using a hierarchical model specification.
+This included a global smoother for hour (i.e.
+average response), along with group-level smoothers for identified cat with a common wiggliness [i.e.
+model GS in @pedersen2019hierarchical].
+This model structure penalises functions which deviate strongly from the average response, so individual with few detections will take the shape of the global response.
 
@@
 
-```{r map, out.width="100%", fig.cap = "Locations of study sites in south-west Victoria, Australia.
+```{r diel-map, fig.cap = "Locations of our study regions in south-west Victoria, Australia.
 The grids of camera-traps are denoted by white dots.
@@
 Native vegetation is indicated by dark green, with hill shading.
-Map tiles by Stamen Design, under CC BY 3.0, map data by OpenStreetMap, under CC BY SA."}
+\\textit{Map tiles by Stamen Design, under CC BY 3.0, map data by OpenStreetMap, under CC BY SA.}"}
 knitr::include_graphics("../figs/fig1_map.png")
@@
 
-## Predator diel activity across vegetation types
+## Spatial variation in predator activity
 
+Increasing difference between the minimum and maximum activity estimate over the 24-hour cycle for each location.
+Red foxes concentrate their activity during particular times of the day–especially in the Glenelg region (a)–whereas feral cats have a relatively consistent activity throughout the daily cycle.
+Both predators exhibited similar spatial patterns in diel activity strength across the Glenelg region, but seemingly opposing patterns in the Otway Ranges.
+
+## Variation in predator activity across vegetation types
 On average, both predator species occupied similar times of the day--mostly nocturnal with activity peaks around sunrise and sunset.
@@
 In the wet forests, ox activity remained fairly constant throughout day and night barely had any diel activity pattern here, cats more nocturnal.
-Deviance explained and Rsq values for the two models were XX.
 
-## Fox-induced changes in feral cat diel activity
+Both invasive predator have an average (i) crepuscular - nocturnal diel activity pattern, with slight deviations across the dry EVC groups and larger deviations in wet forests.
+Unlike feral cats, fox spatial activity was relatively consistent across EVC groups.14
 
+## Feral cat spatiotemporal avoidance of foxes
 We saw evidence of changes in cat diel activity due to fox activity in all three habitat types.
@@
 Slightly higher uncertainty where fox spatial activity is high, this is because we had a largest sample size for fox absences, and fewest at the extreme of fox activity.
+
+In the Glenelg region, there were more feral cat detections wherethere were more fox detections, but diel activity shifted from night to day (a).
+In the Otway Ranges, feral cats shifted todiurnal activity where fox actity was high in dry vegetation types (b), but became more nocturnal where fox actity washigh in the rainforests and wet forests (c).
+
+## Individual variation in feral cat diel activity
+
+Individual heterogeneity in feral catFelis catusdiel activity across the Glenelg region (39 identified indiviuals; a)and wet forests of the western Otway Ranges (94 identified indiviuals; b) in south-west Victoria, Australia.
+As an example,the predicted activity of the seven individuals with the most detections (iii).
+Dotted, vertical lines represent average sunriseand sunset times.
+Shaded areas indicate 95% confidence intervals.
+The global function of feral cat diel activity in theGlenelg region was shrunk to an almost flat line, signalling high individual variation (deviations can therefore be intepreteddirectly as unique diel patterns).
+In the Otway wet forests, feral cats were on average nocturnal, individuals mostly deviatedwith different activity peaks near sunrise and sunset times.
 
@@
 table1 <- table1 %>%
-  select(region, vegetation_group, sites, deployments, fox_count_vr, cat_count_vr)  %>% 
+  select(vegetation_group, region, sites, deployments, fox_count_vr, cat_count_vr)  %>% 
   distinct() %>%
@@
 # make table
-pander(table1, style = "simple", split.table = Inf, caption = "Summary of camera-trap deployments in unique sites within Ecological Vegetation Class groups and independent counts of predator detections across two broad regions in south-west Victoria, Australia.", round = 2,
-              col.names = c("Region", "Vegetation", "Sites", "Deployments",  "Fox counts", "Cat counts"),
+pander(table1, style = "simple", split.table = Inf, caption = "Summary of the number of camera-trap deployments, unique survey sites and 'independent' counts of invasive predator detections across Ecological Vegetation Class groups within the Glenelg and Otway regions, south-west Victoria, Australia.", round = 2,
+              col.names = c("Vegetation", "Region", "Sites", "Deployments",  "Fox counts", "Cat counts"),
 )
@@
 
-```{r veg, out.width="100%", fig.cap = "Fox (A) and feral cat (B) diel activity patterns across Ecological Vegetation Class groups in south-west Victoria, Australia.
+```{r space, out.width="100%", fig.cap = "Relative spatial patterns of invasive predator diel activity strength across the two study regions in south-west Victoria, Australia.
+White crosses depict unique camera-trap sites; colour brightness scales with increasing difference between the minimum and maximum activity estimate over the 24-hour cycle for each location.
+Red foxes \\textit{Vulpes vulpes} concentrate their activity during particular times of the day--especially in the Glenelg region (a)--whereas feral cats \\textit{Felis catus} have a relatively consistent activity throughout the daily cycle.
+Both predators exhibited similar spatial patterns in diel activity strength across the Glenelg region, but seemingly opposing patterns in the Otway Ranges."}
+knitr::include_graphics("../figs/diel_strength_600dpi.png")
+
+```
+
+\newpage
+
+```{r veg, out.width="100%", fig.cap = "Red foxes \\textit{Vulpes vulpes} (a) and feral cat \\textit{Felis catus} (B) activity patterns across Ecological Vegetation Class (EVC) groups in south-west Victoria, Australia.
 Dotted, vertical lines represent average sunrise and sunset times.
-Shaded areas indicate 95\\% confidence intervals."}
+Shaded areas indicate 95\\% confidence intervals.
+Both invasive predator have an average (i) crepuscular - nocturnal diel activity pattern, with slight deviations across the dry EVC groups and larger deviations in wet forests.
+Unlike feral cats, fox spatial activity was relatively consistent across EVC groups."}
 knitr::include_graphics("../figs/predator_veg.png")
@@
 
-```{r count, out.width="100%", fig.cap="Predicted feral cat diel activity (A) and uncertainty (B) across each habitat type in response to foxes (log counts adjusted for survey effort).
+```{r count, out.width="100%", fig.cap="Variation in feral cat \\textit{Felis catus} activity (a) and uncertainty (b) in response to 'independent' red fox \\textit{Vulpes vulpes} detections (log-transformed and survey effort adjusted) across each 'habitat type' in south-west Victoria, Australia.
 Grey vertical lines respresent average sunrise and sunset times.
-Shaded areas indicate 95\\% confidence intervals."}
+In the Glenelg region, there were more feral cat detections where there were more fox detections, but diel activity shifted from night to day (a).
+In the Otway Ranges, feral cats shifted to diurnal activity where fox actity was high in dry vegetation types (b), but became more nocturnal where fox actity was high in the rainforests and wet forests (c)."}
 knitr::include_graphics("../figs/cat_fox_count.png")
+```
+
+\newpage
+
+```{r ind, out.width="100%", fig.cap = "Individual heterogeneity in feral cat \\textit{Felis catus} diel activity across the Glenelg region (39 identified indiviuals; a) and wet forests of the western Otway Ranges (94 identified indiviuals; b) in south-west Victoria, Australia.
+As an example, the predicted activity of the seven individuals with the most detections (iii).
+Dotted, vertical lines represent average sunrise and sunset times.
+Shaded areas indicate 95\\% confidence intervals.
+The global function of feral cat diel activity in the Glenelg region was shrunk to an almost flat line, signalling high individual variation (deviations can therefore be intepreted directly as unique diel patterns).
+In the Otway wet forests, feral cats were on average nocturnal, individuals mostly deviated with different activity peaks near sunrise and sunset times."}
+knitr::include_graphics("../figs/cat_ind.png")
 ```
@@
 ==========
+
+GRAVEYARD
+===================
+
+Diel activity patterns are a product of physiological constraints, species adaptiveness and landscape context.
+These dynamics depend on both environmental conditions and those of other species, but few statistical approaches incorporate both factors, or allow responses to be dynamic across space and time.
+This is key because species interactions and niche preferences are multidimensional.
+
+Animal diel patterns may not only vary across different spatial contexts, but so too may threat responses.
+
+Animals may be naive to threats (e.g.
+alien predators), unable to respond to threats (e.g.
+scarce prey), or the threat may be too minimal to warrant a response.
+Some environments more easily facilitate avoidance (e.g.
+high food and shelter accessibility), and so accounting for landscape context is key.
+
+It is critical to replicate studies across different ecosystem contexts to understand how generalisable behavioural changes are, but this is seldom done.
+Invasive species, which are widespread and highly adaptable to new conditions, are perfect subjects to test generalisability in spatiotemporal responses.
+That and there are clear management incentives.
+
+We expected the activity of each predator species to follow a broadly consistent pattern (e.g.
+lower during the day), but potentially have small shifts in relative activity between sunset and sunrise across the different EVC groups (1).
+Fitting separate responses for each EVC group would throw away shared information, which is particularly costly for the least surveyed groups, and would result in more noisy estimates.

### 2021-09-05 manuscript/manuscript.tex
fox ~ ts, region ~ re

--- 2021-08-05 manuscript/manuscript.tex
@@
 \usepackage{lineno} % add
-  \linenumbers % turns line numbering on
 \providecommand{\tightlist}{%
@@
             pdfauthor={},
-            pdftitle={Joint spatiotemporal models reveal dramatic shifts in mesopredator diel activity across individuals, habitat types and apex predator activity levels},
+            pdftitle={Quantifying spatiotemporal shifts in predator behaviour across habitats, individuals and threats.},
             colorlinks=false,
@@
 
-  \title{Joint spatiotemporal models reveal dramatic shifts in mesopredator diel activity across individuals, habitat types and apex predator activity levels}
+  \title{Quantifying spatiotemporal shifts in predator behaviour across habitats, individuals and threats.}
     \author[UOM]{Matthew W. Rees\corref{1}}
@@
 
-Effective wildlife management relies on understanding species population and behavioural dynamics to accurately predict intervention outcomes.
-These dynamics depend on both environmental conditions and those of other species, but few statistical approaches incorporate both factors, or allow responses to be dynamic across space and time.
-This is key because species interactions and niche preferences are multidimensional.
+Diel activity patterns - the distribution of activity throughout the daily cycle - are an important but overlooked aspect of behavioural ecology.
+Animals must allocate particular times of the day to rest, forage, socialise and reproduce, which are traded-off against threat avoidance (e.g.~predation) and deterrence against resource loss (i.e.~inter and intraspecific competition).
+Diel activity can therefore vary spatially across different environmental conditions (e.g.~food, shelter availability) and species assemblages.
+This is important to understand because constrained diel behaviour can incur fitness costs to individuals and populations (Lamb 2020 PNAS).
 
-Fear is a near universal driver of animal behaviour.
-Fear may be from other individuals of their species, a dominant species (including humans) or an event (e.g.~fire).
-Animals may be naive to threats (e.g.~alien predators), unable to respond to threats (e.g.~scarce prey), or the threat may be too minimal to warrant a response.
-Some environments more easily facilitate avoidance (e.g.~high food and shelter accessibility), and so accounting for landscape context is key.
-A threat response may be routine, or only employed or enhanced when the threat is heightened.
-This likely depends on how constant and predictable the threat is.
+Flexible spatiotemporal behaviour can allow animals to adapt to changing conditions.
+For example, elk in Yellowstone National Park use risky, open parts of the landscape when the activity of reintroduced wolves is low, suggesting coexistence without major fitness costs (Kohl et al.~2019).
+Diel activity decisions are clearly linked to habitat type, and there is growing recognition of individual-level variation in these decisions (Basille et al 2015; time-limited seabirds; urban bears).
+This is important to consider because populations with high individual heterogeneity are likely more adaptable to threats (Wolf and Weissing 2012; Dingemanse and Wolf 2013).
+Most studies examine spatiotemporal behavioural responses by repeating spatial analyses at different time periods (generally night and day), but few consider continuous shifts in diel activity across spatial contexts.
 
-Animals may respond to threats by altering their activity patterns (i.e.~avoidance).
-An animal may avoid locations where the threat is likely to occur (hereafter `spatial avoidance'), avoid times of the day when a threat is likely to occur (hereafter `temporal avoidance'), or avoid both locations and times a threat is heightened (hereafter `spatiotemporal avoidance').
-Such avoidance likely has trade-offs, which may or may not incur costs to individual and population fitness.
-Modelling approaches which do not allow spatiotemporal and dynamic fear reponses can only make limited inference about species interactions.
+Diel activity patterns are increasingly being considered as technology like telemetry devices and camera-traps have become more widely available.
+Telemetry devices like GPS collars and accelerometers provide high resolution data at the individual-level.
+However, resource constraints generally limit these studies to few individuals (Harmsen 2009), and target similar personalities which are susceptible to physical capture.
+Camera-traps are particularly useful because they concurrently survey multiple species--including the elusive individuals--providing inference on ecological differences and species interactions.
+However, individual-level inference would be limited to species with unique markings.
+Distiller et al 2020 recently used camera-traps to individually identify jaguars and estimate sex difference in diel activity patters.
+Despite having information on individual diel activity patterns, studies predominantly estimate at the population-level due to small sample sizes.
 
-Species interactions are usually considered in degrees of overlap between activity patterns.
-Spatiotemporal avoidance is most often considered, using joint species distribution models (including MSOMs).
-However, temporal avoidance is increasingly being used (perhaps just allowed due to widely available GPS and camera technology) with kernel density estimation methods.
-Although, these methods don't allow environmental covariates, and are particularly susceptible to noisy estimates.
-Importantly, some are beginning to combine these methods to allow spatiotemporal responses (Azzou et al 2021; Farris et al 2020).
-However, degrees of overlap in activity patterns only get at routine avoidance, we are generally interested in whether changes in threat levels lead to altered behaviour.
-This is more indicative of species interactions because static levels of overlap may just reflect different niche preferences rather than avoidance.
+A variety of statistical approaches are used to examine how diel activity patterns vary among habitat types, individuals and co-occurring species.
+Kernel density estimation is most often used to estimate hourly activity curves (Linkie \& Ridout, 2011), however this methods is susceptible to noise from sample sizes (Iannaerlli) and does allow interactions with covariates (Frey 2017).
+They remain popular because largely because they estimate a coefficient of overlap between two diel activity curves - either between two species, or one species in two different contexts.
+However, rarely are there only two interacting species or categorical contexts.
+Joint approaches with spatial overlap coefficients are being developed (e.g.~Azzou et al 2021; Farris et al 2020) to account for the spatiotemporal nature of species avoidance strategies.
+However, activity overlap coefficients may just reflect different niche preferences.
+A threat response may not be routine, but only employed or enhanced when the threat is heightened.
+And so, whether an animal \emph{changes} its activity patterns with an increasing threat level (e.g.~higher apex predator activity) is more interesting and relevant to management than arbitrary overlap.
 
-Dynamic behavioural responses are also important from a management perspective because they want to know whether an action will cause a change in animal behaviour / range?).
-For example, after the reintroduction of wolves in Yellowstone, it was widely thought that this had limited elk populations.
-Using GPS data and habitat selection methods, Kohl et al.~2018 found that elk used riskier parts of landscape at times when wolves were least active -- suggesting coexistence without fitness costs.
-However, there are many other large carnivores in this ecosystem which are likely to impact elk behaviour.
-And so, will elk respond to wolves in the same way in different environmental contexts?
-There are little-no examples of replicated spatiotemporal interactions studies.
+Straightforward, integrated methods for examining changes in spatiotemporal animal behaviour are needed.
+Generalised additive models (hereafter `GAMs') are increasingly used to estimate diel activity patterns (e.g xx).
+GAMs learn from data to specify nonlinear relationships between response and explanatory variables.
+Unlike kernel density estimation (Linke and Ridout), GAMs include smoothing penalties to avoid overfitting (making them less susceptible to noisy estimates due to small sample sizes) and allow diel activity to vary with explanatory variables.
+GAMs are also being used to test spatial associations between species (e.g.~my chapter 2) as a dominant species may only suppress subordinate species at high densities (Johnson).
+Despite this potential, diel activity curves are rarely modelled with categorical or continuous covariates, particularly using another species as that covariate (although see Cunningham).
+Further, GAMs can share information across conditions using heirarchical smooths, which estimate an average response, as well as group-level estimates (Pedersen et al.
+2019).
+Group-level estimates are more penalised as they deviate from the average, allowing confidence that these are true deviations across groups (e.g individuals, categorical environmental conditions).
 
-There is contention around the interactions between two invasive predators - feral cats and foxes.
-Foxes are routinely culled, and there is concern that this could lead to a mesopredator release of feral cats.
-Mesopredator release may manifest in both behavioural and numerical changes, which my previous chapter identified - but that behavioral responses (ranging and detectability) were not universal.
-Here, we aim to dig into the behavioral response further - in particular diel activity times.
-
-In this study, we used generalised additive models to assess (1) how predator (fox and cat) diel activity varies between vegetation types, and (2) whether an introduced mesopredator (cat) changes its diel and/or spatial activity in response to the activity of an introduced apex predator (fox).
-Because foxes and cats fill a similar function niche, we expected that, on average, they would have similar diel activity patterns across vegetation types (hypothesis 1).
-However, where fox detections were high, we hypothesised that the subordinate cat to be less active (hypothesis 2) and shift diel activity to times when foxes are least active (hypothesis 3).
-We tested this on a simple predator system where only cats and foxes occur, and fox activity is manipulated through lethal control.
-This allowed focus on the sole interactions between these two predators, and ensured differences in correlations wasn't just due to differences in niche preferences.
+Here we provide a case-study to illustrate how generalised additive models can be used to assess spatiotemporal behaviour of two predator species.
+The Red fox \emph{Vulpes vulpes} (hereafter `fox') and feral cats \emph{Felis catus} (hereafter `cat') have some of the widest distributions of any carnivores, and co-occur across most of their global range.
+Because of their devastating impacts on native prey in their introduced range, understanding differences in their behavioural ecology across space, and how foxes may impact feral cat behaviour is key to effective management.
+Numerous studies have tested spatial and temporal associations between these invasive predators and environmental covariates, but not in a spatially explicit or joint modelling framework.
+We therefore investigated how invasive predator activity varies across (1) space and (2) vegetation types, as well as how feral cat activity varies across fox activity levels (3) and individuals in two distinct study regions.
+We tested this in a simple predator system where only cats and foxes occur, and fox activity is manipulated through lethal control.
+This allowed focus on the sole interactions between these two predators, and ensuring correlations doesn't simply reflect different niche preferences.
 
@@
 We collated camera-trap data on these two predator species from three distinct studies across two regions of south-west Victoria, Australia (Fig.
-1).
-The purpose of each study was to experimentally survey changes in the mammalian community due to fox control, and so our dataset covered a wide range contexts with and without fox-baiting, allowing us to investigate fox-cat interactions across a gradient of apex predator (fox) activity.
+\ref{fig:diel-map}).
+The terrestrial mammalian predator guild is depauperate throughout both regions.
+Dingoes \emph{Canis familiaris} are now absent throughout, as is the tiger quoll \emph{Dasyurus maculatus}, more recently absent in the Otway Ranges (last sighted in 2014 despite extensive camera-trapping).
+Introduced foxes and cats are therefore the only medium-large functional mammalian terrestrial predators across both regions.
 
-The terrestrial mammalian predator guild is depauperate throughout both regions.
-Dingoes \emph{Canis familiaris} are now absent throughout, as is the the tiger quoll \emph{Dasyurus maculatus}, more recently absent in the Otway Ranges (last sighted in 2014 despite extensive camera-trapping).
-Introduced red foxes and feral cats are therefore the only medium-large functional mammalian terrestrial predators across both regions.
-
-In the baited parts of the regions, government land managers conduct ongoing targeted fox control for biodiversity conservation.
+The purpose of each collated study was to experimentally survey changes in the mammalian community due to fox control.
+In broad sections of each region, government land managers conduct ongoing targeted fox control for biodiversity conservation.
 Poison baits containing 3 mg of sodium mono-fluroacetate (1080) are buried at a depth of 10 cm at 1-km intervals along accessible forest tracks and roads.
+This allowed us to investigate fox-cat interactions across a gradient of apex predator (fox) activity and other landscape contexts.
 
 In the Glenelg region, Gunditjmara country, natural vegetation is highly fragmented.
-Here, camera-survey sites were situated in six distinct forest blocks: two are separated by a river, the others are separated by at least \emph{XX} km of pastoral farming and residential property.
+Here, camera-survey sites were situated in six distinct forest blocks: two are separated by a river, the others are separated by at least \textbf{XX} km of pastoral farming and residential property (Fig.
+\ref{fig:map}).
 Foxes in three of the forest blocks have been subject to lethal control since October 2005, with 1080 poison-baits replaced fortnightly (Robley et al.
 2014).
-
 The Otway Ranges, Gadubanud country, is a largely continuous patch of natural vegetation with a strong east-west rainfall and elevation gradient (Swan et al.
@@
 The majority of this baiting began in 2016 - 2017, although fox-baiting commenced in some small sections in 2008.
-
 Alongside fragmentation, the extremely rugged, wet forest and rainforest matrix in the western Otway Ranges is the main difference between the two regions.
-Prescribed fire, primarily for the purpose of asset protection, is another key management action undertaken throughout both regions (except in very wet sections of Otway Ranges).
+Prescribed fire is another key management action undertaken throughout both regions (except in very wet sections of Otway Ranges), primarily for asset protection.
 
@@
 We compiled camera-trap data from three distinct studies across the two regions.
-After removing camera-traps which were operational for less than 14 days, this totalled 3659 deployments of camera-traps at 1232 camera-trap sites.
-Survey durations ranged between 14 and 93 days (mean 47), totalling 171,966 trapnights.
+After removing camera-traps which were operational for less than 14 days, this totalled 3694 deployments of camera-traps at 1232 camera-trap sites.
+Camera-traps were active for an average of 47 days, totalling 172,167 trapnights.
 Camera-trap spacing was variable; the average minimum distance between simultaneously deployed camera-traps was 445, 853 and 1266 metres for each study.
-Camera-trap were deployed in the Glenelg region between 2013 -- 2019, whereas in the Otway Ranges surveys were conducted in 2016 -- 2019.
+Camera-traps were deployed in the Glenelg region between 2013 -- 2019, and in the Otway Ranges between 2016 -- 2019.
 
@@
 (2019).
-
 Using the `reshape' R-package (Wickham 2007), we manipulated the detection table into a dataframe with a row for each hour of the day, for every camera-trap deployment, recording the total number of `independent' fox and feral cat detections within each hour for the entire camera-trap survey duration.
 
-We identified the broad vegetation type (Ecological Vegetation Class (hereafter `EVC') group (DELWP 2020)) at each camera-trap location.
-Six vegetation types were surveyed in both regions, although to varying degrees (Table 1).
+We attached the Ecological Vegetation Class (hereafter `EVC') group (DELWP 2020)) at each camera-trap location.
+\textbf{Six?} vegetation types were surveyed in both regions, although to varying degrees (Table 1).
 In the Otway region, rainforests are interspersed (primarily in low lying gullies) throughout wet and damp forests at a finescale, and we merged these groups (hereafter `wet forests').
-
 For each deployment, we calculated the number of poison fox-baits within a 2.3 km radius around each camera-trap - the average maximum distance foxes in these regions travel from their home range centre (Hradsky et al.
 2017).
+To investigate changes in feral cat diel activity in response to counts of foxes.
+We added an extra column to our dataframe with the number of fox detections divided by the log of the number of survey days to account for differences in camera-trap survey durations (hereafter `adjusted fox counts').
 
@@
 
-We modelled predator counts (response variable) with generalised additive mixed-effects models (hereafter `GAMMs') implemented in the `mgcv' R-package (Wood 2017).
-We first fit these models using the poisson family, but switched to the negative binomial family as overdispersion, but not zero-inflation, was detected using the `DHARMa' R-package (Hartig 2020).
+We modelled the total predator counts for each camera-trap deployment (response variable) with generalised additive mixed-effect models implemented in the `mgcv' R-package (Wood 2017).
+GAMs learn from the data to specify nonlinear relationships between response and explanatory variables.
+A maximum wiggliness is specified for each explanatory variable, with complexity automatically reduced (ultimately to a linear function) through a smoothing penalty to avoid overfitting.
+We used REML estimation and the negative binomial family (as overdispersion, but not zero-inflation, was detected using the `DHARMa' R-package (Hartig 2020)).
 We specified a model offset to account for differences in camera-trap survey duration, and a random intercept for each site to account for repeat sampling.
-We used a duchon spline static spatial smooth to model both spatial autocorrelation and differences in predator activity across space (i.e.~the effect of unmodelled environmental covariates).
-For the fox model, we additionally included a thin plate regression spline for poison-bait density, with a factor smoother for each region, as bait replacement schedules differed between the regions.
+For fox models (n = 2), we included a smooth effect of poison-bait density using a thin plate regression spline basis.
 This formed the base model specification for each model we fitted.
-Models (n = 2) differed in their specification of the cyclical hour smooth in order to explore (1) variations in predator diel activity across EVC groups, as well as (2) whether feral cat diel activity changed across the gradient of observed fox counts (detailed in the sections below).
-We used the double penalty model selection approach, which penalises the null space in addition to the range space (i.e.~shrinking wiggly terms to linear terms) of the spline basis, meaning covariate effects can be entirely removed from the model (Marra and Wood 2011).
+Models differed in their specification of the cyclical hour smooth in order to provide inference on variations of predator diel activity across (1) space (i.e.~coordinates) (2) EVC groups, as across for feral cats across the range of observed fox counts (3), and across each identified individual cat (4), detailed in the sections below.
+This model specification allowed both spatial activity to be higher in each context (i.e.~across coordinates, vegetation types, fox activity levels, individual cats) in addition to variable diel activity patterns.
+Code used to specify each model set is presented in the Supporting Information.
 We plotted models using `ggplot2' and `gratia' R-packages (Wickham 2016, Simpson 2021).
 
-We expected the activity of each predator species to follow a broadly consistent pattern (e.g.~lower during the day), but potentially have small shifts in relative activity between sunset and sunrise across the different EVC groups (1).
-Fitting separate responses for each EVC group would throw away shared information, which is particularly costly for the least surveyed groups, and would result in more noisy estimates.
-We therefore specified a global smoother for hour (i.e.~average response), along with group-level smoothers with shared wiggliness for each EVC group (i.e.~model GS in Pedersen et al.
+\hypertarget{spatial-variation-in-predator-activity}{%
+\subsubsection{Spatial variation in predator activity}\label{spatial-variation-in-predator-activity}}
+
+We fit a model separately for both foxes and cats which included a tensor product interaction between a spatial smooth and hourly smooth.
+This allowed predators to have different activity levels across space (static across the years surveyed), as well as a variation in diel activity across space.
+Space was modelled using camera-trap coordinates and a duchon spline basis (Miller and Wood).
+To compare relative diel activity pattern strengths--how much activity patterns vary across the daily cycle--we plotted the difference between the minimum and maximum activity estimate for every predicted location.
+
+\hypertarget{variation-in-predator-activity-across-vegetation-types}{%
+\subsubsection{Variation in predator activity across vegetation types}\label{variation-in-predator-activity-across-vegetation-types}}
+
+We estimated diel activity across EVC group using a hierarchical model specification.
+This included a global smoother for hour (i.e.~average response), along with group-level smoothers for each EVC group with shared shared wiggliness (i.e.~model GS in Pedersen et al.
 2019).
-This model structure penalises functions which deviate strongly from the average response, increasing confidence that they are true deviations opposed to the result of sampling noise.
+This model structure penalises functions which deviate strongly from the average response.
 
-To investigate changes in feral cat diel activity in response to counts of foxes.
-We added an extra column to our dataframe with the number of fox detections divided by the log of the number of survey days to account for differences in camera-trap survey durations (hereafter `adjusted fox counts').
+\hypertarget{feral-cat-spatiotemporal-avoidance-of-foxes}{%
+\subsubsection{Feral cat spatiotemporal avoidance of foxes}\label{feral-cat-spatiotemporal-avoidance-of-foxes}}
 
-Analysis of stage 1 showed that fox diel activity patterns in wet forests differed relative to all other EVC groups.
-We therefore expected feral cats might to temporally avoid foxes in a different way in wet forests than the dry vegetation groups, but that responses would be similar in all the drier vegetation groups.
+Analysis of fox diel activity across EVC groups (section 4.2.2) showed strong similarity between all vegetation groups except wet forests.
+We therefore hypothesised that cats would temporally avoid foxes by becoming more diurnal in dry vegetation groups, but not in wet forests.
 We therefore modelled fox-induced changes in feral cat diel activity separately for wet and dry vegetation groups.
-We further split dry vegetation groups by region, to examine whether patterns remained consistent across space?
-for spatial replication.
-We refer to the resulting variable as `habitat type', which had three levels: wet forests of the Otway Ranges (`wet\_otways') dry EVC groups of the Otway Ranges (`dry\_otways') and all of the Glenelg region data (`dry\_glenelg').
+We further split dry vegetation groups by region for replication.
+We refer to the resulting variable as `habitat type', which had three levels: (1) wet forests and rainforests in the western Otway Ranges (`wet\_otways'), (2) dry EVC groups of the Otway Ranges (`dry\_otways') and the Glenelg region (`dry\_glenelg').
 
-We used a tensor product of hour (using cyclic cubic regression spline) and log adjusted fox count (using thin plate regression spline) smooths to model feral cat diel activity across the range of observed fox activity, separately for each habitat type.
-By also having double penalty shrinkage turned on, this meant that feral cat activity could vary nonlinearly and be penalised consistently across both dimensions.
+We used a tensor product of hour (using a cyclic cubic regression spline) and log adjusted fox count (using a thin plate regression spline) smooths to model feral cat diel activity across the range of observed fox activity.
+We specified this with a by variable factor smooth to model separate responses for each habitat type.
+We turned on the double penalty model selection, which penalises the null space in addition to the range space (i.e.~shrinking wiggly terms to linear functions) of the spline basis, meaning covariate effects can be entirely removed from the model (Marra and Wood 2011).
+This meant that feral cat activity could vary nonlinearly and be penalised consistently across both dimensions.
 This model specification allows five different scenarios: that there was (1) no effect of hour or foxes on feral cats (2) a static hourly effect only, (3) a spatial response to foxes only, (4) a spatial response to foxes and an unrelated static hourly effect, (5) a spatial response to foxes and a hourly effect which changes across the range of fox counts.
+Additionally, we included a separate smooth for space (using a duchon spline) to account for the effect of unmodelled environmental covariates and spatial autocorrelation -- increasing confidence that only cat responses to foxes were being estimated.
+
+\hypertarget{individual-variation-in-feral-cat-diel-activity}{%
+\subsubsection{Individual variation in feral cat diel activity}\label{individual-variation-in-feral-cat-diel-activity}}
+
+One of the collated datasets identified identified individual cats based on unique pelage patterns across the wet forests of the western Otway four forest blocks in the glenelg region (matt chapter 2).
+We used only this dataset to estimated individual heterogeneity in cat diel activity using a hierarchical model specification.
+This included a global smoother for hour (i.e.~average response), along with group-level smoothers for identified cat with a common wiggliness (i.e.~model GS in Pedersen et al.
+2019).
+This model structure penalises functions which deviate strongly from the average response, so individual with few detections will take the shape of the global response.
 
@@
 \begin{figure}
-\includegraphics[width=1\linewidth]{../figs/fig1_map} \caption{Locations of study sites in south-west Victoria, Australia.
+\includegraphics[width=16.68in]{../figs/fig1_map} \caption{Locations of our study regions in south-west Victoria, Australia.
 The grids of camera-traps are denoted by white dots.
@@
 Native vegetation is indicated by dark green, with hill shading.
-Map tiles by Stamen Design, under CC BY 3.0, map data by OpenStreetMap, under CC BY SA.}\label{fig:map}
+\textit{Map tiles by Stamen Design, under CC BY 3.0, map data by OpenStreetMap, under CC BY SA.}}\label{fig:diel-map}
 \end{figure}
@@
 
-\hypertarget{predator-diel-activity-across-vegetation-types}{%
-\subsection{Predator diel activity across vegetation types}\label{predator-diel-activity-across-vegetation-types}}
+\hypertarget{spatial-variation-in-predator-activity-1}{%
+\subsection{Spatial variation in predator activity}\label{spatial-variation-in-predator-activity-1}}
+
+Increasing difference between the minimum and maximum activity estimate over the 24-hour cycle for each location.
+Red foxes concentrate their activity during particular times of the day--especially in the Glenelg region (a)--whereas feral cats have a relatively consistent activity throughout the daily cycle.
+Both predators exhibited similar spatial patterns in diel activity strength across the Glenelg region, but seemingly opposing patterns in the Otway Ranges.
+
+\hypertarget{variation-in-predator-activity-across-vegetation-types-1}{%
+\subsection{Variation in predator activity across vegetation types}\label{variation-in-predator-activity-across-vegetation-types-1}}
 
@@
 In the wet forests, ox activity remained fairly constant throughout day and night barely had any diel activity pattern here, cats more nocturnal.
-Deviance explained and Rsq values for the two models were XX.
 
-\hypertarget{fox-induced-changes-in-feral-cat-diel-activity}{%
-\subsection{Fox-induced changes in feral cat diel activity}\label{fox-induced-changes-in-feral-cat-diel-activity}}
+Both invasive predator have an average (i) crepuscular - nocturnal diel activity pattern, with slight deviations across the dry EVC groups and larger deviations in wet forests.
+Unlike feral cats, fox spatial activity was relatively consistent across EVC groups.14
+
+\hypertarget{feral-cat-spatiotemporal-avoidance-of-foxes-1}{%
+\subsection{Feral cat spatiotemporal avoidance of foxes}\label{feral-cat-spatiotemporal-avoidance-of-foxes-1}}
 
@@
 
+In the Glenelg region, there were more feral cat detections wherethere were more fox detections, but diel activity shifted from night to day (a).
+In the Otway Ranges, feral cats shifted todiurnal activity where fox actity was high in dry vegetation types (b), but became more nocturnal where fox actity washigh in the rainforests and wet forests (c).
+
+\hypertarget{individual-variation-in-feral-cat-diel-activity-1}{%
+\subsection{Individual variation in feral cat diel activity}\label{individual-variation-in-feral-cat-diel-activity-1}}
+
+Individual heterogeneity in feral catFelis catusdiel activity across the Glenelg region (39 identified indiviuals; a)and wet forests of the western Otway Ranges (94 identified indiviuals; b) in south-west Victoria, Australia.
+As an example,the predicted activity of the seven individuals with the most detections (iii).
+Dotted, vertical lines represent average sunriseand sunset times.
+Shaded areas indicate 95\% confidence intervals.
+The global function of feral cat diel activity in theGlenelg region was shrunk to an almost flat line, signalling high individual variation (deviations can therefore be intepreteddirectly as unique diel patterns).
+In the Otway wet forests, feral cats were on average nocturnal, individuals mostly deviatedwith different activity peaks near sunrise and sunset times.
+
 \newpage
@@
 \begin{longtable}[]{@{}cccccc@{}}
-\caption{Summary of camera-trap deployments in unique sites within Ecological Vegetation Class groups and independent counts of predator detections across two broad regions in south-west Victoria, Australia.}\tabularnewline
+\caption{Summary of the number of camera-trap deployments, unique survey sites and `independent' counts of invasive predator detections across Ecological Vegetation Class groups within the Glenelg and Otway regions, south-west Victoria, Australia.}\tabularnewline
 \toprule
-Region & Vegetation & Sites & Deployments & Fox counts & Cat counts\tabularnewline
+Vegetation & Region & Sites & Deployments & Fox counts & Cat counts\tabularnewline
 \midrule
@@
 \toprule
-Region & Vegetation & Sites & Deployments & Fox counts & Cat counts\tabularnewline
+Vegetation & Region & Sites & Deployments & Fox counts & Cat counts\tabularnewline
 \midrule
 \endhead
-Glenelg & Dry Forest & 25 & 68 & 347 & 9\tabularnewline
-Otway & Dry Forest & 111 & 310 & 341 & 158\tabularnewline
-Glenelg & Heathland & 40 & 119 & 265 & 59\tabularnewline
-Otway & Heathland & 3 & 9 & 8 & 6\tabularnewline
-Glenelg & Heathy Woodland & 154 & 424 & 574 & 96\tabularnewline
-Otway & Heathy Woodland & 82 & 256 & 160 & 66\tabularnewline
-Glenelg & Herb-rich Woodland & 59 & 372 & 860 & 197\tabularnewline
-Otway & Herb-rich Woodland & 2 & 6 & 3 & 2\tabularnewline
-Glenelg & Lowland Forest & 383 & 1046 & 1900 & 290\tabularnewline
-Otway & Lowland Forest & 52 & 162 & 190 & 35\tabularnewline
-Glenelg & Swampy Scrub & 4 & 10 & 19 & 8\tabularnewline
-Otway & Swampy Scrub & 36 & 97 & 64 & 88\tabularnewline
-Otway & Wet Forest & 281 & 780 & 715 & 1187\tabularnewline
+Dry Forest & Glenelg & 25 & 68 & 347 & 9\tabularnewline
+Dry Forest & Otway & 111 & 310 & 341 & 158\tabularnewline
+Heathland & Glenelg & 40 & 119 & 265 & 59\tabularnewline
+Heathland & Otway & 3 & 9 & 8 & 6\tabularnewline
+Heathy Woodland & Glenelg & 154 & 424 & 574 & 96\tabularnewline
+Heathy Woodland & Otway & 82 & 256 & 160 & 66\tabularnewline
+Herb-rich Woodland & Glenelg & 59 & 372 & 860 & 197\tabularnewline
+Herb-rich Woodland & Otway & 2 & 6 & 3 & 2\tabularnewline
+Lowland Forest & Glenelg & 383 & 1046 & 1900 & 290\tabularnewline
+Lowland Forest & Otway & 52 & 162 & 190 & 35\tabularnewline
+Swampy Scrub & Glenelg & 4 & 10 & 19 & 8\tabularnewline
+Swampy Scrub & Otway & 36 & 97 & 64 & 88\tabularnewline
+Wet Forest & Otway & 281 & 780 & 715 & 1187\tabularnewline
 \bottomrule
@@
 \begin{figure}
-\includegraphics[width=1\linewidth]{../figs/predator_veg} \caption{Fox (A) and feral cat (B) diel activity patterns across Ecological Vegetation Class groups in south-west Victoria, Australia.
-Dotted, vertical lines represent average sunrise and sunset times.
-Shaded areas indicate 95\% confidence intervals.}\label{fig:veg}
+\includegraphics[width=1\linewidth]{../figs/diel_strength_600dpi} \caption{Relative spatial patterns of invasive predator diel activity strength across the two study regions in south-west Victoria, Australia.
+White crosses depict unique camera-trap sites; colour brightness scales with increasing difference between the minimum and maximum activity estimate over the 24-hour cycle for each location.
+Red foxes \textit{Vulpes vulpes} concentrate their activity during particular times of the day--especially in the Glenelg region (a)--whereas feral cats \textit{Felis catus} have a relatively consistent activity throughout the daily cycle.
+Both predators exhibited similar spatial patterns in diel activity strength across the Glenelg region, but seemingly opposing patterns in the Otway Ranges.}\label{fig:space}
 \end{figure}
@@
 \begin{figure}
-\includegraphics[width=1\linewidth]{../figs/cat_fox_count} \caption{Predicted feral cat diel activity (A) and uncertainty (B) across each habitat type in response to foxes (log counts adjusted for survey effort).
-Grey vertical lines respresent average sunrise and sunset times.
-Shaded areas indicate 95\% confidence intervals.}\label{fig:count}
+\includegraphics[width=1\linewidth]{../figs/predator_veg} \caption{Red foxes \textit{Vulpes vulpes} (a) and feral cat \textit{Felis catus} (B) activity patterns across Ecological Vegetation Class (EVC) groups in south-west Victoria, Australia.
+Dotted, vertical lines represent average sunrise and sunset times.
+Shaded areas indicate 95\% confidence intervals.
+Both invasive predator have an average (i) crepuscular - nocturnal diel activity pattern, with slight deviations across the dry EVC groups and larger deviations in wet forests.
+Unlike feral cats, fox spatial activity was relatively consistent across EVC groups.}\label{fig:veg}
 \end{figure}
@@
 \begin{figure}
-\includegraphics[width=1\linewidth]{../figs/cat_ind} \caption{Individual variations in feral cat diel activity patterns across the Glenelg region (A) and wet forests of the western Otway Ranges (B) in south-west Victoria, Australia.
+\includegraphics[width=1\linewidth]{../figs/cat_fox_count} \caption{Variation in feral cat \textit{Felis catus} activity (a) and uncertainty (b) in response to 'independent' red fox \textit{Vulpes vulpes} detections (log-transformed and survey effort adjusted) across each 'habitat type' in south-west Victoria, Australia.
+Grey vertical lines respresent average sunrise and sunset times.
+In the Glenelg region, there were more feral cat detections where there were more fox detections, but diel activity shifted from night to day (a).
+In the Otway Ranges, feral cats shifted to diurnal activity where fox actity was high in dry vegetation types (b), but became more nocturnal where fox actity was high in the rainforests and wet forests (c).}\label{fig:count}
+\end{figure}
+
+\newpage
+
+\begin{figure}
+\includegraphics[width=1\linewidth]{../figs/cat_ind} \caption{Individual heterogeneity in feral cat \textit{Felis catus} diel activity across the Glenelg region (39 identified indiviuals; a) and wet forests of the western Otway Ranges (94 identified indiviuals; b) in south-west Victoria, Australia.
+As an example, the predicted activity of the seven individuals with the most detections (iii).
 Dotted, vertical lines represent average sunrise and sunset times.
-Shaded areas indicate 95\% confidence intervals.}\label{fig:ind}
+Shaded areas indicate 95\% confidence intervals.
+The global function of feral cat diel activity in the Glenelg region was shrunk to an almost flat line, signalling high individual variation (deviations can therefore be intepreted directly as unique diel patterns).
+In the Otway wet forests, feral cats were on average nocturnal, individuals mostly deviated with different activity peaks near sunrise and sunset times.}\label{fig:ind}
 \end{figure}
@@
 
+\hypertarget{graveyard}{%
+\section{GRAVEYARD}\label{graveyard}}
+
+Diel activity patterns are a product of physiological constraints, species adaptiveness and landscape context.
+These dynamics depend on both environmental conditions and those of other species, but few statistical approaches incorporate both factors, or allow responses to be dynamic across space and time.
+This is key because species interactions and niche preferences are multidimensional.
+
+Animal diel patterns may not only vary across different spatial contexts, but so too may threat responses.
+
+Animals may be naive to threats (e.g.~alien predators), unable to respond to threats (e.g.~scarce prey), or the threat may be too minimal to warrant a response.
+Some environments more easily facilitate avoidance (e.g.~high food and shelter accessibility), and so accounting for landscape context is key.
+
+It is critical to replicate studies across different ecosystem contexts to understand how generalisable behavioural changes are, but this is seldom done.
+Invasive species, which are widespread and highly adaptable to new conditions, are perfect subjects to test generalisability in spatiotemporal responses.
+That and there are clear management incentives.
+
+We expected the activity of each predator species to follow a broadly consistent pattern (e.g.~lower during the day), but potentially have small shifts in relative activity between sunset and sunrise across the different EVC groups (1).
+Fitting separate responses for each EVC group would throw away shared information, which is particularly costly for the least surveyed groups, and would result in more noisy estimates.
+
 \hypertarget{refs}{}

### 2021-09-05 r_scripts/5_gams_diel_activity.R

@@ -9,6 +9,7 @@ library(patchwork)
 library(viridis)
 library(gratia)
 library(chron)
+library(sf)
 
 rm(list = ls())
 options(scipen = 999) 
@@ -25,8 +26,8 @@ records <- read.csv("derived_data/counts_hour.csv")
 # drop zoi's data 
 records <- filter(records, data_source != "zoi")
 
-# exclude stations left out for less than 2 weeks 
-records <- filter(records, survey_duration >= 14)
+# exclude stations left out for less than 7 days - something musta went wrong with those few cams
+records <- filter(records, survey_duration >= 7)
 hist(records$survey_duration / 23, breaks = 50)
 summary(records$survey_duration)
 
@@ -41,6 +42,13 @@ records$XGROUPNAME <- if_else(records$XGROUPNAME == "Wet or Damp Forests", "Wet 
 records$XGROUPNAME <- if_else(records$XGROUPNAME == "Riparian Scrubs or Swampy Scrubs and Woodlands", "Swampy Scrubs", records$XGROUPNAME)
 records$XGROUPNAME <- substr(records$XGROUPNAME, 1, nchar(records$XGROUPNAME) - 1)
 table(records$XGROUPNAME)
+
+# add xy coordinates
+records_sf <- st_as_sf(records, coords = c("longitude", "latitude"), crs = 4326) %>% 
+  st_transform(crs = 32754)
+records$x <- st_coordinates(records_sf)[,1]
+records$y <- st_coordinates(records_sf)[,2]
+head(records)
 
 
 # ADD FOX COUNTS PER CAMERA -----------------------------------------------
@@ -96,6 +104,7 @@ gam_fox <- bam(fox ~ s(hour, bs = "cc", k = 8) +
                      s(hour, vegetation_group, bs = "fs", xt = list(bs = "cc"), k = 8) + 
                      s(foxbaits, region, bs = "fs", xt = list(bs = "tp"), k = 4) + 
                      s(longitude, latitude, bs = "ds",  m = c(1, 0.5), k = 200) +
+                     s(region, bs = "re") +  
                      s(station, bs = "re") +  
                      offset(log(survey_duration)), 
                     data = records, family = nb, knots = list(hour = c(0, 23)), nthreads = 3, discrete = TRUE)
[…]

### 2021-09-05 r_scripts/6_individual_cat_variation.R

@@ -194,7 +194,6 @@ gam_data_g <- transform(gam_data_g,
 gam_cat_ind_g <- bam(cat ~ s(hour, bs = "cc", k = 8) +   
                            s(hour, individual, bs = "fs", xt = list(bs = "cc"), k = 8) +  
                            s(longitude, latitude, bs = "ds",  m = c(1, 0.5), k = 150) +
-                           s(station, bs = "re") +  
                            offset(log(survey_duration)), 
                      data = gam_data_g, family = nb, knots = list(hour = c(0, 23)), nthreads = 3, discrete = TRUE)
 summary(gam_cat_ind_g)

### 2021-09-28 diel_activity_analysis.Rmd
diel strength % diff

@@
 
-### Spatial variation in the strength of diel patterns
+## Spatial variation in the strength of diel patterns
 
@@
 # predict model results into dataframe
-data_g_plot <- cbind(data_g_plot, predict.gam(gam_fox_sp, newdata = data_g_plot, se.fit = TRUE, type = "link", exclude = c("s(station)", "s(survey_duration)")))
+data_g_plot <- cbind(data_g_plot, predict.gam(gam_fox_sp, newdata = data_g_plot, se.fit = TRUE, type = "response", exclude = c("s(station)", "s(survey_duration)")))
 data_g_plot <- rename(data_g_plot, fox_predicted = fit,  fox_predicted_se = se.fit) # rename
@@
   group_by(location_id) %>% 
-  mutate(diel_diff = max(fox_predicted) - min(fox_predicted))
+  mutate(diel_diff = max(fox_predicted) - min(fox_predicted),          
+         diel_diff_pc = diel_diff / min(fox_predicted) * 100)
 
@@
   geom_tile()+
-  scale_fill_viridis("log(activity)", option = "viridis", limits = c(-5.27, -1.13)) +
-  geom_point(data = records_g, fill = NA, col = "white", size = 0.7, alpha = 0.1, shape = 3) +
+  scale_fill_viridis("Activity", option = "viridis") +
+  geom_point(data = records_g, fill = NA, col = "white", size = 0.7, alpha = 1, shape = 3) +
   theme_bw(10) + 
@@
 # fox plot: spatial difference
-g_fox_plot_diff <- ggplot(aes(x, y, fill = diel_diff),
+g_fox_plot_diff <- ggplot(aes(x, y, fill = diel_diff_pc),
                      data = data_g_plot) +
   geom_tile()+
-  scale_fill_viridis("Relative diel pattern strength", option = "viridis", limits = c(0.5, 1.55), breaks=c(0.5, 1.03, 1.55),labels=c("low", "medium", "high")) +
-  geom_point(data = records_g, fill = NA, col = "white", size = 0.7, alpha = 0.015, shape = 3) +
+  scale_fill_viridis("Relative activity increase (%)", option = "viridis", limits = c(76, 371)) +
+  geom_point(data = records_g, fill = NA, col = "white", size = 0.7, alpha = 1, shape = 3) +
   theme_bw(10) + 
@@
 data_o_plot = data_o_plot[with(data_o_plot, inSide(otways_buffer_df, x, y)),]
-data_o_plot <- cbind(data_o_plot, predict.gam(gam_fox_sp, newdata = data_o_plot, se.fit = TRUE, type = "link", exclude = c("s(station)", "s(survey_duration)")))
+data_o_plot <- cbind(data_o_plot, predict.gam(gam_fox_sp, newdata = data_o_plot, se.fit = TRUE, type = "response", exclude = c("s(station)", "s(survey_duration)")))
 data_o_plot <- rename(data_o_plot, fox_predicted = fit,  fox_predicted_se = se.fit) # rename
@@
   group_by(location_id) %>% 
-  mutate(diel_diff = max(fox_predicted) - min(fox_predicted))
+  mutate(diel_diff = max(fox_predicted) - min(fox_predicted),
+         diel_diff_pc = diel_diff / min(fox_predicted) * 100)
 
@@
   geom_tile()+
-  scale_fill_viridis("log(activity)", option = "viridis", limits = c(-5.27, -1.13)) +
-  geom_point(data = records_o, fill = NA, col = "white", size = 0.7, alpha = 0.1, shape = 3) +
+  scale_fill_viridis("activity", option = "viridis") +
+  geom_point(data = records_o, fill = NA, col = "white", size = 0.7, alpha = 1, shape = 3) +
   theme_bw(10) + 
@@
 # fox plot: spatial difference
-o_fox_plot_diff <- ggplot(aes(x, y, fill = diel_diff),
+o_fox_plot_diff <- ggplot(aes(x, y, fill = diel_diff_pc),
                      data = data_o_plot) +
   geom_tile()+
-  scale_fill_viridis("Relative diel pattern strength", option = "viridis", limits = c(0.5, 1.55), breaks=c(0.5, 1.03, 1.55),labels=c("low", "medium", "high")) +
-  geom_point(data = records_o, fill = NA, col = "white", size = 0.7, alpha = 0.015, shape = 3) +
+  scale_fill_viridis("Relative activity increase (%)", option = "viridis", limits = c(76, 371)) +
+  geom_point(data = records_o, fill = NA, col = "white", size = 0.7, alpha = 1, shape = 3) +
   theme_bw(10) + 
@@
 # predict model results into dataframe
-data_g_plot <- cbind(data_g_plot, predict.gam(gam_cat_sp, newdata = data_g_plot, se.fit = TRUE, type = "link", exclude = c("s(station)", "s(survey_duration)")))
+data_g_plot <- cbind(data_g_plot, predict.gam(gam_cat_sp, newdata = data_g_plot, se.fit = TRUE, type = "response", exclude = c("s(station)", "s(survey_duration)")))
 data_g_plot <- rename(data_g_plot, cat_predicted = fit,  cat_predicted_se = se.fit) # rename
@@
   group_by(location_id) %>% 
-  mutate(diel_diff = max(cat_predicted) - min(cat_predicted))
+  mutate(diel_diff = max(cat_predicted) - min(cat_predicted),          
+         diel_diff_pc = diel_diff / min(cat_predicted) * 100)
 
@@
 # cat plot: spatial difference
-g_cat_plot_diff <- ggplot(aes(x, y, fill = diel_diff),
+g_cat_plot_diff <- ggplot(aes(x, y, fill = diel_diff_pc),
                      data = data_g_plot) +
   geom_tile()+
-  scale_fill_viridis("Relative diel pattern strength", option = "viridis", limits = c(0.5, 1.55), breaks=c(0.5, 1.03, 1.55),labels=c("low", "medium", "high")) +
-  geom_point(data = records_g, fill = NA, col = "white", size = 0.7, alpha = 0.015, shape = 3) +
+  scale_fill_viridis("Relative activity increase (%)", option = "viridis", limits = c(76, 371)) +
+  geom_point(data = records_g, fill = NA, col = "white", size = 0.7, alpha = 1, shape = 3) +
   theme_bw(10) + 
@@
 # predict model results into dataframe
-data_o_plot <- cbind(data_o_plot, predict.gam(gam_cat_sp, newdata = data_o_plot, se.fit = TRUE, type = "link", exclude = c("s(station)", "s(survey_duration)")))
+data_o_plot <- cbind(data_o_plot, predict.gam(gam_cat_sp, newdata = data_o_plot, se.fit = TRUE, type = "response", exclude = c("s(station)", "s(survey_duration)")))
 data_o_plot <- rename(data_o_plot, cat_predicted = fit,  cat_predicted_se = se.fit) # rename
@@
   geom_tile()+
-  scale_fill_viridis("log(activity)", option = "viridis", limits = c(-5.27, -1.13)) +
-  geom_point(data = records_o, fill = NA, col = "white", size = 0.7, alpha = 0.1, shape = 3) +
+  scale_fill_viridis("Activity", option = "viridis") +
+  geom_point(data = records_o, fill = NA, col = "white", size = 0.7, alpha = 1, shape = 3) +
   theme_bw(10) + 
@@
 # cat plot: spatial difference
-o_cat_plot_diff <- ggplot(aes(x, y, fill = diel_diff),
+o_cat_plot_diff <- ggplot(aes(x, y, fill = diel_diff_pc),
                      data = data_o_plot) +
   geom_tile()+
-  scale_fill_viridis("Relative diel pattern strength", option = "viridis", limits = c(0.5, 1.55), breaks=c(0.5, 1.03, 1.55),labels=c("low", "medium", "high")) +
-  geom_point(data = records_o, fill = NA, col = "white", size = 0.7, alpha = 0.015, shape = 3) +
+  scale_fill_viridis("Relative activity increase (%)", option = "viridis", limits = c(76, 371)) +
+  geom_point(data = records_o, fill = NA, col = "white", size = 0.7, alpha = 1, shape = 3) +
   theme_bw(10) + 

### 2021-10-22 r_scripts/1_add_solar_times.R
comments r1

@@ -1,62 +1,0 @@
-## add average anchored solar time to species records
-# using methods from Vazquez et al. 2019 https://doi.org/10.1111/2041-210X.13290
-
-rm(list = ls())
-library(maptools)        
-library(dplyr)
-library(activity) 
-library(lubridate)
-
-# load function to change clock time to radial time for overlap package
-source("r_scripts/function_clock_time_to_radian.R")
-
-# load data
-records <- read.csv("raw_data/predator_records.csv")
-camdata <- read.csv("raw_data/camdata.csv")
-
-# remove stupid X's
-camdata <- subset(camdata, select = -c(X, X1))
-records <- subset(records, select = -c(X))
-
-# add camdata info to records 
-records <- left_join(records, camdata, by = "station_year")
-
-# take a matrix of the coordinates
-coords <- matrix(c(records$longitude, records$latitude), nrow = length(records$latitude))
-# add a coordinate reference system (can't be UTM)
-coords <- sp::SpatialPoints(coords, proj4string=sp::CRS("+proj=longlat +datum=WGS84")) 
-
-# specify detection time as a date class, as well as timezone (use Brisbane because it doesn't have daylight savings)
-records$date_time <- ymd_hms(records$date_time, tz = "Australia/Brisbane")
-
-# calculate the time of sunrise given the location and timing of each species detection 
-sunriset_df <- as.data.frame(sunriset(coords, records$date_time, direction = "sunrise", POSIXct.out=TRUE))
-# convert to radian time 
-sunrise_radian <- ClocktimeToTimeRad(sunriset_df$time) 
-# calculate the time of sunset given the location and timing of each species detection 
-sunsett_df <- as.data.frame(sunriset(coords, records$date_time, direction = "sunset", POSIXct.out=TRUE))
-# convert to radian 
-sunset_radian <- ClocktimeToTimeRad(sunsett_df$time) 
-
-# save these as a separate matrix for the transtime function below 
-anchors <- matrix(c(sunrise_radian, sunset_radian), nrow = length(sunset_radian))
-
-# we also need to convert the detection time to radian time
-det_radian <- ClocktimeToTimeRad(records$date_time)
-
-# we can now calculate the average anchored time (time relative to both sunrise and sunset) for every detection
-det_anchored <- transtime(det_radian, anchors)
-
-# add to dataframe
-records$aa_radian <- det_anchored
-# as well as in clock time (new cols)
[…]

### 2021-10-22 r_scripts/3_remove_repeat_detections.R

@@ -1,37 +1,0 @@
-# REMOVE REPEAT OBSERVATIONS FOR SPECIES WIHIN 30 MINUTES AT A CAMERA FOR APPROXIMATE TEMPORAL INDEPENDENCE
-
-rm(list = ls())
-library(dplyr)
-library(lubridate)
-
-# load records
-records <- read.csv("derived_data/records_solar.csv")
-
-# date_time as date class
-records$date_time <- ymd_hms(records$date_time, tz = "Australia/Brisbane")
-
-## remove records of the same species at a particular cam-trap within 30 minutes
-time_to_independence <- 1800 # number of seconds in 30 minutes
-
-# get time (in seconds) since last visit
-records <- records %>%
-  group_by(station_year, species) %>%
-  arrange(date_time) %>%
-  mutate(seconds_diff = date_time - lag(date_time))
-
-# change NA's to a value greater than the time to independence so they get kept when we filter out 
-records$seconds_diff[is.na(records$seconds_diff)] <- time_to_independence + 1
-
-# filter out records less than 60 minutes
-records <- filter(records, seconds_diff > time_to_independence)
-
-# remove seconds_diff col
-records$seconds_diff <- NULL
-
-# looks good?
-head(records)
-
-# save
-write.csv(records, "derived_data/records_solar_ind.csv")
-
-# END

### 2021-10-22 r_scripts/4_reshape_by_hour.R

@@ -1,75 +1,0 @@
-## FORMAT RECORDS INTO COUNTS PER HOUR FOR EACH CAMERA-TRAP DEPLOYOMENT 
-
-rm(list = ls())
-library(readr)
-library(dplyr)
-library(reshape2)
-library(lubridate)
-library(maptools)     
-
-# load function to change clock time to radial time for overlap package
-source("r_scripts/function_clock_time_to_radian.R")
-
-# load data
-records <- read.csv("derived_data/records_solar_ind.csv")
-camdata <- read.csv("raw_data/camdata.csv")
-
-
-# DATAFRAME 1:  HOUR ----------------------------------------------------
-# need to do this for summing
-records$occ <- 1
-
-# make solar hour an integer, of character class
-records$hour <- as.integer(records$aa_clock)
-
-# melt the dataframe
-gam_data <- melt(records, id.var = colnames(records), measure.var = "occ")
-
-# cast to derive count per year per station_year long format
-gam_data = dcast(gam_data, station_year + hour ~ species, fill = 0, fun = sum)
-
-# make a table with all unique combinations - (use camdata so we don't just take the sites with detections)
-df <- expand.grid(station_year = unique(camdata$station_year), hour = 0:23)
-
-# merge into gam_data so every hour for every station is provided (even with no predator detections) - fills missing values with NA
-gam_data <- left_join(df, gam_data)
-# change NA's to 0's
-gam_data$fox <- ifelse(is.na(gam_data$fox), 0, gam_data$fox)
-gam_data$cat <- ifelse(is.na(gam_data$cat), 0, gam_data$cat)
-unique(is.na(gam_data)) # check it worked
-
-# sort by station, hour
-gam_data <- arrange(gam_data, station_year, hour)
-head(gam_data)
-
-## merge in site covariates
-gam_data <- left_join(gam_data, camdata, by = "station_year")
-head(gam_data)
-
-
-# ADD SUNSET SUNRISE TIMES FOR STATIONS ------------------------------------------
-# make date class
-gam_data$date_start <- ymd(gam_data$date_start, tz = "Australia/Brisbane")
-gam_data$date_end <- ymd(gam_data$date_end, tz = "Australia/Brisbane")
-# take a matrix of the coordinates
-coords <- matrix(c(gam_data$longitude, gam_data$latitude), nrow = length(gam_data$latitude))
-# add a coordinate reference system (can't be UTM)
[…]

### 2021-10-22 r_scripts/5_gams_diel_activity.R

@@ -1,264 +1,0 @@
-## FIT GAMS TO MODEL DIEL ACTIVITY FOR FOXES AND FERAL CATS
-# MATTHEW REES
-
-# SETUP ------------------------------------------------------------
-library(dplyr)
-library(mgcv)
-library(ggplot2)
-library(patchwork)
-library(viridis)
-library(gratia)
-library(chron)
-library(sf)
-
-rm(list = ls())
-options(scipen = 999) 
-
-# Set the default theme for ggplot objects 
-theme_set(theme_bw())
-theme_update(panel.grid = element_blank())
-
-
-# PREPARE DATA ------------------------------------------------------------
-## load gam data - 23 rows for each hour of every deployment
-records <- read.csv("derived_data/counts_hour.csv")
-
-# drop zoi's data 
-records <- filter(records, data_source != "zoi")
-
-# exclude stations left out for less than 7 days - something musta went wrong with those few cams
-records <- filter(records, survey_duration >= 7)
-hist(records$survey_duration / 23, breaks = 50)
-summary(records$survey_duration)
-
-# as we only had three "Riverine Grassy Woodlands or Forests" unique survey sites --> drop em
-records <- filter(records, XGROUPNAME != "Riverine Grassy Woodlands or Forests")
-
-# as we only had 20 "Rainforests" unique survey sites, all of which interspersed in "Wet Forests" in the Otways, reclassify as "Wet Forests". 
-records$XGROUPNAME <- if_else(records$XGROUPNAME == "Rainforests", "Wet or Damp Forests", as.character(records$XGROUPNAME))
-
-# abbreviate EVC group names for plotting
-records$XGROUPNAME <- if_else(records$XGROUPNAME == "Wet or Damp Forests", "Wet Forests", records$XGROUPNAME)
-records$XGROUPNAME <- if_else(records$XGROUPNAME == "Riparian Scrubs or Swampy Scrubs and Woodlands", "Swampy Scrubs", records$XGROUPNAME)
-records$XGROUPNAME <- substr(records$XGROUPNAME, 1, nchar(records$XGROUPNAME) - 1)
-table(records$XGROUPNAME)
-
-# add xy coordinates
-records_sf <- st_as_sf(records, coords = c("longitude", "latitude"), crs = 4326) %>% 
-  st_transform(crs = 32754)
-records$x <- st_coordinates(records_sf)[,1]
-records$y <- st_coordinates(records_sf)[,2]
-head(records)
-
-
[…]

### 2021-10-22 r_scripts/7_coat_cat_variation.R

@@ -1,334 +1,0 @@
-## REFORMAT CAT IDENTIFIED RECORDS FOR DIEL ACTIVITY GAMS
-
-rm(list = ls())
-library(stringr)
-library(maptools)        
-library(dplyr)
-library(activity) 
-library(lubridate)
-library(reshape2)
-library(mgcv)
-library(gratia)
-library(ggplot2)
-library(patchwork)
-
-# load function to change clock time to radial time for overlap package
-source("r_scripts/function_clock_time_to_radian.R")
-
-# Set the default theme for ggplot objects 
-theme_set(theme_bw())
-theme_update(panel.grid = element_blank())
-
-# load cat records
-records <- read.csv("raw_data/cat_id_detections.csv")
-records$X <- NULL
-
-# load camdata - but just matts deployments
-camdata <- read.csv("raw_data/camdata.csv")
-camdata <- filter(camdata, data_source == "matt")
-camdata$X <- NULL
-camdata$X.1 <- NULL
-
-
-
-# ADD SOLAR TIMES ---------------------------------------------------------
-# take a matrix of the coordinates
-coords <- matrix(c(records$longitude, records$latitude), nrow = length(records$latitude))
-# add a coordinate reference system (can't be UTM)
-coords <- sp::SpatialPoints(coords, proj4string=sp::CRS("+proj=longlat +datum=WGS84")) 
-
-# specify detection time as a date class, as well as timezone (use Brisbane because it doesn't have daylight savings)
-records$date_time <- ymd_hms(records$date_time, tz = "Australia/Brisbane")
-
-# calculate the time of sunrise given the location and timing of each species detection 
-sunriset_df <- as.data.frame(sunriset(coords, records$date_time, direction = "sunrise", POSIXct.out=TRUE))
-# convert to radian time 
-sunrise_radian <- ClocktimeToTimeRad(sunriset_df$time) 
-# calculate the time of sunset given the location and timing of each species detection 
-sunsett_df <- as.data.frame(sunriset(coords, records$date_time, direction = "sunset", POSIXct.out=TRUE))
-# convert to radian 
-sunset_radian <- ClocktimeToTimeRad(sunsett_df$time) 
-
-# save these as a separate matrix for the transtime function below 
-anchors <- matrix(c(sunrise_radian, sunset_radian), nrow = length(sunset_radian))
-
-# we also need to convert the detection time to radian time
-det_radian <- ClocktimeToTimeRad(records$date_time)
-
[…]

### 2021-10-22 r_scripts/Modified_function_for_detection_histories_by_minute.R

@@ -1,714 +1,0 @@
-# Get occupancy detection histories by hours ------------------------------
-library(camtrapR)
-
-# modify function detectionHistory in package to get detection histories of 1-hour interval: modify lines 479, 483 (x2) and 549.
-
-# Step 1: define helper functions
-# for detectionHistory functions
-
-checkCamOpColumnNames <- function(cameraOperationMatrix){
-  camopTest <- try(as.Date(colnames(cameraOperationMatrix)), silent = TRUE)
-  if(class(camopTest) == "try-error") stop(paste('could not interpret column names in camOp as Dates. Desired format is YYYY-MM-DD, e.g. "2016-12-31". First column name in your camera operation matrix is "', colnames(cameraOperationMatrix)[1], '"', sep = '' ), call. = FALSE)
-}
-
-
-createDateRangeTable <- function(cam.op,
-                                 subset_species_tmp,
-                                 buffer_tmp,
-                                 stationCol_tmp,
-                                 day1_tmp,
-                                 occasionStartTime_tmp,
-                                 maxNumberDays_tmp,
-                                 timeZone_tmp)
-{
-  
-  cam.tmp.min <- apply(cam.op, MARGIN = 1, function(X){min(which(!is.na(X)))})    # 1st day of each station
-  cam.tmp.max <- apply(cam.op, MARGIN = 1, function(X){max(which(!is.na(X)))})    # last day of each station
-  
-  rec.tmp.min  <- aggregate(as.Date(subset_species_tmp$DateTime2, tz = timeZone_tmp),
-                            list(subset_species_tmp[,stationCol_tmp]),
-                            FUN = min)
-  rec.tmp.max  <- aggregate(as.Date(subset_species_tmp$DateTime2, tz = timeZone_tmp),
-                            list(subset_species_tmp[,stationCol_tmp]),
-                            FUN = max)
-  
-  
-  
-  date_ranges <- data.frame(rec.min = rec.tmp.min[match(rownames(cam.op), rec.tmp.min[,1]), 2],       # first record
-                            rec.max = rec.tmp.max[match(rownames(cam.op), rec.tmp.max[,1]), 2],       # last record
-                            cam.min = as.POSIXct(colnames(cam.op)[cam.tmp.min], tz = timeZone_tmp),   # station setup date
[…]

### 2021-10-22 r_scripts/format_records_hour_gams.R

@@ -1,0 +1,178 @@
+# load r packages
+library(maptools)        
+library(dplyr)
+library(activity) 
+library(lubridate)
+library(reshape2)
+library(sf)
+# note: chron package is loaded further down
+
+# load function to change clock time to radial time for overlap package
+source("r_scripts/function_clock_time_to_radian.R")
+
+
+# LOAD / PREPARE DATA -----------------------------------------------------
+
+# load data
+records <- read.csv("raw_data/predator_records.csv")
+camdata <- read.csv("raw_data/camdata.csv")
+# remove stupid X's
+camdata <- subset(camdata, select = -c(X, X1))
+records <- subset(records, select = -c(X))
+# add camdata info to records - NA for sites without pred detections?
+records <- left_join(records, camdata, by = c("station", "station_year"))
+head(records)
+
+
+# 1) ADD SOLAR TIMES -------------------------------------------------------
+## add average anchored solar time to species records
+# using methods from Vazquez et al. 2019 https://doi.org/10.1111/2041-210X.13290
+
+# take a matrix of the coordinates
+coords <- matrix(c(records$longitude, records$latitude), nrow = length(records$latitude))
+# add a coordinate reference system (can't be UTM)
+coords <- sp::SpatialPoints(coords, proj4string=sp::CRS("+proj=longlat +datum=WGS84")) 
+# specify detection time as a date class, as well as timezone (use Brisbane because it doesn't have daylight savings)
+records$date_time <- ymd_hms(records$date_time, tz = "Australia/Brisbane")
+records$date_start <- ymd(records$date_start, tz = "Australia/Brisbane")
+records$date_end <- ymd(records$date_end, tz = "Australia/Brisbane")
+
+
+## ADD ADJUSTED HOUR FOR EACH SPECIES DETECTION
+# calculate the time of sunrise given the location and timing of each species detection 
+sunriset_df <- as.data.frame(sunriset(coords, records$date_time, direction = "sunrise", POSIXct.out=TRUE))
+# convert to radian time 
+sunrise_radian <- ClocktimeToTimeRad(sunriset_df$time) 
+# calculate the time of sunset given the location and timing of each species detection 
+sunsett_df <- as.data.frame(sunriset(coords, records$date_time, direction = "sunset", POSIXct.out=TRUE))
[…]

### 2021-10-22 r_scripts/lorelograms/Modified_function_for_detection_histories_by_minute.R

@@ -1,0 +1,714 @@
+# Get occupancy detection histories by hours ------------------------------
+library(camtrapR)
+
+# modify function detectionHistory in package to get detection histories of 1-hour interval: modify lines 479, 483 (x2) and 549.
+
+# Step 1: define helper functions
+# for detectionHistory functions
+
+checkCamOpColumnNames <- function(cameraOperationMatrix){
+  camopTest <- try(as.Date(colnames(cameraOperationMatrix)), silent = TRUE)
+  if(class(camopTest) == "try-error") stop(paste('could not interpret column names in camOp as Dates. Desired format is YYYY-MM-DD, e.g. "2016-12-31". First column name in your camera operation matrix is "', colnames(cameraOperationMatrix)[1], '"', sep = '' ), call. = FALSE)
+}
+
+
+createDateRangeTable <- function(cam.op,
+                                 subset_species_tmp,
+                                 buffer_tmp,
+                                 stationCol_tmp,
+                                 day1_tmp,
+                                 occasionStartTime_tmp,
+                                 maxNumberDays_tmp,
+                                 timeZone_tmp)
+{
+  
+  cam.tmp.min <- apply(cam.op, MARGIN = 1, function(X){min(which(!is.na(X)))})    # 1st day of each station
+  cam.tmp.max <- apply(cam.op, MARGIN = 1, function(X){max(which(!is.na(X)))})    # last day of each station
+  
+  rec.tmp.min  <- aggregate(as.Date(subset_species_tmp$DateTime2, tz = timeZone_tmp),
+                            list(subset_species_tmp[,stationCol_tmp]),
+                            FUN = min)
+  rec.tmp.max  <- aggregate(as.Date(subset_species_tmp$DateTime2, tz = timeZone_tmp),
+                            list(subset_species_tmp[,stationCol_tmp]),
+                            FUN = max)
+  
+  
+  
+  date_ranges <- data.frame(rec.min = rec.tmp.min[match(rownames(cam.op), rec.tmp.min[,1]), 2],       # first record
+                            rec.max = rec.tmp.max[match(rownames(cam.op), rec.tmp.max[,1]), 2],       # last record
+                            cam.min = as.POSIXct(colnames(cam.op)[cam.tmp.min], tz = timeZone_tmp),   # station setup date
[…]

### 2021-10-22 r_scripts/lorelograms/lorelograms.R

@@ -1,0 +1,72 @@
+# Identify the minimum interval to approximate independence
+# code and methods adapated from Iannarilli et al. (2019).
+
+#' Load lorelogram package
+if (!require("devtools")) install.packages("devtools", 
+                                           repos = "http://cran.us.r-project.org", 
+                                           dependencies = "Imports")
+install.packages("usethis")
+library(usethis)
+devtools::install_github("FabiolaIannarilli/lorelogram") 
+library(lorelogram)
+library(camtrapR)
+library(dplyr)
+
+# load updated functions from Iannarilli et al. (2019).
+source("r_scripts/Modified_function_for_detection_histories_by_minute.R")
+
+# load data
+records <- read.csv("derived_data/records_solar.csv")
+camdata <- read.csv("raw_data/camdata.csv")
+
+
+## FILTER TO 1 GRID DEPLOYMENT AT A TIME  (memory limit exhausts / R crashes when you try to do any more)
+records <- filter(records, data_source == "matt" & block == "annya")
+camdata <- filter(camdata, data_source == "matt" & block == "annya")
+
+# make camera operation matrix using camtrapR package
+camop <- cameraOperation(CTtable = camdata, stationCol = "station_year", setupCol = "date_start", retrievalCol = "date_end", writecsv = FALSE, hasProblems = FALSE, dateFormat = "%Y-%m-%d")
+write.csv(camop, "derived_data/camop.csv") # have to save and reload to get it in the right format for the function below
+camop_problem <- read.csv("derived_data/camop.csv", header=TRUE, check.names = FALSE) 
+row.names(camop_problem) <- camop_problem[,1]
+camop_problem <- camop_problem[,-1] #to remove first column with station name 
+
+# run modified dethist function - written by Iannarilli et al. (2019)
+DetHist_Minute_cat <- myfunc_detectionHistory_minute(recordTable = records,
+                                                     species = "cat", #change species name
+                                                     camOp = camop_problem,
+                                                     stationCol =  "station_year",
+                                                     speciesCol = "species",
+                                                     recordDateTimeCol = "date_time", 
+                                                     recordDateTimeFormat = "%Y-%m-%d %H:%M:%S",
+                                                     occasionLength = 10, 
+                                                     includeEffort = FALSE,
+                                                     scaleEffort = FALSE, 
+                                                     day1 = "survey", 
+                                                     timeZone="Australia/Brisbane") #adjust time zone
+
[…]

### 2021-10-22 r_scripts/models.R

@@ -1,0 +1,108 @@
+## FIT GAMS TO MODEL DIEL ACTIVITY FOR FOXES AND FERAL CATS
+# MATTHEW REES
+
+library(dplyr)
+library(mgcv)
+
+# load data
+records <- read.csv("derived_data/predator_counts_hour.csv")
+
+
+# Transform variables -----------------------------------------------------
+
+# new variable for fox counts adjusted for (log) survey effort (for presences only - we don't want negative counts)
+records$fox_count_adj <- ifelse(records$fox_count > 0, records$fox_count/log(records$survey_duration), records$fox_count)
+
+# take log 
+records$fox_count_adj <- log(records$fox_count_adj + 1)
+
+## Transform variable class  for GAMs
+records <- transform(records,
+                     hour = as.integer(hour),
+                     cat = as.integer(cat), 
+                     fox = as.integer(fox), 
+                     foxbaits = as.integer(foxbaits), 
+                     fox_pa = factor(fox_pa, ordered = FALSE), 
+                     fox_count_adj = as.numeric(fox_count_adj), 
+                     region = factor(region, ordered = FALSE), 
+                     station = factor(station, ordered = FALSE), 
+                     vegetation_group = factor(XGROUPNAME, ordered = FALSE), 
+                     habitat_type = factor(habitat_type, ordered = FALSE), 
+                     survey_duration = as.integer(survey_duration)
+)
+
+# rename habitat type variable (for plotting)
+records$habitat_type <- recode_factor(records$habitat_type,  "dry_glenelg" = "Dry vegetation (Glenelg region)",
+                                      "dry_otway" = "Dry vegetation (Otway Ranges)",
+                                      "wet_otway" = "Wet forest (Otway Ranges)")
+
+
+
+# Changes in spatial activity across the daily cycle ----------------------
+
+gam_fox_sp <- bam(fox ~ t2(x, y, hour, d = c(2, 1), bs = c("ds", "cc"), k = c(80, 8), m = c(1, 0.5), full = TRUE) +
+                        s(foxbaits, bs = "tp", k = 4) + 
+                        s(station, bs = "re") +  
+                        offset(log(survey_duration)), 
+                   data = records, family = nb, knots = list(hour = c(0, 23)), nthreads = 3, discrete = TRUE)
+
+
[…]

### 2021-10-22 r_scripts/models_individual_cats.R

@@ -1,0 +1,321 @@
+## REFORMAT CAT IDENTIFIED RECORDS FOR DIEL ACTIVITY GAMS
+
+rm(list = ls())
+library(stringr)
+library(maptools)        
+library(dplyr)
+library(activity) 
+library(lubridate)
+library(reshape2)
+library(mgcv)
+library(gratia)
+library(ggplot2)
+library(patchwork)
+
+# load function to change clock time to radial time for overlap package
+source("r_scripts/function_clock_time_to_radian.R")
+
+# Set the default theme for ggplot objects 
+theme_set(theme_bw())
+theme_update(panel.grid = element_blank())
+
+# load cat records
+records <- read.csv("raw_data/cat_id_detections.csv")
+records$X <- NULL
+
+# load camdata - but just matts deployments
+camdata <- read.csv("raw_data/camdata.csv")
+camdata <- filter(camdata, data_source == "matt")
+camdata$X <- NULL
+camdata$X.1 <- NULL
+
+
+
+# ADD SOLAR TIMES ---------------------------------------------------------
+# take a matrix of the coordinates
+coords <- matrix(c(records$longitude, records$latitude), nrow = length(records$latitude))
+# add a coordinate reference system (can't be UTM)
+coords <- sp::SpatialPoints(coords, proj4string=sp::CRS("+proj=longlat +datum=WGS84")) 
+
+# specify detection time as a date class, as well as timezone (use Brisbane because it doesn't have daylight savings)
+records$date_time <- ymd_hms(records$date_time, tz = "Australia/Brisbane")
+
+# calculate the time of sunrise given the location and timing of each species detection 
+sunriset_df <- as.data.frame(sunriset(coords, records$date_time, direction = "sunrise", POSIXct.out=TRUE))
+# convert to radian time 
+sunrise_radian <- ClocktimeToTimeRad(sunriset_df$time) 
+# calculate the time of sunset given the location and timing of each species detection 
+sunsett_df <- as.data.frame(sunriset(coords, records$date_time, direction = "sunset", POSIXct.out=TRUE))
+# convert to radian 
+sunset_radian <- ClocktimeToTimeRad(sunsett_df$time) 
+
+# save these as a separate matrix for the transtime function below 
+anchors <- matrix(c(sunrise_radian, sunset_radian), nrow = length(sunset_radian))
+
+# we also need to convert the detection time to radian time
+det_radian <- ClocktimeToTimeRad(records$date_time)
+
[…]

### 2021-10-22 r_scripts/plot_cat_v_fox.R

@@ -1,0 +1,74 @@
+# PLOT FIGURE
+
+## CAT PLOT - 3D 
+df = expand.grid(hour = 0:23,
+                 habitat_type = levels(records$habitat_type),
+                 longitude = mean(records$longitude),
+                 latitude = mean(records$latitude),
+                 fox_count_adj = seq(min(records$fox_count_adj), max(records$fox_count_adj), length=150),
+                 station = "T052",
+                 survey_duration = mean(records$survey_duration))
+# predict model results into dataframe
+x = sapply(gam_cat_fox$smooth, "[[",  "label")
+df_int <- cbind(df, predict(gam_cat_fox, newdata = df, se.fit = TRUE, type = "link", exclude = c(x[4:5], "s(survey_duration)")))
+# plot
+plot_range <- ggplot(aes(hour, fox_count_adj, fill = fit), data = df_int) +
+  geom_tile() +
+  facet_wrap(~habitat_type, nrow =1) +
+  scale_fill_viridis("Cat activity", option = "viridis", breaks = c(-5, -3), labels = c("Low", "High")) + 
+  ggtitle("", subtitle = "Feral cat") +
+  geom_vline(xintercept = c(6.16,18.34), colour = "black", size = 0.6, linetype="dotted") + 
+  scale_y_continuous(breaks=c(0, 3), labels = c("Low", "High")) +
+  ylab("Fox activity") + 
+  xlab("Hour") 
+
+## CAT PLOT - 2D (same model at min / max fox)
+# min df 
+df_int_min <- df_int %>%
+  group_by(habitat_type) %>%
+  slice_min(fox_count_adj, n = 1) %>%
+  mutate(minmax = "0min")
+# max df 
+df_int_max <- df_int %>%
+  group_by(habitat_type) %>%
+  slice_max(fox_count_adj, n = 1) %>%
+  mutate(minmax = "1max")
+# join df
+df_int_minmax <- rbind(df_int_min, df_int_max)
+# plot
+plot_wet_lowhigh <- ggplot(data=df_int_minmax, aes(x=hour, y=fit, group=minmax, col = minmax)) +
+  geom_line(aes(y=fit, x=hour), lwd = 1.2) +
+  facet_wrap(~habitat_type, nrow =1) +
+  ggtitle("", subtitle = "Feral cat") +
[…]

### 2021-10-22 r_scripts/plot_predator_spatiotemporal.R

@@ -1,0 +1,230 @@
+# PLOT SPATIO-TEMPORAL MODELS
+# 1) DIEL ACTIVITY STRENGTH ACROSS SPACE
+# 2) ANIMATED GIF FOR SPATIOTEMPORAL ACTIVITY
+
+# additional packages
+library(gganimate)
+library(sp)
+library(sf)
+
+# split records
+records_g <- filter(records, region == "glenelg")
+records_o <- filter(records, region == "otways")
+
+# cam locations 
+camdata <- records %>%
+  distinct(., station, .keep_all = TRUE) %>%
+  select(region, station, x, y) 
+# add new variable to match variable we will facet by - region x species
+camdata_fox <- camdata
+camdata_fox$species <- "Red fox"
+camdata_cat <- camdata
+camdata_cat$species <- "Feral cat"
+camdata <- rbind(camdata_fox, camdata_cat)
+camdata$facet <- paste0(camdata$species, "_", camdata$region)
+unique(camdata$facet)
+
+# MAKE SPATIAL DATAFRAMES TO PREDICT INTO ----------------------------------------------
+
+## GLENELG:
+# take just the cams
+records_g_cams <- distinct(records_g, station, .keep_all = TRUE)
+# change it to sf class
+records_g_cams <- st_as_sf(records_g_cams, coords = c("x", "y"), crs = 32754) 
+# make a 4km buffer around each camera
+glenelg_cams_buffer = st_buffer(records_g_cams, 4000)
+# dissolve the buffer
+glenelg_cams_buffer = st_union(glenelg_cams_buffer)
+# convert back to dataframe
+glenelg_buffer_df <- fortify(as_Spatial(glenelg_cams_buffer))%>%
+  transmute(x = long, y = lat, order = order, group = group)
+# split by each group (seperate polygon for each grid)
+buffer_df1 <- glenelg_buffer_df[which(glenelg_buffer_df$group == "ID1.1"),]
+buffer_df2 <- glenelg_buffer_df[which(glenelg_buffer_df$group == "ID1.2"),]
+buffer_df3 <- glenelg_buffer_df[which(glenelg_buffer_df$group == "ID1.3"),]
+buffer_df4 <- glenelg_buffer_df[which(glenelg_buffer_df$group == "ID1.4"),]
+# make a dataframe from this to predict into
+data_g_plot = expand.grid(
+  x = seq(min(glenelg_buffer_df$x), 
[…]

### 2021-10-22 r_scripts/plot_predator_vegetation_type.R

@@ -1,0 +1,44 @@
+
+
+# function to run and plot GAMs for each species
+gam_plot <- function(gam_model, data){
+  ## Plot global smooth (using the gratia package)
+  plot_global <- draw(evaluate_smooth(gam_model, overall_uncertainty = TRUE, "s(hour)"), xlab = "Hour",  title = "", subtitle = "i.   Average") + 
+    geom_vline(xintercept = c(6.16,18.34), colour = "black", size = 0.6, linetype="dotted") 
+  ## Plot the difference smooths using the gratia package
+  plot_difference <- draw(evaluate_smooth(gam_model, "s(hour,vegetation_group)"), xlab = "Hour",  title = "", subtitle = "ii.   Vegetation-specific deviations from the average") + 
+    geom_vline(xintercept = c(6.16,18.34), colour = "black", size = 0.6, linetype="dotted") 
+  ## Plot the actual hourly activity for each vegetation type   
+  # make a new dataframe to predict into
+  df <- expand.grid(hour = 0:23,
+                    vegetation_group = levels(data$vegetation_group),
+                    foxbaits = 0, 
+                    region = c("glenelg", "otways"),
+                    survey_duration = mean(data$survey_duration))
+  # get smooths to exlude
+  x = sapply(gam_model$smooth, "[[",  "label")
+  # predict model estimates (and uncertainty) into this dataframe, excluding the impact of variables other than hour:
+  df <- cbind(df, predict.gam(gam_model, newdata = df, se.fit = TRUE, type = "link", newdata.guaranteed = TRUE, exclude = c(x[3:7], "s(survey_duration)")))
+  plot_veg <- ggplot(data=df, aes(x=hour, y=fit, group=vegetation_group)) +
+    # ylim(-13, -2) +
+    geom_line(aes(y=fit, x=hour), lwd = 0.7) +
+    facet_wrap(~vegetation_group, nrow =1) +
+    geom_ribbon(aes(ymin=(fit-2*se.fit), ymax=(fit+2*se.fit)), alpha=0.2) +
+    geom_vline(xintercept = c(6.16,18.34), colour = "black", size = 0.6, linetype="dotted") + 
+    labs(subtitle = "iii.   Predicted activity", x = "Hour", y = "log(count)") 
+  ## Assemble plots
+  patch <- (plot_global | plot_difference) / plot_veg 
+  patch
+}
+
+# save
[…]

### 2021-10-22 README.Rmd
test

---
output: github_document
---

# diel-activity-invasive-predators

test

### 2021-10-22 r_scripts/plot_predator_spatiotemporal.R
typo

@@ -164,7 +164,7 @@ camdata_o <- filter(camdata, region == "otways")
 plot_diff_g <- ggplot(aes(x, y, fill = diel_diff_pc), data = model_predictions_g) +
    geom_tile() +
    facet_wrap(~species, scales = "free") +
-   scale_fill_viridis("Difference (%) in activity throughout the daiy cycle", option = "viridis") +
+   scale_fill_viridis("Difference (%) in activity throughout the daily cycle", option = "viridis") +
    geom_point(data = camdata_g, fill = NA, col = "white", size = 0.7, alpha = 0.3, shape = 3) +
    theme_bw(10) + 
    ggtitle("Glenelg region") +
@@ -175,7 +175,7 @@ plot_diff_g <- ggplot(aes(x, y, fill = diel_diff_pc), data = model_predictions_g
 plot_diff_o <- ggplot(aes(x, y, fill = diel_diff_pc), data = model_predictions_o) +
   geom_tile() +
   facet_wrap(~species, scales = "free") +
-  scale_fill_viridis("Difference (%) in activity throughout the daiy cycle", option = "viridis") +
+  scale_fill_viridis("Difference (%) in activity throughout the daily cycle", option = "viridis") +
   geom_point(data = camdata_o, fill = NA, col = "white", size = 0.7, alpha = 0.3, shape = 3) +
   theme_bw(10) + 
   ggtitle("Otway Ranges") +

### 2021-10-22 README.Rmd
test

@@
 
-# diel-activity-invasive-predators
+<!-- README.md is generated from README.Rmd. Please edit that file -->
 
-test
+```{r, echo = FALSE}
+knitr::opts_chunk$set(
+  collapse = TRUE,
+  comment = "#>",
+  fig.path = "man/figures/README-"
+)
+```
+
+## Overview
+
+ggplot2 is a system for declaratively creating graphics, based on [The Grammar of Graphics][gg-book].
+You provide the data, tell ggplot2 how to map variables to aesthetics, what graphical primitives to use, and it takes care of the details.
+
+## Installation
+
+```{r, eval = FALSE}
+# The easiest way to get ggplot2 is to install the whole tidyverse:
+install.packages("tidyverse")
+# Alternatively, install just ggplot2:
+install.packages("ggplot2")
+# Or the development version from GitHub:
+# install.packages("devtools")
+devtools::install_github("tidyverse/ggplot2")
+```
+
+## Cheatsheet
+
+<a href="https://github.com/rstudio/cheatsheets/blob/master/data-visualization.pdf"><img src="https://raw.githubusercontent.com/rstudio/cheatsheets/master/pngs/thumbnails/data-visualization-cheatsheet-thumbs.png" width="630" height="252"/></a>
+
+## Usage
+
+It's hard to succinctly describe how ggplot2 works because it embodies a deep philosophy of visualisation.
+However, in most cases you start with `ggplot()`, supply a dataset and aesthetic mapping (with `aes()`).
+You then add on layers (like `geom_point()` or `geom_histogram()`), scales (like `scale_colour_brewer()`), faceting specifications (like `facet_wrap()`) and coordinate systems (like `coord_flip()`).
+
+```{r example}
+library(ggplot2)
+ggplot(mpg, aes(displ, hwy, colour = class)) + 
+  geom_point()
+```
+
+## Lifecycle
+
+[![lifecycle](https://img.shields.io/badge/lifecycle-stable-brightgreen.svg)](https://lifecycle.r-lib.org/articles/stages.html)
+
+ggplot2 is now over 10 years old and is used by hundreds of thousands of people to make millions of plots.
+That means, by-and-large, ggplot2 itself changes relatively little.
+When we do make changes, they will be generally to add new functions or arguments rather than changing the behaviour of existing functions, and if we do make changes to existing behaviour we will do them for compelling reasons.
+
+If you are looking for innovation, look to ggplot2's rich ecosystem of extensions. See a community maintained list at <https://exts.ggplot2.tidyverse.org/gallery/>.
+
+## Learning ggplot2
+
+If you are new to ggplot2 you are better off starting with a systematic introduction, rather than trying to learn from reading individual documentation pages.
+Currently, there are three good places to start:
+
+1.  The [Data Visualisation][r4ds-vis] and
+    [Graphics for communication][r4ds-comm] chapters in 
+    [R for Data Science][r4ds]. R for Data Science is designed to
+    give you a comprehensive introduction to the
+    [tidyverse](https://www.tidyverse.org), and these two chapters will
+    get you up to speed with the essentials of ggplot2 as quickly as 
+    possible.
+
+1.  If you'd like to take an online course, try
+    [Data Visualization in R With ggplot2][oreilly] by Kara Woo.
+
+1.  If you'd like to follow a webinar, try
+    [Plotting Anything with ggplot2](https://youtu.be/h29g21z0a68) by Thomas Lin 
+    Pedersen.
+
+1.  If you want to dive into making common graphics as quickly 
+    as possible, I recommend [The R Graphics Cookbook][cookbook] 
+    by Winston Chang. It provides a set of recipes to solve common
+    graphics problems.
+
+If you've mastered the basics and want to learn more, read [ggplot2: Elegant Graphics for Data Analysis][ggplot2-book].
+It describes the theoretical underpinnings of ggplot2 and shows you how all the pieces fit together.
+This book helps you understand the theory that underpins ggplot2, and will help you create new types of graphics specifically tailored to your needs.
+
+## Getting help
+
+There are two main places to get help with ggplot2:
+
+1.  The [RStudio community][community] is a friendly place to ask any
+    questions about ggplot2.
+
+1.  [Stack Overflow][so] is a great source of answers to common ggplot2
+    questions. It is also a great place to get help, once you have
+    created a reproducible example that illustrates your problem.
+
+[community]: https://community.rstudio.com/
+[ggplot2-book]: https://ggplot2-book.org
+[gg-book]: https://www.amazon.com/Grammar-Graphics-Statistics-Computing/dp/0387245448/ref=as_li_ss_tl
+[so]: https://stackoverflow.com/questions/tagged/ggplot2?sort=frequent&pageSize=50
+[cookbook]: https://r-graphics.org
+[r4ds]: https://r4ds.had.co.nz
+[r4ds-vis]: https://r4ds.had.co.nz/data-visualisation.html
+[r4ds-comm]: https://r4ds.had.co.nz/graphics-for-communication.html
+[oreilly]: https://learning.oreilly.com/videos/data-visualization-in/9781491963661/

### 2021-10-22 README.Rmd
gifs

--- 2021-10-22 README.Rmd
@@
 
-# diel-activity-invasive-predators
+<!-- README.md is generated from README.Rmd. Please edit that file -->
 
-test
+```{r, echo = FALSE}
+knitr::opts_chunk$set(
+  collapse = TRUE,
+  comment = "#>")
+```
+
+## Overview
+
+ggplot2 is a system for declaratively creating graphics, based on [The Grammar of Graphics][gg-book].
+You provide the data, tell ggplot2 how to map variables to aesthetics, what graphical primitives to use, and it takes care of the details.
+
+## Installation
+
+### Glenelg region
+
+```{r glenelg, echo = FALSE}
+knitr::include_graphics("figs/glenelg_predator_activity.gif")
+```
+
+### Otway ranges
+
+```{r otways, echo = FALSE}
+knitr::include_graphics("figs/otways_predator_activity.gif")
+```

### 2021-10-22 README.Rmd
update readme

@@
 
-ggplot2 is a system for declaratively creating graphics, based on [The Grammar of Graphics][gg-book].
-You provide the data, tell ggplot2 how to map variables to aesthetics, what graphical primitives to use, and it takes care of the details.
+This repository contains the data and code of my fourth PhD data chapter.
 
-## Installation
+Here, I've built generalised additive models ('GAMs'), using the ['mgcv' R-package](https://cran.r-project.org/web/packages/mgcv/mgcv.pdf), of invasive predator spatiotemporal activity in south-west Victoria, Australia.
 
-### Glenelg region
+This paper is somewhat split between an exploratory analysis of the different ways GAMs can be used to jointly model spatial and diel activity, and aiming to provide inference on the drivers of invasive predator spatiotemporal activity in temperate, heterogeneous landscapes.
+Particularly whether a subordinate mesopredator (feral cat *Felis catus*) exhibits (signs of) behavioural avoidance of an apex predator (Red fox *Vulpes vulpes*).
+
+There are four main model types, differing in how the hourly smooth was specified:
+
+1. Predator activity across space (tensor product of hour and coordinates)
+2. Predator activity across vegetation types (hierarchical model of hour x vegetation type)
+3. Feral cat activity potential avoidance of foxes (tensor product of hour and coordinates)
+4. Variation in diel activity of feral cats (hierarchical model of hour x individual ID)
+
+## Acknowledgements
+
+The idea to use GAMs for this chapter came from Prof.
+Nick Golding - Nick also wrote me some initial simulation code and helped me get started with GAMs.
+I also learnt most of these methods and gained inspiration from [online course materials](https://noamross.github.io/mgcv-esa-2018/) and [this paper on heirarchical GAMs](https://peerj.com/articles/6876/) from Noam Ross, Gavin Simpson, David Miller and Eric Pedersen.
+
+I compiled camera-trap data from three distinct studies:
+
+- My PhD surveys 2017 - 2019.
+
+Funding and support for these surveys came from the Conservation Ecology Centre, the Victorian  Government Department of Environment, Land, Water, and Planning, Parks Victoria, Arthur Rylah  Institute for Environmental Research, Parks Victoria.
+
+- Glenelg Ark monitoring data 2013 - 2019.
+
+Robley, A., Moloney, P., Stringer, L.
+and Donald, S.
+(2020).
+Glenelg Ark 2005–2019: long-term predator and native mammal response to predator control.
+Arthur Rylah Institute for Environmental Research Technical Report Series No.
+318.
+Department of Environment, Land, Water and Planning, Heidelberg, Victoria.
+
+- Otway Ark monitoring data 2016 - 2018.
+
+Robley, A., Fanson, B.
+and Parks Victoria West Coast District Team.
+(2019).
+The Otway Ark: response of predators and native species 2016–2018.
+Arthur Rylah Institute for Environmental Research Technical Report Series No.
+299.
+Department of Environment, Land, Water and Planning, Heidelberg, Victoria.
+
+**Surveys were carried out the traditional lands of the Gunditjmara, Gadubanud, Kirrae Wurrung,  Gulidjan and Wadawurrung.**
+
+## Repository structure
+
+**Raw data is contained in the "raw_data/" folder**.
+
+- *"predator_records.csv"* has a row for every cat and fox camera-trap detection (no time separation between consecutive detections). 
+- *"camdata.csv"* contains camera-trap deployment information and explanatory variables.
+Note that there are plenty of camera-traps which were deployed but did not detect predators - predator_records and camdata need to be merged to account for absences.
+- *"cat_id_detections.csv"* has a row for every cat detection for surveys in which cats were individually identified (also includes unmarked and unidentifiable cats).
+This is a subset of the data used in the other models, and a subset of the data used in another PhD chapter:  https://github.com/matt-w-rees/invasive-mesopredator-release (minus the Lower Glenelg National Park cats - I didn't have the detection times for these cats on-hand).
+
+**R scripts are contained in the "r_scripts/" folder**.
+
+- script and function to fit lorelograms using adapted code from [Iannarilli et al.
+2019](https://doi.org/10.1111/2041-210X.13308) are contained in the *"/lorelograms"* folder.
+This was an exploratory analysis to choose the best time-to-indepence for consecutive detections of species on each camera-trap (I chose 30 mins).
+This involves creating a minute by minute detection history so is computationally intensive.
+- *"format_records_hour_gams.R"* reformats the raw data files into a usable dataframe for spatiotemporal GAMs.
+This accounts for solar times using [average double anchorage methods](https://doi.org/10.1111/2041-210X.13290), removes non-indepedent detections (within 30 mins) and reshapes into a dataframe where, for each camera-trap, there are 0 - 23 rows for hour of the day, a column for cat and fox filled with the number of 'independent' detections in each hour, as well as columns for deployment information and explanatory variables.
+This dataframe is saved at **"derived_data/predator_counts_hour.csv"**.
+- *"models.R"* uses mgcv to fit GAMs
+- scripts beginning with *"plot_"* contain code used to create the figures with ggplot. These scripts require the *"models"* script to be run first in the same session. 
+- *"individual_cats_reformat_model_plots.R"* reformats "cat_id_detections.csv" into a dataframe, runs heirarchical GAMs for each cat / cat category, and creates the figures (yeah this was a bit of an afterthought).
+
+**Figures for the manuscript are contained in the "figs/" folder**.
+
+## GIFs of Invasive predator spatiotemporal activity:
+
+#### Glenelg region
 
@@
 
-### Otway ranges
+#### Otway Ranges
 

### 2021-10-22 README.Rmd
fieldwork photo
rearrange
unimelb

@@
 
-Funding and support for these surveys came from the Conservation Ecology Centre, the Victorian  Government Department of Environment, Land, Water, and Planning, Parks Victoria, Arthur Rylah  Institute for Environmental Research, Parks Victoria.
+Funding and support for these surveys came from the University of Melbourne, Conservation Ecology Centre, the Victorian  Government Department of Environment, Land, Water, and Planning, Parks Victoria, Arthur Rylah  Institute for Environmental Research, Parks Victoria.
 
@@
 Department of Environment, Land, Water and Planning, Heidelberg, Victoria.
+
+```{r photo, echo = FALSE}
+knitr::include_graphics("figs/steph.jpg")
+```
 

### 2021-12-26 README.Rmd
typo
rm ind

@@
 
-Here, I've built generalised additive models ('GAMs'), using the ['mgcv' R-package](https://cran.r-project.org/web/packages/mgcv/mgcv.pdf), of invasive predator spatiotemporal activity in south-west Victoria, Australia.
+Here I've built generalised additive models ('GAMs'), using the ['mgcv' R-package](https://cran.r-project.org/web/packages/mgcv/mgcv.pdf), of invasive predator spatiotemporal activity in south-west Victoria, Australia.
 
-This paper is somewhat split between an exploratory analysis of the different ways GAMs can be used to jointly model spatial and diel activity, and aiming to provide inference on the drivers of invasive predator spatiotemporal activity in temperate, heterogeneous landscapes.
+This paper aims to provide inference on the drivers of invasive predator spatiotemporal activity in temperate, heterogeneous landscapes.
 Particularly whether a subordinate mesopredator (feral cat *Felis catus*) exhibits (signs of) behavioural avoidance of an apex predator (Red fox *Vulpes vulpes*).
 
-There are four main model types, differing in how the hourly smooth was specified:
+There are three model types, differing in how the hourly smooth was specified:
 
@@
 3. Feral cat activity potential avoidance of foxes (tensor product of hour and coordinates)
-4. Variation in diel activity of feral cats (hierarchical model of hour x individual ID)
 
@@
 
-**Surveys were carried out the traditional lands of the Gunditjmara, Gadubanud, Kirrae Wurrung,  Gulidjan and Wadawurrung.**
+**Surveys were carried out the traditional lands of the Gunditjmara and Eastern Maar.**
 
@@
 Note that there are plenty of camera-traps which were deployed but did not detect predators - predator_records and camdata need to be merged to account for absences.
-- *"cat_id_detections.csv"* has a row for every cat detection for surveys in which cats were individually identified (also includes unmarked and unidentifiable cats).
-This is a subset of the data used in the other models, and a subset of the data used in another PhD chapter:  https://github.com/matt-w-rees/invasive-mesopredator-release (minus the Lower Glenelg National Park cats - I didn't have the detection times for these cats on-hand).
 
@@
 - *"models.R"* uses mgcv to fit GAMs
-- scripts beginning with *"plot_"* contain code used to create the figures with ggplot. These scripts require the *"models"* script to be run first in the same session. 
-- *"individual_cats_reformat_model_plots.R"* reformats "cat_id_detections.csv" into a dataframe, runs heirarchical GAMs for each cat / cat category, and creates the figures (yeah this was a bit of an afterthought).
+- scripts beginning with *"plot_"* contain code used to create the figures. These scripts require the *"models"* script to be run first in the same session. 
+- *"individual_cats_reformat_model_plots.R"* reformats "cat_id_detections.csv" into a dataframe, runs heirarchical GAMs for each cat / cat category, and creates the figures (not included in the paper).
 

### 2021-12-26 r_scripts/models.R

@@ -124,6 +124,7 @@ library(patchwork)
 library(ggpubr)
 library(viridis)
 library(gratia)
+library(forcats)
 
 # Set the default theme for ggplot objects 
 theme_set(theme_bw())

### 2021-12-26 r_scripts/plot_cat_v_fox.R

@@ -7,7 +7,7 @@ df = expand.grid(hour = 0:23,
                  latitude = mean(records$latitude),
                  fox_count_adj = seq(min(records$fox_count_adj), max(records$fox_count_adj), length=150),
                  station = "T052",
-                 survey_duration = mean(records$survey_duration))
+                 survey_duration = 60)
 # predict model results into dataframe
 x = sapply(gam_cat_fox$smooth, "[[",  "label")
 df_int <- cbind(df, predict(gam_cat_fox, newdata = df, se.fit = TRUE, type = "link", exclude = c(x[4:5], "s(survey_duration)")))
@@ -44,7 +44,7 @@ plot_wet_lowhigh <- ggplot(data=df_int_minmax, aes(x=hour, y=fit, group=minmax, 
   scale_color_manual(values = c("dodgerblue", "tomato"), labels = c("Low", "High"), name = "Fox activity") +   
   scale_fill_manual(values = c("dodgerblue", "tomato"), labels = c("Low", "High"), name = "Fox activity") +    
   geom_vline(xintercept = c(6.16,18.34), colour = "black", size = 0.6, linetype="dotted") + 
-  ylab("Effect") + 
+  ylab("log(count)") + 
   xlab("Hour")
 
 
@@ -54,7 +54,7 @@ df = expand.grid(hour = 0:23,
                  habitat_type = levels(records$habitat_type),
                  station = "T052",
                  foxbaits = 0,
-                 survey_duration = mean(records$survey_duration))
+                 survey_duration = 60)
 # predict model results into dataframe
 x = sapply(gam_fox_ht$smooth, "[[",  "label")
 df_int <- cbind(df, predict(gam_fox_ht, newdata = df, se.fit = TRUE, type = "link", exclude = c(x[4:8], "s(survey_duration)")))
@@ -65,7 +65,7 @@ plot_fox <- ggplot(aes(hour, fit), data = df_int) +
   ggtitle("", subtitle = "Red Fox") +
   geom_ribbon(aes(ymin=(fit-2*se.fit), ymax=(fit+2*se.fit)), alpha=0.2) +
   geom_vline(xintercept = c(6.16,18.34), colour = "black", size = 0.6, linetype="dotted") + 
-  ylab("Effect") + 
+  ylab("log(count)") + 
   xlab("Hour")
 
 #  save

### 2021-12-26 r_scripts/plot_cat_v_fox_alternative_med.R

@@ -1,0 +1,55 @@
+# PLOT FIGURE
+
+## CAT PLOT - 3D 
+df = expand.grid(hour = 0:23,
+                 habitat_type = levels(records$habitat_type),
+                 longitude = mean(records$longitude),
+                 latitude = mean(records$latitude),
+                 fox_count_adj = seq(min(records$fox_count_adj), max(records$fox_count_adj), length=150),
+                 station = "T052",
+                 survey_duration = 60)
+# predict model results into dataframe
+x = sapply(gam_cat_fox$smooth, "[[",  "label")
+df_int <- cbind(df, predict(gam_cat_fox, newdata = df, se.fit = TRUE, type = "link", exclude = c(x[4:5], "s(survey_duration)")))
+
+## CAT PLOT - 2D (same model at min / max fox)
+# min df 
+df_int_min <- df_int %>%
+  group_by(habitat_type) %>%
+  slice_min(fox_count_adj, n = 1) %>%
+  mutate(minmax = "0min")
+# median df
+df_int_med <- df_int %>% 
+  filter(abs(fox_count_adj - median(fox_count_adj)) == min(abs(fox_count_adj - median(fox_count_adj)))) %>% 
+  mutate(minmax = "1mid")
+# max df 
+df_int_max <- df_int %>%
+  group_by(habitat_type) %>%
+  slice_max(fox_count_adj, n = 1) %>%
+  mutate(minmax = "2max")
+# join df
+df_int_minmax <- bind_rows(df_int_min, df_int_med, df_int_max)
+# rename facet
+df_int_minmax$minmax <- as.factor(df_int_minmax$minmax)
+df_int_minmax$minmax <- recode_factor(df_int_minmax$minmax,  
+                                      "0min" = "Minimum",
+                                      "1mid" = "Median",
+                                      "2max" = "Maximum")
+# plot
+plot_wet_lowhigh <- ggplot(data=df_int_minmax, aes(x=hour, y=fit, group=minmax, col = minmax)) +
+  geom_line(aes(y=fit, x=hour), lwd = 1.2) +
+  facet_wrap(~minmax ~habitat_type, nrow = 3, scales = "free") +
+  ggtitle("", subtitle = "Feral cat") +
+  geom_ribbon(aes(ymin=(fit-2*se.fit), ymax=(fit+2*se.fit), fill = minmax), alpha=0.2) +
[…]

### 2021-12-26 r_scripts/plot_predator_spatiotemporal.R

@@ -164,8 +164,8 @@ camdata_o <- filter(camdata, region == "otways")
 # plot glenelg
 plot_diff_g <- ggplot(aes(x, y, fill = diel_diff_pc), data = model_predictions_g) +
    geom_tile() +
-   facet_wrap(~species, scales = "free") +
-   scale_fill_viridis("Difference (%) in activity throughout the daily cycle", option = "viridis") +
+  facet_wrap(~factor(species, levels=c('Red fox','Feral cat')), scales = "free") +
+  scale_fill_viridis("Difference (%) in activity throughout the daily cycle", option = "viridis") +
    geom_point(data = camdata_g, fill = NA, col = "white", size = 0.7, alpha = 0.3, shape = 3) +
    theme_bw(10) + 
    ggtitle("Glenelg region") +
@@ -175,7 +175,7 @@ plot_diff_g <- ggplot(aes(x, y, fill = diel_diff_pc), data = model_predictions_g
 # plot otways
 plot_diff_o <- ggplot(aes(x, y, fill = diel_diff_pc), data = model_predictions_o) +
   geom_tile() +
-  facet_wrap(~species, scales = "free") +
+  facet_wrap(~factor(species, levels=c('Red fox','Feral cat')), scales = "free") +
   scale_fill_viridis("Difference (%) in activity throughout the daily cycle", option = "viridis") +
   geom_point(data = camdata_o, fill = NA, col = "white", size = 0.7, alpha = 0.3, shape = 3) +
   theme_bw(10) + 

### 2021-12-26 r_scripts/plot_predator_vegetation_type.R

@@ -14,11 +14,11 @@ gam_plot <- function(gam_model, data){
                     vegetation_group = levels(data$vegetation_group),
                     foxbaits = 0, 
                     region = c("glenelg", "otways"),
-                    survey_duration = mean(data$survey_duration))
+                    survey_duration = 60)
   # get smooths to exlude
   x = sapply(gam_model$smooth, "[[",  "label")
   # predict model estimates (and uncertainty) into this dataframe, excluding the impact of variables other than hour:
-  df <- cbind(df, predict.gam(gam_model, newdata = df, se.fit = TRUE, type = "link", newdata.guaranteed = TRUE, exclude = c(x[3:7], "s(survey_duration)")))
+  df <- cbind(df, predict.gam(gam_model, newdata = df, se.fit = TRUE, type = "link", newdata.guaranteed = TRUE, exclude = c(x[3:7])))
   plot_veg <- ggplot(data=df, aes(x=hour, y=fit, group=vegetation_group)) +
     # ylim(-13, -2) +
     geom_line(aes(y=fit, x=hour), lwd = 0.7) +

### 2022-02-14 manuscript/main_text.Rmd
ms

---
title: "Spatial variation in predator diel activity patterns -- feral cats avoid red foxes in time, not space"
author:
  - name: Matthew W. Rees
    email: matt.wayne.rees@gmail.com
    affiliation: UOM
    footnote: 1
  - name: Jack H. Pascoe
    email: 
    affiliation: CEC
  - name: Mark Le Pla
    email: 
    affiliation: CEC
  - name: Alan Robley
    email: 
    affiliation: ARI
  - name: Emma K. Birnbaum
    email: 
    affiliation: CEC
  - name: Brendan A. Wintle
    email: 
    affiliation: UOM
  - name: Bronwyn A. Hradsky
    email: 
    affiliation: UOM
    footnote: 
address:
  - code: UOM
    address: Quantitative & Applied Ecology Group, School of Ecosystem and Forest Science, The University of Melbourne, Parkville, VIC, Australia
  - code: CEC 
    address: Conservation Ecology Centre, Otway Lighthouse Rd, Cape Otway, VIC, Australia
  - code: ARI 
    address: Department of Environment, Land, Water and Planning, Arthur Rylah Institute for Environmental Research, Heidelberg, Australia
footnote:
  - code: 1
    text: "Corresponding Author"
abstract: 
keywords: "feral cat; diel activity patterns; generalised additive model; invasive predator;  intraguild predator interactions; mesopredator release; red fox; lethal predator control"
journal: "?"
date: "`r Sys.Date()`"
bibliography: /Users/mrees2/Dropbox/personal/matt/github/phd-thesis/index/bib/thesis.bib
header-includes:
    - \usepackage{setspace}\doublespacing
    - \usepackage{float}
    - \floatplacement{figure}{H}
    - \newcommand{\beginsupplement}{\setcounter{table}{0}  \renewcommand{\thetable}{S\arabic{table}} \setcounter{figure}{0} \renewcommand{\thefigure}{S\arabic{figure}} \setcounter{section}{0} \renewcommand{\thesection}{S\arabic{section}}}
    - \usepackage{lineno}
linenumbers: false
csl: journal-of-applied-ecology.csl
geometry: margin=1.1in
fontsize: 12pt
output:
  bookdown::pdf_book:
    base_format: rticles::elsevier_article
editor_options: 
  chunk_output_type: inline
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo      = FALSE,
  warning   = FALSE,
  message   = FALSE,
  error     = FALSE,
  cache     = FALSE,
  fig.path  = "../figs/",
  fig.align = "center",
  out.width = "100%")

library(kableExtra)
library(dplyr)
library(readr)

```

\parskip=12pt

*Article type*: Research Article

*Running title*: Spatiotemporal predator activity

*Word count (abstract, body)*: 330, 4338

*References*: ?

*Figures*: 5

*Tables*: 0

*Authorship*:

*Data accessibility*:   Data and code will be deposited on the Dryad Digital Repository after acceptance and can be viewed here: [https://github.com/matt-w-rees/spatiotemporal-gams-invasive-predators](https://github.com/matt-w-rees/spatiotemporal-gams-invasive-predators).

\newpage

\linenumbers

ABSTRACT {-}
========

Understanding the constraints that apex predators impose on subordinate species is important for anticipating the outcomes of predator management. Subordinate predators may avoid dominant predators in time or space, making it difficult to quantify changes in antipredator behaviours unless joint spatiotemporal analyses are used.

In this study, we tested whether an invasive apex predator (red fox *Vulpes vulpes*) suppresses or alters the spatiotemporal activity of an invasive mesopredator (feral cat *Felis catus*). We surveyed these predators using 3667 camera-trap deployments across two regions of south-eastern Australia; foxes were poison-baited in some landscapes within each region. The simple predator guild in these regions allowed sharp focus on the interactions between these species across experimental gradients of fox (apex predator) activity. We used generalised additive models to quantify overall predator activity across space and fluctuations in predator activity throughout the daily cycle (i.e., diel activity patterns).

When averaged across the study region, red foxes and feral cats had very similar diel activity patterns; however, there was important differentiation at a finer scale. When fox counts at a camera-trap were high, feral cats did not reduce their overall activity but shifted their diel activity patterns to less risky times of the day. In dry habitats of both regions, cats shifted from being nocturnal-crepuscular to mostly diurnal. In wet forest habitat, fox activity was consistent throughout the diel period; but when fox counts were high, cats became more nocturnal, avoiding dawn in particular. Changes in cat diel activity patterns may facilitate spatial coexistence between the two invasive predators, potentially shifting impacts onto different native prey species.

It is well-appreciated that overall predator activity varies spatially and fluctuates throughout the daily cycle. Our study demonstrates that diel activity patterns also vary across space, likely mediated by both landscape context and fear. Apex predator avoidance appears to be dynamic across landscapes of fear—a  key nuance which is overlooked when simply comparing the average activity overlap between two species or the spatial overlap of species occurrence.

\newpage

INTRODUCTION
==========================

Predators shape ecosystems through both predation and the fear of predation [@ritchie2009predator; @creel2008relationships]. Fear-induced behavioural suppression can be as detrimental to subordinate species as predation itself [@schmitz2004trophic; @preisser2005scared]. These non-consumptive effects of apex predator are expected to be strong drivers of mesopredator behaviour, particularly when resource competition is high [@ritchie2009predator]. A strengthening or relaxing of antipredator behaviours by mesopredators can have cascading effects across the entire ecosystem: altering population demographics, species interactions, ecological function and human-wildlife coexistence [@estes2011trophic; @lamb2020ecology; @brown1999ecology; @gaynor2019landscapes; @ripple2004wolves]. Hence, understanding how apex predators constrain the behaviour of subordinate species is important to accurately predict the ecosystem-wide consequences of predator management, such as reintroductions or lethal control [@gaynor2021applied].

Spatial and/or temporal niche partitioning may allow predators to coexist by reducing encounter-rates and resource overlap [@kronfeld2003partitioning]. However, mesopredators may not consistently employ avoidance behaviours because perceived predation risk is temporally and spatially variable, and antipredator behaviours typically involve a trade-off against resource acquisition, such as limiting or relegating activity to suboptimal places or times [@lima1999temporal; @lima1990behavioral]. Therefore, optimal predator avoidance strategies are likely to vary across heterogeneous landscapes where resource availability (e.g., shelter, food) and perceived predation risks differ [@kauffman2007landscape; @willems2009predator; @wirsing2021context]. For example, temporal predator avoidance may be preferable over spatial avoidance if food is constantly available throughout the day, and vice versa. These concepts are unified under the ‘ecology of fear’ [@brown1999ecology] concept, which has gained increasing attention in recent times [@gaynor2019landscapes]. Notably, the mesopredator release hypothesis has recently been expanded from increases in mesopredator abundance following apex predator decline [@soule1988reconstructed] to also include changes in mesopredator behaviour [@brashares2010ecological].

To accurately quantify avoidance within a predator guild, we first need to understand how the overall activity and diel activity patterns of each species varies ‘naturally’ across landscapes, particularly for species with broad distributions. It is widely recognised that the overall activity of different predator species varies across their distributions, but their diel activity patterns are often assumed to be constant. In this paper, we use the term ‘overall activity’ to refer to the number of ‘independent’ predator detections at a site (offset to account for survey duration; analogous to an activity or abundance index), and ‘diel activity pattern’ to refer to fluctuations in relative activity throughout the 24-hour daily cycle. Overall activity is influenced by predator behaviour, population density and the detection process [@anderson2001need], whereas diel activity patterns are a behavioural trait (less likely to be affected by the detection process, if the survey methodology remains consistent).

Despite modern predator survey technologies providing time-stamped detections, detection times are commonly discarded from analyses, probably because joint modelling of overall activity and diel activity patterns is more complicated. When temporal avoidance is tested, it is usually considered in an ad-hoc fashion, by fitting separate models for spatial and temporal avoidance, or by repeating spatial analyses (e.g., resource selection functions) at different time periods [e.g., @smith2019integrating; @basille2015plastic; @kohl2019prey]. However, discretising the daily cycle into categorical periods (e.g., day and night) introduces bias, assuming animals have complete step-changes in behaviour rather than progressive shifts across the daily cycle.  Further, dawn and dusk are particularly important times for many predator species.

Generalised Additive Models (hereafter 'GAMs') are increasingly being used to estimate animal diel activity patterns and offer a flexible framework to jointly consider overall activity. GAMs also have other benefits, including smoothing penalties to reduce overfitting, the ability to capture nonlinear interactions between multiple variables with different units, and the ability to share information across categorical variables through hierarchical specifications [@wood2017generalized; @pedersen2019hierarchical]. However, we are only aware of one study which allowed animal diel activity to interact with predation risk as a continuous variable in a GAM [although without considering overall activity; @cunningham2019temporal].

The red fox *Vulpes vulpes* (hereafter 'fox'; ~6 kg) and feral cat *Felis catus* (hereafter 'cat'; ~4 kg) have devastating impacts on native prey throughout their introduced range, implicated in the extinction of ~10 and 63 species, respectively [@doherty2016invasive]. The impacts of these invasive predators have been particularly extreme on the Australian continent [@woinarski2015ongoing]. Cats are more difficult to manage, and so introduced predator control programs in Australia often target only foxes [particularly through poison-baiting; @reddiex2007control]. As foxes and cats compete for many of the same resources, there is concern that lethal fox control could cause a mesopredator release [@soule1988reconstructed] of feral cats [@glen2005complex; @doherty2017stop; @wayne2017recoveries; @marlow2015cats; @comer2020integrating; @robley2014long]. There is some evidence that feral cats increase in activity [although highly uncertain; @hunter2018not], density (Chapter \@ref(density)) and alter their diets and use of space [Chapter \@ref(density); @molsher2017mesopredator] in response to fox control. Other studies have investigated potential spatial and temporal interactions between these invasive predators [e.g., @roshier2021space], but not in response to fox control, or in a joint spatiotemporal framework that allows flexibility in cat avoidance behaviours in respect to differences in fox activity.

In this study, we explored how the overall activity and diel activity patterns of two competing invasive predators varied across heterogeneous landscapes, in response to (1) space and (2) vegetation types. We then investigated (3) whether cat diel activity patterns change in response to the overall level of fox activity. Our study was conducted in a simple predator system where foxes and cats are the only mammalian carnivores, and fox activity is manipulated using lethal control in some landscapes. This allowed sole focus on the interactions between these two predators, across an experimental gradient of apex predator (fox) activity. We illustrate how GAMs can provide a simple framework to jointly assess spatial and temporal animal activity patterns, as well as avoidance behaviours.

\newpage

MATERIALS AND METHODS
==========================

## Study area and camera-trapping 
We compiled data from multiple smaller-scale camera-trap studies across two regions in south-west Victoria, Australia: the Glenelg region and Otway Ranges (Fig. \@ref(fig:diel-map)). Introduced foxes and cats are the only medium-large functional mammalian terrestrial carnivores here: native dingoes *Canis familiaris* are long-absent throughout, while tiger quolls *Dasyurus maculatus* are long-absent in the Glenelg region and likely functionally extinct in the Otway Ranges (last confirmed sighting in 2014). In broad sections of each region, government land managers conduct ongoing targeted lethal fox control for biodiversity conservation. Poison-baits containing 3 mg of sodium mono-fluroacetate ('1080') are buried at a depth of 12 - 15 cm at 1-km intervals along accessible forest tracks and roads. Different road densities result in variable densities of poison-baits. Managers also frequently implement prescribed fire across both regions, primarily to reduce fuel loads to prevent large wildfires.

### Glenelg region
In the Glenelg region, large patches of natural vegetation are fragmented by pastoral farming and residential properties (Fig. \@ref(fig:diel-map)). Foxes in three distinct forest blocks in this region have been subject to poison-baiting since October 2005, with fortnightly bait replacements [@robley2014long]. These forest blocks, along with three similar, unbaited forest blocks to the north have been simultaneously surveyed annually under the 'Glenelg Ark' fox control program since 2005 [40 sites per block; @robley2020glenelg]. Hair-tubes were used to monitor species from 2005 - 2013 [presented in @robley2014long], replaced by camera-traps from 2013; here we present camera-trap data from 2013 - 2019 [@robley2020glenelg]. We also included a further 425 camera-trap deployments at unique locations from early 2018 (M.W.R., PhD surveys). This totals 2039 camera-trap deployments in the Glenelg region, collected in a control-impact experimental design (foxes had been continuously controlled for at 8 - 14 years in the treatment landscapes at the time of these surveys).

### Otway Ranges
The Otway Ranges is a largely continuous patch of natural vegetation with a strong east-west rainfall gradient (Fig. \@ref(fig:diel-map)). A matrix of cool temperate rainforest and wet forest at high altitudes in the south-west descend into a large heathland directly north, and into dry forests and then heathlands to the north-east. Fox-baiting commenced in small sections of the Otway Ranges in 2008 and large-scale systematic baiting began in 2016 - 2017 under the 'Otway Ark' program [@robley2019otway]. For the first six weeks, poison-baits were replaced weekly, then changing to ongoing monthly bait-replacement. There was a pause in baiting for approximately six months during the second half of 2018. Fox control recommenced in late 2018 with four weeks of fortnightly bait-replacement, before returning to monthly bait-replacement. A large section of the Otway Ranges to the north-west remains unbaited, but is monitored as an experimental non-treatment site [@robley2019otway]. Otway Ark managers survey 372 camera-trap sites annually (sequentially across the region); we present one 'before' baiting survey and two 'after' baiting surveys of each site from 2016 - 2018, totalling 1113 camera-trap deployments [@robley2019otway]. We also include data from an additional before-after control-impact surveys (one 'before' baiting survey and two 'after' bating surveys) in the western section of the Otway Ranges, conducted annually 2017 - 2019 (M.W.R PhD surveys). This added a further 195 sites and 524 camera-trap deployments.

#### Camera-trap set-ups
All camera-trap deployments consisted of a Reconyx (Holmen, Wisconsin) brand camera-trap (white or infrared flash), attached to a tree or a metal picket, facing a lure. The Glenelg Ark and Otway Ark fox monitoring programs positioned camera-traps at least 40 cm above ground on a tree or a metal picket and angled downwards toward a lure approximately 1 - 1.5 m away [@robley2020glenelg; @robley2019otway]. The lures consisted of peanut butter, golden syrup and rolled oats mixed into a small ball, placed within a tea strainer or PVC pipe container and secured either to the ground, or 20 - 60 cm above ground on a wooden stake. The M.W.R. PhD surveys across both regions positioned camera-traps lower on a tree (around 15 - 30 cm above the ground) angled only slightly downwards toward a tuna oil lure approximately 2 - 2.5 m away [detailed in @rees2019unexpectedly]. Camera-traps were active for an average of 47 days (maximum 93 days), totalling 172,052 trap-nights.

## Data preparation
All data analysis was conducted in R version 3.6.3 [@R]. We first used lorelograms to identify the minimum interval to approximate independence [@iannarilli2019lorelograms]; this indicated that discarding repeat detections of a species within 30 minutes was sufficient to reduce temporal autocorrelation. To account for day length variation across space and time, we extracted sunrise and sunset times for each camera-trap deployment using the 'maptools' R-package [@maptools] and adjusted detection times to be relative to sunrise and sunset using the average double anchoring approach described by @vazquez2019comparing. We then built a dataframe consisting of a row for each hour of the day (0 – 23), for every camera-trap deployment (n = 3667), recording the total number of 'independent' fox and feral cat detections within each hour across the camera-trap survey.

## Generalised additive models
We modelled the total number of independent detections of each predator per hour for each camera-trap deployment (response variable) with generalised additive mixed-effect models implemented in the 'mgcv' R-package [@wood2017generalized]. We used the negative binomial family, as overdispersion, but not zero-inflation, was detected with a poisson distribution using the 'DHARMa' R-package [@DHARMa]. We specified the natural log of the number of survey days as a model offset to account for differences in camera-trap survey duration, and a random intercept for each site to account for repeat sampling. For fox models, we also included a smooth effect of poison-bait density with separate responses per region to account for the effect of fox control (all figures in this manuscript are derived from fox models predicted to a no fox-baiting scenario). This formed the base model specification for each model we fitted; models differed in their specification of the cyclical hour smooth to provide inference on variations of predator diel activity across the four questions of interest; this is detailed in the sections below.

### How does predator overall activity and diel activity patterns vary across space? (model 1)
To examine how the overall activity and diel activity patterns of each predator varied across space, we fit a model for each predator which included a tensor product interaction between a spatial smooth and hourly smooth. This allowed predators to have different activity levels across space (static across the years surveyed), as well as variation in diel activity pattern across space. Space was modelled using camera-trap coordinates and a duchon spline basis [@miller2014finite]. To examine how the relative strength of diel activity patterns changed across space, we plotted the percentage increase from the minimum to maximum activity estimate within the daily cycle for each predicted location (hereafter referred to as ‘diel activity pattern strength’).

#### How does predator overall activity and diel activity patterns vary across vegetation types? (model 2)
Predator activity varied across space; we hypothesised that this was partly due to differences in vegetation type, based both on the observed spatial patterns and because vegetation type is a major driver of understorey habitat structure and prey occurrence in these regions [@swan2015predicting; @hradsky2017bayesian]. To test whether the diel activity pattern of each predator varied among vegetation types, we identified the Ecological Vegetation Class group [hereafter 'vegetation type'; standard units for vegetation classification in Victoria; @delwp2020bioregions] for each unique camera-trap site, totalling eight vegetation types. As rainforests are interspersed (primarily in low lying gullies) at fine-scales throughout wet and damp forests in the south-eastern Otway Ranges, we merged them together (hereafter referred to as 'wet forests'). We then estimated predator activity across vegetation types using a hierarchical model specification: a global smoother for hour (i.e., average response) and group-level smoothers with shared wiggliness for the seven vegetation types ['model GS' detailed in @pedersen2019hierarchical]. We also included a random effect to account for differences in overall activity levels between the two regions.

### Do feral cats avoid foxes in space or time? (model 3)
Fox diel activity across vegetation types showed strong similarity between all vegetation types except wet forests. To examine whether cats avoid foxes in space or time, we therefore modelled fox-induced changes in feral cat diel activity separately for wet forest and dry vegetation types. We further split dry vegetation types by region for replication. We refer to the resulting variable as 'habitat type', which had three levels: (i) wet forests and rainforests in the western Otway Ranges ('wet_otways'), (ii) dry vegetation types in the Otway Ranges ('dry_otways') and (iii) dry vegetation types in the Glenelg region ('dry_glenelg'). We hypothesised that cats would avoid foxes in time by becoming more diurnal in dry vegetation types where foxes were mostly nocturnal, but not in wet forests where fox activity showed little variation across the daily cycle.

To investigate changes in feral cat diel activity across the range of observed fox activity, we first quantified fox activity for each camera trap deployment as the total number of fox detections for the deployment divided by the number of survey days, to adjust for survey duration (hereafter 'adjusted fox counts'). We modelled an interaction between hour (cyclical spine) and adjusted fox counts (thin plate regression spline with shrinkage - meaning fox effects could be entirely removed from the model if not supported by sufficient data), allowing cats to have nonlinear responses to both hour and adjusted fox counts. We fit separate tensor product interactions for each habitat type (using a 'by-variable' term). For a direct visual comparison to fox activity, we fit another fox model where a diel curve was estimated separately across each of the three habitat types.

\newpage

```{r diel-map, fig.cap = "Locations of our study regions in south-west Victoria, Australia. The grids of camera-traps are denoted by white dots. The Glenelg region is to the west and Otway region to the east. Native vegetation is indicated by dark green, with hill shading. \\textit{Map tiles by Stamen Design, under CC BY 3.0, map data by OpenStreetMap, under CC BY SA.}"}
knitr::include_graphics("../figs/map_cams.png")
```

\newpage

RESULTS
============

Overall, we collated 5449 and 2202 independent detections of foxes and cats, respectively (separated by at least 30 minutes) from 172,052 camera-trap nights (Table \@ref(tab:diel-tab1)).

## How does predator overall activity and diel activity patterns vary across space? (model 1)
Predator activity varied considerably across space and throughout the 24-hour daily cycle, and there was some variation in the predator diel activity patterns across space. On average, both predators showed similar diel activity patterns: mostly nocturnal with peaks in activity around sunrise and sunset (i.e., crepuscular; Fig. \@ref(fig:diel-veg)i). The main difference between the species was that fox activity peaked just after sunset and they were less likely to be active during the day than cats. Cats also tended to be more active at sunset relative to sunrise.

Diel activity pattern strength also differed between the species. Fox activity was concentrated strongly at particular times of the day, especially in the Glenelg region where activity varied by up to 371% throughout the daily cycle (Fig. \@ref(fig:diel-space)a). Feral cats had relatively more consistent activity throughout the daily cycle and across regions; the maximum difference in cat activity throughout the daily cycle for any given location was 185%.

Variation in diel activity patterns across space, as well as differences in overall activity between the predators, was strongest in the Otway Ranges. For example, overall fox activity (Fig. \@ref(fig:diel-space-o-fox)) and diel activity pattern strength(Fig. \@ref(fig:diel-space)b) were lowest in the south-west Otway Ranges, while feral cat overall activity (Fig. \@ref(fig:diel-space-o-cat)) and diel activity pattern strength (Fig. \@ref(fig:diel-space)b) were highest in that subregion.

## How does predator overall activity and diel activity patterns vary across vegetation types? (model 2)

Overall levels of fox activity were similar across all vegetation types, except wet forests where fox activity was considerably lower. Overall cat activity was more variable across vegetation types; lowest in heathy woodlands and highest in wet forests (Fig. \@ref(fig:diel-veg)b).

Diel activity patterns for foxes were similar across all vegetation type except wet forests; in wet forests, foxes were consistently active throughout the daily cycle (Fig. \@ref(fig:diel-veg)a). On the other hand, cats were nocturnal (and most active) in wet forests, but largely crepuscular in all other vegetation types (Fig. \@ref(fig:diel-veg)b). For both predators, the random effect for region (Glenelg or Otways) in the vegetation models shrank to near-zero=, indicating all variation between the regions was explained by the vegetation covariate and site random intercept.

## Do feral cats avoid foxes in space or time? (model 3)
Cat spatial activity was relatively unaffected by the fox activity in both habitat types of the Otway Ranges; and, if anything, increased with increasing adjusted fox counts in the Glenelg region (Fig. \@ref(fig:diel-cat-fox)), indicating cats did not avoid foxes spatially.

Across all habitat types, feral cat diel activity patterns changed across the gradient of fox activity (Fig. \@ref(fig:diel-cat-fox)). In the Glenelg region and Otway dry habitat types, feral cats had a nocturnal-crepuscular diel activity pattern where fox activity was low, but were most active during the day where fox activity was high. In contrast, in the wet forests of the Otway Ranges, feral cats were more strongly nocturnal when fox activity was high.

\newpage

```{r diel-st-int-o-fox, fig.cap = "Interaction effect of space-time on red fox \\textit{Vulpes vulpes} activity across each hour of the day (0 - 23) in the Otway Ranges, Australia (model 1), as an example. Corresponding plots for feral cats \\textit{Felis catus} in this region, as well as both predators in the Glenelg region are provided in the Supporting Information, as are the marginal effects of space and time. White crosses depict unique camera-trap sites."}
knitr::include_graphics("../figs/spte_diff_avg_o_fox.png")

```

```{r diel-space, fig.cap = "The strength of diel activity patterns of two invasive predators varied within the two study regions in south-west Victoria, Australia (model 1). White crosses depict unique camera-trap sites; colour brightness scales with increasing percentage difference between the minimum and maximum activity estimate over the 24-hour cycle for each location. Red foxes \\textit{Vulpes vulpes} (a, c) concentrated their activity during particular times of the day, especially in the Glenelg region (a) and the drier parts of the Otways (c), whereas feral cat \\textit{Felis catus} activity was relatively consistent activity throughout the daily cycle (b, d)."}
knitr::include_graphics("../figs/diel_strength_600dpi.png")

```

\newpage

```{r diel-veg, fig.cap = "Red foxes \\textit{Vulpes vulpes} (a) and feral cat \\textit{Felis catus} (B) diel activity patterns overall (i) and across different Ecological Vegetation Class (EVC) groups (ii, iii) in south-west Victoria, Australia (model 2). Dotted, vertical lines represent average sunrise and sunset times. Shaded areas indicate 95\\% confidence intervals. Both invasive predators had a crepuscular to nocturnal diel activity pattern on average, with slight deviations across the drier EVC groups and large deviations in wet forests (ii; wet forests shown as pink line). The overall level of activity was relatively consistent across EVC groups for foxes (a – iii), whereas it differed substantially for feral cats (b - iii)."}
knitr::include_graphics("../figs/predator_veg.png")
```

\newpage

```{r diel-cat-fox, fig.cap="Variation in mean feral cat \\textit{Felis catus} activity (a) and associated uncertainty estimates (b) in response to count of ’independent’ red fox \\textit{Vulpes vulpes} detections (log-transformed and survey effort adjusted) across each ’habitat type’ in south-west Victoria, Australia (model 3). Grey vertical lines represent average sunrise and sunset times. In the Glenelg region, there were more feral cat detections where there were more fox detections, but cat peak diel activity shifted from crepuscular night to pre-dawn and midday (a). In the Otway Ranges, feral cat activity also peaked during the day where fox activity was high in dry vegetation types (b), but was more nocturnal where fox activity was high in the rainforests and wet forests (c)."}
knitr::include_graphics("../figs/cat_fox_count.png")
```

\newpage

DISCUSSION
===================

A key question in ecological theory is whether animals are evolutionary hardwired to occupy particular temporal niches, or have circadian rhythms that are responsive to changing environmental conditions and interactions with other species [@daan1981adaptive; @schoener1974compression; @lima1990behavioral]. Here we demonstrate that diel activity patterns are not fixed, but vary across space based on landscape context and fear. In our study, sympatric invasive predators had similar diel activity patterns when averaged across broad regions [i.e., high circular overlap; Fig. \@ref(fig:diel-space-time-marginal); as did @roshier2021space], but behaviours varied considerably within landscapes. Fox daily activity patterns were most strongly tied to the daily cycle in dry habitat types, but showed little diel activity pattern in the wet forests (Fig. \@ref(fig:diel-veg)a). In contrast, cats were mostly nocturnal in wet forests but crepuscular in dry vegetation types (Fig. \@ref(fig:diel-veg)b). Within broad habitat types, cats altered their diel activity patterns at sites with higher fox activity (Fig. \@ref(fig:diel-cat-fox)). Control programs that reducing invasive fox activity are therefore likely to change cat diel activity patterns, which in turn may alter cat impacts on native prey species. Quantifying changes in diel activity patterns provides important context for understanding species interactions, which is key for effective ecosystem management [@gaynor2021applied].

Shifting diel activity patterns may facilitate spatial coexistence of dominant and subordinate species [@carothers1984time]. For cats, altering diel activity patterns to less-preferred times of day may be worthwhile to persist in high-quality habitat. Few studies have demonstrated predator-induced shifts in diel activity [@kronfeld2003partitioning], but notably, ship rats *Rattus norvegicus* were also found to switch from nocturnal to diurnal behaviour in response to fox activity [@fenn1995use]. For cats in our study, a switch to diurnal behaviour where fox activity was high in the dry vegetation types may have been facilitated by the higher abundance of reptiles in these habitat types relative to wet forests [which are mostly diurnal; @woinarski2018reptiles]. At wet forest sites with high fox activity, cats concentrated their activity away from sunrise and sunset towards midnight, despite a diurnal shift appearing to similarly reduce the risk of a fox encounter. In this situation, we expect becoming more nocturnal to be favourable over a diurnal shift because this is when small mammals are active and cats would be least visible to foxes [cats here mostly had black or grey coats]. Overall fox activity was twice as low in the wet forests relative to dry habitat types, and so cats were likely under less pressure to radically alter their diel activity patterns. Understanding how these potential avoidance behaviours impacts native prey is a key research priority to improve invasive predator management.

Cats may have avoided foxes in time, but we saw no sign of spatial avoidance (i.e. no fine-scale negative association between fox and cat overall activity (Fig. \@ref(fig:diel-cat-fox)b:c). Overlap in spatial activity at fine scales (i.e., at the camera-trap sites) may have been facilitated by temporal avoidance [@kronfeld2003partitioning]. However, we considered spatial avoidance across the entire survey duration (averaging 47 days). Cats may indeed avoid foxes spatially, but transiently on the scale of hours to days—after all, how is a cat to know where to avoid a fox without encountering signs of one? Short-term spatial avoidance is quite plausible given foxes mark territories using scats and odours, which cats could tangibly associate with high risk shortly after. Temporary spatial avoidance could be tested using decay curves [e.g., @niedballa2019assessing]. Alternatively, no sign of spatial avoidance between these invasive predators could also be an artefact of the quality of camera set-up and hence detectability. The 3667 camera-traps were deployed by numerous people, and the quality of set-ups differed considerably in terms of detecting predators. Camera-traps that angled only slightly downwards, rather than upwards or strongly downwards, seemed most effective at detecting both predator species (M.W Rees, personal observation). Predator interactions are routinely inferred through spatial associations between species; however such analyses are subject to numerous pitfalls which can make inference unreliable [reviewed in @guillaume2020co].

A distinction of our study from others is that we modelled potential avoidance behaviours in a simple predator guild, where apex predator activity was artificially manipulated. This reduces potential bias from differences in niche preferences and the unmodelled impacts of other predators in the system. We also included replication across different habitat types. However, because our study did not consider associations with prey species, we cannot distinguish whether changes in cat diel activity patterns were the result of direct fox avoidance or indirect associations with shared prey. For example, low fox activity may promote the availability of a preferred shared prey species with a diel pattern which differs from those of cats on average, and as a result, cats might shift diel activity patterns at sites with low fox activity to more closely match those of the more abundant prey species. We would expect introduced European rabbits *Oryctolagus cuniculus* and hares *Lepus europaeus* (which are diurnal) to be particularly likely to induce such a response in dry vegetation types [@stobo2020management; @mcgregor2020short], however, rabbits and hares are rare within the natural vegetation of the landscapes we surveyed (only ever being detected at 40 of 1232 sites) and there is little evidence that predation by foxes suppresses rabbit populations [@norbury2015pests; @scroggie2018invasive]. While of particular interest, whether temporal fox-cat interactions are direct or indirect does not change the resulting impact on native prey, and hence the outcomes of fox management.

Flexible antipredator behaviours make evolutionary sense, but have been rarely demonstrated in terms of spatiotemporal predator avoidance [although, see @relyea2003predators; @brown2013phenotypically; @cunningham2019temporal], because this often requires manipulative experiments or at least more complicated models [@kronfeld2003partitioning]. Our study demonstrates that GAMs offer a powerful tool for modelling continuous shifts in animal activity across both space and time, capable of capturing complex interactions and sharing information across categorical variables. The inbuilt smoothing penalties are another benefit of GAMs over kernel density estimation [@ridout2009estimating], in which noisy data can produce spurious estimates [@iannarilli2019lorelograms; @frey2017investigating].  The alternative approach of simply comparing average diel activity overlap between two species [@ridout2009estimating] would have been misleading for two reasons. Firstly, predator diel activity patterns varied ‘naturally’ across heterogeneous landscapes (requiring avoidance to be tested in wet forests and dry vegetation types separately; Fig. \@ref(fig:diel-veg)). Secondly, apex predator temporal avoidance strategies were not consistently employed, but depended on overall apex predator activity and vegetation type (Fig. \@ref(fig:diel-cat-fox)). Despite their underlying statistical complexity, GAMs in the 'mgcv' R-package are straightforward to fit. Our GAM framework for modelling spatiotemporal activity can be used on any species with time-stamped detections, including datasets with categorical or continuous covariates and hierarchical groupings.

Animal diel activity patterns can be complex, varying across space, habitat types and threat-levels. Despite telling an important story about how animals interact with each other and the environment, detection times are commonly discarded from statistical analyses of camera-trap data. In the rare instances that they are considered, diel activity patterns are predominantly estimated at the population-level, overlooking finer-scale behaviours that can affect fitness, survival and ecosystem-impacts. Our results demonstrate the importance of (a) considering diel activity in regards to species interactions, (b) modelling *changes* in animal behaviour rather than overlap with other species, and (c) testing avoidance behaviours within a joint spatiotemporal framework. Our study adds to the limited body of evidence that top predators can produce a landscape of fear which is powerful enough to reverse the diel activity patterns of subordinate species [@kronfeld2003partitioning].

\newpage

ACKNOWLEDGEMENTS {-}
===================

\newpage

REFERENCES {#references .unnumbered}
==========

<div id="refs"></div>

\newpage

\beginsupplement

# Supporting Information {-}

\newpage

$~$

$~$

$~$

```{r diel-tab1}

# load records
records <- read_csv("/Users/mrees2/Dropbox/personal/matt/github/phd-thesis/index/data/diel/counts_hour_cleaned.csv")
# drop hour
records <- distinct(records, station_year, .keep_all = TRUE)
# get unique sites dataframe
camdata <- distinct(records, station, .keep_all = TRUE)

table1 <- records %>%
 group_by(region, vegetation_group) %>%
 mutate(sites = length(unique(station)),
     deployments = length(unique(station_year)),
     fox_count_vr = sum(fox_count),
     cat_count_vr = sum(cat_count),
     fox_present = sum(fox_pa == "present"),
     cat_present = sum(cat_pa == "present")
     )

table1 <- table1 %>%
 dplyr::select(vegetation_group, region, sites, deployments, fox_count_vr, cat_count_vr) %>% 
 distinct() %>%
 arrange(vegetation_group, region)

# get sums for total row
total <- colSums(Filter(is.numeric, table1))
# add to df
table1 <- rbind(table1, total)
table1[14,1] <- "Total"
table1[14,2] <- ""

# rename region 
table1$region <- factor(table1$region)
table1$region <- gsub("glenelg", "Glenelg", table1$region)
table1$region <- gsub("otway", "Otway", table1$region)

# remove repeat vegetation name
table1[c(2, 4, 6, 8, 10, 12),1] <- ""

# make table
kbl(table1, longtable = T, booktabs = T, digits = 2,
  caption = "Summary of the number of camera-trap deployments, unique survey sites and 'independent' counts of invasive predator detections across Ecological Vegetation Class groups within the Glenelg and Otway regions, south-west Victoria, Australia.",
   col.names = c("Vegetation", "Region", "Sites", "Deployments", "Fox counts", "Cat counts")) %>% 
  kable_styling(font_size = 10) %>% 
  row_spec(14, bold = T)

```

\newpage

```{r diel-tab-fits}

# load records
fits_summaries <- read_csv("../derived_data/model_summaries.csv") %>%
  dplyr::select(-c(X1))

# make table
kbl(fits_summaries, longtable = T, booktabs = T, digits = 4,
  caption = "Generalised additive model summaries for invasive predator spatiotemporal activity in south-west Victoria, Australia.",
     col.names = c("Species", "Model", "EDF", "dev.expl", "r.sq")) %>% 
  kable_styling(font_size = 10) %>% 
  footnote(general = c("EDF - estimated degrees of freedom of all model terms.",
                       "dev.expl - proportion of the null deviance explained by the model. ",
                       "r.sq -  adjusted r-squared value."))

```

\newpage

```{r diel-space-time-marginal, fig.cap = "Marginal effect of time (model 1) on predators across both study regions in south-west Victoria, Australia. White crosses depict unique camera-trap sites"}
knitr::include_graphics("../figs/avg_diel_predator.png")

```

\newpage

```{r diel-space-time-marginal-g}
knitr::include_graphics("../figs/sp_marginal_g.png")

```

```{r diel-space-time-marginal-o, fig.cap = "Marginal effect of space on predators across south-west Victoria, Australia (model 1). White crosses depict unique camera-trap sites"}
knitr::include_graphics("../figs/sp_marginal_o.png")

```

\newpage

```{r diel-st-int-g-fox, fig.cap = "Interaction effect of space-time on feral cat \\textit{Felis catus} activity across each hour of the day (0 - 23) in the Glenelg region, Australia (model 1). White crosses depict unique camera-trap sites. "}
knitr::include_graphics("../figs/spte_diff_avg_g_fox.png")

```

\newpage

```{r diel-st-int-g-cat, fig.cap = "Interaction effect of space-time on feral cat \\textit{Felis catus} activity across each hour of the day (0 - 23) in the Glenelg region, Australia (model 1). White crosses depict unique camera-trap sites. "}
knitr::include_graphics("../figs/spte_diff_avg_g_cat.png")

```

\newpage

```{r diel-st-int-o-cat, fig.cap = "Interaction effect of space-time on feral cat \\textit{Felis catus} activity across each hour of the day (0 - 23) in the Otway Ranges, Australia (model 1). White crosses depict unique camera-trap sites. "}
knitr::include_graphics("../figs/spte_diff_avg_o_cat.png")

```

\newpage

```{r diel-space-g-fox, fig.cap = "Overall spatial activity of red foxes \\textit{Vulpes vulpes} for each hour of the day (0 - 23) in the Glenelg region, Australia (model 1). White crosses depict unique camera-trap sites"}
knitr::include_graphics("../figs/spte_facet_o_fox.png")

```

\newpage

```{r diel-space-g-cat, fig.cap = "Overall spatial activity of feral cats \\textit{Felis catus} for each hour of the day (0 - 23) in the Glenelg region, Australia (model 1). White crosses depict unique camera-trap sites"}
knitr::include_graphics("../figs/spte_facet_g_cat.png")

```

\newpage

```{r diel-space-o-fox, fig.cap = "Overall spatial activity of red foxes \\textit{Vulpes vulpes} for each hour of the day (0 - 23) in the Otway Ranges, Australia (model 1). White crosses depict unique camera-trap sites"}
knitr::include_graphics("../figs/spte_facet_o_fox.png")

```

\newpage

```{r diel-space-o-cat, fig.cap = "Overall spatial activity of feral cats \\textit{Felis catus} for each hour of the day (0 - 23) in the Otway Ranges, Australia (model 1). White crosses depict unique camera-trap sites"}
knitr::include_graphics("../figs/spte_facet_o_cat.png")

```

### 2022-02-14 manuscript/main_text.tex
ms

\documentclass[]{elsarticle} %review=doublespace preprint=single 5p=2 column
%%% Begin My package additions %%%%%%%%%%%%%%%%%%%
\usepackage[hyphens]{url}

  \journal{?} % Sets Journal name

\usepackage{lineno} % add
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

\usepackage{graphicx}
\usepackage{booktabs} % book-quality tables
%%%%%%%%%%%%%%%% end my additions to header

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{fontspec}
  \ifxetex
    \usepackage{xltxtra,xunicode}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\usepackage[margin=1.1in]{geometry}
\bibliographystyle{elsarticle-harv}
\usepackage{longtable}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={Spatial variation in predator diel activity patterns -- feral cats avoid red foxes in time, not space},
            colorlinks=false,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls

\setcounter{secnumdepth}{5}
% Pandoc toggle for numbering sections (defaults to be off)

% Pandoc citation processing

% Pandoc header
\usepackage{setspace}\doublespacing
\usepackage{float}
\floatplacement{figure}{H}
\newcommand{\beginsupplement}{\setcounter{table}{0}  \renewcommand{\thetable}{S\arabic{table}} \setcounter{figure}{0} \renewcommand{\thefigure}{S\arabic{figure}} \setcounter{section}{0} \renewcommand{\thesection}{S\arabic{section}}}
\usepackage{lineno}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}

\begin{document}
\begin{frontmatter}

  \title{Spatial variation in predator diel activity patterns -- feral cats avoid red foxes in time, not space}
    \author[UOM]{Matthew W. Rees\corref{1}}
   \ead{matt.wayne.rees@gmail.com} 
    \author[CEC]{Jack H. Pascoe}

    \author[CEC]{Mark Le Pla}

    \author[ARI]{Alan Robley}

    \author[CEC]{Emma K. Birnbaum}

    \author[UOM]{Brendan A. Wintle}

    \author[UOM]{Bronwyn A. Hradsky}

      \address[UOM]{Quantitative \& Applied Ecology Group, School of Ecosystem and Forest Science, The University of Melbourne, Parkville, VIC, Australia}
    \address[CEC]{Conservation Ecology Centre, Otway Lighthouse Rd, Cape Otway, VIC, Australia}
    \address[ARI]{Department of Environment, Land, Water and Planning, Arthur Rylah Institute for Environmental Research, Heidelberg, Australia}
      \cortext[1]{Corresponding Author}

  \begin{abstract}

  \end{abstract}
   \begin{keyword} feral cat; diel activity patterns; generalised additive model; invasive predator; intraguild predator interactions; mesopredator release; red fox; lethal predator control\end{keyword}
 \end{frontmatter}

\parskip=12pt

\emph{Article type}: Research Article

\emph{Running title}: Spatiotemporal predator activity

\emph{Word count (abstract, body)}: 330, 4338

\emph{References}: ?

\emph{Figures}: 5

\emph{Tables}: 0

\emph{Authorship}:

\emph{Data accessibility}: Data and code will be deposited on the Dryad Digital Repository after acceptance and can be viewed here: \url{https://github.com/matt-w-rees/spatiotemporal-gams-invasive-predators}.

\newpage

\hypertarget{abstract}{%
\section*{ABSTRACT}\label{abstract}}
\addcontentsline{toc}{section}{ABSTRACT}

Understanding the constraints that apex predators impose on subordinate species is important for anticipating the outcomes of predator management. Subordinate predators may avoid dominant predators in time or space, making it difficult to quantify changes in antipredator behaviours unless joint spatiotemporal analyses are used.

In this study, we tested whether an invasive apex predator (red fox \emph{Vulpes vulpes}) suppresses or alters the spatiotemporal activity of an invasive mesopredator (feral cat \emph{Felis catus}). We surveyed these predators using 3667 camera-trap deployments across two regions of south-eastern Australia; foxes were poison-baited in some landscapes within each region. The simple predator guild in these regions allowed sharp focus on the interactions between these species across experimental gradients of fox (apex predator) activity. We used generalised additive models to quantify overall predator activity across space and fluctuations in predator activity throughout the daily cycle (i.e., diel activity patterns).

When averaged across the study region, red foxes and feral cats had very similar diel activity patterns; however, there was important differentiation at a finer scale. When fox counts at a camera-trap were high, feral cats did not reduce their overall activity but shifted their diel activity patterns to less risky times of the day. In dry habitats of both regions, cats shifted from being nocturnal-crepuscular to mostly diurnal. In wet forest habitat, fox activity was consistent throughout the diel period; but when fox counts were high, cats became more nocturnal, avoiding dawn in particular. Changes in cat diel activity patterns may facilitate spatial coexistence between the two invasive predators, potentially shifting impacts onto different native prey species.

It is well-appreciated that overall predator activity varies spatially and fluctuates throughout the daily cycle. Our study demonstrates that diel activity patterns also vary across space, likely mediated by both landscape context and fear. Apex predator avoidance appears to be dynamic across landscapes of fear---a key nuance which is overlooked when simply comparing the average activity overlap between two species or the spatial overlap of species occurrence.

\newpage

\hypertarget{introduction}{%
\section{INTRODUCTION}\label{introduction}}

Predators shape ecosystems through both predation and the fear of predation (Creel \& Christianson 2008; Ritchie \& Johnson 2009). Fear-induced behavioural suppression can be as detrimental to subordinate species as predation itself (Schmitz, Krivan, \& Ovadia 2004; Preisser, Bolnick, \& Benard 2005). These non-consumptive effects of apex predator are expected to be strong drivers of mesopredator behaviour, particularly when resource competition is high (Ritchie \& Johnson 2009). A strengthening or relaxing of antipredator behaviours by mesopredators can have cascading effects across the entire ecosystem: altering population demographics, species interactions, ecological function and human-wildlife coexistence (Brown, Laundré, \& Gurung 1999; Ripple \& Beschta 2004; Estes \emph{et al.} 2011; Gaynor \emph{et al.} 2019; Lamb \emph{et al.} 2020). Hence, understanding how apex predators constrain the behaviour of subordinate species is important to accurately predict the ecosystem-wide consequences of predator management, such as reintroductions or lethal control (Gaynor \emph{et al.} 2021).

Spatial and/or temporal niche partitioning may allow predators to coexist by reducing encounter-rates and resource overlap (Kronfeld-Schor \& Dayan 2003). However, mesopredators may not consistently employ avoidance behaviours because perceived predation risk is temporally and spatially variable, and antipredator behaviours typically involve a trade-off against resource acquisition, such as limiting or relegating activity to suboptimal places or times (Lima \& Dill 1990; Lima \& Bednekoff 1999). Therefore, optimal predator avoidance strategies are likely to vary across heterogeneous landscapes where resource availability (e.g., shelter, food) and perceived predation risks differ (Kauffman \emph{et al.} 2007; Willems \& Hill 2009; Wirsing \emph{et al.} 2021). For example, temporal predator avoidance may be preferable over spatial avoidance if food is constantly available throughout the day, and vice versa. These concepts are unified under the `ecology of fear' (Brown, Laundré, \& Gurung 1999) concept, which has gained increasing attention in recent times (Gaynor \emph{et al.} 2019). Notably, the mesopredator release hypothesis has recently been expanded from increases in mesopredator abundance following apex predator decline (Soulé \emph{et al.} 1988) to also include changes in mesopredator behaviour (Brashares \emph{et al.} 2010).

To accurately quantify avoidance within a predator guild, we first need to understand how the overall activity and diel activity patterns of each species varies `naturally' across landscapes, particularly for species with broad distributions. It is widely recognised that the overall activity of different predator species varies across their distributions, but their diel activity patterns are often assumed to be constant. In this paper, we use the term `overall activity' to refer to the number of `independent' predator detections at a site (offset to account for survey duration; analogous to an activity or abundance index), and `diel activity pattern' to refer to fluctuations in relative activity throughout the 24-hour daily cycle. Overall activity is influenced by predator behaviour, population density and the detection process (Anderson 2001), whereas diel activity patterns are a behavioural trait (less likely to be affected by the detection process, if the survey methodology remains consistent).

Despite modern predator survey technologies providing time-stamped detections, detection times are commonly discarded from analyses, probably because joint modelling of overall activity and diel activity patterns is more complicated. When temporal avoidance is tested, it is usually considered in an ad-hoc fashion, by fitting separate models for spatial and temporal avoidance, or by repeating spatial analyses (e.g., resource selection functions) at different time periods (e.g., Smith \emph{et al.} 2019; Basille \emph{et al.} 2015; Kohl \emph{et al.} 2019). However, discretising the daily cycle into categorical periods (e.g., day and night) introduces bias, assuming animals have complete step-changes in behaviour rather than progressive shifts across the daily cycle. Further, dawn and dusk are particularly important times for many predator species.

Generalised Additive Models (hereafter `GAMs') are increasingly being used to estimate animal diel activity patterns and offer a flexible framework to jointly consider overall activity. GAMs also have other benefits, including smoothing penalties to reduce overfitting, the ability to capture nonlinear interactions between multiple variables with different units, and the ability to share information across categorical variables through hierarchical specifications (Wood 2017; Pedersen \emph{et al.} 2019). However, we are only aware of one study which allowed animal diel activity to interact with predation risk as a continuous variable in a GAM (although without considering overall activity; Cunningham \emph{et al.} 2019).

The red fox \emph{Vulpes vulpes} (hereafter `fox'; \textasciitilde6 kg) and feral cat \emph{Felis catus} (hereafter `cat'; \textasciitilde4 kg) have devastating impacts on native prey throughout their introduced range, implicated in the extinction of \textasciitilde10 and 63 species, respectively (Doherty \emph{et al.} 2016). The impacts of these invasive predators have been particularly extreme on the Australian continent (Woinarski, Burbidge, \& Harrison 2015). Cats are more difficult to manage, and so introduced predator control programs in Australia often target only foxes (particularly through poison-baiting; Reddiex \emph{et al.} 2007). As foxes and cats compete for many of the same resources, there is concern that lethal fox control could cause a mesopredator release (Soulé \emph{et al.} 1988) of feral cats (Glen \& Dickman 2005; Robley \emph{et al.} 2014; Marlow \emph{et al.} 2015; Doherty \& Ritchie 2017; Comer \emph{et al.} 2020; Wayne \emph{et al.} 2017). There is some evidence that feral cats increase in activity (although highly uncertain; Hunter \emph{et al.} 2018), density (Chapter \ref{density}) and alter their diets and use of space (Chapter \ref{density}; Molsher \emph{et al.} 2017) in response to fox control. Other studies have investigated potential spatial and temporal interactions between these invasive predators (e.g., Roshier \& Carter 2021), but not in response to fox control, or in a joint spatiotemporal framework that allows flexibility in cat avoidance behaviours in respect to differences in fox activity.

In this study, we explored how the overall activity and diel activity patterns of two competing invasive predators varied across heterogeneous landscapes, in response to (1) space and (2) vegetation types. We then investigated (3) whether cat diel activity patterns change in response to the overall level of fox activity. Our study was conducted in a simple predator system where foxes and cats are the only mammalian carnivores, and fox activity is manipulated using lethal control in some landscapes. This allowed sole focus on the interactions between these two predators, across an experimental gradient of apex predator (fox) activity. We illustrate how GAMs can provide a simple framework to jointly assess spatial and temporal animal activity patterns, as well as avoidance behaviours.

\newpage

\hypertarget{materials-and-methods}{%
\section{MATERIALS AND METHODS}\label{materials-and-methods}}

\hypertarget{study-area-and-camera-trapping}{%
\subsection{Study area and camera-trapping}\label{study-area-and-camera-trapping}}

We compiled data from multiple smaller-scale camera-trap studies across two regions in south-west Victoria, Australia: the Glenelg region and Otway Ranges (Fig. \ref{fig:diel-map}). Introduced foxes and cats are the only medium-large functional mammalian terrestrial carnivores here: native dingoes \emph{Canis familiaris} are long-absent throughout, while tiger quolls \emph{Dasyurus maculatus} are long-absent in the Glenelg region and likely functionally extinct in the Otway Ranges (last confirmed sighting in 2014). In broad sections of each region, government land managers conduct ongoing targeted lethal fox control for biodiversity conservation. Poison-baits containing 3 mg of sodium mono-fluroacetate (`1080') are buried at a depth of 12 - 15 cm at 1-km intervals along accessible forest tracks and roads. Different road densities result in variable densities of poison-baits. Managers also frequently implement prescribed fire across both regions, primarily to reduce fuel loads to prevent large wildfires.

\hypertarget{glenelg-region}{%
\subsubsection{Glenelg region}\label{glenelg-region}}

In the Glenelg region, large patches of natural vegetation are fragmented by pastoral farming and residential properties (Fig. \ref{fig:diel-map}). Foxes in three distinct forest blocks in this region have been subject to poison-baiting since October 2005, with fortnightly bait replacements (Robley \emph{et al.} 2014). These forest blocks, along with three similar, unbaited forest blocks to the north have been simultaneously surveyed annually under the `Glenelg Ark' fox control program since 2005 (40 sites per block; Robley \emph{et al.} 2020). Hair-tubes were used to monitor species from 2005 - 2013 (presented in Robley \emph{et al.} 2014), replaced by camera-traps from 2013; here we present camera-trap data from 2013 - 2019; Robley \emph{et al.} (2020){]}. We also included a further 425 camera-trap deployments at unique locations from early 2018 (M.W.R., PhD surveys). This totals 2039 camera-trap deployments in the Glenelg region, collected in a control-impact experimental design (foxes had been continuously controlled for at 8 - 14 years in the treatment landscapes at the time of these surveys).

\hypertarget{otway-ranges}{%
\subsubsection{Otway Ranges}\label{otway-ranges}}

The Otway Ranges is a largely continuous patch of natural vegetation with a strong east-west rainfall gradient (Fig. \ref{fig:diel-map}). A matrix of cool temperate rainforest and wet forest at high altitudes in the south-west descend into a large heathland directly north, and into dry forests and then heathlands to the north-east. Fox-baiting commenced in small sections of the Otway Ranges in 2008 and large-scale systematic baiting began in 2016 - 2017 under the `Otway Ark' program (Robley, Moloney, \& Parks Victoria West Coast District Team 2019). For the first six weeks, poison-baits were replaced weekly, then changing to ongoing monthly bait-replacement. There was a pause in baiting for approximately six months during the second half of 2018. Fox control recommenced in late 2018 with four weeks of fortnightly bait-replacement, before returning to monthly bait-replacement. A large section of the Otway Ranges to the north-west remains unbaited, but is monitored as an experimental non-treatment site (Robley, Moloney, \& Parks Victoria West Coast District Team 2019). Otway Ark managers survey 372 camera-trap sites annually (sequentially across the region); we present one `before' baiting survey and two `after' baiting surveys of each site from 2016 - 2018, totalling 1113 camera-trap deployments (Robley, Moloney, \& Parks Victoria West Coast District Team 2019). We also include data from an additional before-after control-impact surveys (one `before' baiting survey and two `after' bating surveys) in the western section of the Otway Ranges, conducted annually 2017 - 2019 (M.W.R PhD surveys). This added a further 195 sites and 524 camera-trap deployments.

\hypertarget{camera-trap-set-ups}{%
\paragraph{Camera-trap set-ups}\label{camera-trap-set-ups}}

All camera-trap deployments consisted of a Reconyx (Holmen, Wisconsin) brand camera-trap (white or infrared flash), attached to a tree or a metal picket, facing a lure. The Glenelg Ark and Otway Ark fox monitoring programs positioned camera-traps at least 40 cm above ground on a tree or a metal picket and angled downwards toward a lure approximately 1 - 1.5 m away (Robley, Moloney, \& Parks Victoria West Coast District Team 2019; Robley \emph{et al.} 2020). The lures consisted of peanut butter, golden syrup and rolled oats mixed into a small ball, placed within a tea strainer or PVC pipe container and secured either to the ground, or 20 - 60 cm above ground on a wooden stake. The M.W.R. PhD surveys across both regions positioned camera-traps lower on a tree (around 15 - 30 cm above the ground) angled only slightly downwards toward a tuna oil lure approximately 2 - 2.5 m away (detailed in Rees \emph{et al.} 2019). Camera-traps were active for an average of 47 days (maximum 93 days), totalling 172,052 trap-nights.

\hypertarget{data-preparation}{%
\subsection{Data preparation}\label{data-preparation}}

All data analysis was conducted in R version 3.6.3 (R Core Team 2020). We first used lorelograms to identify the minimum interval to approximate independence (Iannarilli \emph{et al.} 2019); this indicated that discarding repeat detections of a species within 30 minutes was sufficient to reduce temporal autocorrelation. To account for day length variation across space and time, we extracted sunrise and sunset times for each camera-trap deployment using the `maptools' R-package (Bivand \& Lewin-Koh 2021) and adjusted detection times to be relative to sunrise and sunset using the average double anchoring approach described by Vazquez \emph{et al.} (2019). We then built a dataframe consisting of a row for each hour of the day (0 -- 23), for every camera-trap deployment (n = 3667), recording the total number of `independent' fox and feral cat detections within each hour across the camera-trap survey.

\hypertarget{generalised-additive-models}{%
\subsection{Generalised additive models}\label{generalised-additive-models}}

We modelled the total number of independent detections of each predator per hour for each camera-trap deployment (response variable) with generalised additive mixed-effect models implemented in the `mgcv' R-package (Wood 2017). We used the negative binomial family, as overdispersion, but not zero-inflation, was detected with a poisson distribution using the `DHARMa' R-package (Hartig 2020). We specified the natural log of the number of survey days as a model offset to account for differences in camera-trap survey duration, and a random intercept for each site to account for repeat sampling. For fox models, we also included a smooth effect of poison-bait density with separate responses per region to account for the effect of fox control (all figures in this manuscript are derived from fox models predicted to a no fox-baiting scenario). This formed the base model specification for each model we fitted; models differed in their specification of the cyclical hour smooth to provide inference on variations of predator diel activity across the four questions of interest; this is detailed in the sections below.

\hypertarget{how-does-predator-overall-activity-and-diel-activity-patterns-vary-across-space-model-1}{%
\subsubsection{How does predator overall activity and diel activity patterns vary across space? (model 1)}\label{how-does-predator-overall-activity-and-diel-activity-patterns-vary-across-space-model-1}}

To examine how the overall activity and diel activity patterns of each predator varied across space, we fit a model for each predator which included a tensor product interaction between a spatial smooth and hourly smooth. This allowed predators to have different activity levels across space (static across the years surveyed), as well as variation in diel activity pattern across space. Space was modelled using camera-trap coordinates and a duchon spline basis (Miller \& Wood 2014). To examine how the relative strength of diel activity patterns changed across space, we plotted the percentage increase from the minimum to maximum activity estimate within the daily cycle for each predicted location (hereafter referred to as `diel activity pattern strength').

\hypertarget{how-does-predator-overall-activity-and-diel-activity-patterns-vary-across-vegetation-types-model-2}{%
\paragraph{How does predator overall activity and diel activity patterns vary across vegetation types? (model 2)}\label{how-does-predator-overall-activity-and-diel-activity-patterns-vary-across-vegetation-types-model-2}}

Predator activity varied across space; we hypothesised that this was partly due to differences in vegetation type, based both on the observed spatial patterns and because vegetation type is a major driver of understorey habitat structure and prey occurrence in these regions (Swan \emph{et al.} 2015; Hradsky \emph{et al.} 2017). To test whether the diel activity pattern of each predator varied among vegetation types, we identified the Ecological Vegetation Class group (hereafter `vegetation type'; standard units for vegetation classification in Victoria; Department of Environment, Land, Water \& Planning 2020) for each unique camera-trap site, totalling eight vegetation types. As rainforests are interspersed (primarily in low lying gullies) at fine-scales throughout wet and damp forests in the south-eastern Otway Ranges, we merged them together (hereafter referred to as `wet forests'). We then estimated predator activity across vegetation types using a hierarchical model specification: a global smoother for hour (i.e., average response) and group-level smoothers with shared wiggliness for the seven vegetation types (`model GS' detailed in Pedersen \emph{et al.} 2019). We also included a random effect to account for differences in overall activity levels between the two regions.

\hypertarget{do-feral-cats-avoid-foxes-in-space-or-time-model-3}{%
\subsubsection{Do feral cats avoid foxes in space or time? (model 3)}\label{do-feral-cats-avoid-foxes-in-space-or-time-model-3}}

Fox diel activity across vegetation types showed strong similarity between all vegetation types except wet forests. To examine whether cats avoid foxes in space or time, we therefore modelled fox-induced changes in feral cat diel activity separately for wet forest and dry vegetation types. We further split dry vegetation types by region for replication. We refer to the resulting variable as `habitat type', which had three levels: (i) wet forests and rainforests in the western Otway Ranges (`wet\_otways'), (ii) dry vegetation types in the Otway Ranges (`dry\_otways') and (iii) dry vegetation types in the Glenelg region (`dry\_glenelg'). We hypothesised that cats would avoid foxes in time by becoming more diurnal in dry vegetation types where foxes were mostly nocturnal, but not in wet forests where fox activity showed little variation across the daily cycle.

To investigate changes in feral cat diel activity across the range of observed fox activity, we first quantified fox activity for each camera trap deployment as the total number of fox detections for the deployment divided by the number of survey days, to adjust for survey duration (hereafter `adjusted fox counts'). We modelled an interaction between hour (cyclical spine) and adjusted fox counts (thin plate regression spline with shrinkage - meaning fox effects could be entirely removed from the model if not supported by sufficient data), allowing cats to have nonlinear responses to both hour and adjusted fox counts. We fit separate tensor product interactions for each habitat type (using a `by-variable' term). For a direct visual comparison to fox activity, we fit another fox model where a diel curve was estimated separately across each of the three habitat types.

\newpage

\begin{figure}

{\centering \includegraphics[width=1\linewidth]{../figs/map_cams}

}

\caption{Locations of our study regions in south-west Victoria, Australia. The grids of camera-traps are denoted by white dots. The Glenelg region is to the west and Otway region to the east. Native vegetation is indicated by dark green, with hill shading. \textit{Map tiles by Stamen Design, under CC BY 3.0, map data by OpenStreetMap, under CC BY SA.}}\label{fig:diel-map}
\end{figure}

\newpage

\hypertarget{results}{%
\section{RESULTS}\label{results}}

Overall, we collated 5449 and 2202 independent detections of foxes and cats, respectively (separated by at least 30 minutes) from 172,052 camera-trap nights (Table \ref{tab:diel-tab1}).

\hypertarget{how-does-predator-overall-activity-and-diel-activity-patterns-vary-across-space-model-1-1}{%
\subsection{How does predator overall activity and diel activity patterns vary across space? (model 1)}\label{how-does-predator-overall-activity-and-diel-activity-patterns-vary-across-space-model-1-1}}

Predator activity varied considerably across space and throughout the 24-hour daily cycle, and there was some variation in the predator diel activity patterns across space. On average, both predators showed similar diel activity patterns: mostly nocturnal with peaks in activity around sunrise and sunset (i.e., crepuscular; Fig. \ref{fig:diel-veg}i). The main difference between the species was that fox activity peaked just after sunset and they were less likely to be active during the day than cats. Cats also tended to be more active at sunset relative to sunrise.

Diel activity pattern strength also differed between the species. Fox activity was concentrated strongly at particular times of the day, especially in the Glenelg region where activity varied by up to 371\% throughout the daily cycle (Fig. \ref{fig:diel-space}a). Feral cats had relatively more consistent activity throughout the daily cycle and across regions; the maximum difference in cat activity throughout the daily cycle for any given location was 185\%.

Variation in diel activity patterns across space, as well as differences in overall activity between the predators, was strongest in the Otway Ranges. For example, overall fox activity (Fig. \ref{fig:diel-space-o-fox}) and diel activity pattern strength(Fig. \ref{fig:diel-space}b) were lowest in the south-west Otway Ranges, while feral cat overall activity (Fig. \ref{fig:diel-space-o-cat}) and diel activity pattern strength (Fig. \ref{fig:diel-space}b) were highest in that subregion.

\hypertarget{how-does-predator-overall-activity-and-diel-activity-patterns-vary-across-vegetation-types-model-2-1}{%
\subsection{How does predator overall activity and diel activity patterns vary across vegetation types? (model 2)}\label{how-does-predator-overall-activity-and-diel-activity-patterns-vary-across-vegetation-types-model-2-1}}

Overall levels of fox activity were similar across all vegetation types, except wet forests where fox activity was considerably lower. Overall cat activity was more variable across vegetation types; lowest in heathy woodlands and highest in wet forests (Fig. \ref{fig:diel-veg}b).

Diel activity patterns for foxes were similar across all vegetation type except wet forests; in wet forests, foxes were consistently active throughout the daily cycle (Fig. \ref{fig:diel-veg}a). On the other hand, cats were nocturnal (and most active) in wet forests, but largely crepuscular in all other vegetation types (Fig. \ref{fig:diel-veg}b). For both predators, the random effect for region (Glenelg or Otways) in the vegetation models shrank to near-zero=, indicating all variation between the regions was explained by the vegetation covariate and site random intercept.

\hypertarget{do-feral-cats-avoid-foxes-in-space-or-time-model-3-1}{%
\subsection{Do feral cats avoid foxes in space or time? (model 3)}\label{do-feral-cats-avoid-foxes-in-space-or-time-model-3-1}}

Cat spatial activity was relatively unaffected by the fox activity in both habitat types of the Otway Ranges; and, if anything, increased with increasing adjusted fox counts in the Glenelg region (Fig. \ref{fig:diel-cat-fox}), indicating cats did not avoid foxes spatially.

Across all habitat types, feral cat diel activity patterns changed across the gradient of fox activity (Fig. \ref{fig:diel-cat-fox}). In the Glenelg region and Otway dry habitat types, feral cats had a nocturnal-crepuscular diel activity pattern where fox activity was low, but were most active during the day where fox activity was high. In contrast, in the wet forests of the Otway Ranges, feral cats were more strongly nocturnal when fox activity was high.

\newpage

\begin{figure}

{\centering \includegraphics[width=1\linewidth]{../figs/spte_diff_avg_o_fox}

}

\caption{Interaction effect of space-time on red fox \textit{Vulpes vulpes} activity across each hour of the day (0 - 23) in the Otway Ranges, Australia (model 1), as an example. Corresponding plots for feral cats \textit{Felis catus} in this region, as well as both predators in the Glenelg region are provided in the Supporting Information, as are the marginal effects of space and time. White crosses depict unique camera-trap sites.}\label{fig:diel-st-int-o-fox}
\end{figure}

\begin{figure}

{\centering \includegraphics[width=1\linewidth]{../figs/diel_strength_600dpi}

}

\caption{The strength of diel activity patterns of two invasive predators varied within the two study regions in south-west Victoria, Australia (model 1). White crosses depict unique camera-trap sites; colour brightness scales with increasing percentage difference between the minimum and maximum activity estimate over the 24-hour cycle for each location. Red foxes \textit{Vulpes vulpes} (a, c) concentrated their activity during particular times of the day, especially in the Glenelg region (a) and the drier parts of the Otways (c), whereas feral cat \textit{Felis catus} activity was relatively consistent activity throughout the daily cycle (b, d).}\label{fig:diel-space}
\end{figure}

\newpage

\begin{figure}

{\centering \includegraphics[width=1\linewidth]{../figs/predator_veg}

}

\caption{Red foxes \textit{Vulpes vulpes} (a) and feral cat \textit{Felis catus} (B) diel activity patterns overall (i) and across different Ecological Vegetation Class (EVC) groups (ii, iii) in south-west Victoria, Australia (model 2). Dotted, vertical lines represent average sunrise and sunset times. Shaded areas indicate 95\% confidence intervals. Both invasive predators had a crepuscular to nocturnal diel activity pattern on average, with slight deviations across the drier EVC groups and large deviations in wet forests (ii; wet forests shown as pink line). The overall level of activity was relatively consistent across EVC groups for foxes (a – iii), whereas it differed substantially for feral cats (b - iii).}\label{fig:diel-veg}
\end{figure}

\newpage

\begin{figure}

{\centering \includegraphics[width=1\linewidth]{../figs/cat_fox_count}

}

\caption{Variation in mean feral cat \textit{Felis catus} activity (a) and associated uncertainty estimates (b) in response to count of ’independent’ red fox Vulpes vulpes detections (log-transformed and survey effort adjusted) across each ’habitat type’ in south-west Victoria, Australia (model 3). Grey vertical lines represent average sunrise and sunset times. In the Glenelg region, there were more feral cat detections where there were more fox detections, but cat peak diel activity shifted from crepuscular night to pre-dawn and midday (a). In the Otway Ranges, feral cat activity also peaked during the day where fox activity was high in dry vegetation types (b), but was more nocturnal where fox activity was high in the rainforests and wet forests (c).}\label{fig:diel-cat-fox}
\end{figure}

\newpage

\hypertarget{discussion}{%
\section{DISCUSSION}\label{discussion}}

A key question in ecological theory is whether animals are evolutionary hardwired to occupy particular temporal niches, or have circadian rhythms that are responsive to changing environmental conditions and interactions with other species (Schoener 1974; Daan 1981; Lima \& Dill 1990). Here we demonstrate that diel activity patterns are not fixed, but vary across space based on landscape context and fear. In our study, sympatric invasive predators had similar diel activity patterns when averaged across broad regions (i.e., high circular overlap; Fig. \ref{fig:diel-space-time-marginal}; as did Roshier \& Carter 2021), but behaviours varied considerably within landscapes. Fox daily activity patterns were most strongly tied to the daily cycle in dry habitat types, but showed little diel activity pattern in the wet forests (Fig. \ref{fig:diel-veg}a). In contrast, cats were mostly nocturnal in wet forests but crepuscular in dry vegetation types (Fig. \ref{fig:diel-veg}b). Within broad habitat types, cats altered their diel activity patterns at sites with higher fox activity (Fig. \ref{fig:diel-cat-fox}). Control programs that reducing invasive fox activity are therefore likely to change cat diel activity patterns, which in turn may alter cat impacts on native prey species. Quantifying changes in diel activity patterns provides important context for understanding species interactions, which is key for effective ecosystem management (Gaynor \emph{et al.} 2021).

Shifting diel activity patterns may facilitate spatial coexistence of dominant and subordinate species (Carothers \& Jaksić 1984). For cats, altering diel activity patterns to less-preferred times of day may be worthwhile to persist in high-quality habitat. Few studies have demonstrated predator-induced shifts in diel activity (Kronfeld-Schor \& Dayan 2003), but notably, ship rats \emph{Rattus norvegicus} were also found to switch from nocturnal to diurnal behaviour in response to fox activity (Fenn \& Macdonald 1995). For cats in our study, a switch to diurnal behaviour where fox activity was high in the dry vegetation types may have been facilitated by the higher abundance of reptiles in these habitat types relative to wet forests (which are mostly diurnal; Woinarski \emph{et al.} 2018). At wet forest sites with high fox activity, cats concentrated their activity away from sunrise and sunset towards midnight, despite a diurnal shift appearing to similarly reduce the risk of a fox encounter. In this situation, we expect becoming more nocturnal to be favourable over a diurnal shift because this is when small mammals are active and cats would be least visible to foxes {[}cats here mostly had black or grey coats{]}. Overall fox activity was twice as low in the wet forests relative to dry habitat types, and so cats were likely under less pressure to radically alter their diel activity patterns. Understanding how these potential avoidance behaviours impacts native prey is a key research priority to improve invasive predator management.

Cats may have avoided foxes in time, but we saw no sign of spatial avoidance (i.e.~no fine-scale negative association between fox and cat overall activity (Fig. \ref{fig:diel-cat-fox}b:c). Overlap in spatial activity at fine scales (i.e., at the camera-trap sites) may have been facilitated by temporal avoidance (Kronfeld-Schor \& Dayan 2003). However, we considered spatial avoidance across the entire survey duration (averaging 47 days). Cats may indeed avoid foxes spatially, but transiently on the scale of hours to days---after all, how is a cat to know where to avoid a fox without encountering signs of one? Short-term spatial avoidance is quite plausible given foxes mark territories using scats and odours, which cats could tangibly associate with high risk shortly after. Temporary spatial avoidance could be tested using decay curves (e.g., Niedballa \emph{et al.} 2019). Alternatively, no sign of spatial avoidance between these invasive predators could also be an artefact of the quality of camera set-up and hence detectability. The 3667 camera-traps were deployed by numerous people, and the quality of set-ups differed considerably in terms of detecting predators. Camera-traps that angled only slightly downwards, rather than upwards or strongly downwards, seemed most effective at detecting both predator species (M.W Rees, personal observation). Predator interactions are routinely inferred through spatial associations between species; however such analyses are subject to numerous pitfalls which can make inference unreliable (reviewed in Blanchet, Cazelles, \& Gravel 2020).

A distinction of our study from others is that we modelled potential avoidance behaviours in a simple predator guild, where apex predator activity was artificially manipulated. This reduces potential bias from differences in niche preferences and the unmodelled impacts of other predators in the system. We also included replication across different habitat types. However, because our study did not consider associations with prey species, we cannot distinguish whether changes in cat diel activity patterns were the result of direct fox avoidance or indirect associations with shared prey. For example, low fox activity may promote the availability of a preferred shared prey species with a diel pattern which differs from those of cats on average, and as a result, cats might shift diel activity patterns at sites with low fox activity to more closely match those of the more abundant prey species. We would expect introduced European rabbits \emph{Oryctolagus cuniculus} and hares \emph{Lepus europaeus} (which are diurnal) to be particularly likely to induce such a response in dry vegetation types (McGregor \emph{et al.} 2020; Stobo-Wilson \emph{et al.} 2020), however, rabbits and hares are rare within the natural vegetation of the landscapes we surveyed (only ever being detected at 40 of 1232 sites) and there is little evidence that predation by foxes suppresses rabbit populations (Norbury \& Jones 2015; Scroggie \emph{et al.} 2018). While of particular interest, whether temporal fox-cat interactions are direct or indirect does not change the resulting impact on native prey, and hence the outcomes of fox management.

Flexible antipredator behaviours make evolutionary sense, but have been rarely demonstrated in terms of spatiotemporal predator avoidance (although, see Relyea 2003; Brown \emph{et al.} 2013; Cunningham \emph{et al.} 2019), because this often requires manipulative experiments or at least more complicated models (Kronfeld-Schor \& Dayan 2003). Our study demonstrates that GAMs offer a powerful tool for modelling continuous shifts in animal activity across both space and time, capable of capturing complex interactions and sharing information across categorical variables. The inbuilt smoothing penalties are another benefit of GAMs over kernel density estimation (Ridout \& Linkie 2009), in which noisy data can produce spurious estimates (Frey \emph{et al.} 2017; Iannarilli \emph{et al.} 2019). The alternative approach of simply comparing average diel activity overlap between two species (Ridout \& Linkie 2009) would have been misleading for two reasons. Firstly, predator diel activity patterns varied `naturally' across heterogeneous landscapes (requiring avoidance to be tested in wet forests and dry vegetation types separately; Fig. \ref{fig:diel-veg}). Secondly, apex predator temporal avoidance strategies were not consistently employed, but depended on overall apex predator activity and vegetation type (Fig. \ref{fig:diel-cat-fox}). Despite their underlying statistical complexity, GAMs in the `mgcv' R-package are straightforward to fit. Our GAM framework for modelling spatiotemporal activity can be used on any species with time-stamped detections, including datasets with categorical or continuous covariates and hierarchical groupings.

Animal diel activity patterns can be complex, varying across space, habitat types and threat-levels. Despite telling an important story about how animals interact with each other and the environment, detection times are commonly discarded from statistical analyses of camera-trap data. In the rare instances that they are considered, diel activity patterns are predominantly estimated at the population-level, overlooking finer-scale behaviours that can affect fitness, survival and ecosystem-impacts. Our results demonstrate the importance of (a) considering diel activity in regards to species interactions, (b) modelling \emph{changes} in animal behaviour rather than overlap with other species, and (c) testing avoidance behaviours within a joint spatiotemporal framework. Our study adds to the limited body of evidence that top predators can produce a landscape of fear which is powerful enough to reverse the diel activity patterns of subordinate species (Kronfeld-Schor \& Dayan 2003).

\newpage

\hypertarget{acknowledgements}{%
\section*{ACKNOWLEDGEMENTS}\label{acknowledgements}}
\addcontentsline{toc}{section}{ACKNOWLEDGEMENTS}

\newpage

\hypertarget{references}{%
\section*{REFERENCES}\label{references}}
\addcontentsline{toc}{section}{REFERENCES}

\hypertarget{refs}{}
\leavevmode\hypertarget{ref-anderson2001need}{}%
Anderson, D.R. (2001) The need to get the basics right in wildlife field studies. \emph{Wildlife Society Bulletin}, 1294--1297.

\leavevmode\hypertarget{ref-basille2015plastic}{}%
Basille, M., Fortin, D., Dussault, C., Bastille-Rousseau, G., Ouellet, J.-P. \& Courtois, R. (2015) Plastic response of fearful prey to the spatiotemporal dynamics of predator distribution. \emph{Ecology}, \textbf{96}, 2622--2631.

\leavevmode\hypertarget{ref-maptools}{}%
Bivand, R. \& Lewin-Koh, N. (2021) \emph{Maptools: Tools for Handling Spatial Objects}.

\leavevmode\hypertarget{ref-guillaume2020co}{}%
Blanchet, F.G., Cazelles, K. \& Gravel, D. (2020) Co-occurrence is not evidence of ecological interactions. \emph{Ecology Letters}, \textbf{23}, 1050--1063.

\leavevmode\hypertarget{ref-brashares2010ecological}{}%
Brashares, J.S., Prugh, L.R., Stoner, C.J. \& Epps, C.W. (2010) Ecological and conservation implications of mesopredator release. \emph{Trophic cascades: predators, prey, and the changing dynamics of nature}, 221--240.

\leavevmode\hypertarget{ref-brown2013phenotypically}{}%
Brown, G.E., Ferrari, M.C., Elvidge, C.K., Ramnarine, I. \& Chivers, D.P. (2013) Phenotypically plastic neophobia: A response to variable predation risk. \emph{Proceedings of the Royal Society B: Biological Sciences}, \textbf{280}, 20122712.

\leavevmode\hypertarget{ref-brown1999ecology}{}%
Brown, J.S., Laundré, J.W. \& Gurung, M. (1999) The ecology of fear: Optimal foraging, game theory, and trophic interactions. \emph{Journal of Mammalogy}, \textbf{80}, 385--399.

\leavevmode\hypertarget{ref-carothers1984time}{}%
Carothers, J.H. \& Jaksić, F.M. (1984) Time as a niche difference: The role of interference competition. \emph{Oikos}, 403--406.

\leavevmode\hypertarget{ref-comer2020integrating}{}%
Comer, S., Clausen, L., Cowen, S., Pinder, J., Thomas, A., Burbidge, A.H., Tiller, C., Algar, D. \& Speldewinde, P. (2020) Integrating feral cat \emph{(Felis catus)} control into landscape-scale introduced predator management to improve conservation prospects for threatened fauna: A case study from the south coast of Western Australia. \emph{Wildlife Research}, \textbf{47}, 762--778.

\leavevmode\hypertarget{ref-creel2008relationships}{}%
Creel, S. \& Christianson, D. (2008) Relationships between direct predation and risk effects. \emph{Trends in Ecology \& Evolution}, \textbf{23}, 194--201.

\leavevmode\hypertarget{ref-cunningham2019temporal}{}%
Cunningham, C.X., Scoleri, V., Johnson, C.N., Barmuta, L.A. \& Jones, M.E. (2019) Temporal partitioning of activity: Rising and falling top-predator abundance triggers community-wide shifts in diel activity. \emph{Ecography}, \textbf{42}, 2157--2168.

\leavevmode\hypertarget{ref-daan1981adaptive}{}%
Daan, S. (1981) Adaptive daily strategies in behavior. \emph{Biological rhythms} (ed J. Aschoff), pp. 275--298. Springer US, Boston, MA.

\leavevmode\hypertarget{ref-delwp2020bioregions}{}%
Department of Environment, Land, Water \& Planning. (2020) Bioregions and EVC Benchmarks, \url{https://www.environment.vic.gov.au/biodiversity/bioregions-and-evc-benchmarks}

\leavevmode\hypertarget{ref-doherty2016invasive}{}%
Doherty, T.S., Glen, A.S., Nimmo, D.G., Ritchie, E.G. \& Dickman, C.R. (2016) Invasive predators and global biodiversity loss. \emph{Proceedings of the National Academy of Sciences}, \textbf{113}, 11261--11265.

\leavevmode\hypertarget{ref-doherty2017stop}{}%
Doherty, T.S. \& Ritchie, E.G. (2017) Stop jumping the gun: A call for evidence-based invasive predator management. \emph{Conservation Letters}, \textbf{10}, 15--22.

\leavevmode\hypertarget{ref-estes2011trophic}{}%
Estes, J.A., Terborgh, J., Brashares, J.S., Power, M.E., Berger, J., Bond, W.J., Carpenter, S.R., Essington, T.E., Holt, R.D., Jackson, J.B. \& others. (2011) Trophic downgrading of planet earth. \emph{Science}, \textbf{333}, 301--306.

\leavevmode\hypertarget{ref-fenn1995use}{}%
Fenn, M.G. \& Macdonald, D.W. (1995) Use of middens by red foxes: Risk reverses rhythms of rats. \emph{Journal of Mammalogy}, \textbf{76}, 130--136.

\leavevmode\hypertarget{ref-frey2017investigating}{}%
Frey, S., Fisher, J.T., Burton, A.C. \& Volpe, J.P. (2017) Investigating animal activity patterns and temporal niche partitioning using camera-trap data: Challenges and opportunities. \emph{Remote Sensing in Ecology and Conservation}, \textbf{3}, 123--132.

\leavevmode\hypertarget{ref-gaynor2019landscapes}{}%
Gaynor, K.M., Brown, J.S., Middleton, A.D., Power, M.E. \& Brashares, J.S. (2019) Landscapes of fear: Spatial patterns of risk perception and response. \emph{Trends in Ecology \& Evolution}, \textbf{34}, 355--368.

\leavevmode\hypertarget{ref-gaynor2021applied}{}%
Gaynor, K.M., Cherry, M.J., Gilbert, S.L., Kohl, M.T., Larson, C.L., Newsome, T.M., Prugh, L.R., Suraci, J.P., Young, J.K. \& Smith, J.A. (2021) An applied ecology of fear framework: Linking theory to conservation practice. \emph{Animal Conservation}, \textbf{24}, 308--321.

\leavevmode\hypertarget{ref-glen2005complex}{}%
Glen, A.S. \& Dickman, C.R. (2005) Complex interactions among mammalian carnivores in Australia, and their implications for wildlife management. \emph{Biological Reviews}, \textbf{80}, 387--401.

\leavevmode\hypertarget{ref-DHARMa}{}%
Hartig, F. (2020) \emph{DHARMa: Residual Diagnostics for Hierarchical (Multi-Level / Mixed) Regression Models}.

\leavevmode\hypertarget{ref-hradsky2017bayesian}{}%
Hradsky, B.A., Penman, T.D., Ababei, D., Hanea, A., Ritchie, E.G., York, A. \& Di Stefano, J. (2017) Bayesian networks elucidate interactions between fire and other drivers of terrestrial fauna distributions. \emph{Ecosphere}, \textbf{8}, e01926.

\leavevmode\hypertarget{ref-hunter2018not}{}%
Hunter, D.O., Lagisz, M., Leo, V., Nakagawa, S. \& Letnic, M. (2018) Not all predators are equal: A continent-scale analysis of the effects of predator control on Australian mammals. \emph{Mammal Review}, \textbf{48}, 108--122.

\leavevmode\hypertarget{ref-iannarilli2019lorelograms}{}%
Iannarilli, F., Arnold, T.W., Erb, J. \& Fieberg, J.R. (2019) Using lorelograms to measure and model correlation in binary data: Applications to ecological studies. \emph{Methods in Ecology and Evolution}, \textbf{10}, 2153--2162.

\leavevmode\hypertarget{ref-kauffman2007landscape}{}%
Kauffman, M.J., Varley, N., Smith, D.W., Stahler, D.R., MacNulty, D.R. \& Boyce, M.S. (2007) Landscape heterogeneity shapes predation in a newly restored predator--prey system. \emph{Ecology Letters}, \textbf{10}, 690--700.

\leavevmode\hypertarget{ref-kohl2019prey}{}%
Kohl, M.T., Ruth, T.K., Metz, M.C., Stahler, D.R., Smith, D.W., White, P.J. \& MacNulty, D.R. (2019) Do prey select for vacant hunting domains to minimize a multi-predator threat? \emph{Ecology Letters}, \textbf{22}, 1724--1733.

\leavevmode\hypertarget{ref-kronfeld2003partitioning}{}%
Kronfeld-Schor, N. \& Dayan, T. (2003) Partitioning of time as an ecological resource. \emph{Annual review of ecology, evolution, and systematics}, \textbf{34}, 153--181.

\leavevmode\hypertarget{ref-lamb2020ecology}{}%
Lamb, C.T., Ford, A.T., McLellan, B.N., Proctor, M.F., Mowat, G., Ciarniello, L., Nielsen, S.E. \& Boutin, S. (2020) The ecology of human-carnivore coexistence. \emph{Proceedings of the National Academy of Sciences}, \textbf{117}, 17876--17883.

\leavevmode\hypertarget{ref-lima1999temporal}{}%
Lima, S.L. \& Bednekoff, P.A. (1999) Temporal variation in danger drives antipredator behavior: The predation risk allocation hypothesis. \emph{The American Naturalist}, \textbf{153}, 649--659.

\leavevmode\hypertarget{ref-lima1990behavioral}{}%
Lima, S.L. \& Dill, L.M. (1990) Behavioral decisions made under the risk of predation: A review and prospectus. \emph{Canadian journal of zoology}, \textbf{68}, 619--640.

\leavevmode\hypertarget{ref-marlow2015cats}{}%
Marlow, N.J., Thomas, N.D., Williams, A.A., Macmahon, B., Lawson, J., Hitchen, Y., Angus, J. \& Berry, O. (2015) Cats \emph{(Felis catus)} are more abundant and are the dominant predator of woylies \emph{(Bettongia penicillata)} after sustained fox \emph{(Vulpes vulpes)} control. \emph{Australian Journal of Zoology}, \textbf{63}, 18--27.

\leavevmode\hypertarget{ref-mcgregor2020short}{}%
McGregor, H., Moseby, K., Johnson, C.N. \& Legge, S. (2020) The short-term response of feral cats to rabbit population decline: Are alternative native prey more at risk? \emph{Biological Invasions}, \textbf{22}, 799--811.

\leavevmode\hypertarget{ref-miller2014finite}{}%
Miller, D.L. \& Wood, S.N. (2014) Finite area smoothing with generalized distance splines. \emph{Environmental and ecological statistics}, \textbf{21}, 715--731.

\leavevmode\hypertarget{ref-molsher2017mesopredator}{}%
Molsher, R., Newsome, A.E., Newsome, T.M. \& Dickman, C.R. (2017) Mesopredator management: Effects of red fox control on the abundance, diet and use of space by feral cats. \emph{PLoS One}, \textbf{12}, e0168460.

\leavevmode\hypertarget{ref-niedballa2019assessing}{}%
Niedballa, J., Wilting, A., Sollmann, R., Hofer, H. \& Courtiol, A. (2019) Assessing analytical methods for detecting spatiotemporal interactions between species from camera trapping data. \emph{Remote Sensing in Ecology and Conservation}, \textbf{5}, 272--285.

\leavevmode\hypertarget{ref-norbury2015pests}{}%
Norbury, G. \& Jones, C. (2015) Pests controlling pests: Does predator control lead to greater european rabbit abundance in australasia? \emph{Mammal Review}, \textbf{45}, 79--87.

\leavevmode\hypertarget{ref-pedersen2019hierarchical}{}%
Pedersen, E.J., Miller, D.L., Simpson, G.L. \& Ross, N. (2019) Hierarchical generalized additive models in ecology: An introduction with mgcv. \emph{PeerJ}, \textbf{7}, e6876.

\leavevmode\hypertarget{ref-preisser2005scared}{}%
Preisser, E.L., Bolnick, D.I. \& Benard, M.F. (2005) Scared to death? The effects of intimidation and consumption in predator--prey interactions. \emph{Ecology}, \textbf{86}, 501--509.

\leavevmode\hypertarget{ref-R}{}%
R Core Team. (2020) \emph{R: A Language and Environment for Statistical Computing}. R Foundation for Statistical Computing, Vienna, Austria.

\leavevmode\hypertarget{ref-reddiex2007control}{}%
Reddiex, B., Forsyth, D.M., McDonald-Madden, E., Einoder, L.D., Griffioen, P.A., Chick, R.R. \& Robley, A.J. (2007) Control of pest mammals for biodiversity protection in Australia. I. Patterns of control and monitoring. \emph{Wildlife Research}, \textbf{33}, 691--709.

\leavevmode\hypertarget{ref-rees2019unexpectedly}{}%
Rees, M.W., Pascoe, J.H., Wintle, B.A., Le Pla, M., Birnbaum, E.K. \& Hradsky, B.A. (2019) Unexpectedly high densities of feral cats in a rugged temperate forest. \emph{Biological Conservation}, \textbf{239}, 108287.

\leavevmode\hypertarget{ref-relyea2003predators}{}%
Relyea, R.A. (2003) Predators come and predators go: The reversibility of predator-induced traits. \emph{Ecology}, \textbf{84}, 1840--1848.

\leavevmode\hypertarget{ref-ridout2009estimating}{}%
Ridout, M.S. \& Linkie, M. (2009) Estimating overlap of daily activity patterns from camera trap data. \emph{Journal of Agricultural, Biological, and Environmental Statistics}, \textbf{14}, 322--337.

\leavevmode\hypertarget{ref-ripple2004wolves}{}%
Ripple, W.J. \& Beschta, R.L. (2004) Wolves and the ecology of fear: Can predation risk structure ecosystems? \emph{BioScience}, \textbf{54}, 755--766.

\leavevmode\hypertarget{ref-ritchie2009predator}{}%
Ritchie, E.G. \& Johnson, C.N. (2009) Predator interactions, mesopredator release and biodiversity conservation. \emph{Ecology Letters}, \textbf{12}, 982--998.

\leavevmode\hypertarget{ref-robley2014long}{}%
Robley, A., Gormley, A.M., Forsyth, D.M. \& Triggs, B. (2014) Long-term and large-scale control of the introduced red fox increases native mammal occupancy in Australian forests. \emph{Biological Conservation}, \textbf{180}, 262--269.

\leavevmode\hypertarget{ref-robley2019otway}{}%
Robley, A., Moloney, P. \& Parks Victoria West Coast District Team. (2019) \emph{The Otway Ark: response of predators and native species 2016--2018.} Arthur Rylah Institute for Environmental Research Technical Report Series No. 299. Department of Environment, Land, Water; Planning, Heidelberg, Victoria.

\leavevmode\hypertarget{ref-robley2020glenelg}{}%
Robley, A., Moloney, P., Stringer, L. \& Donald, S. (2020) \emph{Glenelg Ark 2005--2019: long-term predator and native mammal response to predator control.} Arthur Rylah Institute for Environmental Research Technical Report Series No. 318. Department of Environment, Land, Water; Planning, Heidelberg, Victoria.

\leavevmode\hypertarget{ref-roshier2021space}{}%
Roshier, D.A. \& Carter, A. (2021) Space use and interactions of two introduced mesopredators, european red fox and feral cat, in an arid landscape. \emph{Ecosphere}, \textbf{12}, e03628.

\leavevmode\hypertarget{ref-schmitz2004trophic}{}%
Schmitz, O.J., Krivan, V. \& Ovadia, O. (2004) Trophic cascades: The primacy of trait-mediated indirect interactions. \emph{Ecology Letters}, \textbf{7}, 153--163.

\leavevmode\hypertarget{ref-schoener1974compression}{}%
Schoener, T.W. (1974) The compression hypothesis and temporal resource partitioning. \emph{Proceedings of the National Academy of Sciences}, \textbf{71}, 4169--4172.

\leavevmode\hypertarget{ref-scroggie2018invasive}{}%
Scroggie, M.P., Forsyth, D.M., McPhee, S.R., Matthews, J., Stuart, I.G., Stamation, K.A., Lindeman, M. \& Ramsey, D.S. (2018) Invasive prey controlling invasive predators? European rabbit abundance does not determine red fox population dynamics. \emph{Journal of Applied Ecology}, \textbf{55}, 2621--2631.

\leavevmode\hypertarget{ref-smith2019integrating}{}%
Smith, J.A., Donadio, E., Pauli, J.N., Sheriff, M.J. \& Middleton, A.D. (2019) Integrating temporal refugia into landscapes of fear: Prey exploit predator downtimes to forage in risky places. \emph{Oecologia}, \textbf{189}, 883--890.

\leavevmode\hypertarget{ref-soule1988reconstructed}{}%
Soulé, M.E., Bolger, D.T., Alberts, A.C., Wrights, J., Sorice, M. \& Hill, S. (1988) Reconstructed dynamics of rapid extinctions of chaparral-requiring birds in urban habitat islands. \emph{Conservation Biology}, \textbf{2}, 75--92.

\leavevmode\hypertarget{ref-stobo2020management}{}%
Stobo-Wilson, A.M., Brandle, R., Johnson, C.N. \& Jones, M.E. (2020) Management of invasive mesopredators in the Flinders Ranges, South Australia: Effectiveness and implications. \emph{Wildlife Research}, \textbf{47}, 720--730.

\leavevmode\hypertarget{ref-swan2015predicting}{}%
Swan, M., Christie, F., Sitters, H., York, A. \& Di Stefano, J. (2015) Predicting faunal fire responses in heterogeneous landscapes: The role of habitat structure. \emph{Ecological Applications}, \textbf{25}, 2293--2305.

\leavevmode\hypertarget{ref-vazquez2019comparing}{}%
Vazquez, C., Rowcliffe, J.M., Spoelstra, K. \& Jansen, P.A. (2019) Comparing diel activity patterns of wildlife across latitudes and seasons: Time transformations using day length. \emph{Methods in Ecology and Evolution}, \textbf{10}, 2057--2066.

\leavevmode\hypertarget{ref-wayne2017recoveries}{}%
Wayne, A.F., Maxwell, M.A., Ward, C.G., Wayne, J.C., Vellios, C.V. \& Wilson, I.J. (2017) Recoveries and cascading declines of native mammals associated with control of an introduced predator. \emph{Journal of Mammalogy}, \textbf{98}, 489--501.

\leavevmode\hypertarget{ref-willems2009predator}{}%
Willems, E.P. \& Hill, R.A. (2009) Predator-specific landscapes of fear and resource distribution: Effects on spatial range use. \emph{Ecology}, \textbf{90}, 546--555.

\leavevmode\hypertarget{ref-wirsing2021context}{}%
Wirsing, A.J., Heithaus, M.R., Brown, J.S., Kotler, B.P. \& Schmitz, O.J. (2021) The context dependence of non-consumptive predator effects. \emph{Ecology Letters}, \textbf{24}, 113--129.

\leavevmode\hypertarget{ref-woinarski2015ongoing}{}%
Woinarski, J.C.Z., Burbidge, A.A. \& Harrison, P.L. (2015) Ongoing unraveling of a continental fauna: Decline and extinction of Australian mammals since European settlement. \emph{Proceedings of the National Academy of Sciences}, \textbf{112}, 4531--4540.

\leavevmode\hypertarget{ref-woinarski2018reptiles}{}%
Woinarski, J., Murphy, B., Palmer, R., Legge, S., Dickman, C., Doherty, T., Edwards, G., Nankivell, A., Read, J. \& Stokeld, D. (2018) How many reptiles are killed by cats in Australia? \emph{Wildlife Research}, \textbf{45}, 247--266.

\leavevmode\hypertarget{ref-wood2017generalized}{}%
Wood, S.N. (2017) \emph{Generalized Additive Models: An Introduction with R}. CRC press.

\newpage

\setcounter{table}{0}  \renewcommand{\thetable}{S\arabic{table}} \setcounter{figure}{0} \renewcommand{\thefigure}{S\arabic{figure}} \setcounter{section}{0} \renewcommand{\thesection}{S\arabic{section}}

\hypertarget{supporting-information}{%
\section*{Supporting Information}\label{supporting-information}}
\addcontentsline{toc}{section}{Supporting Information}

\newpage

\(~\)

\(~\)

\(~\)

\begingroup\fontsize{10}{12}\selectfont

\begin{longtable}[t]{llrrrr}
\caption{\label{tab:diel-tab1}Summary of the number of camera-trap deployments, unique survey sites and 'independent' counts of invasive predator detections across Ecological Vegetation Class groups within the Glenelg and Otway regions, south-west Victoria, Australia.}\\
\toprule
Vegetation & Region & Sites & Deployments & Fox counts & Cat counts\\
\midrule
Dry Forest & Glenelg & 25 & 69 & 347 & 9\\
 & Otways & 111 & 314 & 341 & 158\\
Heathland & Glenelg & 40 & 119 & 265 & 59\\
 & Otways & 3 & 9 & 8 & 6\\
Heathy Woodland & Glenelg & 154 & 424 & 574 & 96\\
\addlinespace
 & Otways & 82 & 256 & 160 & 66\\
Herb-rich Woodland & Glenelg & 59 & 373 & 863 & 198\\
 & Otways & 2 & 6 & 3 & 2\\
Lowland Forest & Glenelg & 383 & 1046 & 1900 & 290\\
 & Otways & 52 & 163 & 190 & 35\\
\addlinespace
Swampy Scrub & Glenelg & 4 & 10 & 19 & 8\\
 & Otways & 36 & 98 & 64 & 88\\
Wet Forest & Otways & 281 & 780 & 715 & 1187\\
\textbf{Total} & \textbf{} & \textbf{1232} & \textbf{3667} & \textbf{5449} & \textbf{2202}\\
\bottomrule
\end{longtable}
\endgroup{}

\newpage

\begingroup\fontsize{10}{12}\selectfont

\begin{longtable}[t]{llrrr}
\caption{\label{tab:diel-tab-fits}Generalised additive model summaries for invasive predator spatiotemporal activity in south-west Victoria, Australia.}\\
\toprule
Species & Model & EDF & dev.expl & r.sq\\
\midrule
fox & 1\_spatial & 772.6469 & 0.3514 & 0.1170\\
fox & 2\_vegetation\_type & 766.1054 & 0.3498 & 0.1155\\
fox & 3\_habitat\_type & 768.2692 & 0.3487 & 0.1139\\
cat & 1\_spatial & 562.9236 & 0.2314 & 0.0772\\
cat & 2\_vegetation\_type & 566.7562 & 0.2300 & 0.0769\\
\addlinespace
cat & 3\_fox\_by\_habitat\_type & 585.8580 & 0.2317 & 0.0779\\
\bottomrule
\multicolumn{5}{l}{\rule{0pt}{1em}\textit{Note: }}\\
\multicolumn{5}{l}{\rule{0pt}{1em}EDF - estimated degrees of freedom of all model terms.}\\
\multicolumn{5}{l}{\rule{0pt}{1em}dev.expl - proportion of the null deviance explained by the model. }\\
\multicolumn{5}{l}{\rule{0pt}{1em}r.sq -  adjusted r-squared value.}\\
\end{longtable}
\endgroup{}

\newpage

\begin{figure}

{\centering \includegraphics[width=1\linewidth]{../figs/avg_diel_predator}

}

\caption{Marginal effect of time (model 1) on predators across both study regions in south-west Victoria, Australia. White crosses depict unique camera-trap sites}\label{fig:diel-space-time-marginal}
\end{figure}

\newpage

\begin{center}\includegraphics[width=1\linewidth]{../figs/sp_marginal_g} \end{center}

\begin{figure}

{\centering \includegraphics[width=1\linewidth]{../figs/sp_marginal_o}

}

\caption{Marginal effect of space on predators across south-west Victoria, Australia (model 1). White crosses depict unique camera-trap sites}\label{fig:diel-space-time-marginal-o}
\end{figure}

\newpage

\begin{figure}

{\centering \includegraphics[width=1\linewidth]{../figs/spte_diff_avg_g_fox}

}

\caption{Interaction effect of space-time on feral cat \textit{Felis catus} activity across each hour of the day (0 - 23) in the Glenelg region, Australia (model 1). White crosses depict unique camera-trap sites. }\label{fig:diel-st-int-g-fox}
\end{figure}

\newpage

\begin{figure}

{\centering \includegraphics[width=1\linewidth]{../figs/spte_diff_avg_g_cat}

}

\caption{Interaction effect of space-time on feral cat \textit{Felis catus} activity across each hour of the day (0 - 23) in the Glenelg region, Australia (model 1). White crosses depict unique camera-trap sites. }\label{fig:diel-st-int-g-cat}
\end{figure}

\newpage

\begin{figure}

{\centering \includegraphics[width=1\linewidth]{../figs/spte_diff_avg_o_cat}

}

\caption{Interaction effect of space-time on feral cat \textit{Felis catus} activity across each hour of the day (0 - 23) in the Otway Ranges, Australia (model 1). White crosses depict unique camera-trap sites. }\label{fig:diel-st-int-o-cat}
\end{figure}

\newpage

\begin{figure}

{\centering \includegraphics[width=1\linewidth]{../figs/spte_facet_o_fox}

}

\caption{Overall spatial activity of red foxes \textit{Vulpes vulpes} for each hour of the day (0 - 23) in the Glenelg region, Australia (model 1). White crosses depict unique camera-trap sites}\label{fig:diel-space-g-fox}
\end{figure}

\newpage

\begin{figure}

{\centering \includegraphics[width=1\linewidth]{../figs/spte_facet_g_cat}

}

\caption{Overall spatial activity of feral cats \textit{Felis catus} for each hour of the day (0 - 23) in the Glenelg region, Australia (model 1). White crosses depict unique camera-trap sites}\label{fig:diel-space-g-cat}
\end{figure}

\newpage

\begin{figure}

{\centering \includegraphics[width=1\linewidth]{../figs/spte_facet_o_fox}

}

\caption{Overall spatial activity of red foxes \textit{Vulpes vulpes} for each hour of the day (0 - 23) in the Otway Ranges, Australia (model 1). White crosses depict unique camera-trap sites}\label{fig:diel-space-o-fox}
\end{figure}

\newpage

\begin{figure}

{\centering \includegraphics[width=1\linewidth]{../figs/spte_facet_o_cat}

}

\caption{Overall spatial activity of feral cats \textit{Felis catus} for each hour of the day (0 - 23) in the Otway Ranges, Australia (model 1). White crosses depict unique camera-trap sites}\label{fig:diel-space-o-cat}
\end{figure}

\end{document}

### 2022-02-14 r_scripts/plot_cat_v_fox.R

@@ -18,7 +18,7 @@ plot_range <- ggplot(aes(hour, fox_count_adj, fill = fit), data = df_int) +
   scale_fill_viridis("Cat activity", option = "viridis", breaks = c(-5, -3), labels = c("Low", "High")) + 
   ggtitle("", subtitle = "Feral cat") +
   geom_vline(xintercept = c(6.16,18.34), colour = "black", size = 0.6, linetype="dotted") + 
-  scale_y_continuous(breaks=c(0, 3), labels = c("Low", "High")) +
+  scale_y_continuous(breaks=c(min(records$fox_count_adj), max(records$fox_count_adj)), labels = c("Low", "High")) +
   ylab("Fox activity") + 
   xlab("Hour") 
 
@@ -71,6 +71,20 @@ plot_fox <- ggplot(aes(hour, fit), data = df_int) +
 #  save
 png("figs/cat_fox_count.png", width = 9, height = 10, res = 600, units = "in")
 plot_fox / plot_range / plot_wet_lowhigh + plot_annotation(tag_levels = "a")
+dev.off()
+
+
+# for slides
+png("figs/cat_fox_count_pres1.png", width = 7.2, height = 5, res = 600, units = "in")
+plot_fox / plot_spacer() + plot_annotation(tag_levels = "a")
+dev.off()
+  
+png("figs/cat_fox_count_pres2.png", width = 7.2, height = 5.5, res = 600, units = "in")
+plot_fox / plot_range + plot_annotation(tag_levels = "a")
+dev.off()
+
+png("figs/cat_fox_count_pres3.png", width = 7.2, height = 5.5, res = 600, units = "in")
+plot_fox / plot_wet_lowhigh + plot_annotation(tag_levels = "a")
 dev.off()
 
 # END

### 2022-02-14 r_scripts/plot_spatial_only_interaction.R

@@ -317,6 +317,32 @@ plot_facet_o_cat
 dev.off()
 
 
+# PLOT INTERACTION AS GIFS ------------------------------------------------
+# additional packages
+library(gganimate)
+
+# otways
+plot_gif_o <- ggplot(aes(x, y, fill = fit_int), data = model_predictions_o) +
+  geom_tile() +
+  scale_fill_viridis("Change from average diel pattern (log[count])", option = "viridis") +
+  geom_point(data = camdata_o, fill = NA, col = "white", size = 0.7, alpha = 1, shape = 3) +
+  theme_bw(10) + 
+  facet_wrap(~species, scales = "free") + 
+  theme(axis.title = element_blank(),
+        legend.position = "bottom") +  
+  labs(title = 'Hour: {frame_time}:00') +
+  transition_time(hour) +
+  ease_aes('linear')
+
+
+# save gif
+animate(plot_gif_o, units = "in", width = 10, height = 5,  res = 200)
+anim_save("otways_interaction.gif", animation = last_animation(), path = "figs/")
+
+
+
+
+
 # PLOT MARGINAL SPACE -----------------------------------------------------
 
 ## GLENELG

### 2022-04-14 manuscript/main_text.Rmd
post-phd-getting-ready-for-submission

@@
     footnote: 1
+  - name: Brendan A. Wintle
+    email: 
+    affiliation: UOM
   - name: Jack H. Pascoe
@@
     affiliation: CEC
-  - name: Alan Robley
-    email: 
-    affiliation: ARI
   - name: Emma K. Birnbaum
@@
     affiliation: CEC
-  - name: Brendan A. Wintle
-    email: 
-    affiliation: UOM
   - name: Bronwyn A. Hradsky
@@
     address: Conservation Ecology Centre, Otway Lighthouse Rd, Cape Otway, VIC, Australia
-  - code: ARI 
-    address: Department of Environment, Land, Water and Planning, Arthur Rylah Institute for Environmental Research, Heidelberg, Australia
 footnote:
@@
 In broad sections of each region, government land managers conduct ongoing targeted lethal fox control for biodiversity conservation.
-Poison-baits containing 3 mg of sodium mono-fluroacetate ('1080') are buried at a depth of 12 - 15 cm at 1-km intervals along accessible forest tracks and roads.
+Poison-baits containing 3 mg of sodium fluroacetate ('1080') are buried at a depth of 12 - 15 cm at 1-km intervals along accessible forest tracks and roads.
 Different road densities result in variable densities of poison-baits.
@@
 Hair-tubes were used to monitor species from 2005 - 2013 [presented in @robley2014long], replaced by camera-traps from 2013; here we present camera-trap data from 2013 - 2019 [@robley2020glenelg].
-We also included a further 425 camera-trap deployments at unique locations from early 2018 (M.W.R., PhD surveys).
+We also included a further 425 camera-trap deployments at unique locations from early 2018 (M.W.R.
+PhD surveys).
 This totals 2039 camera-trap deployments in the Glenelg region, collected in a control-impact experimental design (foxes had been continuously controlled for at 8 - 14 years in the treatment landscapes at the time of these surveys).
@@
 
-#### Camera-trap set-ups
+### Camera-trap set-ups
 All camera-trap deployments consisted of a Reconyx (Holmen, Wisconsin) brand camera-trap (white or infrared flash), attached to a tree or a metal picket, facing a lure.
@@
 
-#### How does predator overall activity and diel activity patterns vary across vegetation types? (model 2)
+### How does predator overall activity and diel activity patterns vary across vegetation types? (model 2)
 Predator activity varied across space; we hypothesised that this was partly due to differences in vegetation type, based both on the observed spatial patterns and because vegetation type is a major driver of understorey habitat structure and prey occurrence in these regions [@swan2015predicting; @hradsky2017bayesian].
@@
 ============
-
 Overall, we collated 5449 and 2202 independent detections of foxes and cats, respectively (separated by at least 30 minutes) from 172,052 camera-trap nights (Table \@ref(tab:diel-tab1)).
 
-## How does predator overall activity and diel activity patterns vary across space? (model 1)
+### How does predator overall activity and diel activity patterns vary across space? (model 1)
 Predator activity varied considerably across space and throughout the 24-hour daily cycle, and there was some variation in the predator diel activity patterns across space.
@@
 For example, overall fox activity (Fig.
-\@ref(fig:diel-space-o-fox)) and diel activity pattern strength(Fig.
+\@ref(fig:diel-space-o-fox)) and diel activity pattern strength (Fig.
 \@ref(fig:diel-space)b) were lowest in the south-west Otway Ranges, while feral cat overall activity (Fig.
@@
 
-## How does predator overall activity and diel activity patterns vary across vegetation types? (model 2)
+### How does predator overall activity and diel activity patterns vary across vegetation types? (model 2)
 
@@
 \@ref(fig:diel-veg)b).
-For both predators, the random effect for region (Glenelg or Otways) in the vegetation models shrank to near-zero=, indicating all variation between the regions was explained by the vegetation covariate and site random intercept.
+For both predators, the random effect for region (Glenelg or Otways) in the vegetation models shrank to near-zero, indicating all variation between the regions was explained by the vegetation covariate and site random intercept.
 
-## Do feral cats avoid foxes in space or time? (model 3)
+### Do feral cats avoid foxes in space or time? (model 3)
 Cat spatial activity was relatively unaffected by the fox activity in both habitat types of the Otway Ranges; and, if anything, increased with increasing adjusted fox counts in the Glenelg region (Fig.
@@
 \@ref(fig:diel-cat-fox)).
-Control programs that reducing invasive fox activity are therefore likely to change cat diel activity patterns, which in turn may alter cat impacts on native prey species.
+Control programs that reduce invasive fox activity are therefore likely to change cat diel activity patterns, which in turn may alter cat impacts on native prey species.
 Quantifying changes in diel activity patterns provides important context for understanding species interactions, which is key for effective ecosystem management [@gaynor2021applied].
@@
 For cats, altering diel activity patterns to less-preferred times of day may be worthwhile to persist in high-quality habitat.
-Few studies have demonstrated predator-induced shifts in diel activity [@kronfeld2003partitioning], but notably, ship rats *Rattus norvegicus* were also found to switch from nocturnal to diurnal behaviour in response to fox activity [@fenn1995use].
+Few studies have demonstrated predator-induced shifts in diel activity [@kronfeld2003partitioning], but notably, ship rats *Rattus norvegicus* were also found to switch from nocturnal to diurnal behaviour in response to fox activity [@fenn1995use] and a similar nocturnal-diurnal shift was observed in American mink *Neovison vison* following the recolonisation of native predators [@harrington2009impact].
 For cats in our study, a switch to diurnal behaviour where fox activity was high in the dry vegetation types may have been facilitated by the higher abundance of reptiles in these habitat types relative to wet forests [which are mostly diurnal; @woinarski2018reptiles].
 At wet forest sites with high fox activity, cats concentrated their activity away from sunrise and sunset towards midnight, despite a diurnal shift appearing to similarly reduce the risk of a fox encounter.
-In this situation, we expect becoming more nocturnal to be favourable over a diurnal shift because this is when small mammals are active and cats would be least visible to foxes [cats here mostly had black or grey coats].
+In this situation, we expect becoming more nocturnal to be favourable over a diurnal shift because this is when small mammals are active and cats would be least visible to foxes (cats here mostly had black or grey coats).
 Overall fox activity was twice as low in the wet forests relative to dry habitat types, and so cats were likely under less pressure to radically alter their diel activity patterns.
@@
 \@ref(fig:diel-cat-fox)b:c).
-Overlap in spatial activity at fine scales (i.e., at the camera-trap sites) may have been facilitated by temporal avoidance [@kronfeld2003partitioning].
+Overlap in spatial activity at fine scales (i.e., at the level of camera-trap sites) may have been facilitated by temporal avoidance [@kronfeld2003partitioning].
 However, we considered spatial avoidance across the entire survey duration (averaging 47 days).
@@
 In the rare instances that they are considered, diel activity patterns are predominantly estimated at the population-level, overlooking finer-scale behaviours that can affect fitness, survival and ecosystem-impacts.
-Our results demonstrate the importance of (a) considering diel activity in regards to species interactions, (b) modelling *changes* in animal behaviour rather than overlap with other species, and (c) testing avoidance behaviours within a joint spatiotemporal framework.
+Our results demonstrate the importance of (a) considering diel activity in regards to species interactions, (b) modelling changes in animal behaviour rather than overlap with other species, and (c) testing avoidance behaviours within a joint spatiotemporal framework.
 Our study adds to the limited body of evidence that top predators can produce a landscape of fear which is powerful enough to reverse the diel activity patterns of subordinate species [@kronfeld2003partitioning].
@@
 ===================
+
+I had field assistance from: Shauni Omond, Shayne Neal, Asitha Samarawickrama, Shelley Thompson, Erin Harris, Hannah Killian, Lani Watson, Mark Dorman, Jack Davis, Carl Roffey, Bruce Edley, Larissa Oliveira Gonçalves, Ben Lake, Chantelle Geissler, Aviya Naccarella, Emily Gronow, Harley England, David Pitts, Annie Hobby, Louise Falls, Thomas McKinnon, Jimmy Downie, Marney Hradsky, Stephanie Samson, Robin Sinclair, Asmaa Alhusainan, Kelly Forrester, Tammana Wadawani, Emily Gregg, Hannah Edwards, Adam Beck, Vishnu Memnon, Sandy Lu, Pia Lentini, Nick Golding, Emily McColl-Gausden, Nina Page, Maggie Campbell-Jones, Kyle Quinn and Jack Dickson.
+
+Surveys were conducted under University of Melbourne Animal Ethics Committee approval 1714119 and Victorian Government Department of Environment, Land Water and Planning Research Permit 10008273.
 
@@
 fits_summaries <- read_csv("../derived_data/model_summaries.csv") %>%
-  dplyr::select(-c(X1))
+  dplyr::select(-c(...1))
 

### 2022-04-14 manuscript/main_text.tex
post-phd-getting-ready-for-submission

@@
 \usepackage{graphicx}
-\usepackage{booktabs} % book-quality tables
 %%%%%%%%%%%%%%%% end my additions to header
@@
 \bibliographystyle{elsarticle-harv}
-\usepackage{longtable}
+\usepackage{longtable,booktabs,array}
+\usepackage{calc} % for calculating minipage widths
+% Correct order of tables after \paragraph or \subparagraph
+\usepackage{etoolbox}
+\makeatletter
+\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
+\makeatother
+% Allow footnotes in longtable head/foot
+\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
+\makesavenoteenv{longtable}
 \ifxetex
@@
 
+\linenumbers
+
 \hypertarget{abstract}{%
@@
 In broad sections of each region, government land managers conduct ongoing targeted lethal fox control for biodiversity conservation.
-Poison-baits containing 3 mg of sodium mono-fluroacetate (`1080') are buried at a depth of 12 - 15 cm at 1-km intervals along accessible forest tracks and roads.
+Poison-baits containing 3 mg of sodium fluroacetate (`1080') are buried at a depth of 12 - 15 cm at 1-km intervals along accessible forest tracks and roads.
 Different road densities result in variable densities of poison-baits.
@@
 These forest blocks, along with three similar, unbaited forest blocks to the north have been simultaneously surveyed annually under the `Glenelg Ark' fox control program since 2005 (40 sites per block; Robley \emph{et al.} 2020).
-Hair-tubes were used to monitor species from 2005 - 2013 (presented in Robley \emph{et al.} 2014), replaced by camera-traps from 2013; here we present camera-trap data from 2013 - 2019; Robley \emph{et al.} (2020){]}.
-We also included a further 425 camera-trap deployments at unique locations from early 2018 (M.W.R., PhD surveys).
+Hair-tubes were used to monitor species from 2005 - 2013 (presented in Robley \emph{et al.} 2014), replaced by camera-traps from 2013; here we present camera-trap data from 2013 - 2019 (Robley \emph{et al.} 2020).
+We also included a further 425 camera-trap deployments at unique locations from early 2018 (M.W.R.
+PhD surveys).
 This totals 2039 camera-trap deployments in the Glenelg region, collected in a control-impact experimental design (foxes had been continuously controlled for at 8 - 14 years in the treatment landscapes at the time of these surveys).
@@
 \hypertarget{camera-trap-set-ups}{%
-\paragraph{Camera-trap set-ups}\label{camera-trap-set-ups}}
+\subsubsection{Camera-trap set-ups}\label{camera-trap-set-ups}}
 
@@
 \hypertarget{how-does-predator-overall-activity-and-diel-activity-patterns-vary-across-vegetation-types-model-2}{%
-\paragraph{How does predator overall activity and diel activity patterns vary across vegetation types?
+\subsubsection{How does predator overall activity and diel activity patterns vary across vegetation types?
 (model 2)}\label{how-does-predator-overall-activity-and-diel-activity-patterns-vary-across-vegetation-types-model-2}}
@@
 \hypertarget{how-does-predator-overall-activity-and-diel-activity-patterns-vary-across-space-model-1-1}{%
-\subsection{How does predator overall activity and diel activity patterns vary across space?
+\subsubsection{How does predator overall activity and diel activity patterns vary across space?
 (model 1)}\label{how-does-predator-overall-activity-and-diel-activity-patterns-vary-across-space-model-1-1}}
@@
 For example, overall fox activity (Fig.
-\ref{fig:diel-space-o-fox}) and diel activity pattern strength(Fig.
+\ref{fig:diel-space-o-fox}) and diel activity pattern strength (Fig.
 \ref{fig:diel-space}b) were lowest in the south-west Otway Ranges, while feral cat overall activity (Fig.
@@
 \hypertarget{how-does-predator-overall-activity-and-diel-activity-patterns-vary-across-vegetation-types-model-2-1}{%
-\subsection{How does predator overall activity and diel activity patterns vary across vegetation types?
+\subsubsection{How does predator overall activity and diel activity patterns vary across vegetation types?
 (model 2)}\label{how-does-predator-overall-activity-and-diel-activity-patterns-vary-across-vegetation-types-model-2-1}}
@@
 \ref{fig:diel-veg}b).
-For both predators, the random effect for region (Glenelg or Otways) in the vegetation models shrank to near-zero=, indicating all variation between the regions was explained by the vegetation covariate and site random intercept.
+For both predators, the random effect for region (Glenelg or Otways) in the vegetation models shrank to near-zero, indicating all variation between the regions was explained by the vegetation covariate and site random intercept.
 
 \hypertarget{do-feral-cats-avoid-foxes-in-space-or-time-model-3-1}{%
-\subsection{Do feral cats avoid foxes in space or time? (model 3)}\label{do-feral-cats-avoid-foxes-in-space-or-time-model-3-1}}
+\subsubsection{Do feral cats avoid foxes in space or time? (model 3)}\label{do-feral-cats-avoid-foxes-in-space-or-time-model-3-1}}
 
@@
 
-\caption{Variation in mean feral cat \textit{Felis catus} activity (a) and associated uncertainty estimates (b) in response to count of ’independent’ red fox Vulpes vulpes detections (log-transformed and survey effort adjusted) across each ’habitat type’ in south-west Victoria, Australia (model 3).
+\caption{Variation in mean feral cat \textit{Felis catus} activity (a) and associated uncertainty estimates (b) in response to count of ’independent’ red fox \textit{Vulpes vulpes} detections (log-transformed and survey effort adjusted) across each ’habitat type’ in south-west Victoria, Australia (model 3).
 Grey vertical lines represent average sunrise and sunset times.
@@
 \ref{fig:diel-cat-fox}).
-Control programs that reducing invasive fox activity are therefore likely to change cat diel activity patterns, which in turn may alter cat impacts on native prey species.
+Control programs that reduce invasive fox activity are therefore likely to change cat diel activity patterns, which in turn may alter cat impacts on native prey species.
 Quantifying changes in diel activity patterns provides important context for understanding species interactions, which is key for effective ecosystem management (Gaynor \emph{et al.} 2021).
@@
 For cats, altering diel activity patterns to less-preferred times of day may be worthwhile to persist in high-quality habitat.
-Few studies have demonstrated predator-induced shifts in diel activity (Kronfeld-Schor \& Dayan 2003), but notably, ship rats \emph{Rattus norvegicus} were also found to switch from nocturnal to diurnal behaviour in response to fox activity (Fenn \& Macdonald 1995).
+Few studies have demonstrated predator-induced shifts in diel activity (Kronfeld-Schor \& Dayan 2003), but notably, ship rats \emph{Rattus norvegicus} were also found to switch from nocturnal to diurnal behaviour in response to fox activity (Fenn \& Macdonald 1995) and a similar nocturnal-diurnal shift was observed in American mink \emph{Neovison vison} following the recolonisation of native predators (Harrington \emph{et al.} 2009).
 For cats in our study, a switch to diurnal behaviour where fox activity was high in the dry vegetation types may have been facilitated by the higher abundance of reptiles in these habitat types relative to wet forests (which are mostly diurnal; Woinarski \emph{et al.} 2018).
 At wet forest sites with high fox activity, cats concentrated their activity away from sunrise and sunset towards midnight, despite a diurnal shift appearing to similarly reduce the risk of a fox encounter.
-In this situation, we expect becoming more nocturnal to be favourable over a diurnal shift because this is when small mammals are active and cats would be least visible to foxes {[}cats here mostly had black or grey coats{]}.
+In this situation, we expect becoming more nocturnal to be favourable over a diurnal shift because this is when small mammals are active and cats would be least visible to foxes (cats here mostly had black or grey coats).
 Overall fox activity was twice as low in the wet forests relative to dry habitat types, and so cats were likely under less pressure to radically alter their diel activity patterns.
@@
 \ref{fig:diel-cat-fox}b:c).
-Overlap in spatial activity at fine scales (i.e., at the camera-trap sites) may have been facilitated by temporal avoidance (Kronfeld-Schor \& Dayan 2003).
+Overlap in spatial activity at fine scales (i.e., at the level of camera-trap sites) may have been facilitated by temporal avoidance (Kronfeld-Schor \& Dayan 2003).
 However, we considered spatial avoidance across the entire survey duration (averaging 47 days).
@@
 Camera-traps that angled only slightly downwards, rather than upwards or strongly downwards, seemed most effective at detecting both predator species (M.W Rees, personal observation).
-Predator interactions are routinely inferred through spatial associations between species; however such analyses are subject to numerous pitfalls which can make inference unreliable (reviewed in Blanchet, Cazelles, \& Gravel 2020).
-
+Predator interactions are routinely inferred through spatial associations between species; however such analyses are subject to numerous pitfalls which can make inference unreliable (reviewed in Blanchet, Cazelles, \& Gravel 2020).\\
 A distinction of our study from others is that we modelled potential avoidance behaviours in a simple predator guild, where apex predator activity was artificially manipulated.
@@
 In the rare instances that they are considered, diel activity patterns are predominantly estimated at the population-level, overlooking finer-scale behaviours that can affect fitness, survival and ecosystem-impacts.
-Our results demonstrate the importance of (a) considering diel activity in regards to species interactions, (b) modelling \emph{changes} in animal behaviour rather than overlap with other species, and (c) testing avoidance behaviours within a joint spatiotemporal framework.
+Our results demonstrate the importance of (a) considering diel activity in regards to species interactions, (b) modelling changes in animal behaviour rather than overlap with other species, and (c) testing avoidance behaviours within a joint spatiotemporal framework.
 Our study adds to the limited body of evidence that top predators can produce a landscape of fear which is powerful enough to reverse the diel activity patterns of subordinate species (Kronfeld-Schor \& Dayan 2003).
@@
 (2010) Ecological and conservation implications of mesopredator release.
-\emph{Trophic cascades: predators, prey, and the changing dynamics of nature}, 221--240.
+\emph{Trophic cascades: Predators, prey, and the changing dynamics of nature} pp.
+221--240.
+Island Press.
 
@@
 \leavevmode\hypertarget{ref-daan1981adaptive}{}%
-Daan, S. (1981) Adaptive daily strategies in behavior. \emph{Biological rhythms} (ed J. Aschoff), pp. 275--298. Springer US, Boston, MA.
+Daan, S. (1981) Adaptive daily strategies in behavior. \emph{Biological rhythms} pp. 275--298. Springer Publishing.
 
@@
 
+\leavevmode\hypertarget{ref-harrington2009impact}{}%
+Harrington, L.A., Harrington, A.L., Yamaguchi, N., Thom, M.D., Ferreras, P., Windham, T.R.
+\& Macdonald, D.W.
+(2009) The impact of native competitors on an alien invasive: Temporal niche shifts to avoid interspecific aggression.
+\emph{Ecology}, \textbf{90}, 1207--1216.
+
 \leavevmode\hypertarget{ref-DHARMa}{}%
@@
 \leavevmode\hypertarget{ref-kronfeld2003partitioning}{}%
-Kronfeld-Schor, N. \& Dayan, T. (2003) Partitioning of time as an ecological resource. \emph{Annual review of ecology, evolution, and systematics}, \textbf{34}, 153--181.
+Kronfeld-Schor, N. \& Dayan, T. (2003) Partitioning of time as an ecological resource. \emph{Annual Review of Ecology, Evolution, and Systematics}, \textbf{34}, 153--181.
 
@@
 \leavevmode\hypertarget{ref-lima1990behavioral}{}%
-Lima, S.L. \& Dill, L.M. (1990) Behavioral decisions made under the risk of predation: A review and prospectus. \emph{Canadian journal of zoology}, \textbf{68}, 619--640.
+Lima, S.L. \& Dill, L.M. (1990) Behavioral decisions made under the risk of predation: A review and prospectus. \emph{Canadian Journal of Zoology}, \textbf{68}, 619--640.
 
@@
 \leavevmode\hypertarget{ref-miller2014finite}{}%
-Miller, D.L. \& Wood, S.N. (2014) Finite area smoothing with generalized distance splines. \emph{Environmental and ecological statistics}, \textbf{21}, 715--731.
+Miller, D.L. \& Wood, S.N. (2014) Finite area smoothing with generalized distance splines. \emph{Environmental and Ecological Statistics}, \textbf{21}, 715--731.
 
@@
 \leavevmode\hypertarget{ref-norbury2015pests}{}%
-Norbury, G. \& Jones, C. (2015) Pests controlling pests: Does predator control lead to greater european rabbit abundance in australasia? \emph{Mammal Review}, \textbf{45}, 79--87.
+Norbury, G. \& Jones, C. (2015) Pests controlling pests: Does predator control lead to greater European rabbit abundance in Australasia? \emph{Mammal Review}, \textbf{45}, 79--87.
 
@@
 \leavevmode\hypertarget{ref-woinarski2018reptiles}{}%
-Woinarski, J., Murphy, B., Palmer, R., Legge, S., Dickman, C., Doherty, T., Edwards, G., Nankivell, A., Read, J.
+Woinarski, J.C.Z., Murphy, B., Palmer, R., Legge, S., Dickman, C., Doherty, T., Edwards, G., Nankivell, A., Read, J.
 \& Stokeld, D.

### 2022-04-15 manuscript/diel-manuscript-supp.Rmd
submission1-JAE

--- 2022-04-14 manuscript/main_text.Rmd
@@
 keywords: "feral cat; diel activity patterns; generalised additive model; invasive predator;  intraguild predator interactions; mesopredator release; red fox; lethal predator control"
-journal: "?"
+journal: "Journal of Animal Ecology"
 date: "`r Sys.Date()`"
@@
 linenumbers: false
-csl: journal-of-applied-ecology.csl
+csl: journal-of-animal-ecology.csl
 geometry: margin=1.1in
@@
 
-*Article type*: Research Article
-
-*Running title*: Spatiotemporal predator activity
-
-*Word count (abstract, body)*: 330, 4338
-
-*References*: ?
-
-*Figures*: 5
-
-*Tables*: 0
-
-*Authorship*:
-
-*Data accessibility*:   Data and code will be deposited on the Dryad Digital Repository after acceptance and can be viewed here: [https://github.com/matt-w-rees/spatiotemporal-gams-invasive-predators](https://github.com/matt-w-rees/spatiotemporal-gams-invasive-predators).
-
 \newpage
@@
 
+1.
 Understanding the constraints that apex predators impose on subordinate species is important for anticipating the outcomes of predator management.
@@
 
-In this study, we tested whether an invasive apex predator (red fox *Vulpes vulpes*) suppresses or alters the spatiotemporal activity of an invasive mesopredator (feral cat *Felis catus*).
+2. In this study, we tested whether an invasive apex predator (red fox *Vulpes vulpes*) suppresses or alters the spatiotemporal activity of an invasive mesopredator (feral cat *Felis catus*).
+
+3.
 We surveyed these predators using 3667 camera-trap deployments across two regions of south-eastern Australia; foxes were poison-baited in some landscapes within each region.
@@
 
+4.
 When averaged across the study region, red foxes and feral cats had very similar diel activity patterns; however, there was important differentiation at a finer scale.
@@
 
+5.
 It is well-appreciated that overall predator activity varies spatially and fluctuates throughout the daily cycle.
 Our study demonstrates that diel activity patterns also vary across space, likely mediated by both landscape context and fear.
-Apex predator avoidance appears to be dynamic across landscapes of fear—a  key nuance which is overlooked when simply comparing the average activity overlap between two species or the spatial overlap of species occurrence.
+Apex predator avoidance appears to be dynamic across landscapes of fear&mdash;a key nuance which is overlooked when simply comparing the average activity overlap between two species or the spatial overlap of species occurrence.
 
@@
 As foxes and cats compete for many of the same resources, there is concern that lethal fox control could cause a mesopredator release [@soule1988reconstructed] of feral cats [@glen2005complex; @doherty2017stop; @wayne2017recoveries; @marlow2015cats; @comer2020integrating; @robley2014long].
-There is some evidence that feral cats increase in activity [although highly uncertain; @hunter2018not], density (Chapter \@ref(density)) and alter their diets and use of space [Chapter \@ref(density); @molsher2017mesopredator] in response to fox control.
+There is some evidence that feral cats increase in activity [although highly uncertain; @hunter2018not], density (M.W.R, unpublished data) and alter their diets and use of space [@molsher2017mesopredator] in response to fox control.
 Other studies have investigated potential spatial and temporal interactions between these invasive predators [e.g., @roshier2021space], but not in response to fox control, or in a joint spatiotemporal framework that allows flexibility in cat avoidance behaviours in respect to differences in fox activity.
@@
 
-I had field assistance from: Shauni Omond, Shayne Neal, Asitha Samarawickrama, Shelley Thompson, Erin Harris, Hannah Killian, Lani Watson, Mark Dorman, Jack Davis, Carl Roffey, Bruce Edley, Larissa Oliveira Gonçalves, Ben Lake, Chantelle Geissler, Aviya Naccarella, Emily Gronow, Harley England, David Pitts, Annie Hobby, Louise Falls, Thomas McKinnon, Jimmy Downie, Marney Hradsky, Stephanie Samson, Robin Sinclair, Asmaa Alhusainan, Kelly Forrester, Tammana Wadawani, Emily Gregg, Hannah Edwards, Adam Beck, Vishnu Memnon, Sandy Lu, Pia Lentini, Nick Golding, Emily McColl-Gausden, Nina Page, Maggie Campbell-Jones, Kyle Quinn and Jack Dickson.
+We acknowledge the Gunditjmara and Eastern Maar peoples on whose traditional lands this study took place.
+We thank Nick Golding, who inspired the use of GAMs in this paper and provided statistical advice.
+We are grateful to Alan Robley, the Department of Environment, Land, Water, and Planning (DELWP) Heywood team and Parks Victoria West Coast District Team for sharing camera-trapping datasets.
+The M.W.R PhD surveys were conducted under University of Melbourne Animal Ethics Committee approval 1714119.5 and Victorian Government DELWP Research Permit 10008273, with the assistance of Shauni Omond, Shayne Neal, Asitha Samarawickrama, Shelley Thompson, Erin Harris, Hannah Killian, Lani Watson, Mark Dorman, Jack Davis, Carl Roffey, Bruce Edley, Larissa Oliveira Gonçalves, Ben Lake, Chantelle Geissler, Aviya Naccarella, Emily Gronow, Harley England, David Pitts, Annie Hobby, Louise Falls, Thomas McKinnon, Jimmy Downie, Marney Hradsky, Stephanie Samson, Robin Sinclair, Asmaa Alhusainan, Kelly Forrester, Tammana Wadawani, Emily Gregg, Hannah Edwards, Adam Beck, Vishnu Memnon, Sandy Lu, Pia Lentini, Nick Golding, Emily McColl-Gausden, Nina Page, Maggie Campbell-Jones, Kyle Quinn and Jack Dickson.
+This study was generously supported by the Conservation Ecology Centre, the Victorian Government DELWP, Parks Victoria, Holsworth Wildlife Research Endowment – Equity Trustees Charitable Foundation and Ecological Society of Australia, Australian Research Council Linkage Project LP170101134, and the Australian Government’s National Environmental Science Program through the Threatened Species Recovery Hub.
+M.W.R also received funding from an Australian Government Research Training Program Scholarship and a Victorian Government DELWP top-up scholarship.
 
-Surveys were conducted under University of Melbourne Animal Ethics Committee approval 1714119 and Victorian Government Department of Environment, Land Water and Planning Research Permit 10008273.
+CONFLICT OF INTEREST {-}
+===================
+The authors declare no conflicts of interest.
+
+AUTHOR CONTRIBUTIONS {-}
+===================
+M.W.R conceived the ideas and designed the methodology; M.W.R, J.H.P, M.LP, E.K.B and B.A.H collected some of the data; M.W.R analysed the data with input from B.A.H and B.A.W, and led the writing of the manuscript.
+All authors contributed critically to the drafts and gave final approval for publication.
+
+DATA AVAILABILITY {-}
+===================
+Data and code will be deposited on the Dryad Digital Repository after acceptance and can be viewed here: [https://github.com/matt-w-rees/spatiotemporal-gams-invasive-predators](https://github.com/matt-w-rees/spatiotemporal-gams-invasive-predators).
 
@@
 \newpage
+\beginsupplement
+\setcounter{page}{1}
+\nolinenumbers
 
-\beginsupplement
+# SUPPORTING INFORMATION {-}
 
-# Supporting Information {-}
+*Spatial variation in predator diel activity patterns -- feral cats avoid red foxes in time, not space*
+
+MW Rees, BA Wintle, JH Pascoe, M Le Pla, EK Birnbaum & BA. Hradsky
 

### 2022-04-15 manuscript/diel-manuscript-supp.tex
submission1-JAE

--- 2022-04-14 manuscript/main_text.tex
@@
 
-  \journal{?} % Sets Journal name
+  \journal{Journal of Animal Ecology} % Sets Journal name
 
@@
    \ead{matt.wayne.rees@gmail.com} 
+    \author[UOM]{Brendan A. Wintle}
+
     \author[CEC]{Jack H. Pascoe}
@@
 
-    \author[ARI]{Alan Robley}
-
     \author[CEC]{Emma K. Birnbaum}
-
-    \author[UOM]{Brendan A. Wintle}
 
@@
     \address[CEC]{Conservation Ecology Centre, Otway Lighthouse Rd, Cape Otway, VIC, Australia}
-    \address[ARI]{Department of Environment, Land, Water and Planning, Arthur Rylah Institute for Environmental Research, Heidelberg, Australia}
       \cortext[1]{Corresponding Author}
@@
 
-\emph{Article type}: Research Article
-
-\emph{Running title}: Spatiotemporal predator activity
-
-\emph{Word count (abstract, body)}: 330, 4338
-
-\emph{References}: ?
-
-\emph{Figures}: 5
-
-\emph{Tables}: 0
-
-\emph{Authorship}:
-
-\emph{Data accessibility}: Data and code will be deposited on the Dryad Digital Repository after acceptance and can be viewed here: \url{https://github.com/matt-w-rees/spatiotemporal-gams-invasive-predators}.
-
 \newpage
@@
 
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\item
   Understanding the constraints that apex predators impose on subordinate species is important for anticipating the outcomes of predator management.
 Subordinate predators may avoid dominant predators in time or space, making it difficult to quantify changes in antipredator behaviours unless joint spatiotemporal analyses are used.
-
+\item
   In this study, we tested whether an invasive apex predator (red fox \emph{Vulpes vulpes}) suppresses or alters the spatiotemporal activity of an invasive mesopredator (feral cat \emph{Felis catus}).
+\item
   We surveyed these predators using 3667 camera-trap deployments across two regions of south-eastern Australia; foxes were poison-baited in some landscapes within each region.
@@
 We used generalised additive models to quantify overall predator activity across space and fluctuations in predator activity throughout the daily cycle (i.e., diel activity patterns).
-
+\item
   When averaged across the study region, red foxes and feral cats had very similar diel activity patterns; however, there was important differentiation at a finer scale.
@@
 Changes in cat diel activity patterns may facilitate spatial coexistence between the two invasive predators, potentially shifting impacts onto different native prey species.
-
+\item
   It is well-appreciated that overall predator activity varies spatially and fluctuates throughout the daily cycle.
@@
 Apex predator avoidance appears to be dynamic across landscapes of fear---a key nuance which is overlooked when simply comparing the average activity overlap between two species or the spatial overlap of species occurrence.
+\end{enumerate}
 
@@
 
-Predators shape ecosystems through both predation and the fear of predation (Creel \& Christianson 2008; Ritchie \& Johnson 2009).
-Fear-induced behavioural suppression can be as detrimental to subordinate species as predation itself (Schmitz, Krivan, \& Ovadia 2004; Preisser, Bolnick, \& Benard 2005).
-These non-consumptive effects of apex predator are expected to be strong drivers of mesopredator behaviour, particularly when resource competition is high (Ritchie \& Johnson 2009).
-A strengthening or relaxing of antipredator behaviours by mesopredators can have cascading effects across the entire ecosystem: altering population demographics, species interactions, ecological function and human-wildlife coexistence (Brown, Laundré, \& Gurung 1999; Ripple \& Beschta 2004; Estes \emph{et al.} 2011; Gaynor \emph{et al.} 2019; Lamb \emph{et al.} 2020).
-Hence, understanding how apex predators constrain the behaviour of subordinate species is important to accurately predict the ecosystem-wide consequences of predator management, such as reintroductions or lethal control (Gaynor \emph{et al.} 2021).
-
-Spatial and/or temporal niche partitioning may allow predators to coexist by reducing encounter-rates and resource overlap (Kronfeld-Schor \& Dayan 2003).
-However, mesopredators may not consistently employ avoidance behaviours because perceived predation risk is temporally and spatially variable, and antipredator behaviours typically involve a trade-off against resource acquisition, such as limiting or relegating activity to suboptimal places or times (Lima \& Dill 1990; Lima \& Bednekoff 1999).
-Therefore, optimal predator avoidance strategies are likely to vary across heterogeneous landscapes where resource availability (e.g., shelter, food) and perceived predation risks differ (Kauffman \emph{et al.} 2007; Willems \& Hill 2009; Wirsing \emph{et al.} 2021).
+Predators shape ecosystems through both predation and the fear of predation (Creel \& Christianson, 2008; Ritchie \& Johnson, 2009).
+Fear-induced behavioural suppression can be as detrimental to subordinate species as predation itself (Preisser et al., 2005; Schmitz et al., 2004).
+These non-consumptive effects of apex predator are expected to be strong drivers of mesopredator behaviour, particularly when resource competition is high (Ritchie \& Johnson, 2009).
+A strengthening or relaxing of antipredator behaviours by mesopredators can have cascading effects across the entire ecosystem: altering population demographics, species interactions, ecological function and human-wildlife coexistence (Brown et al., 1999; Estes et al., 2011; Gaynor et al., 2019; Lamb et al., 2020; Ripple \& Beschta, 2004).
+Hence, understanding how apex predators constrain the behaviour of subordinate species is important to accurately predict the ecosystem-wide consequences of predator management, such as reintroductions or lethal control (Gaynor et al., 2021).
+
+Spatial and/or temporal niche partitioning may allow predators to coexist by reducing encounter-rates and resource overlap (Kronfeld-Schor \& Dayan, 2003).
+However, mesopredators may not consistently employ avoidance behaviours because perceived predation risk is temporally and spatially variable, and antipredator behaviours typically involve a trade-off against resource acquisition, such as limiting or relegating activity to suboptimal places or times (Lima \& Bednekoff, 1999; Lima \& Dill, 1990).
+Therefore, optimal predator avoidance strategies are likely to vary across heterogeneous landscapes where resource availability (e.g., shelter, food) and perceived predation risks differ (Kauffman et al., 2007; Willems \& Hill, 2009; Wirsing et al., 2021).
 For example, temporal predator avoidance may be preferable over spatial avoidance if food is constantly available throughout the day, and vice versa.
-These concepts are unified under the `ecology of fear' (Brown, Laundré, \& Gurung 1999) concept, which has gained increasing attention in recent times (Gaynor \emph{et al.} 2019).
-Notably, the mesopredator release hypothesis has recently been expanded from increases in mesopredator abundance following apex predator decline (Soulé \emph{et al.} 1988) to also include changes in mesopredator behaviour (Brashares \emph{et al.} 2010).
+These concepts are unified under the `ecology of fear' (Brown et al., 1999) concept, which has gained increasing attention in recent times (Gaynor et al., 2019).
+Notably, the mesopredator release hypothesis has recently been expanded from increases in mesopredator abundance following apex predator decline (Soulé et al., 1988) to also include changes in mesopredator behaviour (Brashares et al., 2010).
 
@@
 In this paper, we use the term `overall activity' to refer to the number of `independent' predator detections at a site (offset to account for survey duration; analogous to an activity or abundance index), and `diel activity pattern' to refer to fluctuations in relative activity throughout the 24-hour daily cycle.
-Overall activity is influenced by predator behaviour, population density and the detection process (Anderson 2001), whereas diel activity patterns are a behavioural trait (less likely to be affected by the detection process, if the survey methodology remains consistent).
+Overall activity is influenced by predator behaviour, population density and the detection process (Anderson, 2001), whereas diel activity patterns are a behavioural trait (less likely to be affected by the detection process, if the survey methodology remains consistent).
 
 Despite modern predator survey technologies providing time-stamped detections, detection times are commonly discarded from analyses, probably because joint modelling of overall activity and diel activity patterns is more complicated.
-When temporal avoidance is tested, it is usually considered in an ad-hoc fashion, by fitting separate models for spatial and temporal avoidance, or by repeating spatial analyses (e.g., resource selection functions) at different time periods (e.g., Smith \emph{et al.} 2019; Basille \emph{et al.} 2015; Kohl \emph{et al.} 2019).
+When temporal avoidance is tested, it is usually considered in an ad-hoc fashion, by fitting separate models for spatial and temporal avoidance, or by repeating spatial analyses (e.g., resource selection functions) at different time periods (e.g., Smith et al., 2019; Basille et al., 2015; Kohl et al., 2019).
 However, discretising the daily cycle into categorical periods (e.g., day and night) introduces bias, assuming animals have complete step-changes in behaviour rather than progressive shifts across the daily cycle.
@@
 Generalised Additive Models (hereafter `GAMs') are increasingly being used to estimate animal diel activity patterns and offer a flexible framework to jointly consider overall activity.
-GAMs also have other benefits, including smoothing penalties to reduce overfitting, the ability to capture nonlinear interactions between multiple variables with different units, and the ability to share information across categorical variables through hierarchical specifications (Wood 2017; Pedersen \emph{et al.} 2019).
-However, we are only aware of one study which allowed animal diel activity to interact with predation risk as a continuous variable in a GAM (although without considering overall activity; Cunningham \emph{et al.} 2019).
-
-The red fox \emph{Vulpes vulpes} (hereafter `fox'; \textasciitilde6 kg) and feral cat \emph{Felis catus} (hereafter `cat'; \textasciitilde4 kg) have devastating impacts on native prey throughout their introduced range, implicated in the extinction of \textasciitilde10 and 63 species, respectively (Doherty \emph{et al.} 2016).
-The impacts of these invasive predators have been particularly extreme on the Australian continent (Woinarski, Burbidge, \& Harrison 2015).
-Cats are more difficult to manage, and so introduced predator control programs in Australia often target only foxes (particularly through poison-baiting; Reddiex \emph{et al.} 2007).
-As foxes and cats compete for many of the same resources, there is concern that lethal fox control could cause a mesopredator release (Soulé \emph{et al.} 1988) of feral cats (Glen \& Dickman 2005; Robley \emph{et al.} 2014; Marlow \emph{et al.} 2015; Doherty \& Ritchie 2017; Comer \emph{et al.} 2020; Wayne \emph{et al.} 2017).
-There is some evidence that feral cats increase in activity (although highly uncertain; Hunter \emph{et al.} 2018), density (Chapter \ref{density}) and alter their diets and use of space (Chapter \ref{density}; Molsher \emph{et al.} 2017) in response to fox control.
-Other studies have investigated potential spatial and temporal interactions between these invasive predators (e.g., Roshier \& Carter 2021), but not in response to fox control, or in a joint spatiotemporal framework that allows flexibility in cat avoidance behaviours in respect to differences in fox activity.
+GAMs also have other benefits, including smoothing penalties to reduce overfitting, the ability to capture nonlinear interactions between multiple variables with different units, and the ability to share information across categorical variables through hierarchical specifications (Pedersen et al., 2019; Wood, 2017).
+However, we are only aware of one study which allowed animal diel activity to interact with predation risk as a continuous variable in a GAM (although without considering overall activity; Cunningham et al., 2019).
+
+The red fox \emph{Vulpes vulpes} (hereafter `fox'; \textasciitilde6 kg) and feral cat \emph{Felis catus} (hereafter `cat'; \textasciitilde4 kg) have devastating impacts on native prey throughout their introduced range, implicated in the extinction of \textasciitilde10 and 63 species, respectively (Doherty et al., 2016).
+The impacts of these invasive predators have been particularly extreme on the Australian continent (Woinarski et al., 2015).
+Cats are more difficult to manage, and so introduced predator control programs in Australia often target only foxes (particularly through poison-baiting; Reddiex et al., 2007).
+As foxes and cats compete for many of the same resources, there is concern that lethal fox control could cause a mesopredator release (Soulé et al., 1988) of feral cats (Comer et al., 2020; Doherty \& Ritchie, 2017; Glen \& Dickman, 2005; Marlow et al., 2015; Robley et al., 2014; Wayne et al., 2017).
+There is some evidence that feral cats increase in activity (although highly uncertain; Hunter et al., 2018), density (M.W.R, unpublished data) and alter their diets and use of space (Molsher et al., 2017) in response to fox control.
+Other studies have investigated potential spatial and temporal interactions between these invasive predators (e.g., Roshier \& Carter, 2021), but not in response to fox control, or in a joint spatiotemporal framework that allows flexibility in cat avoidance behaviours in respect to differences in fox activity.
 
@@
 \ref{fig:diel-map}).
-Foxes in three distinct forest blocks in this region have been subject to poison-baiting since October 2005, with fortnightly bait replacements (Robley \emph{et al.} 2014).
-These forest blocks, along with three similar, unbaited forest blocks to the north have been simultaneously surveyed annually under the `Glenelg Ark' fox control program since 2005 (40 sites per block; Robley \emph{et al.} 2020).
-Hair-tubes were used to monitor species from 2005 - 2013 (presented in Robley \emph{et al.} 2014), replaced by camera-traps from 2013; here we present camera-trap data from 2013 - 2019 (Robley \emph{et al.} 2020).
+Foxes in three distinct forest blocks in this region have been subject to poison-baiting since October 2005, with fortnightly bait replacements (Robley et al., 2014).
+These forest blocks, along with three similar, unbaited forest blocks to the north have been simultaneously surveyed annually under the `Glenelg Ark' fox control program since 2005 (40 sites per block; Robley et al., 2020).
+Hair-tubes were used to monitor species from 2005 - 2013 (presented in Robley et al., 2014), replaced by camera-traps from 2013; here we present camera-trap data from 2013 - 2019 (Robley et al., 2020).
 We also included a further 425 camera-trap deployments at unique locations from early 2018 (M.W.R.
@@
 A matrix of cool temperate rainforest and wet forest at high altitudes in the south-west descend into a large heathland directly north, and into dry forests and then heathlands to the north-east.
-Fox-baiting commenced in small sections of the Otway Ranges in 2008 and large-scale systematic baiting began in 2016 - 2017 under the `Otway Ark' program (Robley, Moloney, \& Parks Victoria West Coast District Team 2019).
+Fox-baiting commenced in small sections of the Otway Ranges in 2008 and large-scale systematic baiting began in 2016 - 2017 under the `Otway Ark' program (Robley et al., 2019).
 For the first six weeks, poison-baits were replaced weekly, then changing to ongoing monthly bait-replacement.
@@
 Fox control recommenced in late 2018 with four weeks of fortnightly bait-replacement, before returning to monthly bait-replacement.
-A large section of the Otway Ranges to the north-west remains unbaited, but is monitored as an experimental non-treatment site (Robley, Moloney, \& Parks Victoria West Coast District Team 2019).
-Otway Ark managers survey 372 camera-trap sites annually (sequentially across the region); we present one `before' baiting survey and two `after' baiting surveys of each site from 2016 - 2018, totalling 1113 camera-trap deployments (Robley, Moloney, \& Parks Victoria West Coast District Team 2019).
+A large section of the Otway Ranges to the north-west remains unbaited, but is monitored as an experimental non-treatment site (Robley et al., 2019).
+Otway Ark managers survey 372 camera-trap sites annually (sequentially across the region); we present one `before' baiting survey and two `after' baiting surveys of each site from 2016 - 2018, totalling 1113 camera-trap deployments (Robley et al., 2019).
 We also include data from an additional before-after control-impact surveys (one `before' baiting survey and two `after' bating surveys) in the western section of the Otway Ranges, conducted annually 2017 - 2019 (M.W.R PhD surveys).
@@
 All camera-trap deployments consisted of a Reconyx (Holmen, Wisconsin) brand camera-trap (white or infrared flash), attached to a tree or a metal picket, facing a lure.
-The Glenelg Ark and Otway Ark fox monitoring programs positioned camera-traps at least 40 cm above ground on a tree or a metal picket and angled downwards toward a lure approximately 1 - 1.5 m away (Robley, Moloney, \& Parks Victoria West Coast District Team 2019; Robley \emph{et al.} 2020).
+The Glenelg Ark and Otway Ark fox monitoring programs positioned camera-traps at least 40 cm above ground on a tree or a metal picket and angled downwards toward a lure approximately 1 - 1.5 m away (Robley et al., 2019, 2020).
 The lures consisted of peanut butter, golden syrup and rolled oats mixed into a small ball, placed within a tea strainer or PVC pipe container and secured either to the ground, or 20 - 60 cm above ground on a wooden stake.
 The M.W.R.
-PhD surveys across both regions positioned camera-traps lower on a tree (around 15 - 30 cm above the ground) angled only slightly downwards toward a tuna oil lure approximately 2 - 2.5 m away (detailed in Rees \emph{et al.} 2019).
+PhD surveys across both regions positioned camera-traps lower on a tree (around 15 - 30 cm above the ground) angled only slightly downwards toward a tuna oil lure approximately 2 - 2.5 m away (detailed in Rees et al., 2019).
 Camera-traps were active for an average of 47 days (maximum 93 days), totalling 172,052 trap-nights.
@@
 
-All data analysis was conducted in R version 3.6.3 (R Core Team 2020).
-We first used lorelograms to identify the minimum interval to approximate independence (Iannarilli \emph{et al.} 2019); this indicated that discarding repeat detections of a species within 30 minutes was sufficient to reduce temporal autocorrelation.
-To account for day length variation across space and time, we extracted sunrise and sunset times for each camera-trap deployment using the `maptools' R-package (Bivand \& Lewin-Koh 2021) and adjusted detection times to be relative to sunrise and sunset using the average double anchoring approach described by Vazquez \emph{et al.} (2019).
+All data analysis was conducted in R version 3.6.3 (R Core Team, 2020).
+We first used lorelograms to identify the minimum interval to approximate independence (Iannarilli et al., 2019); this indicated that discarding repeat detections of a species within 30 minutes was sufficient to reduce temporal autocorrelation.
+To account for day length variation across space and time, we extracted sunrise and sunset times for each camera-trap deployment using the `maptools' R-package (Bivand \& Lewin-Koh, 2021) and adjusted detection times to be relative to sunrise and sunset using the average double anchoring approach described by Vazquez et al.
+(2019).
 We then built a dataframe consisting of a row for each hour of the day (0 -- 23), for every camera-trap deployment (n = 3667), recording the total number of `independent' fox and feral cat detections within each hour across the camera-trap survey.
@@
 
-We modelled the total number of independent detections of each predator per hour for each camera-trap deployment (response variable) with generalised additive mixed-effect models implemented in the `mgcv' R-package (Wood 2017).
-We used the negative binomial family, as overdispersion, but not zero-inflation, was detected with a poisson distribution using the `DHARMa' R-package (Hartig 2020).
+We modelled the total number of independent detections of each predator per hour for each camera-trap deployment (response variable) with generalised additive mixed-effect models implemented in the `mgcv' R-package (Wood, 2017).
+We used the negative binomial family, as overdispersion, but not zero-inflation, was detected with a poisson distribution using the `DHARMa' R-package (Hartig, 2020).
 We specified the natural log of the number of survey days as a model offset to account for differences in camera-trap survey duration, and a random intercept for each site to account for repeat sampling.
@@
 This allowed predators to have different activity levels across space (static across the years surveyed), as well as variation in diel activity pattern across space.
-Space was modelled using camera-trap coordinates and a duchon spline basis (Miller \& Wood 2014).
+Space was modelled using camera-trap coordinates and a duchon spline basis (Miller \& Wood, 2014).
 To examine how the relative strength of diel activity patterns changed across space, we plotted the percentage increase from the minimum to maximum activity estimate within the daily cycle for each predicted location (hereafter referred to as `diel activity pattern strength').
@@
 
-Predator activity varied across space; we hypothesised that this was partly due to differences in vegetation type, based both on the observed spatial patterns and because vegetation type is a major driver of understorey habitat structure and prey occurrence in these regions (Swan \emph{et al.} 2015; Hradsky \emph{et al.} 2017).
-To test whether the diel activity pattern of each predator varied among vegetation types, we identified the Ecological Vegetation Class group (hereafter `vegetation type'; standard units for vegetation classification in Victoria; Department of Environment, Land, Water \& Planning 2020) for each unique camera-trap site, totalling eight vegetation types.
+Predator activity varied across space; we hypothesised that this was partly due to differences in vegetation type, based both on the observed spatial patterns and because vegetation type is a major driver of understorey habitat structure and prey occurrence in these regions (Hradsky et al., 2017; Swan et al., 2015).
+To test whether the diel activity pattern of each predator varied among vegetation types, we identified the Ecological Vegetation Class group (hereafter ``vegetation type''; standard units for vegetation classification in Victoria; Department of Environment, Land, Water \& Planning, 2020) for each unique camera-trap site, totalling eight vegetation types.
 As rainforests are interspersed (primarily in low lying gullies) at fine-scales throughout wet and damp forests in the south-eastern Otway Ranges, we merged them together (hereafter referred to as `wet forests').
-We then estimated predator activity across vegetation types using a hierarchical model specification: a global smoother for hour (i.e., average response) and group-level smoothers with shared wiggliness for the seven vegetation types (`model GS' detailed in Pedersen \emph{et al.} 2019).
+We then estimated predator activity across vegetation types using a hierarchical model specification: a global smoother for hour (i.e., average response) and group-level smoothers with shared wiggliness for the seven vegetation types (``model GS'' detailed in Pedersen et al., 2019).
 We also included a random effect to account for differences in overall activity levels between the two regions.
@@
 
-A key question in ecological theory is whether animals are evolutionary hardwired to occupy particular temporal niches, or have circadian rhythms that are responsive to changing environmental conditions and interactions with other species (Schoener 1974; Daan 1981; Lima \& Dill 1990).
+A key question in ecological theory is whether animals are evolutionary hardwired to occupy particular temporal niches, or have circadian rhythms that are responsive to changing environmental conditions and interactions with other species (Daan, 1981; Lima \& Dill, 1990; Schoener, 1974).
 Here we demonstrate that diel activity patterns are not fixed, but vary across space based on landscape context and fear.
 In our study, sympatric invasive predators had similar diel activity patterns when averaged across broad regions (i.e., high circular overlap; Fig.
-\ref{fig:diel-space-time-marginal}; as did Roshier \& Carter 2021), but behaviours varied considerably within landscapes.
+\ref{fig:diel-space-time-marginal}; as did Roshier \& Carter, 2021), but behaviours varied considerably within landscapes.
 Fox daily activity patterns were most strongly tied to the daily cycle in dry habitat types, but showed little diel activity pattern in the wet forests (Fig.
@@
 Control programs that reduce invasive fox activity are therefore likely to change cat diel activity patterns, which in turn may alter cat impacts on native prey species.
-Quantifying changes in diel activity patterns provides important context for understanding species interactions, which is key for effective ecosystem management (Gaynor \emph{et al.} 2021).
-
-Shifting diel activity patterns may facilitate spatial coexistence of dominant and subordinate species (Carothers \& Jaksić 1984).
+Quantifying changes in diel activity patterns provides important context for understanding species interactions, which is key for effective ecosystem management (Gaynor et al., 2021).
+
+Shifting diel activity patterns may facilitate spatial coexistence of dominant and subordinate species (Carothers \& Jaksić, 1984).
 For cats, altering diel activity patterns to less-preferred times of day may be worthwhile to persist in high-quality habitat.
-Few studies have demonstrated predator-induced shifts in diel activity (Kronfeld-Schor \& Dayan 2003), but notably, ship rats \emph{Rattus norvegicus} were also found to switch from nocturnal to diurnal behaviour in response to fox activity (Fenn \& Macdonald 1995) and a similar nocturnal-diurnal shift was observed in American mink \emph{Neovison vison} following the recolonisation of native predators (Harrington \emph{et al.} 2009).
-For cats in our study, a switch to diurnal behaviour where fox activity was high in the dry vegetation types may have been facilitated by the higher abundance of reptiles in these habitat types relative to wet forests (which are mostly diurnal; Woinarski \emph{et al.} 2018).
+Few studies have demonstrated predator-induced shifts in diel activity (Kronfeld-Schor \& Dayan, 2003), but notably, ship rats \emph{Rattus norvegicus} were also found to switch from nocturnal to diurnal behaviour in response to fox activity (Fenn \& Macdonald, 1995) and a similar nocturnal-diurnal shift was observed in American mink \emph{Neovison vison} following the recolonisation of native predators (Harrington et al., 2009).
+For cats in our study, a switch to diurnal behaviour where fox activity was high in the dry vegetation types may have been facilitated by the higher abundance of reptiles in these habitat types relative to wet forests (which are mostly diurnal; Woinarski et al., 2018).
 At wet forest sites with high fox activity, cats concentrated their activity away from sunrise and sunset towards midnight, despite a diurnal shift appearing to similarly reduce the risk of a fox encounter.
@@
 \ref{fig:diel-cat-fox}b:c).
-Overlap in spatial activity at fine scales (i.e., at the level of camera-trap sites) may have been facilitated by temporal avoidance (Kronfeld-Schor \& Dayan 2003).
+Overlap in spatial activity at fine scales (i.e., at the level of camera-trap sites) may have been facilitated by temporal avoidance (Kronfeld-Schor \& Dayan, 2003).
 However, we considered spatial avoidance across the entire survey duration (averaging 47 days).
@@
 Short-term spatial avoidance is quite plausible given foxes mark territories using scats and odours, which cats could tangibly associate with high risk shortly after.
-Temporary spatial avoidance could be tested using decay curves (e.g., Niedballa \emph{et al.} 2019).
+Temporary spatial avoidance could be tested using decay curves (e.g., Niedballa et al., 2019).
 Alternatively, no sign of spatial avoidance between these invasive predators could also be an artefact of the quality of camera set-up and hence detectability.
@@
 Camera-traps that angled only slightly downwards, rather than upwards or strongly downwards, seemed most effective at detecting both predator species (M.W Rees, personal observation).
-Predator interactions are routinely inferred through spatial associations between species; however such analyses are subject to numerous pitfalls which can make inference unreliable (reviewed in Blanchet, Cazelles, \& Gravel 2020).\\
+Predator interactions are routinely inferred through spatial associations between species; however such analyses are subject to numerous pitfalls which can make inference unreliable (reviewed in Blanchet et al., 2020).
+
 A distinction of our study from others is that we modelled potential avoidance behaviours in a simple predator guild, where apex predator activity was artificially manipulated.
@@
 For example, low fox activity may promote the availability of a preferred shared prey species with a diel pattern which differs from those of cats on average, and as a result, cats might shift diel activity patterns at sites with low fox activity to more closely match those of the more abundant prey species.
-We would expect introduced European rabbits \emph{Oryctolagus cuniculus} and hares \emph{Lepus europaeus} (which are diurnal) to be particularly likely to induce such a response in dry vegetation types (McGregor \emph{et al.} 2020; Stobo-Wilson \emph{et al.} 2020), however, rabbits and hares are rare within the natural vegetation of the landscapes we surveyed (only ever being detected at 40 of 1232 sites) and there is little evidence that predation by foxes suppresses rabbit populations (Norbury \& Jones 2015; Scroggie \emph{et al.} 2018).
+We would expect introduced European rabbits \emph{Oryctolagus cuniculus} and hares \emph{Lepus europaeus} (which are diurnal) to be particularly likely to induce such a response in dry vegetation types (McGregor et al., 2020; Stobo-Wilson et al., 2020), however, rabbits and hares are rare within the natural vegetation of the landscapes we surveyed (only ever being detected at 40 of 1232 sites) and there is little evidence that predation by foxes suppresses rabbit populations (Norbury \& Jones, 2015; Scroggie et al., 2018).
 While of particular interest, whether temporal fox-cat interactions are direct or indirect does not change the resulting impact on native prey, and hence the outcomes of fox management.
 
-Flexible antipredator behaviours make evolutionary sense, but have been rarely demonstrated in terms of spatiotemporal predator avoidance (although, see Relyea 2003; Brown \emph{et al.} 2013; Cunningham \emph{et al.} 2019), because this often requires manipulative experiments or at least more complicated models (Kronfeld-Schor \& Dayan 2003).
+Flexible antipredator behaviours make evolutionary sense, but have been rarely demonstrated in terms of spatiotemporal predator avoidance (although, see Relyea, 2003; Brown et al., 2013; Cunningham et al., 2019), because this often requires manipulative experiments or at least more complicated models (Kronfeld-Schor \& Dayan, 2003).
 Our study demonstrates that GAMs offer a powerful tool for modelling continuous shifts in animal activity across both space and time, capable of capturing complex interactions and sharing information across categorical variables.
-The inbuilt smoothing penalties are another benefit of GAMs over kernel density estimation (Ridout \& Linkie 2009), in which noisy data can produce spurious estimates (Frey \emph{et al.} 2017; Iannarilli \emph{et al.} 2019).
-The alternative approach of simply comparing average diel activity overlap between two species (Ridout \& Linkie 2009) would have been misleading for two reasons.
+The inbuilt smoothing penalties are another benefit of GAMs over kernel density estimation (Ridout \& Linkie, 2009), in which noisy data can produce spurious estimates (Frey et al., 2017; Iannarilli et al., 2019).
+The alternative approach of simply comparing average diel activity overlap between two species (Ridout \& Linkie, 2009) would have been misleading for two reasons.
 Firstly, predator diel activity patterns varied `naturally' across heterogeneous landscapes (requiring avoidance to be tested in wet forests and dry vegetation types separately; Fig.
@@
 Our results demonstrate the importance of (a) considering diel activity in regards to species interactions, (b) modelling changes in animal behaviour rather than overlap with other species, and (c) testing avoidance behaviours within a joint spatiotemporal framework.
-Our study adds to the limited body of evidence that top predators can produce a landscape of fear which is powerful enough to reverse the diel activity patterns of subordinate species (Kronfeld-Schor \& Dayan 2003).
+Our study adds to the limited body of evidence that top predators can produce a landscape of fear which is powerful enough to reverse the diel activity patterns of subordinate species (Kronfeld-Schor \& Dayan, 2003).
 
@@
 
+We acknowledge the Gunditjmara and Eastern Maar peoples on whose traditional lands this study took place.
+We thank Nick Golding, who inspired the use of GAMs in this paper and provided statistical advice.
+We are grateful to Alan Robley, the Department of Environment, Land, Water, and Planning (DELWP) Heywood team and Parks Victoria West Coast District Team for sharing camera-trapping datasets.
+The M.W.R PhD surveys were conducted under University of Melbourne Animal Ethics Committee approval 1714119.5 and Victorian Government DELWP Research Permit 10008273, with the assistance of Shauni Omond, Shayne Neal, Asitha Samarawickrama, Shelley Thompson, Erin Harris, Hannah Killian, Lani Watson, Mark Dorman, Jack Davis, Carl Roffey, Bruce Edley, Larissa Oliveira Gonçalves, Ben Lake, Chantelle Geissler, Aviya Naccarella, Emily Gronow, Harley England, David Pitts, Annie Hobby, Louise Falls, Thomas McKinnon, Jimmy Downie, Marney Hradsky, Stephanie Samson, Robin Sinclair, Asmaa Alhusainan, Kelly Forrester, Tammana Wadawani, Emily Gregg, Hannah Edwards, Adam Beck, Vishnu Memnon, Sandy Lu, Pia Lentini, Nick Golding, Emily McColl-Gausden, Nina Page, Maggie Campbell-Jones, Kyle Quinn and Jack Dickson.
+This study was generously supported by the Conservation Ecology Centre, the Victorian Government DELWP, Parks Victoria, Holsworth Wildlife Research Endowment -- Equity Trustees Charitable Foundation and Ecological Society of Australia, Australian Research Council Linkage Project LP170101134, and the Australian Government's National Environmental Science Program through the Threatened Species Recovery Hub.
+M.W.R also received funding from an Australian Government Research Training Program Scholarship and a Victorian Government DELWP top-up scholarship.
+
+\hypertarget{conflict-of-interest}{%
+\section*{CONFLICT OF INTEREST}\label{conflict-of-interest}}
+\addcontentsline{toc}{section}{CONFLICT OF INTEREST}
+
+The authors declare no conflicts of interest.
+
+\hypertarget{author-contributions}{%
+\section*{AUTHOR CONTRIBUTIONS}\label{author-contributions}}
+\addcontentsline{toc}{section}{AUTHOR CONTRIBUTIONS}
+
+M.W.R conceived the ideas and designed the methodology; M.W.R, J.H.P, M.LP, E.K.B and B.A.H collected some of the data; M.W.R analysed the data with input from B.A.H and B.A.W, and led the writing of the manuscript.
+All authors contributed critically to the drafts and gave final approval for publication.
+
+\hypertarget{data-availability}{%
+\section*{DATA AVAILABILITY}\label{data-availability}}
+\addcontentsline{toc}{section}{DATA AVAILABILITY}
+
+Data and code will be deposited on the Dryad Digital Repository after acceptance and can be viewed here: \url{https://github.com/matt-w-rees/spatiotemporal-gams-invasive-predators}.
+
 \newpage
@@
 \leavevmode\hypertarget{ref-anderson2001need}{}%
-Anderson, D.R. (2001) The need to get the basics right in wildlife field studies. \emph{Wildlife Society Bulletin}, 1294--1297.
+Anderson, D. R. (2001). The need to get the basics right in wildlife field studies. \emph{Wildlife Society Bulletin}, 1294--1297. \url{https://doi.org/https://doi.org/10.2307/3784156}
 
 \leavevmode\hypertarget{ref-basille2015plastic}{}%
-Basille, M., Fortin, D., Dussault, C., Bastille-Rousseau, G., Ouellet, J.-P.
-\& Courtois, R.
-(2015) Plastic response of fearful prey to the spatiotemporal dynamics of predator distribution.
-\emph{Ecology}, \textbf{96}, 2622--2631.
+Basille, M., Fortin, D., Dussault, C., Bastille-Rousseau, G., Ouellet, J.-P., \& Courtois, R.
+(2015).
+Plastic response of fearful prey to the spatiotemporal dynamics of predator distribution.
+\emph{Ecology}, \emph{96}(10), 2622--2631.
+\url{https://doi.org/https://doi.org/10.1890/14-1706.1}
 
 \leavevmode\hypertarget{ref-maptools}{}%
-Bivand, R. \& Lewin-Koh, N. (2021) \emph{Maptools: Tools for Handling Spatial Objects}.
+Bivand, R., \& Lewin-Koh, N. (2021). \emph{Maptools: Tools for handling spatial objects}. \url{https://CRAN.R-project.org/package=maptools}
 
 \leavevmode\hypertarget{ref-guillaume2020co}{}%
-Blanchet, F.G., Cazelles, K. \& Gravel, D. (2020) Co-occurrence is not evidence of ecological interactions. \emph{Ecology Letters}, \textbf{23}, 1050--1063.
+Blanchet, F.
+G., Cazelles, K., \& Gravel, D.
+(2020).
+Co-occurrence is not evidence of ecological interactions.
+\emph{Ecology Letters}, \emph{23}(7), 1050--1063.
+\url{https://doi.org/https://doi.org/10.1111/ele.13525}
 
 \leavevmode\hypertarget{ref-brashares2010ecological}{}%
-Brashares, J.S., Prugh, L.R., Stoner, C.J.
-\& Epps, C.W.
-(2010) Ecological and conservation implications of mesopredator release.
-\emph{Trophic cascades: Predators, prey, and the changing dynamics of nature} pp.
-221--240.
+Brashares, J.
+S., Prugh, L.
+R., Stoner, C.
+J., \& Epps, C.
+W.
+(2010).
+Ecological and conservation implications of mesopredator release.
+In \emph{Trophic cascades: Predators, prey, and the changing dynamics of nature} (pp.
+221--240).
 Island Press.
@@
 \leavevmode\hypertarget{ref-brown2013phenotypically}{}%
-Brown, G.E., Ferrari, M.C., Elvidge, C.K., Ramnarine, I.
-\& Chivers, D.P.
-(2013) Phenotypically plastic neophobia: A response to variable predation risk.
-\emph{Proceedings of the Royal Society B: Biological Sciences}, \textbf{280}, 20122712.
+Brown, G.
+E., Ferrari, M.
+C., Elvidge, C.
+K., Ramnarine, I., \& Chivers, D.
+P.
+(2013).
+Phenotypically plastic neophobia: A response to variable predation risk.
+\emph{Proceedings of the Royal Society B: Biological Sciences}, \emph{280}(1756), 20122712.
+\url{https://doi.org/https://doi.org/10.1098/rspb.2012.2712}
 
 \leavevmode\hypertarget{ref-brown1999ecology}{}%
-Brown, J.S., Laundré, J.W. \& Gurung, M. (1999) The ecology of fear: Optimal foraging, game theory, and trophic interactions. \emph{Journal of Mammalogy}, \textbf{80}, 385--399.
+Brown, J.
+S., Laundré, J.
+W., \& Gurung, M.
+(1999).
+The ecology of fear: Optimal foraging, game theory, and trophic interactions.
+\emph{Journal of Mammalogy}, \emph{80}(2), 385--399.
+\url{https://doi.org/https://doi.org/10.2307/1383287}
 
 \leavevmode\hypertarget{ref-carothers1984time}{}%
-Carothers, J.H. \& Jaksić, F.M. (1984) Time as a niche difference: The role of interference competition. \emph{Oikos}, 403--406.
+Carothers, J. H., \& Jaksić, F. M. (1984). Time as a niche difference: The role of interference competition. \emph{Oikos}, 403--406. \url{https://doi.org/https://doi.org/10.2307/3544413}
 
 \leavevmode\hypertarget{ref-comer2020integrating}{}%
-Comer, S., Clausen, L., Cowen, S., Pinder, J., Thomas, A., Burbidge, A.H., Tiller, C., Algar, D.
-\& Speldewinde, P.
-(2020) Integrating feral cat \emph{(Felis catus)} control into landscape-scale introduced predator management to improve conservation prospects for threatened fauna: A case study from the south coast of Western Australia.
-\emph{Wildlife Research}, \textbf{47}, 762--778.
+Comer, S., Clausen, L., Cowen, S., Pinder, J., Thomas, A., Burbidge, A.
+H., Tiller, C., Algar, D., \& Speldewinde, P.
+(2020).
+Integrating feral cat \emph{(Felis catus)} control into landscape-scale introduced predator management to improve conservation prospects for threatened fauna: A case study from the south coast of Western Australia.
+\emph{Wildlife Research}, \emph{47}(8), 762--778.
+\url{https://doi.org/https://doi.org/10.1071/WR19217}
 
 \leavevmode\hypertarget{ref-creel2008relationships}{}%
-Creel, S. \& Christianson, D. (2008) Relationships between direct predation and risk effects. \emph{Trends in Ecology \& Evolution}, \textbf{23}, 194--201.
+Creel, S., \& Christianson, D.
+(2008).
+Relationships between direct predation and risk effects.
+\emph{Trends in Ecology \& Evolution}, \emph{23}(4), 194--201.
+\url{https://doi.org/https://doi.org/10.1016/j.tree.2007.12.004}
 
 \leavevmode\hypertarget{ref-cunningham2019temporal}{}%
-Cunningham, C.X., Scoleri, V., Johnson, C.N., Barmuta, L.A.
-\& Jones, M.E.
-(2019) Temporal partitioning of activity: Rising and falling top-predator abundance triggers community-wide shifts in diel activity.
-\emph{Ecography}, \textbf{42}, 2157--2168.
+Cunningham, C.
+X., Scoleri, V., Johnson, C.
+N., Barmuta, L.
+A., \& Jones, M.
+E.
+(2019).
+Temporal partitioning of activity: Rising and falling top-predator abundance triggers community-wide shifts in diel activity.
+\emph{Ecography}, \emph{42}(12), 2157--2168.
+\url{https://doi.org/https://doi.org/10.1111/ecog.04485}
 
 \leavevmode\hypertarget{ref-daan1981adaptive}{}%
-Daan, S. (1981) Adaptive daily strategies in behavior. \emph{Biological rhythms} pp. 275--298. Springer Publishing.
+Daan, S. (1981). Adaptive daily strategies in behavior. In \emph{Biological rhythms} (pp. 275--298). Springer Publishing. \url{https://doi.org/10.1007/978-1-4615-6552-9_15}
 
 \leavevmode\hypertarget{ref-delwp2020bioregions}{}%
-Department of Environment, Land, Water \& Planning. (2020) Bioregions and EVC Benchmarks, \url{https://www.environment.vic.gov.au/biodiversity/bioregions-and-evc-benchmarks}
+Department of Environment, Land, Water \& Planning.
+(2020).
+\emph{Bioregions and EVC Benchmarks}.
+Victorian Government Department of Environment, Land, Water; Planning, Melbourne.
+Accessed June 2021).
+\url{https://www.environment.vic.gov.au/biodiversity/bioregions-and-evc-benchmarks}
 
 \leavevmode\hypertarget{ref-doherty2016invasive}{}%
-Doherty, T.S., Glen, A.S., Nimmo, D.G., Ritchie, E.G.
-\& Dickman, C.R.
-(2016) Invasive predators and global biodiversity loss.
-\emph{Proceedings of the National Academy of Sciences}, \textbf{113}, 11261--11265.
+Doherty, T.
+S., Glen, A.
+S., Nimmo, D.
+G., Ritchie, E.
+G., \& Dickman, C.
+R.
+(2016).
+Invasive predators and global biodiversity loss.
+\emph{Proceedings of the National Academy of Sciences}, \emph{113}(40), 11261--11265.
+\url{https://doi.org/10.1073/pnas.1602480113}
 
 \leavevmode\hypertarget{ref-doherty2017stop}{}%
-Doherty, T.S. \& Ritchie, E.G. (2017) Stop jumping the gun: A call for evidence-based invasive predator management. \emph{Conservation Letters}, \textbf{10}, 15--22.
+Doherty, T.
+S., \& Ritchie, E.
+G.
+(2017).
+Stop jumping the gun: A call for evidence-based invasive predator management.
+\emph{Conservation Letters}, \emph{10}(1), 15--22.
+\url{https://doi.org/https://doi.org/10.1111/conl.12251}
 
 \leavevmode\hypertarget{ref-estes2011trophic}{}%
-Estes, J.A., Terborgh, J., Brashares, J.S., Power, M.E., Berger, J., Bond, W.J., Carpenter, S.R., Essington, T.E., Holt, R.D., Jackson, J.B.
-\& others.
-(2011) Trophic downgrading of planet earth.
-\emph{Science}, \textbf{333}, 301--306.
+Estes, J.
+A., Terborgh, J., Brashares, J.
+S., Power, M.
+E., Berger, J., Bond, W.
+J., Carpenter, S.
+R., Essington, T.
+E., Holt, R.
+D., Jackson, J.
+B., \& others.
+(2011).
+Trophic downgrading of planet earth.
+\emph{Science}, \emph{333}(6040), 301--306.
+\url{https://doi.org/https://doi.org/10.1126/science.1205106}
 
 \leavevmode\hypertarget{ref-fenn1995use}{}%
-Fenn, M.G. \& Macdonald, D.W. (1995) Use of middens by red foxes: Risk reverses rhythms of rats. \emph{Journal of Mammalogy}, \textbf{76}, 130--136.
+Fenn, M.
+G., \& Macdonald, D.
+W.
+(1995).
+Use of middens by red foxes: Risk reverses rhythms of rats.
+\emph{Journal of Mammalogy}, \emph{76}(1), 130--136.
+\url{https://doi.org/https://doi.org/10.2307/1382321}
 
 \leavevmode\hypertarget{ref-frey2017investigating}{}%
-Frey, S., Fisher, J.T., Burton, A.C.
-\& Volpe, J.P.
-(2017) Investigating animal activity patterns and temporal niche partitioning using camera-trap data: Challenges and opportunities.
-\emph{Remote Sensing in Ecology and Conservation}, \textbf{3}, 123--132.
+Frey, S., Fisher, J.
+T., Burton, A.
+C., \& Volpe, J.
+P.
+(2017).
+Investigating animal activity patterns and temporal niche partitioning using camera-trap data: Challenges and opportunities.
+\emph{Remote Sensing in Ecology and Conservation}, \emph{3}(3), 123--132.
+\url{https://doi.org/https://doi.org/10.1002/rse2.60}
 
 \leavevmode\hypertarget{ref-gaynor2019landscapes}{}%
-Gaynor, K.M., Brown, J.S., Middleton, A.D., Power, M.E.
-\& Brashares, J.S.
-(2019) Landscapes of fear: Spatial patterns of risk perception and response.
-\emph{Trends in Ecology \& Evolution}, \textbf{34}, 355--368.
+Gaynor, K.
+M., Brown, J.
+S., Middleton, A.
+D., Power, M.
+E., \& Brashares, J.
+S.
+(2019).
+Landscapes of fear: Spatial patterns of risk perception and response.
+\emph{Trends in Ecology \& Evolution}, \emph{34}(4), 355--368.
+\url{https://doi.org/https://doi.org/10.1016/j.tree.2019.01.004}
 
 \leavevmode\hypertarget{ref-gaynor2021applied}{}%
-Gaynor, K.M., Cherry, M.J., Gilbert, S.L., Kohl, M.T., Larson, C.L., Newsome, T.M., Prugh, L.R., Suraci, J.P., Young, J.K.
-\& Smith, J.A.
-(2021) An applied ecology of fear framework: Linking theory to conservation practice.
-\emph{Animal Conservation}, \textbf{24}, 308--321.
+Gaynor, K.
+M., Cherry, M.
+J., Gilbert, S.
+L., Kohl, M.
+T., Larson, C.
+L., Newsome, T.
+M., Prugh, L.
+R., Suraci, J.
+P., Young, J.
+K., \& Smith, J.
+A.
+(2021).
+An applied ecology of fear framework: Linking theory to conservation practice.
+\emph{Animal Conservation}, \emph{24}(3), 308--321.
+\url{https://doi.org/https://doi.org/10.1111/acv.12629}
 
 \leavevmode\hypertarget{ref-glen2005complex}{}%
-Glen, A.S. \& Dickman, C.R. (2005) Complex interactions among mammalian carnivores in Australia, and their implications for wildlife management. \emph{Biological Reviews}, \textbf{80}, 387--401.
+Glen, A.
+S., \& Dickman, C.
+R.
+(2005).
+Complex interactions among mammalian carnivores in Australia, and their implications for wildlife management.
+\emph{Biological Reviews}, \emph{80}(3), 387--401.
+\url{https://doi.org/https://doi.org/10.1017/S1464793105006718}
 
 \leavevmode\hypertarget{ref-harrington2009impact}{}%
-Harrington, L.A., Harrington, A.L., Yamaguchi, N., Thom, M.D., Ferreras, P., Windham, T.R.
-\& Macdonald, D.W.
-(2009) The impact of native competitors on an alien invasive: Temporal niche shifts to avoid interspecific aggression.
-\emph{Ecology}, \textbf{90}, 1207--1216.
+Harrington, L.
+A., Harrington, A.
+L., Yamaguchi, N., Thom, M.
+D., Ferreras, P., Windham, T.
+R., \& Macdonald, D.
+W.
+(2009).
+The impact of native competitors on an alien invasive: Temporal niche shifts to avoid interspecific aggression.
+\emph{Ecology}, \emph{90}(5), 1207--1216.
+\url{https://doi.org/https://doi.org/10.1890/08-0302.1}
 
 \leavevmode\hypertarget{ref-DHARMa}{}%
-Hartig, F. (2020) \emph{DHARMa: Residual Diagnostics for Hierarchical (Multi-Level / Mixed) Regression Models}.
+Hartig, F. (2020). \emph{DHARMa: Residual diagnostics for hierarchical (multi-level / mixed) regression models}. \url{https://CRAN.R-project.org/package=DHARMa}
 
 \leavevmode\hypertarget{ref-hradsky2017bayesian}{}%
-Hradsky, B.A., Penman, T.D., Ababei, D., Hanea, A., Ritchie, E.G., York, A.
-\& Di Stefano, J.
-(2017) Bayesian networks elucidate interactions between fire and other drivers of terrestrial fauna distributions.
-\emph{Ecosphere}, \textbf{8}, e01926.
+Hradsky, B.
+A., Penman, T.
+D., Ababei, D., Hanea, A., Ritchie, E.
+G., York, A., \& Di Stefano, J.
+(2017).
+Bayesian networks elucidate interactions between fire and other drivers of terrestrial fauna distributions.
+\emph{Ecosphere}, \emph{8}(8), e01926.
+\url{https://esajournals.onlinelibrary.wiley.com/doi/abs/10.1002/ecs2.1926}
 
 \leavevmode\hypertarget{ref-hunter2018not}{}%
-Hunter, D.O., Lagisz, M., Leo, V., Nakagawa, S.
-\& Letnic, M.
-(2018) Not all predators are equal: A continent-scale analysis of the effects of predator control on Australian mammals.
-\emph{Mammal Review}, \textbf{48}, 108--122.
+Hunter, D.
+O., Lagisz, M., Leo, V., Nakagawa, S., \& Letnic, M.
+(2018).
+Not all predators are equal: A continent-scale analysis of the effects of predator control on Australian mammals.
+\emph{Mammal Review}, \emph{48}(2), 108--122.
+\url{https://doi.org/https://doi.org/10.1111/mam.12115}
 
 \leavevmode\hypertarget{ref-iannarilli2019lorelograms}{}%
-Iannarilli, F., Arnold, T.W., Erb, J.
-\& Fieberg, J.R.
-(2019) Using lorelograms to measure and model correlation in binary data: Applications to ecological studies.
-\emph{Methods in Ecology and Evolution}, \textbf{10}, 2153--2162.
+Iannarilli, F., Arnold, T.
+W., Erb, J., \& Fieberg, J.
+R.
+(2019).
+Using lorelograms to measure and model correlation in binary data: Applications to ecological studies.
+\emph{Methods in Ecology and Evolution}, \emph{10}(12), 2153--2162.
+\url{https://doi.org/https://doi.org/10.1111/2041-210X.13308}
 
 \leavevmode\hypertarget{ref-kauffman2007landscape}{}%
-Kauffman, M.J., Varley, N., Smith, D.W., Stahler, D.R., MacNulty, D.R.
-\& Boyce, M.S.
-(2007) Landscape heterogeneity shapes predation in a newly restored predator--prey system.
-\emph{Ecology Letters}, \textbf{10}, 690--700.
+Kauffman, M.
+J., Varley, N., Smith, D.
+W., Stahler, D.
+R., MacNulty, D.
+R., \& Boyce, M.
+S.
+(2007).
+Landscape heterogeneity shapes predation in a newly restored predator--prey system.
+\emph{Ecology Letters}, \emph{10}(8), 690--700.
+\url{https://doi.org/https://doi.org/10.1111/j.1461-0248.2007.01059.x}
 
 \leavevmode\hypertarget{ref-kohl2019prey}{}%
-Kohl, M.T., Ruth, T.K., Metz, M.C., Stahler, D.R., Smith, D.W., White, P.J.
-\& MacNulty, D.R.
-(2019) Do prey select for vacant hunting domains to minimize a multi-predator threat?
-\emph{Ecology Letters}, \textbf{22}, 1724--1733.
+Kohl, M.
+T., Ruth, T.
+K., Metz, M.
+C., Stahler, D.
+R., Smith, D.
+W., White, P.
+J., \& MacNulty, D.
+R.
+(2019).
+Do prey select for vacant hunting domains to minimize a multi-predator threat?
+\emph{Ecology Letters}, \emph{22}(11), 1724--1733.
+\url{https://doi.org/https://doi.org/10.1111/ele.13319}
 
 \leavevmode\hypertarget{ref-kronfeld2003partitioning}{}%
-Kronfeld-Schor, N. \& Dayan, T. (2003) Partitioning of time as an ecological resource. \emph{Annual Review of Ecology, Evolution, and Systematics}, \textbf{34}, 153--181.
+Kronfeld-Schor, N., \& Dayan, T.
+(2003).
+Partitioning of time as an ecological resource.
+\emph{Annual Review of Ecology, Evolution, and Systematics}, \emph{34}(1), 153--181.
+\url{https://doi.org/https://doi.org/10.1146/annurev.ecolsys.34.011802.132435}
 
 \leavevmode\hypertarget{ref-lamb2020ecology}{}%
-Lamb, C.T., Ford, A.T., McLellan, B.N., Proctor, M.F., Mowat, G., Ciarniello, L., Nielsen, S.E.
-\& Boutin, S.
-(2020) The ecology of human-carnivore coexistence.
-\emph{Proceedings of the National Academy of Sciences}, \textbf{117}, 17876--17883.
+Lamb, C.
+T., Ford, A.
+T., McLellan, B.
+N., Proctor, M.
+F., Mowat, G., Ciarniello, L., Nielsen, S.
+E., \& Boutin, S.
+(2020).
+The ecology of human-carnivore coexistence.
+\emph{Proceedings of the National Academy of Sciences}, \emph{117}(30), 17876--17883.
+\url{https://doi.org/10.1073/pnas.1922097117}
 
 \leavevmode\hypertarget{ref-lima1999temporal}{}%
-Lima, S.L. \& Bednekoff, P.A. (1999) Temporal variation in danger drives antipredator behavior: The predation risk allocation hypothesis. \emph{The American Naturalist}, \textbf{153}, 649--659.
+Lima, S.
+L., \& Bednekoff, P.
+A.
+(1999).
+Temporal variation in danger drives antipredator behavior: The predation risk allocation hypothesis.
+\emph{The American Naturalist}, \emph{153}(6), 649--659.
+\url{https://doi.org/https://doi.org/10.1086/303202}
 
 \leavevmode\hypertarget{ref-lima1990behavioral}{}%
-Lima, S.L. \& Dill, L.M. (1990) Behavioral decisions made under the risk of predation: A review and prospectus. \emph{Canadian Journal of Zoology}, \textbf{68}, 619--640.
+Lima, S.
+L., \& Dill, L.
+M.
+(1990).
+Behavioral decisions made under the risk of predation: A review and prospectus.
+\emph{Canadian Journal of Zoology}, \emph{68}(4), 619--640.
+\url{https://doi.org/https://doi.org/10.1139/z90-092}
 
 \leavevmode\hypertarget{ref-marlow2015cats}{}%
-Marlow, N.J., Thomas, N.D., Williams, A.A., Macmahon, B., Lawson, J., Hitchen, Y., Angus, J.
-\& Berry, O.
-(2015) Cats \emph{(Felis catus)} are more abundant and are the dominant predator of woylies \emph{(Bettongia penicillata)} after sustained fox \emph{(Vulpes vulpes)} control.
-\emph{Australian Journal of Zoology}, \textbf{63}, 18--27.
+Marlow, N.
+J., Thomas, N.
+D., Williams, A.
+A., Macmahon, B., Lawson, J., Hitchen, Y., Angus, J., \& Berry, O.
+(2015).
+Cats \emph{(Felis catus)} are more abundant and are the dominant predator of woylies \emph{(Bettongia penicillata)} after sustained fox \emph{(Vulpes vulpes)} control.
+\emph{Australian Journal of Zoology}, \emph{63}(1), 18--27.
 
 \leavevmode\hypertarget{ref-mcgregor2020short}{}%
-McGregor, H., Moseby, K., Johnson, C.N.
-\& Legge, S.
-(2020) The short-term response of feral cats to rabbit population decline: Are alternative native prey more at risk?
-\emph{Biological Invasions}, \textbf{22}, 799--811.
+McGregor, H., Moseby, K., Johnson, C.
+N., \& Legge, S.
+(2020).
+The short-term response of feral cats to rabbit population decline: Are alternative native prey more at risk?
+\emph{Biological Invasions}, \emph{22}(2), 799--811.
+\url{https://doi.org/https://doi.org/10.1007/s10530-019-02131-5}
 
 \leavevmode\hypertarget{ref-miller2014finite}{}%
-Miller, D.L. \& Wood, S.N. (2014) Finite area smoothing with generalized distance splines. \emph{Environmental and Ecological Statistics}, \textbf{21}, 715--731.
+Miller, D. L., \& Wood, S. N. (2014). Finite area smoothing with generalized distance splines. \emph{Environmental and Ecological Statistics}, \emph{21}(4), 715--731.
 
 \leavevmode\hypertarget{ref-molsher2017mesopredator}{}%
-Molsher, R., Newsome, A.E., Newsome, T.M.
-\& Dickman, C.R.
-(2017) Mesopredator management: Effects of red fox control on the abundance, diet and use of space by feral cats.
-\emph{PLoS One}, \textbf{12}, e0168460.
+Molsher, R., Newsome, A.
+E., Newsome, T.
+M., \& Dickman, C.
+R.
+(2017).
+Mesopredator management: Effects of red fox control on the abundance, diet and use of space by feral cats.
+\emph{PLoS One}, \emph{12}(1), e0168460.
+\url{https://doi.org/https://doi.org/10.1371/journal.pone.0168460}
 
 \leavevmode\hypertarget{ref-niedballa2019assessing}{}%
-Niedballa, J., Wilting, A., Sollmann, R., Hofer, H.
-\& Courtiol, A.
-(2019) Assessing analytical methods for detecting spatiotemporal interactions between species from camera trapping data.
-\emph{Remote Sensing in Ecology and Conservation}, \textbf{5}, 272--285.
+Niedballa, J., Wilting, A., Sollmann, R., Hofer, H., \& Courtiol, A.
+(2019).
+Assessing analytical methods for detecting spatiotemporal interactions between species from camera trapping data.
+\emph{Remote Sensing in Ecology and Conservation}, \emph{5}(3), 272--285.
+\url{https://doi.org/https://doi.org/10.1002/rse2.107}
 
 \leavevmode\hypertarget{ref-norbury2015pests}{}%
-Norbury, G. \& Jones, C. (2015) Pests controlling pests: Does predator control lead to greater European rabbit abundance in Australasia? \emph{Mammal Review}, \textbf{45}, 79--87.
+Norbury, G., \& Jones, C.
+(2015).
+Pests controlling pests: Does predator control lead to greater European rabbit abundance in Australasia?
+\emph{Mammal Review}, \emph{45}(2), 79--87.
+\url{https://doi.org/https://doi.org/10.1111/mam.12034}
 
 \leavevmode\hypertarget{ref-pedersen2019hierarchical}{}%
-Pedersen, E.J., Miller, D.L., Simpson, G.L. \& Ross, N. (2019) Hierarchical generalized additive models in ecology: An introduction with mgcv. \emph{PeerJ}, \textbf{7}, e6876.
+Pedersen, E.
+J., Miller, D.
+L., Simpson, G.
+L., \& Ross, N.
+(2019).
+Hierarchical generalized additive models in ecology: An introduction with mgcv.
+\emph{PeerJ}, \emph{7}, e6876.
+\url{https://doi.org/https://doi.org/10.7717/peerj.6876}
 
 \leavevmode\hypertarget{ref-preisser2005scared}{}%
-Preisser, E.L., Bolnick, D.I. \& Benard, M.F. (2005) Scared to death? The effects of intimidation and consumption in predator--prey interactions. \emph{Ecology}, \textbf{86}, 501--509.
+Preisser, E.
+L., Bolnick, D.
+I., \& Benard, M.
+F.
+(2005).
+Scared to death?
+The effects of intimidation and consumption in predator--prey interactions.
+\emph{Ecology}, \emph{86}(2), 501--509.
+\url{https://doi.org/https://doi.org/10.1890/04-0719}
 
 \leavevmode\hypertarget{ref-R}{}%
-R Core Team. (2020) \emph{R: A Language and Environment for Statistical Computing}. R Foundation for Statistical Computing, Vienna, Austria.
+R Core Team. (2020). \emph{R: A language and environment for statistical computing}. R Foundation for Statistical Computing. \url{https://www.R-project.org/}
 
 \leavevmode\hypertarget{ref-reddiex2007control}{}%
-Reddiex, B., Forsyth, D.M., McDonald-Madden, E., Einoder, L.D., Griffioen, P.A., Chick, R.R.
-\& Robley, A.J.
-(2007) Control of pest mammals for biodiversity protection in Australia.
+Reddiex, B., Forsyth, D.
+M., McDonald-Madden, E., Einoder, L.
+D., Griffioen, P.
+A., Chick, R.
+R., \& Robley, A.
+J.
+(2007).
+Control of pest mammals for biodiversity protection in Australia.
 I.
 Patterns of control and monitoring.
-\emph{Wildlife Research}, \textbf{33}, 691--709.
+\emph{Wildlife Research}, \emph{33}(8), 691--709.
+\url{https://doi.org/https://doi.org/10.1071/WR05102}
 
 \leavevmode\hypertarget{ref-rees2019unexpectedly}{}%
-Rees, M.W., Pascoe, J.H., Wintle, B.A., Le Pla, M., Birnbaum, E.K.
-\& Hradsky, B.A.
-(2019) Unexpectedly high densities of feral cats in a rugged temperate forest.
-\emph{Biological Conservation}, \textbf{239}, 108287.
+Rees, M.
+W., Pascoe, J.
+H., Wintle, B.
+A., Le Pla, M., Birnbaum, E.
+K., \& Hradsky, B.
+A.
+(2019).
+Unexpectedly high densities of feral cats in a rugged temperate forest.
+\emph{Biological Conservation}, \emph{239}, 108287.
+\url{https://doi.org/https://doi.org/10.1016/j.biocon.2019.108287}
 
 \leavevmode\hypertarget{ref-relyea2003predators}{}%
-Relyea, R.A. (2003) Predators come and predators go: The reversibility of predator-induced traits. \emph{Ecology}, \textbf{84}, 1840--1848.
+Relyea, R.
+A.
+(2003).
+Predators come and predators go: The reversibility of predator-induced traits.
+\emph{Ecology}, \emph{84}(7), 1840--1848.
+\url{https://doi.org/https://doi.org/10.1890/0012-9658(2003)084\%5B1840:PCAPGT\%5D2.0.CO;2}
 
 \leavevmode\hypertarget{ref-ridout2009estimating}{}%
-Ridout, M.S. \& Linkie, M. (2009) Estimating overlap of daily activity patterns from camera trap data. \emph{Journal of Agricultural, Biological, and Environmental Statistics}, \textbf{14}, 322--337.
+Ridout, M.
+S., \& Linkie, M.
+(2009).
+Estimating overlap of daily activity patterns from camera trap data.
+\emph{Journal of Agricultural, Biological, and Environmental Statistics}, \emph{14}(3), 322--337.
+\url{https://doi.org/https://doi.org/10.1198/jabes.2009.08038}
 
 \leavevmode\hypertarget{ref-ripple2004wolves}{}%
-Ripple, W.J. \& Beschta, R.L. (2004) Wolves and the ecology of fear: Can predation risk structure ecosystems? \emph{BioScience}, \textbf{54}, 755--766.
+Ripple, W.
+J., \& Beschta, R.
+L.
+(2004).
+Wolves and the ecology of fear: Can predation risk structure ecosystems?
+\emph{BioScience}, \emph{54}(8), 755--766.
+\url{https://doi.org/https://doi.org/10.1641/0006-3568(2004)054\%5B0755:WATEOF\%5D2.0.CO;2}
 
 \leavevmode\hypertarget{ref-ritchie2009predator}{}%
-Ritchie, E.G. \& Johnson, C.N. (2009) Predator interactions, mesopredator release and biodiversity conservation. \emph{Ecology Letters}, \textbf{12}, 982--998.
+Ritchie, E.
+G., \& Johnson, C.
+N.
+(2009).
+Predator interactions, mesopredator release and biodiversity conservation.
+\emph{Ecology Letters}, \emph{12}(9), 982--998.
+\url{https://doi.org/https://doi.org/10.1111/j.1461-0248.2009.01347.x}
 
 \leavevmode\hypertarget{ref-robley2014long}{}%
-Robley, A., Gormley, A.M., Forsyth, D.M.
-\& Triggs, B.
-(2014) Long-term and large-scale control of the introduced red fox increases native mammal occupancy in Australian forests.
-\emph{Biological Conservation}, \textbf{180}, 262--269.
+Robley, A., Gormley, A.
+M., Forsyth, D.
+M., \& Triggs, B.
+(2014).
+Long-term and large-scale control of the introduced red fox increases native mammal occupancy in Australian forests.
+\emph{Biological Conservation}, \emph{180}, 262--269.
+\url{https://doi.org/https://doi.org/10.1016/j.biocon.2014.10.017}
 
 \leavevmode\hypertarget{ref-robley2019otway}{}%
-Robley, A., Moloney, P.
-\& Parks Victoria West Coast District Team.
-(2019) \emph{The Otway Ark: response of predators and native species 2016--2018.} Arthur Rylah Institute for Environmental Research Technical Report Series No.
+Robley, A., Moloney, P., \& Parks Victoria West Coast District Team.
+(2019).
+\emph{The Otway Ark: response of predators and native species 2016--2018.} Arthur Rylah Institute for Environmental Research Technical Report Series No.
 299.
@@
 \leavevmode\hypertarget{ref-robley2020glenelg}{}%
-Robley, A., Moloney, P., Stringer, L.
-\& Donald, S.
-(2020) \emph{Glenelg Ark 2005--2019: long-term predator and native mammal response to predator control.} Arthur Rylah Institute for Environmental Research Technical Report Series No.
+Robley, A., Moloney, P., Stringer, L., \& Donald, S.
+(2020).
+\emph{Glenelg Ark 2005--2019: long-term predator and native mammal response to predator control.} Arthur Rylah Institute for Environmental Research Technical Report Series No.
 318.
@@
 \leavevmode\hypertarget{ref-roshier2021space}{}%
-Roshier, D.A. \& Carter, A. (2021) Space use and interactions of two introduced mesopredators, european red fox and feral cat, in an arid landscape. \emph{Ecosphere}, \textbf{12}, e03628.
+Roshier, D.
+A., \& Carter, A.
+(2021).
+Space use and interactions of two introduced mesopredators, european red fox and feral cat, in an arid landscape.
+\emph{Ecosphere}, \emph{12}(7), e03628.
+\url{https://doi.org/https://doi.org/10.1002/ecs2.3628}
 
 \leavevmode\hypertarget{ref-schmitz2004trophic}{}%
-Schmitz, O.J., Krivan, V. \& Ovadia, O. (2004) Trophic cascades: The primacy of trait-mediated indirect interactions. \emph{Ecology Letters}, \textbf{7}, 153--163.
+Schmitz, O.
+J., Krivan, V., \& Ovadia, O.
+(2004).
+Trophic cascades: The primacy of trait-mediated indirect interactions.
+\emph{Ecology Letters}, \emph{7}(2), 153--163.
+\url{https://doi.org/https://doi.org/10.1111/j.1461-0248.2003.00560.x}
 
 \leavevmode\hypertarget{ref-schoener1974compression}{}%
-Schoener, T.W. (1974) The compression hypothesis and temporal resource partitioning. \emph{Proceedings of the National Academy of Sciences}, \textbf{71}, 4169--4172.
+Schoener, T.
+W.
+(1974).
+The compression hypothesis and temporal resource partitioning.
+\emph{Proceedings of the National Academy of Sciences}, \emph{71}(10), 4169--4172.
+\url{https://doi.org/10.1073/pnas.71.10.4169}
 
 \leavevmode\hypertarget{ref-scroggie2018invasive}{}%
-Scroggie, M.P., Forsyth, D.M., McPhee, S.R., Matthews, J., Stuart, I.G., Stamation, K.A., Lindeman, M.
-\& Ramsey, D.S.
-(2018) Invasive prey controlling invasive predators?
+Scroggie, M.
+P., Forsyth, D.
+M., McPhee, S.
+R., Matthews, J., Stuart, I.
+G., Stamation, K.
+A., Lindeman, M., \& Ramsey, D.
+S.
+(2018).
+Invasive prey controlling invasive predators?
 European rabbit abundance does not determine red fox population dynamics.
-\emph{Journal of Applied Ecology}, \textbf{55}, 2621--2631.
+\emph{Journal of Applied Ecology}, \emph{55}(6), 2621--2631.
+\url{https://doi.org/https://doi.org/10.1111/1365-2664.13253}
 
 \leavevmode\hypertarget{ref-smith2019integrating}{}%
-Smith, J.A., Donadio, E., Pauli, J.N., Sheriff, M.J.
-\& Middleton, A.D.
-(2019) Integrating temporal refugia into landscapes of fear: Prey exploit predator downtimes to forage in risky places.
-\emph{Oecologia}, \textbf{189}, 883--890.
+Smith, J.
+A., Donadio, E., Pauli, J.
+N., Sheriff, M.
+J., \& Middleton, A.
+D.
+(2019).
+Integrating temporal refugia into landscapes of fear: Prey exploit predator downtimes to forage in risky places.
+\emph{Oecologia}, \emph{189}(4), 883--890.
+\url{https://doi.org/https://doi.org/10.1007/s00442-019-04381-5}
 
 \leavevmode\hypertarget{ref-soule1988reconstructed}{}%
-Soulé, M.E., Bolger, D.T., Alberts, A.C., Wrights, J., Sorice, M.
-\& Hill, S.
-(1988) Reconstructed dynamics of rapid extinctions of chaparral-requiring birds in urban habitat islands.
-\emph{Conservation Biology}, \textbf{2}, 75--92.
+Soulé, M.
+E., Bolger, D.
+T., Alberts, A.
+C., Wrights, J., Sorice, M., \& Hill, S.
+(1988).
+Reconstructed dynamics of rapid extinctions of chaparral-requiring birds in urban habitat islands.
+\emph{Conservation Biology}, \emph{2}(1), 75--92.
+\url{https://doi.org/https://doi.org/10.1111/j.1523-1739.1988.tb00337.x}
 
 \leavevmode\hypertarget{ref-stobo2020management}{}%
-Stobo-Wilson, A.M., Brandle, R., Johnson, C.N.
-\& Jones, M.E.
-(2020) Management of invasive mesopredators in the Flinders Ranges, South Australia: Effectiveness and implications.
-\emph{Wildlife Research}, \textbf{47}, 720--730.
+Stobo-Wilson, A.
+M., Brandle, R., Johnson, C.
+N., \& Jones, M.
+E.
+(2020).
+Management of invasive mesopredators in the Flinders Ranges, South Australia: Effectiveness and implications.
+\emph{Wildlife Research}, \emph{47}(8), 720--730.
+\url{https://doi.org/https://doi.org/10.1071/WR19237}
 
 \leavevmode\hypertarget{ref-swan2015predicting}{}%
-Swan, M., Christie, F., Sitters, H., York, A.
-\& Di Stefano, J.
-(2015) Predicting faunal fire responses in heterogeneous landscapes: The role of habitat structure.
-\emph{Ecological Applications}, \textbf{25}, 2293--2305.
+Swan, M., Christie, F., Sitters, H., York, A., \& Di Stefano, J.
+(2015).
+Predicting faunal fire responses in heterogeneous landscapes: The role of habitat structure.
+\emph{Ecological Applications}, \emph{25}(8), 2293--2305.
+\url{https://doi.org/https://doi.org/10.1890/14-1533.1}
 
 \leavevmode\hypertarget{ref-vazquez2019comparing}{}%
-Vazquez, C., Rowcliffe, J.M., Spoelstra, K.
-\& Jansen, P.A.
-(2019) Comparing diel activity patterns of wildlife across latitudes and seasons: Time transformations using day length.
-\emph{Methods in Ecology and Evolution}, \textbf{10}, 2057--2066.
+Vazquez, C., Rowcliffe, J.
+M., Spoelstra, K., \& Jansen, P.
+A.
+(2019).
+Comparing diel activity patterns of wildlife across latitudes and seasons: Time transformations using day length.
+\emph{Methods in Ecology and Evolution}, \emph{10}(12), 2057--2066.
+\url{https://doi.org/https://doi.org/10.1111/2041-210X.13290}
 
 \leavevmode\hypertarget{ref-wayne2017recoveries}{}%
-Wayne, A.F., Maxwell, M.A., Ward, C.G., Wayne, J.C., Vellios, C.V.
-\& Wilson, I.J.
-(2017) Recoveries and cascading declines of native mammals associated with control of an introduced predator.
-\emph{Journal of Mammalogy}, \textbf{98}, 489--501.
+Wayne, A.
+F., Maxwell, M.
+A., Ward, C.
+G., Wayne, J.
+C., Vellios, C.
+V., \& Wilson, I.
+J.
+(2017).
+Recoveries and cascading declines of native mammals associated with control of an introduced predator.
+\emph{Journal of Mammalogy}, \emph{98}(2), 489--501.
+\url{https://doi.org/10.1093/jmammal/gyw237}
 
 \leavevmode\hypertarget{ref-willems2009predator}{}%
-Willems, E.P. \& Hill, R.A. (2009) Predator-specific landscapes of fear and resource distribution: Effects on spatial range use. \emph{Ecology}, \textbf{90}, 546--555.
+Willems, E.
+P., \& Hill, R.
+A.
+(2009).
+Predator-specific landscapes of fear and resource distribution: Effects on spatial range use.
+\emph{Ecology}, \emph{90}(2), 546--555.
+\url{https://doi.org/https://doi.org/10.1890/08-0765.1}
 
 \leavevmode\hypertarget{ref-wirsing2021context}{}%
-Wirsing, A.J., Heithaus, M.R., Brown, J.S., Kotler, B.P. \& Schmitz, O.J. (2021) The context dependence of non-consumptive predator effects. \emph{Ecology Letters}, \textbf{24}, 113--129.
+Wirsing, A.
+J., Heithaus, M.
+R., Brown, J.
+S., Kotler, B.
+P., \& Schmitz, O.
+J.
+(2021).
+The context dependence of non-consumptive predator effects.
+\emph{Ecology Letters}, \emph{24}(1), 113--129.
+\url{https://doi.org/https://doi.org/10.1111/ele.13614}
 
 \leavevmode\hypertarget{ref-woinarski2015ongoing}{}%
-Woinarski, J.C.Z., Burbidge, A.A.
-\& Harrison, P.L.
-(2015) Ongoing unraveling of a continental fauna: Decline and extinction of Australian mammals since European settlement.
-\emph{Proceedings of the National Academy of Sciences}, \textbf{112}, 4531--4540.
+Woinarski, J.
+C.
+Z., Burbidge, A.
+A., \& Harrison, P.
+L.
+(2015).
+Ongoing unraveling of a continental fauna: Decline and extinction of Australian mammals since European settlement.
+\emph{Proceedings of the National Academy of Sciences}, \emph{112}(15), 4531--4540.
+\url{https://doi.org/10.1073/pnas.1417301112}
 
 \leavevmode\hypertarget{ref-woinarski2018reptiles}{}%
-Woinarski, J.C.Z., Murphy, B., Palmer, R., Legge, S., Dickman, C., Doherty, T., Edwards, G., Nankivell, A., Read, J.
-\& Stokeld, D.
-(2018) How many reptiles are killed by cats in Australia?
-\emph{Wildlife Research}, \textbf{45}, 247--266.
+Woinarski, J.
+C.
+Z., Murphy, B., Palmer, R., Legge, S., Dickman, C., Doherty, T., Edwards, G., Nankivell, A., Read, J., \& Stokeld, D.
+(2018).
+How many reptiles are killed by cats in Australia?
+\emph{Wildlife Research}, \emph{45}(3), 247--266.
+\url{https://doi.org/https://doi.org/10.1071/WR17160}
 
 \leavevmode\hypertarget{ref-wood2017generalized}{}%
-Wood, S.N. (2017) \emph{Generalized Additive Models: An Introduction with R}. CRC press.
-
-\newpage
-
+Wood, S. N. (2017). \emph{Generalized additive models: An introduction with r}. CRC press.
+
+\newpage
 \setcounter{table}{0}  \renewcommand{\thetable}{S\arabic{table}} \setcounter{figure}{0} \renewcommand{\thefigure}{S\arabic{figure}} \setcounter{section}{0} \renewcommand{\thesection}{S\arabic{section}}
+\setcounter{page}{1}
+\nolinenumbers
 
 \hypertarget{supporting-information}{%
-\section*{Supporting Information}\label{supporting-information}}
-\addcontentsline{toc}{section}{Supporting Information}
+\section*{SUPPORTING INFORMATION}\label{supporting-information}}
+\addcontentsline{toc}{section}{SUPPORTING INFORMATION}
+
+\emph{Spatial variation in predator diel activity patterns -- feral cats avoid red foxes in time, not space}
+
+MW Rees, BA Wintle, JH Pascoe, M Le Pla, EK Birnbaum \& BA. Hradsky
 

### 2023-07-28 manuscript/manuscript.Rmd
changes following oikos review 1

--- 2021-08-05 manuscript/manuscript.Rmd
@@
 ---
-title: Joint spatiotemporal models reveal dramatic shifts in mesopredator diel activity across individuals, habitat types and apex predator activity levels
+title: Spatial variation in predator diel activity patterns
 author:
@@
     email: matt.wayne.rees@gmail.com
+    affiliation: UOM,CSIRO
+    correspondingauthor: true
+  - name: Brendan A. Wintle
     affiliation: UOM
-    footnote: 1
   - name: Jack H. Pascoe
-    email: 
+    affiliation: UOM,CEC
+  - name: Mark Le Pla
+    affiliation: UOM,CEC
+  - name: Emma K. Birnbaum
     affiliation: CEC
-  - name: Mark Le Pla
-    email: 
-    affiliation: CEC
-  - name: Emma K. Birnbaum
-    email: 
-    affiliation: CEC
-    footnote: 
-  - name: Alan Robley
-    email: 
-    affiliation: ARI
-  - name: Nick Golding
-    email: 
-    affiliation: CU,TKI
-  - name: Brendan A. Wintle
-    email: 
+  - name: Bronwyn A. Hradsky
     affiliation: UOM
-  - name: Bronwyn A. Hradsky
-    email: 
-    affiliation: UOM
-    footnote: 
 address:
   - code: UOM
-    address: Quantitative & Applied Ecology Group, School of Ecosystem and Forest Science, The University of Melbourne, Parkville, VIC, Australia
+    address: Quantitative and Applied Ecology Group, School of Ecosystem and Forest Sciences, The University of Melbourne, Parkville, VIC, 3010, Australia
+  - code: CSIRO
+    address: Health and Biosecurity, Commonwealth Scientific and Industrial Research Organisation, Ecosciences Precinct, Dutton Park, QLD, Australia
   - code: CEC 
-    address: Conservation Ecology Centre, Otway Lighthouse Rd, Cape Otway, VIC, Australia
-  - code: ARI 
-    address: Department of Environment, Land, Water and Planning, Arthur Rylah Institute for Environmental Research, Heidelberg, Australia
-  - code: CU 
-    address: Curtin University, Bentley, WA, Australia
-  - code: TKI 
-    address: Telethon Kids Institute, Perth Children's Hospital, Nedlands, WA, Australia
-footnote:
-  - code: 1
-    text: "Corresponding Author"
+    address: Conservation Ecology Centre, Lighthouse Rd, Cape Otway, VIC, 3233, Australia
 abstract: |
-  1. Background.
-  2. Methods.
-  3. Results.
-  4. *Synthesis and applications.*
-
-keywords: "daily activity patterns; camera trap; *Felis catus*; invasive predator; interspecific competition; mesopredator release; diel activity; hierarchical generalised additive models; species interactions; *Vulpes vulpes*."
-journal: "Ecography"
+  *Open research:* Data and code will be deposited on Dryad upon acceptance and can currently be found at this link: https://github.com/matt-w-rees/spatiotemporal-gams-invasive-predators.
+keywords: 
+  - diel activity patterns 
+  - generalised additive model
+  - feral cat
+  - invasive predator
+  - intraguild predator interactions
+  - lethal predator control
+  - mesopredator release
+  - red fox
+  - spatiotemporal model
+journal: "Oikos; Article"
 date: "`r Sys.Date()`"
-bibliography: mybibfile.bib
+classoption: preprint, 3p, authoryear
+bibliography: "thesis-copy.bib"
 header-includes:
   - \usepackage{setspace}\doublespacing
+  - \usepackage{lineno}
   - \usepackage{float}
   - \floatplacement{figure}{H}
-linenumbers: true
-#numbersections: true
-csl: ecography.csl
-geometry: margin=1.1in
+  - \newcommand{\beginsupplement}{\setcounter{table}{0}  \renewcommand{\thetable}{S\arabic{table}} \setcounter{figure}{0} \renewcommand{\thefigure}{S\arabic{figure}} \setcounter{section}{0} \renewcommand{\thesection}{S\arabic{section}}}
+  - \usepackage{pdflscape}
+  - \newcommand{\blandscape}{\begin{landscape}}
+  - \newcommand{\elandscape}{\end{landscape}}
+linenumbers: false
+numbersections: true
+# Use a CSL with `citation_package = "default"`
+csl: "ecography.csl"
 fontsize: 12pt
@@
     base_format: rticles::elsevier_article
-editor_options: 
-  chunk_output_type: console
+    keep_tex: true
+    citation_package: default
 ---
+
+```{r setup, include = FALSE}
+knitr::opts_chunk$set(
+  echo      = FALSE,
+  warning   = FALSE,
+  message   = FALSE,
+  error     = FALSE,
+  cache     = FALSE,
+  fig.path  = "../figs/",
+  fig.align = "center",
+  out.width = "80%")
+
+library(kableExtra)
+library(dplyr)
+library(readr)
+
+```
+
+```{=tex}
 \parskip=12pt 
+\newpage
+\setcounter{page}{1}
+```
+
+# Spatial variation in predator diel activity patterns {.unnumbered}
+
+$~$
+
+\linenumbers
+
+# ABSTRACT {.unnumbered}
+
+Understanding the constraints dominant predators impose on subordinate species is important for predicting ecosystem dynamics and anticipating outcomes of predator management.
+Subordinate predators may avoid dominant predators in time or space, making it difficult to quantify antipredator behaviours unless joint spatiotemporal analyses are used.
+Here, we tested whether an invasive dominant predator (red fox *Vulpes vulpes*) alters the spatiotemporal activity of an invasive subordinate predator (feral cat *Felis catus*).
+We collated records of both species from 3,667 camera-traps deployed experimentally across two regions of south-eastern Australia with simplified predator guilds; foxes were poison-baited in some landscapes within each region.
+We used generalised additive models to quantify changes in predator spatiotemporal activity across geographic space, vegetation types, human footprint and (artificially manipulated) gradients of dominant predator activity.
+Foxes and cats had similar diel activity patterns when averaged across all sites, however there was important differentiation at a finer scale---cats did not reduce their spatial activity but shifted diel patterns when localised fox activity was high.
+Cats were crepuscular on average.
+However, across dry vegetation types of both regions (where foxes were nocturnal), cats shifted to diurnal behaviour with increasing fox activity.
+In contrast, fox activity was relatively consistent throughout the daily cycle in the wet forest; here cats avoided dawn when fox activity was high.
+Changes in cat diel activity patterns may facilitate spatial coexistence between these two invasive predators, potentially shifting feral cat impacts onto different native prey.
+It is well-appreciated that predator activity varies spatially and fluctuates throughout the daily cycle.
+However, our study demonstrates that diel activity patterns also vary across space, likely mediated by both landscape-context and fear.
+dominant predator avoidance appears to be dynamic---a key nuance which is overlooked when simply comparing the average activity overlap between two species.
 
@@
 
-```{r setup, include = FALSE}
+# INTRODUCTION
 
-knitr::opts_chunk$set(echo = FALSE, cache = FALSE, warning = FALSE, message = FALSE)
+Top predators can drive ecosystem structure and dynamics by suppressing populations of rival predators and prey species [@estes2011trophic; @prugh2009rise].
+Dominant predators can suppress subordinate species directly through antagonistic interactions and predation, as well as indirectly through fear [@creel2008relationships; @allen2022fear].
+Fear-induced behavioural suppression can be as detrimental to subordinate species as predation itself, by limiting resource acquisition and breeding success [@schmitz2004trophic; @preisser2005scared; @brown1999ecology].
+Understanding how top predators&mdash;including humans&mdash;constrain the behaviour of subordinate species is therefore important to accurately predict consequences of predator management, such as reintroductions or lethal control [@suraci2016fear; @gaynor2021applied].
 
-```
+Spatial and/or temporal niche partitioning may allow subordinate species to coexist with top predators by reducing encounter-rates and resource overlap [@kronfeld2003partitioning].
+However, subordinate species may not consistently employ avoidance behaviours because perceived predation risk is spatiotemporally variable, and antipredator behaviours typically involve a trade-off against resource acquisition, such as limiting or relegating activity to suboptimal places or times [@lima1999temporal; @lima1990behavioral].
+Therefore, optimal predator avoidance strategies likely vary across heterogeneous landscapes where resource availability (e.g., shelter, food) and perceived predation risks differ [@kauffman2007landscape; @willems2009predator; @wirsing2021context; @davies2021spatial].
+For example, temporal predator avoidance may be preferable over spatial avoidance if food is constantly available throughout the day, and vice versa.
+These theories are unified under the ecology of fear concept [@brown1999ecology], which has gained increasing attention in recent times [@gaynor2019landscapes].
+However, there remains a lack of robust, empirical evidence documenting dynamic changes in antipredator behaviour.
 
-INTRODUCTION
-==========================
+Animal diel activity patterns are increasingly being considered alongside spatial activity and distributional patterns.
+However, spatial and temporal (in regard to the daily cycle) activity patterns are still often considered in an ad-hoc fashion.
+For example, by fitting separate models for spatial and temporal overlap, or by repeating spatial analyses (e.g., resource selection functions) at different time periods such night and day [e.g., @basille2015plastic; @kohl2019prey; @smith2019integrating; @wooster2022predator].
+Such ad-hoc approaches make it difficult to assess the relative importance, as well as dynamic changes in spatial and temporal processes [@suraci2022beyond].
 
-Effective wildlife management relies on understanding species population and behavioural dynamics to accurately predict intervention outcomes.
-These dynamics depend on both environmental conditions and those of other species, but few statistical approaches incorporate both factors, or allow responses to be dynamic across space and time.
-This is key because species interactions and niche preferences are multidimensional.
+There are many approaches to modelling intraguild spatiotemporal interactions, such as time-to-encounter models and coefficients of activity overlap [@frey2017investigating].
+Time-to-encounter approaches consider fine-scale avoidance or attraction by using time intervals between species detections as the response variable, although do not estimate diel activity patterns [@harmsen2009spatial; @niedballa2019assessing; @fancourt2019introduced].
+Multispecies models have been recently developed to assess overlap in spatial and diel activity between two species [@azzou2021sparse], as well as different diel activity patterns based on the presence or absence of another species at a site [@kellner2022two].
+Multispecies models are data-hungry and difficult to fit, but can account for imperfect detection and different types of predictor variables [e.g., @parsons2022effect].
+Given avoidance behaviours are likely dynamic [@gaynor2019landscapes], there is need for statistical approaches which jointly assess continuous shifts in spatial and diel activity patterns of one species in response to another, particularly when combined with experimental data [@suraci2022beyond].
 
-Fear is a near universal driver of animal behaviour.
-Fear may be from other individuals of their species, a dominant species (including humans) or an event (e.g.
-fire).
-Animals may be naive to threats (e.g.
-alien predators), unable to respond to threats (e.g.
-scarce prey), or the threat may be too minimal to warrant a response.
-Some environments more easily facilitate avoidance (e.g.
-high food and shelter accessibility), and so accounting for landscape context is key.
-A threat response may be routine, or only employed or enhanced when the threat is heightened.
-This likely depends on how constant and predictable the threat is.
+Generalised Additive Models (hereafter 'GAMs') are increasingly being used to estimate animal diel activity patterns and offer a flexible framework to jointly consider spatial activity.
+GAMs are useful to model diel activity patterns as they can implement cyclical splines&mdash;joining the end with the start of the day&mdash;as well as complexity penalties to reduce overfitting.
+If complex nonlinear (i.e., wiggly) effects are not statistically supported, smoothing penalties in GAMs 'shrink' them to a linear (i.e., no wiggliness) effect, or even to a horizontal line - removing the entire effect of the predictor variable and therefore  the need for additional tests of statistical significance (such as p-values and information criteria).
+GAMS can also capture nonlinear interactions between multiple variables with different units (i.e., 'tensor products'), and offer the ability to share information across categorical variables through hierarchical specifications [@wood2017generalized; @pedersen2019hierarchical].
+However, we are only aware of one study which allowed animal diel activity to interact with predation risk as a continuous variable in a GAM [although without considering overall activity, @cunningham2019temporal].
 
-Animals may respond to threats by altering their activity patterns (i.e.
-avoidance).
-An animal may avoid locations where the threat is likely to occur (hereafter ‘spatial avoidance’), avoid times of the day when a threat is likely to occur (hereafter ‘temporal avoidance’), or avoid both locations and times a threat is heightened (hereafter ‘spatiotemporal avoidance’).
-Such avoidance likely has trade-offs, which may or may not incur costs to individual and population fitness.
-Modelling approaches which do not allow spatiotemporal and dynamic fear reponses can only make limited inference about species interactions.
+The red fox *Vulpes vulpes* (hereafter 'fox') and feral cat *Felis catus* (hereafter 'cat') have devastating impacts on native prey throughout their introduced range, implicated in the extinction of \~10 and 63 species, respectively [@doherty2016invasive].
+The impacts of these invasive predators are particularly extreme on the Australian continent [@woinarski2015ongoing].
+Cats are more difficult to manage, and so introduced predator control programs in Australia often target only foxes [particularly through poison-baiting, @reddiex2007control].
+As foxes and cats compete for many of the same resources [@fleming2022distinctive], there is concern that lethal fox control could cause mesopredator release [@soule1988reconstructed] of cats [@glen2005complex; @doherty2017stop; @wayne2017recoveries].
+However, there is limited evidence that spatial cat activity increases in response to fox control [@hunter2018not], although there are signs of changes in behaviour and population density [@molsher2017mesopredator; @rees2023mesopredator].
+Other studies have investigated potential spatial and temporal interactions between these invasive predators [e.g., @roshier2021space], but not in response to fox control, or in a joint spatiotemporal framework that allows flexibility in cat avoidance behaviours with respect to differences in fox activity.
+Conservation managers are therefore uncertain as to whether fox control may increase the impacts of feral cats on native prey, through increases in cat density or behavioural changes which improve hunting success.
 
-Species interactions are usually considered in degrees of overlap between activity patterns.
-Spatiotemporal avoidance is most often considered, using joint species distribution models (including MSOMs).
-However, temporal avoidance is increasingly being used (perhaps just allowed due to widely available GPS and camera technology) with kernel density estimation methods.
-Although, these methods don’t allow environmental covariates, and are particularly susceptible to noisy estimates.
-Importantly, some are beginning to combine these methods to allow spatiotemporal responses (Azzou et al 2021; Farris et al 2020).
-However, degrees of overlap in activity patterns only get at routine avoidance, we are generally interested in whether changes in threat levels lead to altered behaviour.
-This is more indicative of species interactions because static levels of overlap may just reflect different niche preferences rather than avoidance.
-
-Dynamic behavioural responses are also important from a management perspective because they want to know whether an action will cause a change in animal behaviour /  range?).
-For example, after the reintroduction of wolves in Yellowstone, it was widely thought that this had limited elk populations.
-Using GPS data and habitat selection methods,  Kohl et al.
-2018 found that elk used riskier parts of landscape at times when wolves were least active -- suggesting coexistence without fitness costs.
-However, there are many other large carnivores in this ecosystem which are likely to impact elk behaviour.
-And so, will elk respond to wolves in the same way in different environmental contexts?
-There are little-no examples of replicated spatiotemporal interactions studies.
-
-There is contention around the interactions between two invasive predators - feral cats and foxes.
-Foxes are routinely culled, and there is concern that this could lead to a mesopredator release of feral cats.
-Mesopredator release may manifest in both behavioural and numerical changes, which my previous chapter identified - but that behavioral responses (ranging and detectability) were not universal.
-Here, we aim to dig into the behavioral response further - in particular diel activity times.
-
-In this study, we used generalised additive models to assess (1) how predator (fox and cat) diel activity varies between vegetation types, and (2) whether an introduced mesopredator (cat) changes its diel and/or spatial activity in response to the activity of an introduced apex predator (fox).
-Because foxes and cats fill a similar function niche, we expected that, on average, they would have similar diel activity patterns across vegetation types (hypothesis 1).
-However, where fox detections were high, we hypothesised that the subordinate cat to be less active (hypothesis 2) and shift diel activity to times when foxes are least active (hypothesis 3).
-We tested this on a simple predator system where only cats and foxes occur, and fox activity is manipulated through lethal control.
-This allowed focus on the sole interactions between these two predators, and ensured differences in correlations wasn’t just due to differences in niche preferences.
+In this study, we first explored (1) how predator diel activity patterns vary across geographic space ('model 1'), and (2) the role of vegetation type and human footprint in predator activity patterns ('model 2').
+Using this knowledge to partition our dataset, we then (3) tested whether the activity of a dominant predator (fox) affected the spatiotemporal activity of a subordinate predator (cat; 'model 3').
+In line with behavioural avoidance or exclusion, we predicted that cat diel activity pattens would shift towards less-risky times of day (i.e., times when fox activity was lower), and/or spatial cat activity would be lower at sites where fox activity was relatively high.
+We illustrate how GAMs can provide a simple framework to jointly assess spatial and temporal animal activity patterns, as well as potential species interactions.
 
@@
 
-MATERIALS AND METHODS
-==========================
+# MATERIALS AND METHODS
 
-## Study area
+## Study area and camera-trapping
+Our study was conducted in a simple predator system where foxes and cats are the only medium-large (> 1 kg) mammalian predators, and fox activity is manipulated using lethal control in some landscapes.
+This situation allowed sole focus on the interactions between these two predators, across an experimental gradient of dominant predator (fox) activity.
+We compiled data from multiple smaller-scale camera-trap studies, each designed to experimentally assess mammal responses to fox control.
+Overall, we collated 5,449 and 2,202 independent detections of foxes and cats, respectively (separated by at least 30 minutes) from 172,052 camera-trap nights (Table \@ref(tab:diel-tab1)).
 
-We collated camera-trap data on these two predator species from three distinct studies across two regions of south-west Victoria, Australia (Fig.
-1).
-The purpose of each study was to experimentally survey changes in the mammalian community due to fox control, and so our dataset covered a wide range contexts with and without fox-baiting, allowing us to investigate fox-cat interactions across a gradient of apex predator (fox) activity.
+We collated camera-trap data across two regions in south-west Victoria, Australia: the Glenelg region and Otway Ranges (Fig.
+\@ref(fig:diel-map)).
+Here, dingoes *Canis familiaris* are long-absent throughout, while tiger quolls *Dasyurus maculatus* are long-absent in the Glenelg region and likely functionally extinct in the Otway Ranges (last confirmed sighting in 2014).
+In broad sections of each region, government land managers conduct ongoing targeted lethal fox control for biodiversity conservation (see sections below for region-specific details).
+Poison-baits containing 3 mg of sodium fluroacetate (compound 1080) are buried at a depth of 12 - 15 cm at 1-km intervals along accessible forest tracks and roads.
+Different road densities therefore result in variable densities of poison-baits.
+Foxes readily dig these poison-baits often resulting in population suppression, whereas it is generally accepted that cats are unlikely to consume buried fox-baits and there is no evidence buried 1080 fox-baits suppresses cat populations [@algar2004feral; @moseby2011use; @risbey1997control; @hunter2018not].
+Managers also frequently implement prescribed fire across both regions, primarily to reduce fuel loads to prevent large wildfires.
 
-The terrestrial mammalian predator guild is depauperate throughout both regions.
-Dingoes *Canis familiaris* are now absent throughout, as is the the tiger quoll *Dasyurus maculatus*, more recently absent in the Otway Ranges (last sighted in 2014 despite extensive camera-trapping).
-Introduced red foxes and feral cats are therefore the only medium-large functional mammalian terrestrial predators across both regions.
+### Glenelg region
+In the Glenelg region, large patches of natural vegetation are fragmented by pastoral farming and residential properties (Fig.
+\@ref(fig:diel-map)).
+Foxes in three distinct forest blocks in this region have been subject to poison-baiting since October 2005, with fortnightly bait replacements [@robley2014long].
+These forest blocks, along with three similar, unbaited forest blocks to the north have been simultaneously surveyed annually under the 'Glenelg Ark' fox control program since 2005 [40 sites per block, @robley2020glenelg].
+Hair-tubes were used to monitor species from 2005 - 2013 [presented in @robley2014long], replaced by camera-traps from 2013; here we present camera-trap data from 2013 - 2019 [@robley2020glenelg].
+We also included a further 425 camera-trap deployments at unique locations from early 2018 [M.W.R PhD Surveys', @rees2023data].
+This totals 2,039 camera-trap deployments in the Glenelg region, collected in a control-impact experimental design (foxes had been continuously controlled for at 8 - 14 years in the treatment landscapes at the time of these surveys).
 
-In the baited parts of the regions, government land managers conduct ongoing targeted fox control for biodiversity conservation.
-Poison baits containing 3 mg of sodium mono-fluroacetate (1080) are buried at a depth of 10 cm at 1-km intervals along accessible forest tracks and roads.
+### Otway Ranges
+The Otway Ranges is a largely continuous patch of natural vegetation with a strong east-west rainfall gradient (Fig.
+\@ref(fig:diel-map)).
+A matrix of cool temperate rainforest and wet forest at high altitudes in the south-west descend into a large heathland directly north, and into dry forests and then heathlands to the north-east.
+Fox-baiting commenced in small sections of the Otway Ranges in 2008 and large-scale systematic baiting began in 2016 - 2017 under the 'Otway Ark' program [@robley2019otway].
+For the first six weeks, poison-baits were replaced weekly, then changing to ongoing monthly bait-replacement.
+There was a pause in baiting for approximately six months during the second half of 2018.
+Fox control recommenced in late 2018 with four weeks of fortnightly bait-replacement, before returning to monthly bait-replacement.
+A large section of the Otway Ranges to the north-west remains unbaited, but is monitored as an experimental non-treatment site [@robley2019otway].
+Otway Ark managers survey 372 camera-trap sites annually (sequentially across the region); we present one 'before' baiting survey and two 'after' baiting surveys of each site from 2016 - 2018, totalling 1,113 camera-trap deployments [@robley2019otway].
+We also include data from an additional before-after control-impact surveys (one 'before' baiting survey and two 'after' baiting surveys) in the western section of the Otway Ranges, conducted annually 2017 - 2019 [@rees2023data].
+This added a further 195 sites and 524 camera-trap deployments (Table \@ref(tab:tab-sum)).
 
-In the Glenelg region, Gunditjmara country, natural vegetation is highly fragmented.
-Here, camera-survey sites were situated in six distinct forest blocks: two are separated by a river, the others are separated by at least *XX* km of pastoral farming and residential property.
-Foxes in three of the forest blocks have been subject to lethal control since October 2005, with 1080 poison-baits replaced fortnightly [@ROBLEY2014262].
-
-The Otway Ranges, Gadubanud country, is a largely continuous patch of natural vegetation with a strong east-west rainfall and elevation gradient [@swan2021species].
-Foxes are controlled through 1080 poison-baits with monthly replacement across most of the Otway Ranges, but a large section to the north-west remains unbaited.
-The majority of this baiting began in 2016 - 2017, although fox-baiting commenced in some small sections in 2008.
-
-Alongside fragmentation, the extremely rugged, wet forest and rainforest matrix in the western Otway Ranges is the main difference between the two regions.
-Prescribed fire, primarily for the purpose of asset protection, is another key management action undertaken throughout both regions (except in very wet sections of Otway Ranges).
-
-## Camera-trap surveys
-
-We compiled camera-trap data from three distinct studies across the two regions.
-After removing camera-traps which were operational for less than 14 days, this totalled 3659 deployments of camera-traps at 1232 camera-trap sites.
-Survey durations ranged between 14 and 93 days (mean 47), totalling 171,966 trapnights.
-Camera-trap spacing was variable; the average minimum distance between simultaneously deployed camera-traps was 445, 853 and 1266 metres for each study.
-Camera-trap were deployed in the Glenelg region between 2013 – 2019, whereas in the Otway Ranges surveys were conducted in 2016 – 2019.
-
-All camera-trap deployments consisted of a Reconyx (Holmen, Wisconsin) brand camera-trap (both white and infrared flashes), attached to a tree or a metal picket, facing a lure.
-Camera-traps were set-up in two ways depending on the key aim of the project, either targeted toward predators or prey species.
-One study across both regions positioned camera-traps lower on a tree (around 15 - 30 cm above the ground – angled only slightly downwards) facing a tuna oil lure approximately 2 - 2.5 m away [detailed in @REES2019108287].
-The remaining camera-traps were positioned higher on a tree or a metal picket (at least 40 cm above ground) and angled downwards more strongly - facing a lure approximately 1 - 1.5 m away.
-These lures consisted of peanut butter, golden syrup and rolled oats mixed into a small ball, placed within a tea strainer or PVC pipe container and secured either to the ground, or 20 - 60 cm above ground on a wooden stake.
-All set-ups were effective in detecting both predator species.
+### Camera-trap set-ups
+All camera-trap deployments consisted of a Reconyx (Holmen, Wisconsin) brand camera-trap (white or infrared flash), attached to a tree or a metal picket, facing a lure.
+All camera-trap sites were positioned in the forest interior at least 30 m away from roads.
+The Glenelg Ark and Otway Ark fox monitoring programs positioned camera-traps at least 40 cm above ground on a tree or a metal picket and angled downwards toward a lure approximately 1 - 1.5 m away [@robley2020glenelg; @robley2019otway].
+The lures consisted of peanut butter, golden syrup and rolled oats mixed into a small ball, placed within a tea strainer or PVC pipe container and secured either to the ground, or 20 - 60 cm above ground on a wooden stake.
+The M.W.R PhD surveys across both regions positioned camera-traps lower on a tree (around 15 - 30 cm above the ground) angled only slightly downwards toward a tuna oil lure approximately 2 - 2.5 m away [detailed in @rees2023mesopredator].
+Camera-traps were active for an average of 47 days (maximum 93 days), totalling 172,052 trap-nights.
+Camera-trap spacing averaged 867 m for Glenelg Ark, 1246 m for Otway Ark, and 444 m for the M.W.R PhD surveys (Table \@ref(tab:tab-sum)).
 
 ## Data preparation
-We first created a table of species detections and deployment information (coordinates, dates) for every camera-trap deployment.
-All data analysis was conducted in R version 3.6.3 [@R].
-We used lorelograms to identify the minimum interval to approximate independence [@https://doi.org/10.1111/2041-210X.13308]; this indicated 30 minutes discarding repeat detections of a species within 30 minutes was sufficient in reducing temporal autocorrelation.
-To account for day length variation across space and time, we extracted sunrise and sunset times for each camera-trap deployment using the 'maptools' R-package [@maptools] and adjusted detection times to be relative to sunrise and sunset using the average double anchoring approach described by @https://doi.org/10.1111/2041-210X.13290.
-
-Using the 'reshape' R-package [@reshape], we manipulated the detection table into a dataframe with a row for each hour of the day, for every camera-trap deployment, recording the total number of 'independent' fox and feral cat detections within each hour for the entire camera-trap survey duration.
-
-We identified the broad vegetation type (Ecological Vegetation Class (hereafter 'EVC') group [@delwp2020]) at each camera-trap location.
-Six vegetation types were surveyed in both regions, although to varying degrees (Table 1).
-In the Otway region, rainforests are interspersed (primarily in low lying gullies) throughout wet and damp forests at a finescale, and we merged these groups (hereafter 'wet forests').
-
-For each deployment, we calculated the number of poison fox-baits within a 2.3 km radius around each camera-trap - the average maximum distance foxes in these regions travel from their home range centre [@hradsky2017human].
+Analyses were conducted in R version 4.1.3 [@R].
+We first used lorelograms to identify the minimum interval to approximate independence [@iannarilli2019lorelograms]; this indicated that discarding repeat detections of a species within 30 minutes was sufficient to reduce temporal autocorrelation.
+To account for day length variation across space and time, we extracted sunrise and sunset times for each camera-trap deployment using the 'maptools' R-package [@maptools] and adjusted detection times to be relative to sunrise and sunset using the average double anchoring approach described by @vazquez2019comparing.
+We then built a dataframe consisting of a row for each hour of the day (0 -- 23), for every camera-trap deployment (n = 3,667), recording the total number of 'independent' fox and feral cat detections within each hour across the camera-trap survey.
 
 ## Generalised additive models 
-We modelled predator counts (response variable) with generalised additive mixed-effects models (hereafter 'GAMMs') implemented in the 'mgcv' R-package [@wood2017].
-We first fit these models using the poisson family, but switched to the negative binomial family as overdispersion, but not zero-inflation, was detected using the 'DHARMa' R-package [@DHARMa].
-We specified a model offset to account for differences in camera-trap survey duration, and a random intercept for each site to account for repeat sampling.
-We used a duchon spline static spatial smooth to model both spatial autocorrelation and differences in predator activity across space (i.e.
-the effect of unmodelled environmental covariates).
-For the fox model, we additionally included a thin plate regression spline for poison-bait density, with a factor smoother for each region, as bait replacement schedules differed between the regions.
-This formed the base model specification for each model we fitted.
-Models (n = 2) differed in their specification of the cyclical hour smooth in order to explore (1) variations in predator diel activity across EVC groups, as well as (2) whether feral cat diel activity changed across the gradient of observed fox counts (detailed in the sections below).
-We used the double penalty model selection approach, which penalises the null space in addition to the range space (i.e.
-shrinking wiggly terms to linear terms) of the spline basis, meaning covariate effects can be entirely removed from the model [@MARRA20112372].
-We plotted models using 'ggplot2' and 'gratia' R-packages [@ggplot2; @gratia].
+We modelled the total number of independent detections of each predator per hour for each camera-trap deployment (response variable) with generalised additive mixed-effect models implemented in the 'mgcv' R-package [@wood2017generalized].
+We used the negative binomial family, as overdispersion, but not zero-inflation, was detected with a poisson distribution using the 'DHARMa' R-package [@DHARMa].
+We specified the natural log of the number of survey days as a model offset to account for differences in camera-trap survey duration, and a random intercept for each site to account for repeat sampling for models 2 and 3 (as pseudoreplication from repeat sampling in model 1 was explicitly accounted for using the spatial term).
+For fox models, we also included a smooth effect of poison-bait density [number of baits within a 2.3 km radius around each camera-trap - the average maximum distance adult foxes travel from their home-range centre in these regions, @hradsky2017human] with separate responses per region to account for the effect of fox control.
+These specifications were included in each model we fitted; models differed in their specification of the cyclical hour smooth to provide inference on variations of predator diel activity across the three questions of interest - this is detailed in the sections below.
+We used the 'ggplot2' [@ggplot2] and 'gratia' [@gratia] R-packages to plot models.
 
-We expected the activity of each predator species to follow a broadly consistent pattern (e.g.
-lower during the day), but potentially have small shifts in relative activity between sunset and sunrise across the different EVC groups (1).
-Fitting separate responses for each EVC group would throw away shared information, which is particularly costly for the least surveyed groups, and would result in more noisy estimates.
-We therefore specified a global smoother for hour (i.e.
-average response), along with group-level smoothers with shared wiggliness for each EVC group [i.e.
-model GS in @pedersen2019hierarchical].
-This model structure penalises functions which deviate strongly from the average response, increasing confidence that they are true deviations opposed to the result of sampling noise.
+In this paper, we use the term 'spatial activity' to refer to the number of 'independent' predator detections at a site (offset to account for survey duration; analogous to an activity or abundance index), and 'diel activity pattern' to refer to fluctuations in relative activity throughout the 24-hour daily cycle.
+The first two models we fit explored how predator spatiotemporal activity varied across these regions, informing the best structure for our third model which tested the activity of foxes limits the spatiotemporal activity of cats (detailed in the sections below).
 
-To investigate changes in feral cat  diel activity in response to counts of foxes.
-We added an extra column to our dataframe with the number of fox detections divided by the log of the number of survey days to account for differences in camera-trap survey durations (hereafter 'adjusted fox counts').
+### How does predator diel activity patterns vary across geographic space? (model 1)
+To examine how the spatial activity and diel activity patterns of each predator varied across space, we fit a model for each predator which included a tensor product interaction between a spatial smooth (thin plate regression spline) and hourly smooth.
+This model specification allowed predators to have different activity levels across space (static across the years surveyed), as well as variation in diel activity pattern across space.
+Space was modelled using camera-trap coordinates and a duchon spline basis [@miller2014finite].
 
-Analysis of stage 1 showed that fox diel activity patterns in wet forests differed relative to all other EVC groups.
-We therefore expected feral cats might to temporally avoid foxes in a different way in wet forests than the dry vegetation groups, but that responses would be similar in all the drier vegetation groups.
-We therefore modelled fox-induced changes in feral cat diel activity separately for wet and dry vegetation groups.
-We further split dry vegetation groups by region, to examine whether patterns remained consistent across space?
-for spatial replication.
-We refer to the resulting variable as 'habitat type', which had three levels: wet forests of the Otway Ranges ('wet_otways') dry EVC groups of the Otway Ranges ('dry_otways') and all of the Glenelg region data ('dry_glenelg').
+### How does vegetation type and human footprint affect predator activity patterns? (model 2)
+The spatial term in model 1 estimated differences in predator activity across geographic space; we hypothesised that this was partly due to differences in vegetation type, based both on the observed spatial patterns and because vegetation type is a major driver of understorey habitat structure and prey occurrence in these regions [@swan2015predicting; @hradsky2017bayesian].
+To test whether the diel activity pattern of each predator varied among vegetation types, we identified the Ecological Vegetation Class group [hereafter 'vegetation type' - standard units for vegetation classification in Victoria, @delwp2020bioregions] for each unique camera-trap site, totalling eight vegetation types.
+As rainforests are interspersed (primarily in low lying gullies) at fine-scales throughout wet and damp forests in the south-eastern Otway Ranges, we merged them together (hereafter referred to as 'wet forests').
+We then estimated predator activity across vegetation types using a hierarchical model specification: a global smoother for hour (i.e., average response) and group-level smoothers with shared wiggliness for the seven vegetation types ['model GS' detailed in @pedersen2019hierarchical].
+We also fit a tensor product interaction term for hour and the human footprint index [@venter2016sixteen; @venter2018last]; as human footprint is known to impact animal activity patterns [@gaynor2018influence; @van2023influence]
 
-We used a tensor product of hour (using cyclic cubic regression spline) and log adjusted fox count (using thin plate regression spline) smooths to model feral cat diel activity across the range of observed fox activity, separately for each habitat type.
-By also having double penalty shrinkage turned on, this meant that feral cat activity could vary nonlinearly and be penalised consistently across both dimensions.
-This model specification allows five different scenarios: that there was (1) no effect of hour or foxes on feral cats (2) a static hourly effect only, (3) a spatial response to foxes only, (4) a spatial response to foxes and an unrelated static hourly effect, (5) a spatial response to foxes and a hourly effect which changes across the range of fox counts.
+### Do feral cats avoid foxes in space or time? (model 3)
+Fox diel activity across vegetation types showed strong similarity between all vegetation types except wet forests.
+To examine whether cats avoid foxes in space or time, we therefore modelled fox-induced changes in feral cat diel activity separately for wet forest and dry vegetation types.
+We further split dry vegetation types by region for further replication.
+We refer to the resulting variable as 'habitat type', which had three levels: (i) wet forests and rainforests in the western Otway Ranges ('wet_otways'), (ii) dry vegetation types in the Otway Ranges ('dry_otways') and (iii) dry vegetation types in the Glenelg region ('dry_glenelg').
+We further hypothesised that cats would avoid foxes in time by becoming more diurnal in dry vegetation types where foxes were mostly nocturnal, but not in wet forests where fox activity showed little variation across the daily cycle.
+
+To investigate changes in feral cat diel activity across the range of observed fox activity, we first quantified fox activity for each camera trap deployment as the total number of fox detections for the deployment divided by the number of survey days, to adjust for differences in survey duration (hereafter 'adjusted fox counts').
+We modelled an interaction between hour and adjusted fox counts (as a thin plate regression spline with shrinkage - so complexity penalties could entirely remove this effect from the model if not supported), allowing cats to have nonlinear responses to both hour and adjusted fox counts.
+We fit separate tensor product interactions for each habitat type (using a 'by-variable' term).
+For a direct visual comparison to fox activity, we fit another fox model where a diel activity curve was estimated separately across each of the three habitat types.
 
@@
 
-```{r map, out.width="100%", fig.cap = "Locations of study sites in south-west Victoria, Australia.
+(ref:fig1cap) Locations of our study regions in south-west Victoria, Australia.
 The grids of camera-traps are denoted by white dots.
-The Glenelg region is to the west and Otway region to the east.
+The Glenelg region (38°05'54"S 141°44'41"E) is to the west and Otway Ranges (38°57'82"S 141°68'41"E) to the east.
 Native vegetation is indicated by dark green, with hill shading.
-Map tiles by Stamen Design, under CC BY 3.0, map data by OpenStreetMap, under CC BY SA."}
-knitr::include_graphics("../figs/fig1_map.png")
+*Map tiles by Stamen Design, under CC BY 3.0, map data by OpenStreetMap, under CC BY SA.*.
+
+```{r diel-map, fig.cap = '(ref:fig1cap)'}
+knitr::include_graphics("../figs/map_cams.png")
 ```
@@
 
-RESULTS
-============
+# RESULTS
 
-## Predator diel activity across vegetation types
+## Lethal suppression of foxes  
+In each model, there was a strong effect of poison fox-baiting on fox spatial activity, particularly in the Glenelg region (Supporting Information Fig.
+\@ref(fig:fox-baits)b).
+Predicted fox counts on the response scale were reduced by a factor of 12, 24, and 28 in the Glenelg region, as well as a factor of 13, 13, and 8 in the Otway Ranges, for models 1, 2, and 3, respectively.
+For example, model 2 predicted fox counts in the Glenelg region to range from 0.097 (95% CI: 0.077 - 0.122) in sites with no fox-baiting, to 0.004 (95% CI: 0.002 - 0.012) at sites with the maximum poison-bait density; in the Otways ranging from 0.041 (95% CI: 0.033 - 0.05) - 0.008 (95% CI: 0.003 - 0.020) across the gradient of poison-bait density.
 
-On average, both predator species occupied similar times of the day--mostly nocturnal with activity peaks around sunrise and sunset.
-The main difference was that fox activity peaked just after sunset rather than on sunset.
-For foxes, this diel pattern was largely exhibited across all vegetation types except wet forest.
-In wet forests and rainforests, the hour of the day had little effect on foxes (especially considering diel activity curve here is being drawn towards a nocturnal pattern).
+## How does predator diel activity patterns vary across geographic space? (model 1)
+Predator activity varied considerably across space and throughout the 24-hour daily cycle, and there was some variation in the predator diel activity patterns across space.
+On average, both predators showed similar diel activity patterns, with peaks in activity around sunrise and sunset (i.e., crepuscular), although were more likely to be active during the night than the day (Fig.
+\@ref(fig:diel-st-int)i).
+On average, the main difference between the species was that fox activity peaked just after sunset and they were less likely to be active during the day than cats.
+Cats also tended to be more active at sunset relative to sunrise (Fig.
+\@ref(fig:diel-st-int)i).
 
-lowest activity during the middle of the day, peaking at sunrise and sunset times, while remaining relatively high at night and dawn. 
-Foxes had a stronger diel activity patterns than cats, much less likely to be out during the day.
+The interaction between hourly and spatial smooths revealed deviations from marginal diel activity pattern in space, particularly across the two regions, and within the Otway Ranges (Fig.
+\@ref(fig:diel-st-int)).
+For example, at midday, foxes were less active in the Glenelg region and more active across much of the Otway Ranges (except for the north-east) than the average diel activity pattern.
+Relative to foxes, cats had slightly more complex (i.e., wiggly) deviations in space from their average diel activity pattern.
+Most notably, cats were less active during sunset and more active at midnight in the Otway Ranges relative to the Glenelg region (Fig.
+\@ref(fig:diel-st-int)).
 
-Diel activity patterns were similar across the different vegetation types for each predator,, except for wet forests, where more substantial changes were seen.
-In the wet forests, ox activity remained fairly constant throughout day and night barely had any diel activity pattern here, cats more nocturnal.
-Deviance explained and Rsq values for the two models were XX.
+## How does vegetation type and human footprint affect predator activity patterns? (model 2)
+Vegetation type had a strong influence on the spatial and diel activity patterns of foxes and cats, whereas human footprint only affected fox spatial activity.
+Spatial levels of fox activity were similar across all vegetation types, except wet forests where fox activity was considerably lower.
+Spatial cat activity was relatively more variable across vegetation types; lowest in heathy woodlands and highest in wet forests (Fig.
+\@ref(fig:diel-veg)b).
+Diel activity patterns for foxes were similar across all vegetation type except wet forests; in wet forests, foxes were consistently active throughout the daily cycle (Fig.
+\@ref(fig:diel-veg)a).
+On the other hand, cats were mostly nocturnal (and most active) in wet forests, but largely crepuscular in all other vegetation types (Fig.
+\@ref(fig:diel-veg)a).
 
-## Fox-induced changes in feral cat diel activity
+There was no evidence that human footprint affected diel activity patterns for either predator species; the interaction term between hour and human footprint index was removed from both models.
+Fox spatial activity increased linearly with the human footprint index (Supporting Information Fig.
+\@ref(fig:fox-hfi)a), although human footprint did not impact cat spatial activity (the marginal effect of human footprint was also removed from the cat model).
 
-We saw evidence of changes in cat diel activity due to fox activity in all three habitat types.
-Similar responses in dry habitats of both regions -- cat spatial activity increased with fox spatial activity, but became more diurnal.
-In Otway wet forests, no real change in spatial activity of cats due to foxes, but cats became more strongly nocturnal.  
-Deviance explained and Rsq value was XX.
-Slightly higher uncertainty where fox spatial activity is high, this is because we had a largest sample size for fox absences, and fewest at the extreme of fox activity.
+## Do feral cats avoid foxes in space or time? (model 3)
+The model indicated that cats avoided foxes temporally rather than spatially (Fig.
+\@ref(fig:diel-cat-fox)).
+Fox activity did not impact spatial cat activity in the Otway Ranges, whereas cat spatial activity slightly increased with adjusted fox counts in the Glenelg region.
+
+Across all habitat types, feral cat diel activity patterns changed across gradients of fox activity (Fig.
+\@ref(fig:diel-cat-fox)).
+In the Glenelg region and Otway dry habitat types, feral cats had a nocturnal-crepuscular diel activity pattern where fox activity was low, but were most active during the day where fox activity was high.
+In contrast, in the wet forests of the Otway Ranges, feral cats were more strongly nocturnal when fox activity was high, avoiding dawn in particular.
 
@@
 
-```{r tab1}
-library(pander)
-library(tidyverse)
+```{r diel-st-int, out.width = "70%", fig.cap = "Marginal diel activity (i) and space-time interaction effect in the Glenelg region (ii) and Otway Ranges (iii) at midnight (0), sunrise (6), midday (12) and sunset (18) times for introduced red foxes \\textit{Vulpes vulpes} and feral cats \\textit{Felis catus}.
+These effects were estimated by a Generalised Additive Model with a tensor product interaction between spatial and hourly smooths, fit to camera-trap data ('model 1').
+Dotted, vertical lines represent average sunrise and sunset times, shaded areas indicate 95\\% confidence intervals (i); black dots depict camera-trap sites (ii; iii)."}
+knitr::include_graphics("../figs/spatial_interaction.png")
+
+```
+
+\newpage
+
+```{r diel-veg, fig.cap = "Red fox \\textit{Vulpes vulpes} (a) and feral cat \\textit{Felis catus} (b) diel activity patterns overall (i) and across different Ecological Vegetation Class (EVC) groups (ii, iii) in south-west Victoria, Australia.
+These effects were estimated by a hierarchical Generalised Additive Model with a global smoother for hour (i.e., average response) and group-level smoothers with shared wiggliness for vegetation type (EVC groups), fit to camera-trap data ('model 2').
+Dotted, vertical lines represent average sunrise and sunset times.
+Shaded areas indicate 95\\% confidence intervals.
+Both invasive predators had a crepuscular to nocturnal diel activity pattern on average, with slight deviations across the drier EVC groups and large deviations in wet forests (ii; wet forests shown as pink line).
+The overall level of activity was relatively consistent across EVC groups for foxes (a – iii), whereas it differed substantially for feral cats (b - iii)."}
+knitr::include_graphics("../figs/predator_veg_leg.png")
+```
+
+\newpage
+
+```{r diel-cat-fox, fig.cap="Red fox \\textit{Vulpes vulpes} diel activity patterns (a), as well as variation in feral cat \\textit{Felis catus} diel activity patterns in response to counts of 'independent' fox detections (log-transformed and survey effort adjusted) across each habitat type in south-west Victoria, Australia (b; c).
+Effects were estimated by a Generalised Additive Model with a tensor product interaction between hour (cyclical spine) and fox counts adjusted for survey length at each camera-trap ('fox activity'; 'model 3'), fit to camera-trap data.
+Panels B and C present the same model in two different ways: (b) cat diel activity across the entire continuum of fox activity, (c) cat diel activity patterns at the minimum and maximum fox activity, with associated 95\\% confidence intervals (shaded bands).
+Grey vertical lines represent average sunrise and sunset times.
+In the Glenelg region, there were more feral cat detections where there were more fox detections, but cat peak diel activity shifted from crepuscular night to pre-dawn and midday (a).
+In the Otway Ranges, feral cat activity also peaked during the day where fox activity was high in dry vegetation types (b), but was more nocturnal where fox activity was high in the rainforests and wet forests (c)."}
+knitr::include_graphics("../figs/cat_fox_count.png")
+```
+
+\newpage
+
+# DISCUSSION
+A key question in ecological theory is whether animals are evolutionarily hardwired to occupy particular temporal niches, or have circadian rhythms that are responsive to changing environmental conditions and interactions with other species [@daan1981adaptive; @schoener1974compression; @lima1990behavioral].
+Here we demonstrate that diel activity patterns are not fixed, but vary across space based on landscape context and fear.
+In our study, sympatric invasive predators had similar diel activity patterns when averaged across broad regions [i.e., high circular overlap, as did @roshier2021space], but behaviours varied considerably within landscapes.
+Foxes were nocturnal in dry habitat types, but showed little variation in activity through the daily cycle in wet forests (Fig.
+\@ref(fig:diel-veg)a).
+In contrast, cats were mostly nocturnal in wet forests but crepuscular in dry vegetation types (Fig.
+\@ref(fig:diel-veg)b).
+Within broad habitat types, cats altered their diel activity patterns at sites with higher fox activity towards less-risky times of day (i.e., times when fox activity was lower), however we found no signs of foxes suppressing cat spatial activity (Fig.
+\@ref(fig:diel-cat-fox)).
+Control programs that reduce invasive fox activity may therefore shift cat diel activity patterns, which in turn may alter cat impacts on native prey species.
+Quantifying changes in diel activity patterns provides important context for understanding species interactions, which is key for effective ecosystem management [@gaynor2021applied].
+
+Shifting diel activity patterns may facilitate spatial coexistence of dominant and subordinate species [@carothers1984time].
+For cats, altering diel activity patterns to less-preferred times of day may be worthwhile to persist in high-quality habitat.
+There have been an increasing number of studies which document slight changes in diel activity peaks of subordinate species in response to dominant predators, but few studies have demonstrated predator-induced reversals in diel activity [@kronfeld2003partitioning].
+Notably, ship rats *Rattus norvegicus* were also found to switch from nocturnal to diurnal behaviour in response to fox activity [@fenn1995use] and a similar nocturnal-diurnal shift was observed in American mink *Neovison vison* following the recolonisation of native predators [@harrington2009impact].
+Our study is particularly unique in that we replicated this behaviour across two separate regions with experimental manipulation, although there was considerable uncertainty in cat diel activity patterns where fox activity was highest (because there were fewer sites with high fox activity relative to low).
+
+For cats in our study, a switch to diurnal behaviour where fox activity was high in the dry vegetation types may have been facilitated by the higher abundance of reptiles in these habitat types relative to wet forests [which are mostly diurnal, @woinarski2018reptiles].
+At wet forest sites with high fox activity, cats concentrated their activity away from sunrise and sunset towards midnight, despite a diurnal shift appearing to similarly reduce the risk of a fox encounter.
+In this situation, we expect becoming more nocturnal to be favourable over a diurnal shift because this is when small mammals are active and cats would be least visible to foxes [cats here mostly had black or grey coats, @rees2023mesopredator].
+Spatial fox activity was twice as low in the wet forests relative to dry habitat types, and so cats were likely under less pressure to radically alter their diel activity patterns.
+Understanding how these potential avoidance behaviours impacts native prey is a key research priority to improve invasive predator management.
+
+Cats appeared to avoid foxes in time, but we saw no evidence of spatial avoidance (i.e.
+no fine-scale negative association between fox and cat spatial activity.
+In fact, there was some evidence that cat and fox spatial activity was positively correlated in the Glenelg region, suggesting temporal partitioning may facilitate predator co-abundance in quality habitat.
+Nonetheless, cats may have still avoided foxes spatially, but transiently on the scale of hours to days, or at finer spatial scales than most camera-trap studies can detect [and which telemetry studies can highlight, e.g., @kohl2019prey; @smith2019integrating].
+Short-term spatial avoidance is quite plausible given foxes mark territories using scats and odours, which cats could tangibly associate with high risk shortly after.
+Subordinate predators may prefer temporal avoidance of dominant predators (changing diel patterns or transient site avoidance) over longer-term site avoidance when both species have similar resource requirements and share many prey species.
+
+Many species increase nightly activity to avoid humans [@gaynor2018influence; @van2023influence], however we effect of human footprint on either predator species on our study.
+This is somewhat unsurprising given there was little variation in human activity among our sites; we studied predators within rugged conservation reserves largely free from human pressures (our camera-trap sites encompassed only the lower 32% of the total human footprint index range).
+On the other hand, foxes within our study region often select for human-modified features more during night than day [@hradsky2017human].
+Subordinate predators may still prefer to avoid dominant (animal) predators over humans, even when this decision leads to higher mortality rates [@prugh2023fear].
+
+A distinction of our study from others is that we modelled potential avoidance behaviours in a simple predator guild, where dominant predator activity was artificially manipulated (suppressed by up to 21-fold).
+The simple predator guild removes bias from unmodelled impacts of other predators in the system [@levi2012wolves].
+Artificial manipulation of the dominant predator also reduces potential bias from differences in niche preferences between the predator species.
+We also included replication across different habitat types.
+However, because our study did not consider associations with prey species, we cannot distinguish whether changes in cat diel activity patterns were the result of direct fox avoidance or indirect associations with shared prey.
+For example, low fox activity may promote the availability of a preferred shared prey species with a diel pattern which differs from those of cats on average, and as a result, cats might shift diel activity patterns at sites with low fox activity to more closely match those of the more abundant prey species.
+We would expect introduced European rabbits *Oryctolagus cuniculus* and hares *Lepus europaeus* (which are diurnal) to be particularly likely to induce such a response in dry vegetation types [@stobo2020management; @mcgregor2020short], however, rabbits and hares are rare within the natural vegetation of the landscapes we surveyed (only ever being detected at 40 of 1,232 sites) and there is little evidence that predation by foxes suppresses rabbit populations [@norbury2015pests; @scroggie2018invasive].
+
+Flexible antipredator behaviours make evolutionary sense, but have been rarely demonstrated in terms of spatiotemporal predator avoidance, because this often requires manipulative experiments or at least more complicated models [although, see @relyea2003predators; @brown2013phenotypically; @cunningham2019temporal].
+Our study demonstrates that GAMs offer a powerful tool for modelling continuous shifts in animal activity across both space and time, capable of capturing complex interactions and sharing information across categorical variables.
+We are only aware of one study which allowed animal diel activity to interact with predation risk as a continuous variable in a GAM [although without allowing for nonlinear shifts via a tensor product or jointly considering spatial activity, @cunningham2019temporal].
+Further, complexity penalties are particularly beneficial in assessing spatiotemporal responses, because they are less prone to spurious estimates from small sample sizes or noisy data, and allow multiple hypotheses to be tested within one model without the need for ad-hoc significance testing (e.g., running a separate model for each scenario and comparing with Information Criteria).
+For example, our model specification allowed cats to respond to fox activity with (a) nonlinear shifts in both spatial and diel activity (as was observed in the Glenelg region), (b) nonlinear shifts in spatial but not diel activity, (c) nonlinear shifts in diel but not spatial activity (as was observed in both habitat types of the Otway Ranges), as well as their respective linear counterparts and also different linear-nonlinear combinations of the first scenario ('model 3').
+**ADD CLOSER**
+
+The alternative approach of simply comparing average diel activity overlap between two species [@ridout2009estimating] would have been misleading for two reasons.
+Firstly, predator diel activity patterns varied 'naturally' across heterogeneous landscapes (requiring avoidance to be tested in wet forests and dry vegetation types separately, Fig.
+\@ref(fig:diel-veg)).
+Secondly, dominant predator temporal avoidance strategies were not consistently employed, but depended on spatial dominant predator activity and habitat type (Fig.
+\@ref(fig:diel-cat-fox)).
+Despite their underlying statistical complexity, GAMs in the 'mgcv' R-package are straightforward to fit.
+Our GAM framework for modelling spatiotemporal activity can be used on any species with time-stamped detections, including datasets with categorical or continuous covariates and hierarchical groupings.
+However, like many other spatiotemporal approaches such as time-to-encounter models, a key limitation is that our GAM approach did not separate the observation processes from the state process.
+Integrating GAM-like functions into recently developed multispecies spatiotemporal models [e.g., @kellner2022two] would be highly advantageous.
+
+Animal diel activity patterns can be complex, varying across space, habitat types and threat-levels [@mccann2017temporal].
+Despite telling an important story about how animals interact with each other and the environment, detection times are commonly discarded from statistical analyses of camera-trap data.
+When considered, diel activity patterns are predominantly estimated at the population-level, overlooking finer-scale behaviours that can affect fitness, survival and ecosystem-impacts.
+Our results demonstrate the importance of (a) considering diel activity in regards to species interactions, (b) modelling changes in animal behaviour rather than overlap with other species, and (c) testing avoidance behaviours within a joint spatiotemporal framework.
+Our study adds to the growing body of evidence that dominant predators may cause fear which is powerful enough to reverse the diel activity patterns of subordinate species [@kronfeld2003partitioning], and crucially, demonstrates that these effects are habitat-dependent.
+
+\newpage
+
+# REFERENCES {.unnumbered}
+
+::: {#refs}
+:::
+
+```{=tex}
+\newpage
+\beginsupplement
+\setcounter{page}{1}
+```
+# SUPPORTING INFORMATION {.unnumbered}
+
+\newpage
+
+```{=tex}
+\newpage
+\blandscape
+```
+
+```{r tab-sum}
+
+cam_dist <- read.csv("../raw_data/survey_summary_distances.csv")
+
+cam_dist <- select(cam_dist, !c(X, sess))
+
+#cam_dist2 <- cam_dist %>%
+#  group_by(data_source) %>%
+#  mutate(mean = mean(mean_mindist))
+
+cam_dist <- arrange(cam_dist, region, data_source, treatment)
+
+# make table
+kbl(cam_dist, booktabs = T, longtable = F, digits = 0,
+  caption = "Summary of experimental camera-trap survey designs around fox control in south-west Victoria, Australia.",
+   col.names = c("Region", "Source", "Block", "Treatment", "Surveys", "Year from", "Year to", "Sites", "Min. spacing (m)", "Mean spacing (m)", "Max spacing (m)")) %>% 
+  kable_styling(font_size = 10, latex_options = "scale_down") 
+```
+
+```{=tex}
+\thispagestyle{empty}
+\elandscape
+\newpage
+```
+
+```{r diel-tab1}
 
 # load records
-records <- read_csv("../derived_data/counts_hour_cleaned.csv")
+records <- read_csv("/Users/ree140/Dropbox/personal/matt/github/phd-thesis/index/data/diel/counts_hour_cleaned.csv")
 # drop hour
@@
 table1 <- table1 %>%
-  select(region, vegetation_group, sites, deployments, fox_count_vr, cat_count_vr)  %>% 
+ dplyr::select(vegetation_group, region, sites, deployments, fox_count_vr, cat_count_vr) %>% 
  distinct() %>%
@@
 
+# get sums for total row
+total <- colSums(Filter(is.numeric, table1))
+# add to df
+table1 <- rbind(table1, total)
+table1[14,1] <- "Total"
+table1[14,2] <- ""
+
 # rename region 
 table1$region <- factor(table1$region)
-levels(table1$region) <- c("Glenelg", "Otway")
+table1$region <- gsub("glenelg", "Glenelg", table1$region)
+table1$region <- gsub("otway", "Otway", table1$region)
+
+# remove repeat vegetation name
+table1[c(2, 4, 6, 8, 10, 12),1] <- ""
 
 # make table
-pander(table1, style = "simple", split.table = Inf, caption = "Summary of camera-trap deployments in unique sites within Ecological Vegetation Class groups and independent counts of predator detections across two broad regions in south-west Victoria, Australia.", round = 2,
-              col.names = c("Region", "Vegetation", "Sites", "Deployments",  "Fox counts", "Cat counts"),
-)
+kbl(table1, longtable = T, booktabs = T, digits = 2,
+  caption = "Summary of the number of camera-trap deployments, unique survey sites and 'independent' counts of invasive predator detections across Ecological Vegetation Class groups within the Glenelg and Otway regions, south-west Victoria, Australia.",
+   col.names = c("Vegetation", "Region", "Sites", "Deployments", "Fox counts", "Cat counts")) %>% 
+  kable_styling(font_size = 10) %>% 
+  row_spec(14, bold = T)
 
@@
 
-```{r veg, out.width="100%", fig.cap = "Fox (A) and feral cat (B) diel activity patterns across Ecological Vegetation Class groups in south-west Victoria, Australia.
-Dotted, vertical lines represent average sunrise and sunset times.
-Shaded areas indicate 95\\% confidence intervals."}
-knitr::include_graphics("../figs/predator_veg.png")
+```{r diel-tab-fits}
+
+# load records
+fits_summaries <- read_csv("../derived_data/model_summaries.csv") %>%
+  dplyr::select(-c(...1))
+
+# make table
+kbl(fits_summaries, longtable = T, booktabs = T, digits = 4,
+  caption = "Generalised additive model summaries for invasive predator spatiotemporal activity in south-west Victoria, Australia.",
+     col.names = c("Species", "Model", "EDF", "dev.expl", "r.sq")) %>% 
+  kable_styling(font_size = 10) %>% 
+  footnote(general = c("EDF - estimated degrees of freedom of all model terms.",
+                       "dev.expl - proportion of the null deviance explained by the model. ",
+                       "r.sq -  adjusted r-squared value."))
+
 ```
@@
 
-```{r count, out.width="100%", fig.cap="Predicted feral cat diel activity (A) and uncertainty (B) across each habitat type in response to foxes (log counts adjusted for survey effort).
-Grey vertical lines respresent average sunrise and sunset times.
+```{r fox-baits, fig.cap = "Effects of 1080 poison-bait density on the spatial activity of red foxes \\textit{Vulpes vulpes}, derived from each model (model 1: spatial, model 2: vegetation group / human footprint, model 3: habitat type).
 Shaded areas indicate 95\\% confidence intervals."}
-knitr::include_graphics("../figs/cat_fox_count.png")
+knitr::include_graphics("../figs/foxbaiting_effects.png")
 ```
 
-\newpage
-
-```{r ind, out.width="100%", fig.cap = "Individual variations in feral cat diel activity patterns across the Glenelg region (A) and wet forests of the western Otway Ranges (B) in south-west Victoria, Australia.
-Dotted, vertical lines represent average sunrise and sunset times.
+```{r fox-hfi, fig.cap = "Effects human footprint index on the spatial activity of red foxes \\textit{Vulpes vulpes}.
+(model 2).
+The interaction effect of human foot print and fox diel activity pattern was removed from this model.
+Both the marginal and interactive effect of human footprint was removed for feral cats \\textit{Felis catus}).
 Shaded areas indicate 95\\% confidence intervals."}
-knitr::include_graphics("../figs/cat_ind.png")
+knitr::include_graphics("../figs/fox_hfi.png")
 ```
-
-\newpage
-
-DISCUSSION
-===================
-
-Foxes likely impacted the spatiotemporal behaviour of feral cats. One of the few studies to actually demonstrate and replicate this across ecosystems. 
-Happens in different ways - context key.
-Spatial only, as well as temporal only models would have had opposite inference - high overlap.
-Joint spatiotemporal analysis is key to understand predator interactions.
-
-Our approach is simple and flexible. 
-Small sample sizes are usually a problem. We demonstrate hierarchical model specification to share information - ability to pool all available data. 
-Our approach is more aligned with null hypothesis testing - small sample sizes shrink to a flat line (or to the average), in contrast to circular overlap. 
-Could be easily adapted to other contexts. E.g. instead of vegetation type, could do individual variation in diel activity.
-
-Combined spatiotemporal modelling - differences to other approaches. 
-Cunningham paper most similar to our approach - different in diel activity only, and linear responses.
-We used a tensor product for fox counts because predator interactions may not be linear [@https://doi.org/10.1111/j.1365-2664.2009.01650.x].
-
-Limitations of our approach. 
-Detectability ignored. 
-Individual heterogeneity ignored.
-Correlation does not = causation. 
-Does not (clearly) seperate numerical responses from behavioural (Borchers SCR paper)
-
-Invasive predators are useful species to test this on, as they are extremely adaptable to different conditions.
-Implications for prey.
-
-Conclusions.
-
-\newpage
-
-ACKNOWLEDGEMENTS
-===================
-We acknowledge and pay respect to the Gadubanud and Gunditjmara people on whose traditional lands this study took place.
-This experiment was a collaborative effort between the Glenelg Ark (Department of Environment, Land Water and Planning) and Otway Ark (Parks Victoria) working groups.
-We are extremely thankful to those who led and assisted with these camera-trap deployments and image processing.
-University of Melbourne led surveys were conducted under University of Melbourne Animal Ethics Committee approval 1714119 and Victorian Government Department of Environment, Land Water and Planning Research Permit 10008273.
-Our study was generously supported by the Conservation Ecology Centre, the Victorian Government Department of Environment, Land Water and Planning, Arthur Rylah Institute for Environmental Research, Parks Victoria and the Australian Government’s National Environmental Science Program through the Threatened Species Recovery Hub, and ARC Linkage Project LP170101134.
-MR also receives support from an Australian Government Research Training Program Scholarship.
-
-OPEN RESEARCH
-===================
-
-Raw data and code are on Github link xx.  
-Data will be deposited on the Dryad Digital Repository after acceptance.
-
-\newpage
-
-REFERENCES {#references .unnumbered}
-==========

### 2023-07-28 manuscript/manuscript.tex
changes following oikos review 1

--- 2021-08-05 manuscript/manuscript.tex
@@
-\documentclass[]{elsarticle} %review=doublespace preprint=single 5p=2 column
+\documentclass[preprint, 3p, authoryear]{elsarticle} %review=doublespace preprint=single 5p=2 column
 %%% Begin My package additions %%%%%%%%%%%%%%%%%%%
+
 \usepackage[hyphens]{url}
 
-  \journal{Ecography} % Sets Journal name
-
-\usepackage{lineno} % add
-  \linenumbers % turns line numbering on
-\providecommand{\tightlist}{%
-  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
+  \journal{Oikos; Article} % Sets Journal name
 
 \usepackage{graphicx}
-\usepackage{booktabs} % book-quality tables
 %%%%%%%%%%%%%%%% end my additions to header
@@
 \usepackage{amssymb,amsmath}
+% TODO: Currently lineno needs to be loaded after amsmath because of conflict
+% https://github.com/latex-lineno/lineno/issues/5
+\usepackage{lineno} % add
 \usepackage{ifxetex,ifluatex}
@@
 \IfFileExists{microtype.sty}{\usepackage{microtype}}{}
-\usepackage[margin=1.1in]{geometry}
-\bibliographystyle{elsarticle-harv}
-\usepackage{longtable}
+
 \ifxetex
@@
             pdfauthor={},
-            pdftitle={Joint spatiotemporal models reveal dramatic shifts in mesopredator diel activity across individuals, habitat types and apex predator activity levels},
+            pdftitle={Spatial variation in predator diel activity patterns},
             colorlinks=false,
@@
             pdfborder={0 0 0}}
-\urlstyle{same}  % don't use monospace font for urls
 
@@
 
+% tightlist command for lists without linebreak
+\providecommand{\tightlist}{%
+  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
+
+% From pandoc table feature
+\usepackage{longtable,booktabs,array}
+\usepackage{calc} % for calculating minipage widths
+% Correct order of tables after \paragraph or \subparagraph
+\usepackage{etoolbox}
+\makeatletter
+\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
+\makeatother
+% Allow footnotes in longtable head/foot
+\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
+\makesavenoteenv{longtable}
+
 % Pandoc citation processing
+\newlength{\cslhangindent}
+\setlength{\cslhangindent}{1.5em}
+\newlength{\csllabelwidth}
+\setlength{\csllabelwidth}{3em}
+\newlength{\cslentryspacingunit} % times entry-spacing
+\setlength{\cslentryspacingunit}{\parskip}
+% for Pandoc 2.8 to 2.10.1
+\newenvironment{cslreferences}%
+  {}%
+  {\par}
+% For Pandoc 2.11+
+\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
+ {% don't indent paragraphs
+  \setlength{\parindent}{0pt}
+  % turn on hanging indent if param 1 is 1
+  \ifodd #1
+  \let\oldpar\par
+  \def\par{\hangindent=\cslhangindent\oldpar}
+  \fi
+  % set entry spacing
+  \setlength{\parskip}{#2\cslentryspacingunit}
+ }%
+ {}
+\usepackage{calc}
+\newcommand{\CSLBlock}[1]{#1\hfill\break}
+\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
+\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
+\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
 
-% Pandoc header
 \usepackage{setspace}\doublespacing
+\usepackage{lineno}
 \usepackage{float}
 \floatplacement{figure}{H}
+\newcommand{\beginsupplement}{\setcounter{table}{0}  \renewcommand{\thetable}{S\arabic{table}} \setcounter{figure}{0} \renewcommand{\thefigure}{S\arabic{figure}} \setcounter{section}{0} \renewcommand{\thesection}{S\arabic{section}}}
+\usepackage{pdflscape}
+\newcommand{\blandscape}{\begin{landscape}}
+\newcommand{\elandscape}{\end{landscape}}
+\usepackage{booktabs}
+\usepackage{longtable}
+\usepackage{array}
+\usepackage{multirow}
+\usepackage{wrapfig}
+\usepackage{float}
+\usepackage{colortbl}
+\usepackage{pdflscape}
+\usepackage{tabu}
+\usepackage{threeparttable}
+\usepackage{threeparttablex}
+\usepackage[normalem]{ulem}
+\usepackage{makecell}
+\usepackage{xcolor}
 
 \begin{document}
+
 \begin{frontmatter}
 
-  \title{Joint spatiotemporal models reveal dramatic shifts in mesopredator diel activity across individuals, habitat types and apex predator activity levels}
-    \author[UOM]{Matthew W. Rees\corref{1}}
+  \title{Spatial variation in predator diel activity patterns}
+    \author[UOM,CSIRO]{Matthew W. Rees%
+  \corref{cor1}%
+  }
    \ead{matt.wayne.rees@gmail.com} 
-    \author[CEC]{Jack H. Pascoe}
+    \author[UOM]{Brendan A. Wintle%
+  %
+  }
 
-    \author[CEC]{Mark Le Pla}
+    \author[UOM,CEC]{Jack H. Pascoe%
+  %
+  }
 
-    \author[CEC]{Emma K. Birnbaum}
+    \author[UOM,CEC]{Mark Le Pla%
+  %
+  }
 
-    \author[ARI]{Alan Robley}
+    \author[CEC]{Emma K. Birnbaum%
+  %
+  }
 
-    \author[CU,TKI]{Nick Golding}
+    \author[UOM]{Bronwyn A. Hradsky%
+  %
+  }
 
-    \author[UOM]{Brendan A. Wintle}
-
-    \author[UOM]{Bronwyn A. Hradsky}
-
-      \address[UOM]{Quantitative \& Applied Ecology Group, School of Ecosystem and Forest Science, The University of Melbourne, Parkville, VIC, Australia}
-    \address[CEC]{Conservation Ecology Centre, Otway Lighthouse Rd, Cape Otway, VIC, Australia}
-    \address[ARI]{Department of Environment, Land, Water and Planning, Arthur Rylah Institute for Environmental Research, Heidelberg, Australia}
-    \address[CU]{Curtin University, Bentley, WA, Australia}
-    \address[TKI]{Telethon Kids Institute, Perth Children's Hospital, Nedlands, WA, Australia}
-      \cortext[1]{Corresponding Author}
+      \affiliation[UOM]{Quantitative and Applied Ecology Group, School of Ecosystem and Forest Sciences, The University of Melbourne, Parkville, VIC, 3010, Australia}
+    \affiliation[CSIRO]{Health and Biosecurity, Commonwealth Scientific and Industrial Research Organisation, Ecosciences Precinct, Dutton Park, QLD, Australia}
+    \affiliation[CEC]{Conservation Ecology Centre, Lighthouse Rd, Cape Otway, VIC, 3233, Australia}
+    \cortext[cor1]{Corresponding author}
 
   \begin{abstract}
-  \begin{enumerate}
-  \def\labelenumi{\arabic{enumi}.}
-  \tightlist
-  \item
-    Background.
-  \item
-    Methods.
-  \item
-    Results.
-  \item
-    \emph{Synthesis and applications.}
-  \end{enumerate}
+  \emph{Open research:} Data and code will be deposited on Dryad upon acceptance and can currently be found at this link: https://github.com/matt-w-rees/spatiotemporal-gams-invasive-predators.
   \end{abstract}
-   \begin{keyword} daily activity patterns; camera trap; \emph{Felis catus}; invasive predator; interspecific competition; mesopredator release; diel activity; hierarchical generalised additive models; species interactions; \emph{Vulpes vulpes}.\end{keyword}
+    \begin{keyword}
+    diel activity patterns \sep generalised additive model \sep feral cat \sep invasive predator \sep intraguild predator interactions \sep lethal predator control \sep mesopredator release \sep red fox \sep
+    spatiotemporal model
+  \end{keyword}
+
  \end{frontmatter}
@@
 \parskip=12pt 
+\newpage
+\setcounter{page}{1}
+
+\hypertarget{spatial-variation-in-predator-diel-activity-patterns}{%
+\section*{Spatial variation in predator diel activity patterns}\label{spatial-variation-in-predator-diel-activity-patterns}}
+\addcontentsline{toc}{section}{Spatial variation in predator diel activity patterns}
+
+\(~\)
+
+\linenumbers
+
+\hypertarget{abstract}{%
+\section*{ABSTRACT}\label{abstract}}
+\addcontentsline{toc}{section}{ABSTRACT}
+
+Understanding the constraints dominant predators impose on subordinate species is important for predicting ecosystem dynamics and anticipating outcomes of predator management.
+Subordinate predators may avoid dominant predators in time or space, making it difficult to quantify antipredator behaviours unless joint spatiotemporal analyses are used.
+Here, we tested whether an invasive dominant predator (red fox \emph{Vulpes vulpes}) alters the spatiotemporal activity of an invasive subordinate predator (feral cat \emph{Felis catus}).
+We collated records of both species from 3,667 camera-traps deployed experimentally across two regions of south-eastern Australia with simplified predator guilds; foxes were poison-baited in some landscapes within each region.
+We used generalised additive models to quantify changes in predator spatiotemporal activity across geographic space, vegetation types, human footprint and (artificially manipulated) gradients of dominant predator activity.
+Foxes and cats had similar diel activity patterns when averaged across all sites, however there was important differentiation at a finer scale---cats did not reduce their spatial activity but shifted diel patterns when localised fox activity was high.
+Cats were crepuscular on average.
+However, across dry vegetation types of both regions (where foxes were nocturnal), cats shifted to diurnal behaviour with increasing fox activity.
+In contrast, fox activity was relatively consistent throughout the daily cycle in the wet forest; here cats avoided dawn when fox activity was high.
+Changes in cat diel activity patterns may facilitate spatial coexistence between these two invasive predators, potentially shifting feral cat impacts onto different native prey.
+It is well-appreciated that predator activity varies spatially and fluctuates throughout the daily cycle.
+However, our study demonstrates that diel activity patterns also vary across space, likely mediated by both landscape-context and fear.
+dominant predator avoidance appears to be dynamic---a key nuance which is overlooked when simply comparing the average activity overlap between two species.
 
@@
 
-Effective wildlife management relies on understanding species population and behavioural dynamics to accurately predict intervention outcomes.
-These dynamics depend on both environmental conditions and those of other species, but few statistical approaches incorporate both factors, or allow responses to be dynamic across space and time.
-This is key because species interactions and niche preferences are multidimensional.
+Top predators can drive ecosystem structure and dynamics by suppressing populations of rival predators and prey species (Prugh et al.
+2009, Estes et al.
+2011).
+Dominant predators can suppress subordinate species directly through antagonistic interactions and predation, as well as indirectly through fear (Creel and Christianson 2008, Allen et al.
+2022).
+Fear-induced behavioural suppression can be as detrimental to subordinate species as predation itself, by limiting resource acquisition and breeding success (Brown et al.
+1999, Schmitz et al.
+2004, Preisser et al.
+2005).
+Understanding how top predators---including humans---constrain the behaviour of subordinate species is therefore important to accurately predict consequences of predator management, such as reintroductions or lethal control (Suraci et al.
+2016, Gaynor et al.
+2021).
 
-Fear is a near universal driver of animal behaviour.
-Fear may be from other individuals of their species, a dominant species (including humans) or an event (e.g.~fire).
-Animals may be naive to threats (e.g.~alien predators), unable to respond to threats (e.g.~scarce prey), or the threat may be too minimal to warrant a response.
-Some environments more easily facilitate avoidance (e.g.~high food and shelter accessibility), and so accounting for landscape context is key.
-A threat response may be routine, or only employed or enhanced when the threat is heightened.
-This likely depends on how constant and predictable the threat is.
+Spatial and/or temporal niche partitioning may allow subordinate species to coexist with top predators by reducing encounter-rates and resource overlap (Kronfeld-Schor and Dayan 2003).
+However, subordinate species may not consistently employ avoidance behaviours because perceived predation risk is spatiotemporally variable, and antipredator behaviours typically involve a trade-off against resource acquisition, such as limiting or relegating activity to suboptimal places or times (Lima and Dill 1990, Lima and Bednekoff 1999).
+Therefore, optimal predator avoidance strategies likely vary across heterogeneous landscapes where resource availability (e.g., shelter, food) and perceived predation risks differ (Kauffman et al.
+2007, Willems and Hill 2009, Davies et al.
+2021, Wirsing et al.
+2021).
+For example, temporal predator avoidance may be preferable over spatial avoidance if food is constantly available throughout the day, and vice versa.
+These theories are unified under the ecology of fear concept (Brown et al.
+1999), which has gained increasing attention in recent times (Gaynor et al.
+2019).
+However, there remains a lack of robust, empirical evidence documenting dynamic changes in antipredator behaviour.
 
-Animals may respond to threats by altering their activity patterns (i.e.~avoidance).
-An animal may avoid locations where the threat is likely to occur (hereafter `spatial avoidance'), avoid times of the day when a threat is likely to occur (hereafter `temporal avoidance'), or avoid both locations and times a threat is heightened (hereafter `spatiotemporal avoidance').
-Such avoidance likely has trade-offs, which may or may not incur costs to individual and population fitness.
-Modelling approaches which do not allow spatiotemporal and dynamic fear reponses can only make limited inference about species interactions.
+Animal diel activity patterns are increasingly being considered alongside spatial activity and distributional patterns.
+However, spatial and temporal (in regard to the daily cycle) activity patterns are still often considered in an ad-hoc fashion.
+For example, by fitting separate models for spatial and temporal overlap, or by repeating spatial analyses (e.g., resource selection functions) at different time periods such night and day (e.g., Basille et al.
+2015, Kohl et al.
+2019, Smith et al.
+2019, Wooster et al.
+2022).
+Such ad-hoc approaches make it difficult to assess the relative importance, as well as dynamic changes in spatial and temporal processes (Suraci et al.
+2022).
 
-Species interactions are usually considered in degrees of overlap between activity patterns.
-Spatiotemporal avoidance is most often considered, using joint species distribution models (including MSOMs).
-However, temporal avoidance is increasingly being used (perhaps just allowed due to widely available GPS and camera technology) with kernel density estimation methods.
-Although, these methods don't allow environmental covariates, and are particularly susceptible to noisy estimates.
-Importantly, some are beginning to combine these methods to allow spatiotemporal responses (Azzou et al 2021; Farris et al 2020).
-However, degrees of overlap in activity patterns only get at routine avoidance, we are generally interested in whether changes in threat levels lead to altered behaviour.
-This is more indicative of species interactions because static levels of overlap may just reflect different niche preferences rather than avoidance.
+There are many approaches to modelling intraguild spatiotemporal interactions, such as time-to-encounter models and coefficients of activity overlap (Frey et al.
+2017).
+Time-to-encounter approaches consider fine-scale avoidance or attraction by using time intervals between species detections as the response variable, although do not estimate diel activity patterns (Harmsen et al.
+2009, Fancourt et al.
+2019, Niedballa et al.
+2019).
+Multispecies models have been recently developed to assess overlap in spatial and diel activity between two species (Ait Kaci Azzou et al.
+2021), as well as different diel activity patterns based on the presence or absence of another species at a site (Kellner et al.
+2022).
+Multispecies models are data-hungry and difficult to fit, but can account for imperfect detection and different types of predictor variables (e.g., Parsons et al.
+2022).
+Given avoidance behaviours are likely dynamic (Gaynor et al.
+2019), there is need for statistical approaches which jointly assess continuous shifts in spatial and diel activity patterns of one species in response to another, particularly when combined with experimental data (Suraci et al.
+2022).
 
-Dynamic behavioural responses are also important from a management perspective because they want to know whether an action will cause a change in animal behaviour / range?).
-For example, after the reintroduction of wolves in Yellowstone, it was widely thought that this had limited elk populations.
-Using GPS data and habitat selection methods, Kohl et al.~2018 found that elk used riskier parts of landscape at times when wolves were least active -- suggesting coexistence without fitness costs.
-However, there are many other large carnivores in this ecosystem which are likely to impact elk behaviour.
-And so, will elk respond to wolves in the same way in different environmental contexts?
-There are little-no examples of replicated spatiotemporal interactions studies.
+Generalised Additive Models (hereafter `GAMs') are increasingly being used to estimate animal diel activity patterns and offer a flexible framework to jointly consider spatial activity.
+GAMs are useful to model diel activity patterns as they can implement cyclical splines---joining the end with the start of the day---as well as complexity penalties to reduce overfitting.
+If complex nonlinear (i.e., wiggly) effects are not statistically supported, smoothing penalties in GAMs `shrink' them to a linear (i.e., no wiggliness) effect, or even to a horizontal line - removing the entire effect of the predictor variable and therefore the need for additional tests of statistical significance (such as p-values and information criteria).
+GAMS can also capture nonlinear interactions between multiple variables with different units (i.e., `tensor products'), and offer the ability to share information across categorical variables through hierarchical specifications (Wood 2017, Pedersen et al.
+2019).
+However, we are only aware of one study which allowed animal diel activity to interact with predation risk as a continuous variable in a GAM (although without considering overall activity, Cunningham et al.
+2019).
 
-There is contention around the interactions between two invasive predators - feral cats and foxes.
-Foxes are routinely culled, and there is concern that this could lead to a mesopredator release of feral cats.
-Mesopredator release may manifest in both behavioural and numerical changes, which my previous chapter identified - but that behavioral responses (ranging and detectability) were not universal.
-Here, we aim to dig into the behavioral response further - in particular diel activity times.
+The red fox \emph{Vulpes vulpes} (hereafter `fox') and feral cat \emph{Felis catus} (hereafter `cat') have devastating impacts on native prey throughout their introduced range, implicated in the extinction of \textasciitilde10 and 63 species, respectively (Doherty et al.
+2016).
+The impacts of these invasive predators are particularly extreme on the Australian continent (Woinarski et al.
+2015).
+Cats are more difficult to manage, and so introduced predator control programs in Australia often target only foxes (particularly through poison-baiting, Reddiex et al.
+2007).
+As foxes and cats compete for many of the same resources (Fleming et al.
+2022), there is concern that lethal fox control could cause mesopredator release (Soulé et al.
+1988) of cats (Glen and Dickman 2005, Doherty and Ritchie 2017, Wayne et al.
+2017).
+However, there is limited evidence that spatial cat activity increases in response to fox control (Hunter et al.
+2018), although there are signs of changes in behaviour and population density (Molsher et al.
+2017, Rees et al.
+2023a).
+Other studies have investigated potential spatial and temporal interactions between these invasive predators (e.g., Roshier and Carter 2021), but not in response to fox control, or in a joint spatiotemporal framework that allows flexibility in cat avoidance behaviours with respect to differences in fox activity.
+Conservation managers are therefore uncertain as to whether fox control may increase the impacts of feral cats on native prey, through increases in cat density or behavioural changes which improve hunting success.
 
-In this study, we used generalised additive models to assess (1) how predator (fox and cat) diel activity varies between vegetation types, and (2) whether an introduced mesopredator (cat) changes its diel and/or spatial activity in response to the activity of an introduced apex predator (fox).
-Because foxes and cats fill a similar function niche, we expected that, on average, they would have similar diel activity patterns across vegetation types (hypothesis 1).
-However, where fox detections were high, we hypothesised that the subordinate cat to be less active (hypothesis 2) and shift diel activity to times when foxes are least active (hypothesis 3).
-We tested this on a simple predator system where only cats and foxes occur, and fox activity is manipulated through lethal control.
-This allowed focus on the sole interactions between these two predators, and ensured differences in correlations wasn't just due to differences in niche preferences.
+In this study, we first explored (1) how predator diel activity patterns vary across geographic space (`model 1'), and (2) the role of vegetation type and human footprint in predator activity patterns (`model 2').
+Using this knowledge to partition our dataset, we then (3) tested whether the activity of a dominant predator (fox) affected the spatiotemporal activity of a subordinate predator (cat; `model 3').
+In line with behavioural avoidance or exclusion, we predicted that cat diel activity pattens would shift towards less-risky times of day (i.e., times when fox activity was lower), and/or spatial cat activity would be lower at sites where fox activity was relatively high.
+We illustrate how GAMs can provide a simple framework to jointly assess spatial and temporal animal activity patterns, as well as potential species interactions.
 
@@
 
-\hypertarget{study-area}{%
-\subsection{Study area}\label{study-area}}
+\hypertarget{study-area-and-camera-trapping}{%
+\subsection{Study area and camera-trapping}\label{study-area-and-camera-trapping}}
 
-We collated camera-trap data on these two predator species from three distinct studies across two regions of south-west Victoria, Australia (Fig.
-1).
-The purpose of each study was to experimentally survey changes in the mammalian community due to fox control, and so our dataset covered a wide range contexts with and without fox-baiting, allowing us to investigate fox-cat interactions across a gradient of apex predator (fox) activity.
+Our study was conducted in a simple predator system where foxes and cats are the only medium-large (\textgreater{} 1 kg) mammalian predators, and fox activity is manipulated using lethal control in some landscapes.
+This situation allowed sole focus on the interactions between these two predators, across an experimental gradient of dominant predator (fox) activity.
+We compiled data from multiple smaller-scale camera-trap studies, each designed to experimentally assess mammal responses to fox control.
+Overall, we collated 5,449 and 2,202 independent detections of foxes and cats, respectively (separated by at least 30 minutes) from 172,052 camera-trap nights (Table \ref{tab:diel-tab1}).
 
-The terrestrial mammalian predator guild is depauperate throughout both regions.
-Dingoes \emph{Canis familiaris} are now absent throughout, as is the the tiger quoll \emph{Dasyurus maculatus}, more recently absent in the Otway Ranges (last sighted in 2014 despite extensive camera-trapping).
-Introduced red foxes and feral cats are therefore the only medium-large functional mammalian terrestrial predators across both regions.
+We collated camera-trap data across two regions in south-west Victoria, Australia: the Glenelg region and Otway Ranges (Fig.
+\ref{fig:diel-map}).
+Here, dingoes \emph{Canis familiaris} are long-absent throughout, while tiger quolls \emph{Dasyurus maculatus} are long-absent in the Glenelg region and likely functionally extinct in the Otway Ranges (last confirmed sighting in 2014).
+In broad sections of each region, government land managers conduct ongoing targeted lethal fox control for biodiversity conservation (see sections below for region-specific details).
+Poison-baits containing 3 mg of sodium fluroacetate (compound 1080) are buried at a depth of 12 - 15 cm at 1-km intervals along accessible forest tracks and roads.
+Different road densities therefore result in variable densities of poison-baits.
+Foxes readily dig these poison-baits often resulting in population suppression, whereas it is generally accepted that cats are unlikely to consume buried fox-baits and there is no evidence buried 1080 fox-baits suppresses cat populations (Risbey et al.
+1997, Algar and Burrows 2004, Moseby et al.
+2011, Hunter et al.
+2018).
+Managers also frequently implement prescribed fire across both regions, primarily to reduce fuel loads to prevent large wildfires.
 
-In the baited parts of the regions, government land managers conduct ongoing targeted fox control for biodiversity conservation.
-Poison baits containing 3 mg of sodium mono-fluroacetate (1080) are buried at a depth of 10 cm at 1-km intervals along accessible forest tracks and roads.
+\hypertarget{glenelg-region}{%
+\subsubsection{Glenelg region}\label{glenelg-region}}
 
-In the Glenelg region, Gunditjmara country, natural vegetation is highly fragmented.
-Here, camera-survey sites were situated in six distinct forest blocks: two are separated by a river, the others are separated by at least \emph{XX} km of pastoral farming and residential property.
-Foxes in three of the forest blocks have been subject to lethal control since October 2005, with 1080 poison-baits replaced fortnightly (Robley et al.
+In the Glenelg region, large patches of natural vegetation are fragmented by pastoral farming and residential properties (Fig.
+\ref{fig:diel-map}).
+Foxes in three distinct forest blocks in this region have been subject to poison-baiting since October 2005, with fortnightly bait replacements (Robley et al.
 2014).
+These forest blocks, along with three similar, unbaited forest blocks to the north have been simultaneously surveyed annually under the `Glenelg Ark' fox control program since 2005 (40 sites per block, Robley et al.
+2020).
+Hair-tubes were used to monitor species from 2005 - 2013 (presented in Robley et al.
+2014), replaced by camera-traps from 2013; here we present camera-trap data from 2013 - 2019 (Robley et al.
+2020).
+We also included a further 425 camera-trap deployments at unique locations from early 2018 (M.W.R PhD Surveys', Rees et al.
+2023b).
+This totals 2,039 camera-trap deployments in the Glenelg region, collected in a control-impact experimental design (foxes had been continuously controlled for at 8 - 14 years in the treatment landscapes at the time of these surveys).
 
-The Otway Ranges, Gadubanud country, is a largely continuous patch of natural vegetation with a strong east-west rainfall and elevation gradient (Swan et al.
-2021).
-Foxes are controlled through 1080 poison-baits with monthly replacement across most of the Otway Ranges, but a large section to the north-west remains unbaited.
-The majority of this baiting began in 2016 - 2017, although fox-baiting commenced in some small sections in 2008.
+\hypertarget{otway-ranges}{%
+\subsubsection{Otway Ranges}\label{otway-ranges}}
 
-Alongside fragmentation, the extremely rugged, wet forest and rainforest matrix in the western Otway Ranges is the main difference between the two regions.
-Prescribed fire, primarily for the purpose of asset protection, is another key management action undertaken throughout both regions (except in very wet sections of Otway Ranges).
+The Otway Ranges is a largely continuous patch of natural vegetation with a strong east-west rainfall gradient (Fig.
+\ref{fig:diel-map}).
+A matrix of cool temperate rainforest and wet forest at high altitudes in the south-west descend into a large heathland directly north, and into dry forests and then heathlands to the north-east.
+Fox-baiting commenced in small sections of the Otway Ranges in 2008 and large-scale systematic baiting began in 2016 - 2017 under the `Otway Ark' program (Robley et al.
+2019).
+For the first six weeks, poison-baits were replaced weekly, then changing to ongoing monthly bait-replacement.
+There was a pause in baiting for approximately six months during the second half of 2018.
+Fox control recommenced in late 2018 with four weeks of fortnightly bait-replacement, before returning to monthly bait-replacement.
+A large section of the Otway Ranges to the north-west remains unbaited, but is monitored as an experimental non-treatment site (Robley et al.
+2019).
+Otway Ark managers survey 372 camera-trap sites annually (sequentially across the region); we present one `before' baiting survey and two `after' baiting surveys of each site from 2016 - 2018, totalling 1,113 camera-trap deployments (Robley et al.
+2019).
+We also include data from an additional before-after control-impact surveys (one `before' baiting survey and two `after' baiting surveys) in the western section of the Otway Ranges, conducted annually 2017 - 2019 (Rees et al.
+2023b).
+This added a further 195 sites and 524 camera-trap deployments (Table \ref{tab:tab-sum}).
 
-\hypertarget{camera-trap-surveys}{%
-\subsection{Camera-trap surveys}\label{camera-trap-surveys}}
+\hypertarget{camera-trap-set-ups}{%
+\subsubsection{Camera-trap set-ups}\label{camera-trap-set-ups}}
 
-We compiled camera-trap data from three distinct studies across the two regions.
-After removing camera-traps which were operational for less than 14 days, this totalled 3659 deployments of camera-traps at 1232 camera-trap sites.
-Survey durations ranged between 14 and 93 days (mean 47), totalling 171,966 trapnights.
-Camera-trap spacing was variable; the average minimum distance between simultaneously deployed camera-traps was 445, 853 and 1266 metres for each study.
-Camera-trap were deployed in the Glenelg region between 2013 -- 2019, whereas in the Otway Ranges surveys were conducted in 2016 -- 2019.
-
-All camera-trap deployments consisted of a Reconyx (Holmen, Wisconsin) brand camera-trap (both white and infrared flashes), attached to a tree or a metal picket, facing a lure.
-Camera-traps were set-up in two ways depending on the key aim of the project, either targeted toward predators or prey species.
-One study across both regions positioned camera-traps lower on a tree (around 15 - 30 cm above the ground -- angled only slightly downwards) facing a tuna oil lure approximately 2 - 2.5 m away (detailed in Rees et al.
-2019).
-The remaining camera-traps were positioned higher on a tree or a metal picket (at least 40 cm above ground) and angled downwards more strongly - facing a lure approximately 1 - 1.5 m away.
-These lures consisted of peanut butter, golden syrup and rolled oats mixed into a small ball, placed within a tea strainer or PVC pipe container and secured either to the ground, or 20 - 60 cm above ground on a wooden stake.
-All set-ups were effective in detecting both predator species.
+All camera-trap deployments consisted of a Reconyx (Holmen, Wisconsin) brand camera-trap (white or infrared flash), attached to a tree or a metal picket, facing a lure.
+All camera-trap sites were positioned in the forest interior at least 30 m away from roads.
+The Glenelg Ark and Otway Ark fox monitoring programs positioned camera-traps at least 40 cm above ground on a tree or a metal picket and angled downwards toward a lure approximately 1 - 1.5 m away (Robley et al.
+2019, 2020).
+The lures consisted of peanut butter, golden syrup and rolled oats mixed into a small ball, placed within a tea strainer or PVC pipe container and secured either to the ground, or 20 - 60 cm above ground on a wooden stake.
+The M.W.R PhD surveys across both regions positioned camera-traps lower on a tree (around 15 - 30 cm above the ground) angled only slightly downwards toward a tuna oil lure approximately 2 - 2.5 m away (detailed in Rees et al.
+2023a).
+Camera-traps were active for an average of 47 days (maximum 93 days), totalling 172,052 trap-nights.
+Camera-trap spacing averaged 867 m for Glenelg Ark, 1246 m for Otway Ark, and 444 m for the M.W.R PhD surveys (Table \ref{tab:tab-sum}).
 
@@
 
-We first created a table of species detections and deployment information (coordinates, dates) for every camera-trap deployment.
-All data analysis was conducted in R version 3.6.3 (R Core Team 2020).
-We used lorelograms to identify the minimum interval to approximate independence (Iannarilli et al.
-2019); this indicated 30 minutes discarding repeat detections of a species within 30 minutes was sufficient in reducing temporal autocorrelation.
+Analyses were conducted in R version 4.1.3 (R Core Team 2020).
+We first used lorelograms to identify the minimum interval to approximate independence (Iannarilli et al.
+2019); this indicated that discarding repeat detections of a species within 30 minutes was sufficient to reduce temporal autocorrelation.
 To account for day length variation across space and time, we extracted sunrise and sunset times for each camera-trap deployment using the `maptools' R-package (Bivand and Lewin-Koh 2021) and adjusted detection times to be relative to sunrise and sunset using the average double anchoring approach described by Vazquez et al.
 (2019).
-
-Using the `reshape' R-package (Wickham 2007), we manipulated the detection table into a dataframe with a row for each hour of the day, for every camera-trap deployment, recording the total number of `independent' fox and feral cat detections within each hour for the entire camera-trap survey duration.
-
-We identified the broad vegetation type (Ecological Vegetation Class (hereafter `EVC') group (DELWP 2020)) at each camera-trap location.
-Six vegetation types were surveyed in both regions, although to varying degrees (Table 1).
-In the Otway region, rainforests are interspersed (primarily in low lying gullies) throughout wet and damp forests at a finescale, and we merged these groups (hereafter `wet forests').
-
-For each deployment, we calculated the number of poison fox-baits within a 2.3 km radius around each camera-trap - the average maximum distance foxes in these regions travel from their home range centre (Hradsky et al.
-2017).
+We then built a dataframe consisting of a row for each hour of the day (0 -- 23), for every camera-trap deployment (n = 3,667), recording the total number of `independent' fox and feral cat detections within each hour across the camera-trap survey.
 
@@
 
-We modelled predator counts (response variable) with generalised additive mixed-effects models (hereafter `GAMMs') implemented in the `mgcv' R-package (Wood 2017).
-We first fit these models using the poisson family, but switched to the negative binomial family as overdispersion, but not zero-inflation, was detected using the `DHARMa' R-package (Hartig 2020).
-We specified a model offset to account for differences in camera-trap survey duration, and a random intercept for each site to account for repeat sampling.
-We used a duchon spline static spatial smooth to model both spatial autocorrelation and differences in predator activity across space (i.e.~the effect of unmodelled environmental covariates).
-For the fox model, we additionally included a thin plate regression spline for poison-bait density, with a factor smoother for each region, as bait replacement schedules differed between the regions.
-This formed the base model specification for each model we fitted.
-Models (n = 2) differed in their specification of the cyclical hour smooth in order to explore (1) variations in predator diel activity across EVC groups, as well as (2) whether feral cat diel activity changed across the gradient of observed fox counts (detailed in the sections below).
-We used the double penalty model selection approach, which penalises the null space in addition to the range space (i.e.~shrinking wiggly terms to linear terms) of the spline basis, meaning covariate effects can be entirely removed from the model (Marra and Wood 2011).
-We plotted models using `ggplot2' and `gratia' R-packages (Wickham 2016, Simpson 2021).
+We modelled the total number of independent detections of each predator per hour for each camera-trap deployment (response variable) with generalised additive mixed-effect models implemented in the `mgcv' R-package (Wood 2017).
+We used the negative binomial family, as overdispersion, but not zero-inflation, was detected with a poisson distribution using the `DHARMa' R-package (Hartig 2020).
+We specified the natural log of the number of survey days as a model offset to account for differences in camera-trap survey duration, and a random intercept for each site to account for repeat sampling for models 2 and 3 (as pseudoreplication from repeat sampling in model 1 was explicitly accounted for using the spatial term).
+For fox models, we also included a smooth effect of poison-bait density (number of baits within a 2.3 km radius around each camera-trap - the average maximum distance adult foxes travel from their home-range centre in these regions, Hradsky et al.
+2017a) with separate responses per region to account for the effect of fox control.
+These specifications were included in each model we fitted; models differed in their specification of the cyclical hour smooth to provide inference on variations of predator diel activity across the three questions of interest - this is detailed in the sections below.
+We used the `ggplot2' (Wickham 2016) and `gratia' (Simpson 2021) R-packages to plot models.
 
-We expected the activity of each predator species to follow a broadly consistent pattern (e.g.~lower during the day), but potentially have small shifts in relative activity between sunset and sunrise across the different EVC groups (1).
-Fitting separate responses for each EVC group would throw away shared information, which is particularly costly for the least surveyed groups, and would result in more noisy estimates.
-We therefore specified a global smoother for hour (i.e.~average response), along with group-level smoothers with shared wiggliness for each EVC group (i.e.~model GS in Pedersen et al.
+In this paper, we use the term `spatial activity' to refer to the number of `independent' predator detections at a site (offset to account for survey duration; analogous to an activity or abundance index), and `diel activity pattern' to refer to fluctuations in relative activity throughout the 24-hour daily cycle.
+The first two models we fit explored how predator spatiotemporal activity varied across these regions, informing the best structure for our third model which tested the activity of foxes limits the spatiotemporal activity of cats (detailed in the sections below).
+
+\hypertarget{how-does-predator-diel-activity-patterns-vary-across-geographic-space-model-1}{%
+\subsubsection{How does predator diel activity patterns vary across geographic space? (model 1)}\label{how-does-predator-diel-activity-patterns-vary-across-geographic-space-model-1}}
+
+To examine how the spatial activity and diel activity patterns of each predator varied across space, we fit a model for each predator which included a tensor product interaction between a spatial smooth (thin plate regression spline) and hourly smooth.
+This model specification allowed predators to have different activity levels across space (static across the years surveyed), as well as variation in diel activity pattern across space.
+Space was modelled using camera-trap coordinates and a duchon spline basis (Miller and Wood 2014).
+
+\hypertarget{how-does-vegetation-type-and-human-footprint-affect-predator-activity-patterns-model-2}{%
+\subsubsection{How does vegetation type and human footprint affect predator activity patterns? (model 2)}\label{how-does-vegetation-type-and-human-footprint-affect-predator-activity-patterns-model-2}}
+
+The spatial term in model 1 estimated differences in predator activity across geographic space; we hypothesised that this was partly due to differences in vegetation type, based both on the observed spatial patterns and because vegetation type is a major driver of understorey habitat structure and prey occurrence in these regions (Swan et al.
+2015, Hradsky et al.
+2017b).
+To test whether the diel activity pattern of each predator varied among vegetation types, we identified the Ecological Vegetation Class group (hereafter `vegetation type' - standard units for vegetation classification in Victoria, Department of Environment, Land, Water \& Planning 2020) for each unique camera-trap site, totalling eight vegetation types.
+As rainforests are interspersed (primarily in low lying gullies) at fine-scales throughout wet and damp forests in the south-eastern Otway Ranges, we merged them together (hereafter referred to as `wet forests').
+We then estimated predator activity across vegetation types using a hierarchical model specification: a global smoother for hour (i.e., average response) and group-level smoothers with shared wiggliness for the seven vegetation types (`model GS' detailed in Pedersen et al.
 2019).
-This model structure penalises functions which deviate strongly from the average response, increasing confidence that they are true deviations opposed to the result of sampling noise.
+We also fit a tensor product interaction term for hour and the human footprint index (Venter et al.
+2016, 2018); as human footprint is known to impact animal activity patterns (Gaynor et al.
+2018, Van Scoyoc et al.
+2023)
 
-To investigate changes in feral cat diel activity in response to counts of foxes.
-We added an extra column to our dataframe with the number of fox detections divided by the log of the number of survey days to account for differences in camera-trap survey durations (hereafter `adjusted fox counts').
+\hypertarget{do-feral-cats-avoid-foxes-in-space-or-time-model-3}{%
+\subsubsection{Do feral cats avoid foxes in space or time? (model 3)}\label{do-feral-cats-avoid-foxes-in-space-or-time-model-3}}
 
-Analysis of stage 1 showed that fox diel activity patterns in wet forests differed relative to all other EVC groups.
-We therefore expected feral cats might to temporally avoid foxes in a different way in wet forests than the dry vegetation groups, but that responses would be similar in all the drier vegetation groups.
-We therefore modelled fox-induced changes in feral cat diel activity separately for wet and dry vegetation groups.
-We further split dry vegetation groups by region, to examine whether patterns remained consistent across space?
-for spatial replication.
-We refer to the resulting variable as `habitat type', which had three levels: wet forests of the Otway Ranges (`wet\_otways') dry EVC groups of the Otway Ranges (`dry\_otways') and all of the Glenelg region data (`dry\_glenelg').
+Fox diel activity across vegetation types showed strong similarity between all vegetation types except wet forests.
+To examine whether cats avoid foxes in space or time, we therefore modelled fox-induced changes in feral cat diel activity separately for wet forest and dry vegetation types.
+We further split dry vegetation types by region for further replication.
+We refer to the resulting variable as `habitat type', which had three levels: (i) wet forests and rainforests in the western Otway Ranges (`wet\_otways'), (ii) dry vegetation types in the Otway Ranges (`dry\_otways') and (iii) dry vegetation types in the Glenelg region (`dry\_glenelg').
+We further hypothesised that cats would avoid foxes in time by becoming more diurnal in dry vegetation types where foxes were mostly nocturnal, but not in wet forests where fox activity showed little variation across the daily cycle.
 
-We used a tensor product of hour (using cyclic cubic regression spline) and log adjusted fox count (using thin plate regression spline) smooths to model feral cat diel activity across the range of observed fox activity, separately for each habitat type.
-By also having double penalty shrinkage turned on, this meant that feral cat activity could vary nonlinearly and be penalised consistently across both dimensions.
-This model specification allows five different scenarios: that there was (1) no effect of hour or foxes on feral cats (2) a static hourly effect only, (3) a spatial response to foxes only, (4) a spatial response to foxes and an unrelated static hourly effect, (5) a spatial response to foxes and a hourly effect which changes across the range of fox counts.
+To investigate changes in feral cat diel activity across the range of observed fox activity, we first quantified fox activity for each camera trap deployment as the total number of fox detections for the deployment divided by the number of survey days, to adjust for differences in survey duration (hereafter `adjusted fox counts').
+We modelled an interaction between hour and adjusted fox counts (as a thin plate regression spline with shrinkage - so complexity penalties could entirely remove this effect from the model if not supported), allowing cats to have nonlinear responses to both hour and adjusted fox counts.
+We fit separate tensor product interactions for each habitat type (using a `by-variable' term).
+For a direct visual comparison to fox activity, we fit another fox model where a diel activity curve was estimated separately across each of the three habitat types.
 
@@
 \begin{figure}
-\includegraphics[width=1\linewidth]{../figs/fig1_map} \caption{Locations of study sites in south-west Victoria, Australia.
+
+{\centering \includegraphics[width=0.8\linewidth]{../figs/map_cams}
+
+}
+
+\caption{Locations of our study regions in south-west Victoria, Australia.
 The grids of camera-traps are denoted by white dots.
-The Glenelg region is to the west and Otway region to the east.
+The Glenelg region (38°05'54''S 141°44'41''E) is to the west and Otway Ranges (38°57'82''S 141°68'41''E) to the east.
 Native vegetation is indicated by dark green, with hill shading.
-Map tiles by Stamen Design, under CC BY 3.0, map data by OpenStreetMap, under CC BY SA.}\label{fig:map}
+\emph{Map tiles by Stamen Design, under CC BY 3.0, map data by OpenStreetMap, under CC BY SA.}.}\label{fig:diel-map}
 \end{figure}
@@
 
-\hypertarget{predator-diel-activity-across-vegetation-types}{%
-\subsection{Predator diel activity across vegetation types}\label{predator-diel-activity-across-vegetation-types}}
+\hypertarget{lethal-suppression-of-foxes}{%
+\subsection{Lethal suppression of foxes}\label{lethal-suppression-of-foxes}}
 
-On average, both predator species occupied similar times of the day--mostly nocturnal with activity peaks around sunrise and sunset.
-The main difference was that fox activity peaked just after sunset rather than on sunset.
-For foxes, this diel pattern was largely exhibited across all vegetation types except wet forest.
-In wet forests and rainforests, the hour of the day had little effect on foxes (especially considering diel activity curve here is being drawn towards a nocturnal pattern).
+In each model, there was a strong effect of poison fox-baiting on fox spatial activity, particularly in the Glenelg region (Supporting Information Fig.
+\ref{fig:fox-baits}b).
+Predicted fox counts on the response scale were reduced by a factor of 12, 24, and 28 in the Glenelg region, as well as a factor of 13, 13, and 8 in the Otway Ranges, for models 1, 2, and 3, respectively.
+For example, model 2 predicted fox counts in the Glenelg region to range from 0.097 (95\% CI: 0.077 - 0.122) in sites with no fox-baiting, to 0.004 (95\% CI: 0.002 - 0.012) at sites with the maximum poison-bait density; in the Otways ranging from 0.041 (95\% CI: 0.033 - 0.05) - 0.008 (95\% CI: 0.003 - 0.020) across the gradient of poison-bait density.
 
-lowest activity during the middle of the day, peaking at sunrise and sunset times, while remaining relatively high at night and dawn.
-Foxes had a stronger diel activity patterns than cats, much less likely to be out during the day.
+\hypertarget{how-does-predator-diel-activity-patterns-vary-across-geographic-space-model-1-1}{%
+\subsection{How does predator diel activity patterns vary across geographic space? (model 1)}\label{how-does-predator-diel-activity-patterns-vary-across-geographic-space-model-1-1}}
 
-Diel activity patterns were similar across the different vegetation types for each predator,, except for wet forests, where more substantial changes were seen.
-In the wet forests, ox activity remained fairly constant throughout day and night barely had any diel activity pattern here, cats more nocturnal.
-Deviance explained and Rsq values for the two models were XX.
+Predator activity varied considerably across space and throughout the 24-hour daily cycle, and there was some variation in the predator diel activity patterns across space.
+On average, both predators showed similar diel activity patterns, with peaks in activity around sunrise and sunset (i.e., crepuscular), although were more likely to be active during the night than the day (Fig.
+\ref{fig:diel-st-int}i).
+On average, the main difference between the species was that fox activity peaked just after sunset and they were less likely to be active during the day than cats.
+Cats also tended to be more active at sunset relative to sunrise (Fig.
+\ref{fig:diel-st-int}i).
 
-\hypertarget{fox-induced-changes-in-feral-cat-diel-activity}{%
-\subsection{Fox-induced changes in feral cat diel activity}\label{fox-induced-changes-in-feral-cat-diel-activity}}
+The interaction between hourly and spatial smooths revealed deviations from marginal diel activity pattern in space, particularly across the two regions, and within the Otway Ranges (Fig.
+\ref{fig:diel-st-int}).
+For example, at midday, foxes were less active in the Glenelg region and more active across much of the Otway Ranges (except for the north-east) than the average diel activity pattern.
+Relative to foxes, cats had slightly more complex (i.e., wiggly) deviations in space from their average diel activity pattern.
+Most notably, cats were less active during sunset and more active at midnight in the Otway Ranges relative to the Glenelg region (Fig.
+\ref{fig:diel-st-int}).
 
-We saw evidence of changes in cat diel activity due to fox activity in all three habitat types.
-Similar responses in dry habitats of both regions -- cat spatial activity increased with fox spatial activity, but became more diurnal.
-In Otway wet forests, no real change in spatial activity of cats due to foxes, but cats became more strongly nocturnal.\\
-Deviance explained and Rsq value was XX.
-Slightly higher uncertainty where fox spatial activity is high, this is because we had a largest sample size for fox absences, and fewest at the extreme of fox activity.
+\hypertarget{how-does-vegetation-type-and-human-footprint-affect-predator-activity-patterns-model-2-1}{%
+\subsection{How does vegetation type and human footprint affect predator activity patterns? (model 2)}\label{how-does-vegetation-type-and-human-footprint-affect-predator-activity-patterns-model-2-1}}
 
-\newpage
+Vegetation type had a strong influence on the spatial and diel activity patterns of foxes and cats, whereas human footprint only affected fox spatial activity.
+Spatial levels of fox activity were similar across all vegetation types, except wet forests where fox activity was considerably lower.
+Spatial cat activity was relatively more variable across vegetation types; lowest in heathy woodlands and highest in wet forests (Fig.
+\ref{fig:diel-veg}b).
+Diel activity patterns for foxes were similar across all vegetation type except wet forests; in wet forests, foxes were consistently active throughout the daily cycle (Fig.
+\ref{fig:diel-veg}a).
+On the other hand, cats were mostly nocturnal (and most active) in wet forests, but largely crepuscular in all other vegetation types (Fig.
+\ref{fig:diel-veg}a).
 
-\begin{longtable}[]{@{}cccccc@{}}
-\caption{Summary of camera-trap deployments in unique sites within Ecological Vegetation Class groups and independent counts of predator detections across two broad regions in south-west Victoria, Australia.}\tabularnewline
-\toprule
-Region & Vegetation & Sites & Deployments & Fox counts & Cat counts\tabularnewline
-\midrule
-\endfirsthead
-\toprule
-Region & Vegetation & Sites & Deployments & Fox counts & Cat counts\tabularnewline
-\midrule
-\endhead
-Glenelg & Dry Forest & 25 & 68 & 347 & 9\tabularnewline
-Otway & Dry Forest & 111 & 310 & 341 & 158\tabularnewline
-Glenelg & Heathland & 40 & 119 & 265 & 59\tabularnewline
-Otway & Heathland & 3 & 9 & 8 & 6\tabularnewline
-Glenelg & Heathy Woodland & 154 & 424 & 574 & 96\tabularnewline
-Otway & Heathy Woodland & 82 & 256 & 160 & 66\tabularnewline
-Glenelg & Herb-rich Woodland & 59 & 372 & 860 & 197\tabularnewline
-Otway & Herb-rich Woodland & 2 & 6 & 3 & 2\tabularnewline
-Glenelg & Lowland Forest & 383 & 1046 & 1900 & 290\tabularnewline
-Otway & Lowland Forest & 52 & 162 & 190 & 35\tabularnewline
-Glenelg & Swampy Scrub & 4 & 10 & 19 & 8\tabularnewline
-Otway & Swampy Scrub & 36 & 97 & 64 & 88\tabularnewline
-Otway & Wet Forest & 281 & 780 & 715 & 1187\tabularnewline
-\bottomrule
-\end{longtable}
+There was no evidence that human footprint affected diel activity patterns for either predator species; the interaction term between hour and human footprint index was removed from both models.
+Fox spatial activity increased linearly with the human footprint index (Supporting Information Fig.
+\ref{fig:fox-hfi}a), although human footprint did not impact cat spatial activity (the marginal effect of human footprint was also removed from the cat model).
+
+\hypertarget{do-feral-cats-avoid-foxes-in-space-or-time-model-3-1}{%
+\subsection{Do feral cats avoid foxes in space or time? (model 3)}\label{do-feral-cats-avoid-foxes-in-space-or-time-model-3-1}}
+
+The model indicated that cats avoided foxes temporally rather than spatially (Fig.
+\ref{fig:diel-cat-fox}).
+Fox activity did not impact spatial cat activity in the Otway Ranges, whereas cat spatial activity slightly increased with adjusted fox counts in the Glenelg region.
+
+Across all habitat types, feral cat diel activity patterns changed across gradients of fox activity (Fig.
+\ref{fig:diel-cat-fox}).
+In the Glenelg region and Otway dry habitat types, feral cats had a nocturnal-crepuscular diel activity pattern where fox activity was low, but were most active during the day where fox activity was high.
+In contrast, in the wet forests of the Otway Ranges, feral cats were more strongly nocturnal when fox activity was high, avoiding dawn in particular.
 
@@
 \begin{figure}
-\includegraphics[width=1\linewidth]{../figs/predator_veg} \caption{Fox (A) and feral cat (B) diel activity patterns across Ecological Vegetation Class groups in south-west Victoria, Australia.
-Dotted, vertical lines represent average sunrise and sunset times.
-Shaded areas indicate 95\% confidence intervals.}\label{fig:veg}
+
+{\centering \includegraphics[width=0.7\linewidth]{../figs/spatial_interaction}
+
+}
+
+\caption{Marginal diel activity (i) and space-time interaction effect in the Glenelg region (ii) and Otway Ranges (iii) at midnight (0), sunrise (6), midday (12) and sunset (18) times for introduced red foxes \textit{Vulpes vulpes} and feral cats \textit{Felis catus}.
+These effects were estimated by a Generalised Additive Model with a tensor product interaction between spatial and hourly smooths, fit to camera-trap data ('model 1').
+Dotted, vertical lines represent average sunrise and sunset times, shaded areas indicate 95\% confidence intervals (i); black dots depict camera-trap sites (ii; iii).}\label{fig:diel-st-int}
 \end{figure}
@@
 \begin{figure}
-\includegraphics[width=1\linewidth]{../figs/cat_fox_count} \caption{Predicted feral cat diel activity (A) and uncertainty (B) across each habitat type in response to foxes (log counts adjusted for survey effort).
-Grey vertical lines respresent average sunrise and sunset times.
-Shaded areas indicate 95\% confidence intervals.}\label{fig:count}
+
+{\centering \includegraphics[width=0.8\linewidth]{../figs/predator_veg_leg}
+
+}
+
+\caption{Red fox \textit{Vulpes vulpes} (a) and feral cat \textit{Felis catus} (b) diel activity patterns overall (i) and across different Ecological Vegetation Class (EVC) groups (ii, iii) in south-west Victoria, Australia.
+These effects were estimated by a hierarchical Generalised Additive Model with a global smoother for hour (i.e., average response) and group-level smoothers with shared wiggliness for vegetation type (EVC groups), fit to camera-trap data ('model 2').
+Dotted, vertical lines represent average sunrise and sunset times.
+Shaded areas indicate 95\% confidence intervals.
+Both invasive predators had a crepuscular to nocturnal diel activity pattern on average, with slight deviations across the drier EVC groups and large deviations in wet forests (ii; wet forests shown as pink line).
+The overall level of activity was relatively consistent across EVC groups for foxes (a – iii), whereas it differed substantially for feral cats (b - iii).}\label{fig:diel-veg}
 \end{figure}
@@
 \begin{figure}
-\includegraphics[width=1\linewidth]{../figs/cat_ind} \caption{Individual variations in feral cat diel activity patterns across the Glenelg region (A) and wet forests of the western Otway Ranges (B) in south-west Victoria, Australia.
-Dotted, vertical lines represent average sunrise and sunset times.
-Shaded areas indicate 95\% confidence intervals.}\label{fig:ind}
+
+{\centering \includegraphics[width=0.8\linewidth]{../figs/cat_fox_count}
+
+}
+
+\caption{Red fox \textit{Vulpes vulpes} diel activity patterns (a), as well as variation in feral cat \textit{Felis catus} diel activity patterns in response to counts of 'independent' fox detections (log-transformed and survey effort adjusted) across each habitat type in south-west Victoria, Australia (b; c).
+Effects were estimated by a Generalised Additive Model with a tensor product interaction between hour (cyclical spine) and fox counts adjusted for survey length at each camera-trap ('fox activity'; 'model 3'), fit to camera-trap data.
+Panels B and C present the same model in two different ways: (b) cat diel activity across the entire continuum of fox activity, (c) cat diel activity patterns at the minimum and maximum fox activity, with associated 95\% confidence intervals (shaded bands).
+Grey vertical lines represent average sunrise and sunset times.
+In the Glenelg region, there were more feral cat detections where there were more fox detections, but cat peak diel activity shifted from crepuscular night to pre-dawn and midday (a).
+In the Otway Ranges, feral cat activity also peaked during the day where fox activity was high in dry vegetation types (b), but was more nocturnal where fox activity was high in the rainforests and wet forests (c).}\label{fig:diel-cat-fox}
 \end{figure}
@@
 
-Foxes likely impacted the spatiotemporal behaviour of feral cats. One of the few studies to actually demonstrate and replicate this across ecosystems.
-Happens in different ways - context key.
-Spatial only, as well as temporal only models would have had opposite inference - high overlap.
-Joint spatiotemporal analysis is key to understand predator interactions.
+A key question in ecological theory is whether animals are evolutionarily hardwired to occupy particular temporal niches, or have circadian rhythms that are responsive to changing environmental conditions and interactions with other species (Schoener 1974, Daan 1981, Lima and Dill 1990).
+Here we demonstrate that diel activity patterns are not fixed, but vary across space based on landscape context and fear.
+In our study, sympatric invasive predators had similar diel activity patterns when averaged across broad regions (i.e., high circular overlap, as did Roshier and Carter 2021), but behaviours varied considerably within landscapes.
+Foxes were nocturnal in dry habitat types, but showed little variation in activity through the daily cycle in wet forests (Fig.
+\ref{fig:diel-veg}a).
+In contrast, cats were mostly nocturnal in wet forests but crepuscular in dry vegetation types (Fig.
+\ref{fig:diel-veg}b).
+Within broad habitat types, cats altered their diel activity patterns at sites with higher fox activity towards less-risky times of day (i.e., times when fox activity was lower), however we found no signs of foxes suppressing cat spatial activity (Fig.
+\ref{fig:diel-cat-fox}).
+Control programs that reduce invasive fox activity may therefore shift cat diel activity patterns, which in turn may alter cat impacts on native prey species.
+Quantifying changes in diel activity patterns provides important context for understanding species interactions, which is key for effective ecosystem management (Gaynor et al.
+2021).
 
-Our approach is simple and flexible.
-Small sample sizes are usually a problem. We demonstrate hierarchical model specification to share information - ability to pool all available data.
-Our approach is more aligned with null hypothesis testing - small sample sizes shrink to a flat line (or to the average), in contrast to circular overlap.
-Could be easily adapted to other contexts. E.g. instead of vegetation type, could do individual variation in diel activity.
+Shifting diel activity patterns may facilitate spatial coexistence of dominant and subordinate species (Carothers and Jaksić 1984).
+For cats, altering diel activity patterns to less-preferred times of day may be worthwhile to persist in high-quality habitat.
+There have been an increasing number of studies which document slight changes in diel activity peaks of subordinate species in response to dominant predators, but few studies have demonstrated predator-induced reversals in diel activity (Kronfeld-Schor and Dayan 2003).
+Notably, ship rats \emph{Rattus norvegicus} were also found to switch from nocturnal to diurnal behaviour in response to fox activity (Fenn and Macdonald 1995) and a similar nocturnal-diurnal shift was observed in American mink \emph{Neovison vison} following the recolonisation of native predators (Harrington et al.
+2009).
+Our study is particularly unique in that we replicated this behaviour across two separate regions with experimental manipulation, although there was considerable uncertainty in cat diel activity patterns where fox activity was highest (because there were fewer sites with high fox activity relative to low).
 
-Combined spatiotemporal modelling - differences to other approaches.
-Cunningham paper most similar to our approach - different in diel activity only, and linear responses.
-We used a tensor product for fox counts because predator interactions may not be linear (Johnson and VanDerWal 2009).
+For cats in our study, a switch to diurnal behaviour where fox activity was high in the dry vegetation types may have been facilitated by the higher abundance of reptiles in these habitat types relative to wet forests (which are mostly diurnal, Woinarski et al.
+2018).
+At wet forest sites with high fox activity, cats concentrated their activity away from sunrise and sunset towards midnight, despite a diurnal shift appearing to similarly reduce the risk of a fox encounter.
+In this situation, we expect becoming more nocturnal to be favourable over a diurnal shift because this is when small mammals are active and cats would be least visible to foxes (cats here mostly had black or grey coats, Rees et al.
+2023a).
+Spatial fox activity was twice as low in the wet forests relative to dry habitat types, and so cats were likely under less pressure to radically alter their diel activity patterns.
+Understanding how these potential avoidance behaviours impacts native prey is a key research priority to improve invasive predator management.
 
-Limitations of our approach.
-Detectability ignored.
-Individual heterogeneity ignored.
-Correlation does not = causation.
-Does not (clearly) seperate numerical responses from behavioural (Borchers SCR paper)
+Cats appeared to avoid foxes in time, but we saw no evidence of spatial avoidance (i.e.~no fine-scale negative association between fox and cat spatial activity.
+In fact, there was some evidence that cat and fox spatial activity was positively correlated in the Glenelg region, suggesting temporal partitioning may facilitate predator co-abundance in quality habitat.
+Nonetheless, cats may have still avoided foxes spatially, but transiently on the scale of hours to days, or at finer spatial scales than most camera-trap studies can detect (and which telemetry studies can highlight, e.g., Kohl et al.
+2019, Smith et al.
+2019).
+Short-term spatial avoidance is quite plausible given foxes mark territories using scats and odours, which cats could tangibly associate with high risk shortly after.
+Subordinate predators may prefer temporal avoidance of dominant predators (changing diel patterns or transient site avoidance) over longer-term site avoidance when both species have similar resource requirements and share many prey species.
 
-Invasive predators are useful species to test this on, as they are extremely adaptable to different conditions.
-Implications for prey.
+Many species increase nightly activity to avoid humans (Gaynor et al.
+2018, Van Scoyoc et al.
+2023), however we effect of human footprint on either predator species on our study.
+This is somewhat unsurprising given there was little variation in human activity among our sites; we studied predators within rugged conservation reserves largely free from human pressures (our camera-trap sites encompassed only the lower 32\% of the total human footprint index range).
+On the other hand, foxes within our study region often select for human-modified features more during night than day (Hradsky et al.
+2017a).
+Subordinate predators may still prefer to avoid dominant (animal) predators over humans, even when this decision leads to higher mortality rates (Prugh et al.
+2023).
 
-Conclusions.
+A distinction of our study from others is that we modelled potential avoidance behaviours in a simple predator guild, where dominant predator activity was artificially manipulated (suppressed by up to 21-fold).
+The simple predator guild removes bias from unmodelled impacts of other predators in the system (Levi and Wilmers 2012).
+Artificial manipulation of the dominant predator also reduces potential bias from differences in niche preferences between the predator species.
+We also included replication across different habitat types.
+However, because our study did not consider associations with prey species, we cannot distinguish whether changes in cat diel activity patterns were the result of direct fox avoidance or indirect associations with shared prey.
+For example, low fox activity may promote the availability of a preferred shared prey species with a diel pattern which differs from those of cats on average, and as a result, cats might shift diel activity patterns at sites with low fox activity to more closely match those of the more abundant prey species.
+We would expect introduced European rabbits \emph{Oryctolagus cuniculus} and hares \emph{Lepus europaeus} (which are diurnal) to be particularly likely to induce such a response in dry vegetation types (McGregor et al.
+2020, Stobo-Wilson et al.
+2020), however, rabbits and hares are rare within the natural vegetation of the landscapes we surveyed (only ever being detected at 40 of 1,232 sites) and there is little evidence that predation by foxes suppresses rabbit populations (Norbury and Jones 2015, Scroggie et al.
+2018).
 
-\newpage
+Flexible antipredator behaviours make evolutionary sense, but have been rarely demonstrated in terms of spatiotemporal predator avoidance, because this often requires manipulative experiments or at least more complicated models (although, see Relyea 2003, Brown et al.
+2013, Cunningham et al.
+2019).
+Our study demonstrates that GAMs offer a powerful tool for modelling continuous shifts in animal activity across both space and time, capable of capturing complex interactions and sharing information across categorical variables.
+We are only aware of one study which allowed animal diel activity to interact with predation risk as a continuous variable in a GAM (although without allowing for nonlinear shifts via a tensor product or jointly considering spatial activity, Cunningham et al.
+2019).
+Further, complexity penalties are particularly beneficial in assessing spatiotemporal responses, because they are less prone to spurious estimates from small sample sizes or noisy data, and allow multiple hypotheses to be tested within one model without the need for ad-hoc significance testing (e.g., running a separate model for each scenario and comparing with Information Criteria).
+For example, our model specification allowed cats to respond to fox activity with (a) nonlinear shifts in both spatial and diel activity (as was observed in the Glenelg region), (b) nonlinear shifts in spatial but not diel activity, (c) nonlinear shifts in diel but not spatial activity (as was observed in both habitat types of the Otway Ranges), as well as their respective linear counterparts and also different linear-nonlinear combinations of the first scenario (`model 3').
+\textbf{ADD CLOSER}
 
-\hypertarget{acknowledgements}{%
-\section{ACKNOWLEDGEMENTS}\label{acknowledgements}}
+The alternative approach of simply comparing average diel activity overlap between two species (Ridout and Linkie 2009) would have been misleading for two reasons.
+Firstly, predator diel activity patterns varied `naturally' across heterogeneous landscapes (requiring avoidance to be tested in wet forests and dry vegetation types separately, Fig.
+\ref{fig:diel-veg}).
+Secondly, dominant predator temporal avoidance strategies were not consistently employed, but depended on spatial dominant predator activity and habitat type (Fig.
+\ref{fig:diel-cat-fox}).
+Despite their underlying statistical complexity, GAMs in the `mgcv' R-package are straightforward to fit.
+Our GAM framework for modelling spatiotemporal activity can be used on any species with time-stamped detections, including datasets with categorical or continuous covariates and hierarchical groupings.
+However, like many other spatiotemporal approaches such as time-to-encounter models, a key limitation is that our GAM approach did not separate the observation processes from the state process.
+Integrating GAM-like functions into recently developed multispecies spatiotemporal models (e.g., Kellner et al.
+2022) would be highly advantageous.
 
-We acknowledge and pay respect to the Gadubanud and Gunditjmara people on whose traditional lands this study took place.
-This experiment was a collaborative effort between the Glenelg Ark (Department of Environment, Land Water and Planning) and Otway Ark (Parks Victoria) working groups.
-We are extremely thankful to those who led and assisted with these camera-trap deployments and image processing.
-University of Melbourne led surveys were conducted under University of Melbourne Animal Ethics Committee approval 1714119 and Victorian Government Department of Environment, Land Water and Planning Research Permit 10008273.
-Our study was generously supported by the Conservation Ecology Centre, the Victorian Government Department of Environment, Land Water and Planning, Arthur Rylah Institute for Environmental Research, Parks Victoria and the Australian Government's National Environmental Science Program through the Threatened Species Recovery Hub, and ARC Linkage Project LP170101134.
-MR also receives support from an Australian Government Research Training Program Scholarship.
-
-\hypertarget{open-research}{%
-\section{OPEN RESEARCH}\label{open-research}}
-
-Raw data and code are on Github link xx.\\
-Data will be deposited on the Dryad Digital Repository after acceptance.
+Animal diel activity patterns can be complex, varying across space, habitat types and threat-levels (McCann et al.
+2017).
+Despite telling an important story about how animals interact with each other and the environment, detection times are commonly discarded from statistical analyses of camera-trap data.
+When considered, diel activity patterns are predominantly estimated at the population-level, overlooking finer-scale behaviours that can affect fitness, survival and ecosystem-impacts.
+Our results demonstrate the importance of (a) considering diel activity in regards to species interactions, (b) modelling changes in animal behaviour rather than overlap with other species, and (c) testing avoidance behaviours within a joint spatiotemporal framework.
+Our study adds to the growing body of evidence that dominant predators may cause fear which is powerful enough to reverse the diel activity patterns of subordinate species (Kronfeld-Schor and Dayan 2003), and crucially, demonstrates that these effects are habitat-dependent.
 
@@
 \hypertarget{refs}{}
-\leavevmode\hypertarget{ref-maptools}{}%
-Bivand, R. and Lewin-Koh, N. 2021. Maptools: Tools for handling spatial objects.
+\begin{CSLReferences}{1}{0}
+\leavevmode\vadjust pre{\hypertarget{ref-azzou2021sparse}{}}%
+Ait Kaci Azzou, S. et al. 2021. \href{https://doi.org/10.1111/ecog.05411}{A sparse observation model to quantify species distributions and their overlap in space and time}. - Ecography 44: 928--940.
 
-\leavevmode\hypertarget{ref-delwp2020}{}%
-DELWP 2020. Bioregions and evc benchmarks.
+\leavevmode\vadjust pre{\hypertarget{ref-algar2004feral}{}}%
+Algar, D. and Burrows, N. 2004. Feral cat control research: Western shield review--february 2003. - Conservation Science Western Australia in press.
 
-\leavevmode\hypertarget{ref-DHARMa}{}%
-Hartig, F. 2020. DHARMa: Residual diagnostics for hierarchical (multi-level / mixed) regression models.
+\leavevmode\vadjust pre{\hypertarget{ref-allen2022fear}{}}%
+Allen, M.
+C.
+et al.
+2022.
+\href{https://doi.org/10.1073/pnas.2112404119}{Fear of predators in free-living wildlife reduces population growth over generations}.
+- Proceedings of the National Academy of Sciences 119: e2112404119.
 
-\leavevmode\hypertarget{ref-hradsky2017human}{}%
-Hradsky, B. A. et al. 2017. Human-modified habitats facilitate forest-dwelling populations of an invasive predator, vulpes vulpes. - Scientific Reports 7: 1--12.
+\leavevmode\vadjust pre{\hypertarget{ref-basille2015plastic}{}}%
+Basille, M. et al. 2015. \href{https://doi.org/10.1890/14-1706.1}{Plastic response of fearful prey to the spatiotemporal dynamics of predator distribution}. - Ecology 96: 2622--2631.
 
-\leavevmode\hypertarget{ref-https:ux2fux2fdoi.orgux2f10.1111ux2f2041-210X.13308}{}%
-Iannarilli, F. et al. 2019. Using lorelograms to measure and model correlation in binary data: Applications to ecological studies. - Methods in Ecology and Evolution 10: 2153--2162.
+\leavevmode\vadjust pre{\hypertarget{ref-maptools}{}}%
+Bivand, R. and Lewin-Koh, N. 2021. \href{https://CRAN.R-project.org/package=maptools}{Maptools: Tools for handling spatial objects}.
 
-\leavevmode\hypertarget{ref-https:ux2fux2fdoi.orgux2f10.1111ux2fj.1365-2664.2009.01650.x}{}%
-Johnson, C. N. and VanDerWal, J. 2009. Evidence that dingoes limit abundance of a mesopredator in eastern australian forests. - Journal of Applied Ecology 46: 641--646.
+\leavevmode\vadjust pre{\hypertarget{ref-brown1999ecology}{}}%
+Brown, J. S. et al. 1999. \href{https://doi.org/10.2307/1383287}{The ecology of fear: Optimal foraging, game theory, and trophic interactions}. - Journal of Mammalogy 80: 385--399.
 
-\leavevmode\hypertarget{ref-MARRA20112372}{}%
-Marra, G. and Wood, S. N. 2011. Practical variable selection for generalized additive models. - Computational Statistics \& Data Analysis 55: 2372--2387.
+\leavevmode\vadjust pre{\hypertarget{ref-brown2013phenotypically}{}}%
+Brown, G.
+E.
+et al.
+2013.
+\href{https://doi.org/10.1098/rspb.2012.2712}{Phenotypically plastic neophobia: A response to variable predation risk}.
+- Proceedings of the Royal Society B: Biological Sciences 280: 20122712.
 
-\leavevmode\hypertarget{ref-pedersen2019hierarchical}{}%
-Pedersen, E. J. et al. 2019. Hierarchical generalized additive models in ecology: An introduction with mgcv. - PeerJ 7: e6876.
+\leavevmode\vadjust pre{\hypertarget{ref-carothers1984time}{}}%
+Carothers, J. H. and Jaksić, F. M. 1984. \href{https://doi.org/10.2307/3544413}{Time as a niche difference: The role of interference competition}. - Oikos: 403--406.
 
-\leavevmode\hypertarget{ref-R}{}%
-R Core Team 2020. R: A language and environment for statistical computing. - R Foundation for Statistical Computing.
+\leavevmode\vadjust pre{\hypertarget{ref-creel2008relationships}{}}%
+Creel, S. and Christianson, D. 2008. \href{https://doi.org/10.1016/j.tree.2007.12.004}{Relationships between direct predation and risk effects}. - Trends in Ecology \& Evolution 23: 194--201.
 
-\leavevmode\hypertarget{ref-REES2019108287}{}%
-Rees, M. W. et al. 2019. Unexpectedly high densities of feral cats in a rugged temperate forest. - Biological Conservation 239: 108287.
+\leavevmode\vadjust pre{\hypertarget{ref-cunningham2019temporal}{}}%
+Cunningham, C.
+X.
+et al.
+2019.
+\href{https://doi.org/10.1111/ecog.04485}{Temporal partitioning of activity: Rising and falling top-predator abundance triggers community-wide shifts in diel activity}.
+- Ecography 42: 2157--2168.
 
-\leavevmode\hypertarget{ref-ROBLEY2014262}{}%
-Robley, A. et al. 2014. Long-term and large-scale control of the introduced red fox increases native mammal occupancy in australian forests. - Biological Conservation 180: 262--269.
+\leavevmode\vadjust pre{\hypertarget{ref-daan1981adaptive}{}}%
+Daan, S. 1981. \href{https://doi.org/10.1007/978-1-4615-6552-9_15}{Adaptive daily strategies in behavior}. - In: Biological rhythms. Springer Publishing, pp. 275--298.
 
-\leavevmode\hypertarget{ref-gratia}{}%
-Simpson, G. L. 2021. Gratia: Graceful 'ggplot'-based graphics and other functions for gams fitted using 'mgcv'.
+\leavevmode\vadjust pre{\hypertarget{ref-davies2021spatial}{}}%
+Davies, A. B. et al. 2021. \href{https://doi.org/10.1002/ecy.3319}{Spatial heterogeneity facilitates carnivore coexistence}. - Ecology 102: e03319.
 
-\leavevmode\hypertarget{ref-swan2021species}{}%
-Swan, M. et al. 2021. Species distribution models for conservation planning in fire-prone landscapes. - Biodiversity and Conservation 30: 1119--1136.
+\leavevmode\vadjust pre{\hypertarget{ref-delwp2020bioregions}{}}%
+Department of Environment, Land, Water \& Planning 2020. \href{https://www.environment.vic.gov.au/biodiversity/bioregions-and-evc-benchmarks}{{Bioregions and EVC Benchmarks}}.
 
-\leavevmode\hypertarget{ref-https:ux2fux2fdoi.orgux2f10.1111ux2f2041-210X.13290}{}%
-Vazquez, C. et al. 2019. Comparing diel activity patterns of wildlife across latitudes and seasons: Time transformations using day length. - Methods in Ecology and Evolution 10: 2057--2066.
+\leavevmode\vadjust pre{\hypertarget{ref-doherty2017stop}{}}%
+Doherty, T. S. and Ritchie, E. G. 2017. \href{https://doi.org/10.1111/conl.12251}{Stop jumping the gun: A call for evidence-based invasive predator management}. - Conservation Letters 10: 15--22.
 
-\leavevmode\hypertarget{ref-reshape}{}%
-Wickham, H. 2007. Reshaping data with the reshape package. - Journal of Statistical Software 21: 1--20.
+\leavevmode\vadjust pre{\hypertarget{ref-doherty2016invasive}{}}%
+Doherty, T. S. et al. 2016. \href{https://doi.org/10.1073/pnas.1602480113}{Invasive predators and global biodiversity loss}. - Proceedings of the National Academy of Sciences 113: 11261--11265.
 
-\leavevmode\hypertarget{ref-ggplot2}{}%
-Wickham, H. 2016. Ggplot2: Elegant graphics for data analysis. - Springer-Verlag New York.
+\leavevmode\vadjust pre{\hypertarget{ref-estes2011trophic}{}}%
+Estes, J. A. et al. 2011. \href{https://doi.org/10.1126/science.1205106}{Trophic downgrading of planet earth}. - Science 333: 301--306.
 
-\leavevmode\hypertarget{ref-wood2017}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-fancourt2019introduced}{}}%
+Fancourt, B.
+A.
+et al.
+2019.
+\href{https://doi.org/10.1111/1365-2664.13514}{Do introduced apex predators suppress introduced mesopredators?
+A multiscale spatiotemporal study of dingoes and feral cats in {A}ustralia suggests not}.
+- Journal of Applied Ecology 56: 2584--2595.
+
+\leavevmode\vadjust pre{\hypertarget{ref-fenn1995use}{}}%
+Fenn, M. G. and Macdonald, D. W. 1995. \href{https://doi.org/10.2307/1382321}{Use of middens by red foxes: Risk reverses rhythms of rats}. - Journal of Mammalogy 76: 130--136.
+
+\leavevmode\vadjust pre{\hypertarget{ref-fleming2022distinctive}{}}%
+Fleming, P. A. et al. 2022. \href{https://doi.org/10.1098/rsos.220792}{Distinctive diets of eutherian predators in australia}. - Royal Society Open Science in press.
+
+\leavevmode\vadjust pre{\hypertarget{ref-frey2017investigating}{}}%
+Frey, S.
+et al.
+2017.
+\href{https://doi.org/10.1002/rse2.60}{Investigating animal activity patterns and temporal niche partitioning using camera-trap data: Challenges and opportunities}.
+- Remote Sensing in Ecology and Conservation 3: 123--132.
+
+\leavevmode\vadjust pre{\hypertarget{ref-gaynor2018influence}{}}%
+Gaynor, K. M. et al. 2018. \href{https://doi.org/10.1126/science.aar7121}{The influence of human disturbance on wildlife nocturnality}. - Science 360: 1232--1235.
+
+\leavevmode\vadjust pre{\hypertarget{ref-gaynor2019landscapes}{}}%
+Gaynor, K. M. et al. 2019. \href{https://doi.org/10.1016/j.tree.2019.01.004}{Landscapes of fear: Spatial patterns of risk perception and response}. - Trends in Ecology \& Evolution 34: 355--368.
+
+\leavevmode\vadjust pre{\hypertarget{ref-gaynor2021applied}{}}%
+Gaynor, K. M. et al. 2021. \href{https://doi.org/10.1111/acv.12629}{An applied ecology of fear framework: Linking theory to conservation practice}. - Animal Conservation 24: 308--321.
+
+\leavevmode\vadjust pre{\hypertarget{ref-glen2005complex}{}}%
+Glen, A.
+S.
+and Dickman, C.
+R.
+2005.
+\href{https://doi.org/10.1017/S1464793105006718}{Complex interactions among mammalian carnivores in {{A}ustralia}, and their implications for wildlife management}.
+- Biological Reviews 80: 387--401.
+
+\leavevmode\vadjust pre{\hypertarget{ref-harmsen2009spatial}{}}%
+Harmsen, B.
+J.
+et al.
+2009.
+\href{https://doi.org/10.1644/08-MAMM-A-140R.1}{{Spatial and Temporal Interactions of Sympatric Jaguars (Panthera onca) and Pumas (Puma concolor) in a Neotropical Forest}}.
+- Journal of Mammalogy 90: 612--620.
+
+\leavevmode\vadjust pre{\hypertarget{ref-harrington2009impact}{}}%
+Harrington, L.
+A.
+et al.
+2009.
+\href{https://doi.org/10.1890/08-0302.1}{The impact of native competitors on an alien invasive: Temporal niche shifts to avoid interspecific aggression}.
+- Ecology 90: 1207--1216.
+
+\leavevmode\vadjust pre{\hypertarget{ref-DHARMa}{}}%
+Hartig, F. 2020. \href{https://CRAN.R-project.org/package=DHARMa}{DHARMa: Residual diagnostics for hierarchical (multi-level / mixed) regression models}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-hradsky2017human}{}}%
+Hradsky, B.
+A.
+et al.
+2017a.
+\href{https://doi.org/10.1038/s41598-017-12464-7}{Human-modified habitats facilitate forest-dwelling populations of an invasive predator, \emph{{V}ulpes vulpes}}.
+- Scientific Reports 7: 1--12.
+
+\leavevmode\vadjust pre{\hypertarget{ref-hradsky2017bayesian}{}}%
+Hradsky, B.
+A.
+et al.
+2017b.
+\href{https://esajournals.onlinelibrary.wiley.com/doi/abs/10.1002/ecs2.1926}{Bayesian networks elucidate interactions between fire and other drivers of terrestrial fauna distributions}.
+- Ecosphere 8: e01926.
+
+\leavevmode\vadjust pre{\hypertarget{ref-hunter2018not}{}}%
+Hunter, D.
+O.
+et al.
+2018.
+\href{https://doi.org/10.1111/mam.12115}{Not all predators are equal: A continent-scale analysis of the effects of predator control on {{A}ustralian} mammals}.
+- Mammal Review 48: 108--122.
+
+\leavevmode\vadjust pre{\hypertarget{ref-iannarilli2019lorelograms}{}}%
+Iannarilli, F.
+et al.
+2019.
+\href{https://doi.org/10.1111/2041-210X.13308}{Using lorelograms to measure and model correlation in binary data: Applications to ecological studies}.
+- Methods in Ecology and Evolution 10: 2153--2162.
+
+\leavevmode\vadjust pre{\hypertarget{ref-kauffman2007landscape}{}}%
+Kauffman, M.
+J.
+et al.
+2007.
+\href{https://doi.org/10.1111/j.1461-0248.2007.01059.x}{Landscape heterogeneity shapes predation in a newly restored predator--prey system}.
+- Ecology Letters 10: 690--700.
+
+\leavevmode\vadjust pre{\hypertarget{ref-kellner2022two}{}}%
+Kellner, K.
+F.
+et al.
+2022.
+\href{https://doi.org/10.1007/s13253-021-00482-y}{A two-species occupancy model with a continuous-time detection process reveals spatial and temporal interactions}.
+- Journal of Agricultural, Biological and Environmental Statistics 27: 321--338.
+
+\leavevmode\vadjust pre{\hypertarget{ref-kohl2019prey}{}}%
+Kohl, M. T. et al. 2019. \href{https://doi.org/10.1111/ele.13319}{Do prey select for vacant hunting domains to minimize a multi-predator threat?} - Ecology Letters 22: 1724--1733.
+
+\leavevmode\vadjust pre{\hypertarget{ref-kronfeld2003partitioning}{}}%
+Kronfeld-Schor, N.
+and Dayan, T.
+2003.
+\href{https://doi.org/10.1146/annurev.ecolsys.34.011802.132435}{Partitioning of time as an ecological resource}.
+- Annual Review of Ecology, Evolution, and Systematics 34: 153--181.
+
+\leavevmode\vadjust pre{\hypertarget{ref-levi2012wolves}{}}%
+Levi, T. and Wilmers, C. C. 2012. \href{https://doi.org/10.1890/11-0165.1}{Wolves--coyotes--foxes: A cascade among carnivores}. - Ecology 93: 921--929.
+
+\leavevmode\vadjust pre{\hypertarget{ref-lima1990behavioral}{}}%
+Lima, S. L. and Dill, L. M. 1990. \href{https://doi.org/10.1139/z90-092}{Behavioral decisions made under the risk of predation: A review and prospectus}. - Canadian Journal of Zoology 68: 619--640.
+
+\leavevmode\vadjust pre{\hypertarget{ref-lima1999temporal}{}}%
+Lima, S.
+L.
+and Bednekoff, P.
+A.
+1999.
+\href{https://doi.org/10.1086/303202}{Temporal variation in danger drives antipredator behavior: The predation risk allocation hypothesis}.
+- The American Naturalist 153: 649--659.
+
+\leavevmode\vadjust pre{\hypertarget{ref-mccann2017temporal}{}}%
+McCann, N. P. et al. 2017. \href{https://doi.org/10.1111/ecog.02742}{Temporal scaling in analysis of animal activity}. - Ecography 40: 1436--1444.
+
+\leavevmode\vadjust pre{\hypertarget{ref-mcgregor2020short}{}}%
+McGregor, H.
+et al.
+2020.
+\href{https://doi.org/10.1007/s10530-019-02131-5}{The short-term response of feral cats to rabbit population decline: Are alternative native prey more at risk?} - Biological Invasions 22: 799--811.
+
+\leavevmode\vadjust pre{\hypertarget{ref-miller2014finite}{}}%
+Miller, D. L. and Wood, S. N. 2014. \href{https://doi.org/10.1007/s10651-014-0277-4}{Finite area smoothing with generalized distance splines}. - Environmental and Ecological Statistics 21: 715--731.
+
+\leavevmode\vadjust pre{\hypertarget{ref-molsher2017mesopredator}{}}%
+Molsher, R.
+et al.
+2017.
+\href{https://doi.org/10.1371/journal.pone.0168460}{Mesopredator management: Effects of red fox control on the abundance, diet and use of space by feral cats}.
+- PLoS One 12: e0168460.
+
+\leavevmode\vadjust pre{\hypertarget{ref-moseby2011use}{}}%
+Moseby, K.
+et al.
+2011.
+\href{https://doi.org/10.1071/WR10236}{The use of poison baits to control feral cats and red foxes in arid {South Australia} II.
+Bait type, placement, lures and non-target uptake}.
+- Wildlife Research 38: 350--358.
+
+\leavevmode\vadjust pre{\hypertarget{ref-niedballa2019assessing}{}}%
+Niedballa, J.
+et al.
+2019.
+\href{https://doi.org/10.1002/rse2.107}{Assessing analytical methods for detecting spatiotemporal interactions between species from camera trapping data}.
+- Remote Sensing in Ecology and Conservation 5: 272--285.
+
+\leavevmode\vadjust pre{\hypertarget{ref-norbury2015pests}{}}%
+Norbury, G.
+and Jones, C.
+2015.
+\href{https://doi.org/10.1111/mam.12034}{Pests controlling pests: Does predator control lead to greater {E}uropean rabbit abundance in {A}ustralasia?} - Mammal Review 45: 79--87.
+
+\leavevmode\vadjust pre{\hypertarget{ref-parsons2022effect}{}}%
+Parsons, A. W. et al. 2022. \href{https://doi.org/10.1002/ecs2.3993}{The effect of urbanization on spatiotemporal interactions between gray foxes and coyotes}. - Ecosphere 13: e3993.
+
+\leavevmode\vadjust pre{\hypertarget{ref-pedersen2019hierarchical}{}}%
+Pedersen, E. J. et al. 2019. \href{https://doi.org/10.7717/peerj.6876}{Hierarchical generalized additive models in ecology: An introduction with mgcv}. - PeerJ 7: e6876.
+
+\leavevmode\vadjust pre{\hypertarget{ref-preisser2005scared}{}}%
+Preisser, E. L. et al. 2005. \href{https://doi.org/10.1890/04-0719}{Scared to death? The effects of intimidation and consumption in predator--prey interactions}. - Ecology 86: 501--509.
+
+\leavevmode\vadjust pre{\hypertarget{ref-prugh2009rise}{}}%
+Prugh, L. R. et al. 2009. \href{https://doi.org/10.1525/bio.2009.59.9.9}{The rise of the mesopredator}. - Bioscience 59: 779--791.
+
+\leavevmode\vadjust pre{\hypertarget{ref-prugh2023fear}{}}%
+Prugh, L. R. et al. 2023. \href{https://doi.org/10.1126/science.adf2472}{Fear of large carnivores amplifies human-caused mortality for mesopredators}. - Science 380: 754--758.
+
+\leavevmode\vadjust pre{\hypertarget{ref-R}{}}%
+R Core Team 2020. \href{https://www.R-project.org/}{R: A language and environment for statistical computing}. - R Foundation for Statistical Computing.
+
+\leavevmode\vadjust pre{\hypertarget{ref-reddiex2007control}{}}%
+Reddiex, B.
+et al.
+2007.
+\href{https://doi.org/10.1071/WR05102}{Control of pest mammals for biodiversity protection in {{A}ustralia}.
+I.
+Patterns of control and monitoring}.
+- Wildlife Research 33: 691--709.
+
+\leavevmode\vadjust pre{\hypertarget{ref-rees2023mesopredator}{}}%
+Rees, M.
+W.
+et al.
+2023a.
+\href{https://doi.org/10.1111/1365-2664.14402}{Mesopredator release among invasive predators: Controlling red foxes can increase feral cat density and alter their behaviour}.
+- Journal of Applied Ecology 60: 1100--1114.
+
+\leavevmode\vadjust pre{\hypertarget{ref-rees2023data}{}}%
+Rees, M.
+W.
+et al.
+2023b.
+\href{https://doi.org/10.5061/dryad.69p8cz95w}{Data from: Mesopredator release among invasive predators: Controlling red foxes can increase feral cat density and alter their behaviour}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-relyea2003predators}{}}%
+Relyea, R.
+A.
+2003.
+\href{https://doi.org/10.1890/0012-9658(2003)084\%5B1840:PCAPGT\%5D2.0.CO;2}{Predators come and predators go: The reversibility of predator-induced traits}.
+- Ecology 84: 1840--1848.
+
+\leavevmode\vadjust pre{\hypertarget{ref-ridout2009estimating}{}}%
+Ridout, M.
+S.
+and Linkie, M.
+2009.
+\href{https://doi.org/10.1198/jabes.2009.08038}{Estimating overlap of daily activity patterns from camera trap data}.
+- Journal of Agricultural, Biological, and Environmental Statistics 14: 322--337.
+
+\leavevmode\vadjust pre{\hypertarget{ref-risbey1997control}{}}%
+Risbey, D. A. et al. 1997. \href{https://doi.org/10.1071/WR96051}{Control of feral cats for nature conservation. I. Field tests of four baiting methods}. - Wildlife Research 24: 319--326.
+
+\leavevmode\vadjust pre{\hypertarget{ref-robley2014long}{}}%
+Robley, A.
+et al.
+2014.
+\href{https://doi.org/10.1016/j.biocon.2014.10.017}{Long-term and large-scale control of the introduced red fox increases native mammal occupancy in {{A}ustralian} forests}.
+- Biological Conservation 180: 262--269.
+
+\leavevmode\vadjust pre{\hypertarget{ref-robley2019otway}{}}%
+Robley, A. et al. 2019. {The Otway Ark: response of predators and native species 2016--2018.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-robley2020glenelg}{}}%
+Robley, A. et al. 2020. {Glenelg Ark 2005--2019: long-term predator and native mammal response to predator control.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-roshier2021space}{}}%
+Roshier, D.
+A.
+and Carter, A.
+2021.
+\href{https://doi.org/10.1002/ecs2.3628}{Space use and interactions of two introduced mesopredators, european red fox and feral cat, in an arid landscape}.
+- Ecosphere 12: e03628.
+
+\leavevmode\vadjust pre{\hypertarget{ref-schmitz2004trophic}{}}%
+Schmitz, O. J. et al. 2004. \href{https://doi.org/10.1111/j.1461-0248.2003.00560.x}{Trophic cascades: The primacy of trait-mediated indirect interactions}. - Ecology Letters 7: 153--163.
+
+\leavevmode\vadjust pre{\hypertarget{ref-schoener1974compression}{}}%
+Schoener, T. W. 1974. \href{https://doi.org/10.1073/pnas.71.10.4169}{The compression hypothesis and temporal resource partitioning}. - Proceedings of the National Academy of Sciences 71: 4169--4172.
+
+\leavevmode\vadjust pre{\hypertarget{ref-scroggie2018invasive}{}}%
+Scroggie, M.
+P.
+et al.
+2018.
+\href{https://doi.org/10.1111/1365-2664.13253}{Invasive prey controlling invasive predators?
+European rabbit abundance does not determine red fox population dynamics}.
+- Journal of Applied Ecology 55: 2621--2631.
+
+\leavevmode\vadjust pre{\hypertarget{ref-gratia}{}}%
+Simpson, G. L. 2021. \href{https://gavinsimpson.github.io/gratia/}{{g}ratia: Graceful 'ggplot'-based graphics and other functions for GAMs fitted using 'mgcv'}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-smith2019integrating}{}}%
+Smith, J.
+A.
+et al.
+2019.
+\href{https://doi.org/10.1007/s00442-019-04381-5}{Integrating temporal refugia into landscapes of fear: Prey exploit predator downtimes to forage in risky places}.
+- Oecologia 189: 883--890.
+
+\leavevmode\vadjust pre{\hypertarget{ref-soule1988reconstructed}{}}%
+Soulé, M.
+E.
+et al.
+1988.
+\href{https://doi.org/10.1111/j.1523-1739.1988.tb00337.x}{Reconstructed dynamics of rapid extinctions of chaparral-requiring birds in urban habitat islands}.
+- Conservation Biology 2: 75--92.
+
+\leavevmode\vadjust pre{\hypertarget{ref-stobo2020management}{}}%
+Stobo-Wilson, A.
+M.
+et al.
+2020.
+\href{https://doi.org/10.1071/WR19237}{Management of invasive mesopredators in the {Flinders Ranges, South {A}ustralia}: Effectiveness and implications}.
+- Wildlife Research 47: 720--730.
+
+\leavevmode\vadjust pre{\hypertarget{ref-suraci2016fear}{}}%
+Suraci, J. P. et al. 2016. \href{https://doi.org/10.1038/ncomms10698}{Fear of large carnivores causes a trophic cascade}. - Nature communications 7: 1--7.
+
+\leavevmode\vadjust pre{\hypertarget{ref-suraci2022beyond}{}}%
+Suraci, J. P. et al. 2022. \href{https://doi.org/10.1111/oik.09004}{Beyond spatial overlap: Harnessing new technologies to resolve the complexities of predator--prey interactions}. - Oikos: e09004.
+
+\leavevmode\vadjust pre{\hypertarget{ref-swan2015predicting}{}}%
+Swan, M. et al. 2015. \href{https://doi.org/10.1890/14-1533.1}{Predicting faunal fire responses in heterogeneous landscapes: The role of habitat structure}. - Ecological Applications 25: 2293--2305.
+
+\leavevmode\vadjust pre{\hypertarget{ref-van2023influence}{}}%
+Van Scoyoc, A. et al. 2023. \href{https://doi.org/10.1111/1365-2656.13892}{The influence of human activity on predator-prey spatiotemporal overlap}. - The Journal of Animal Ecology 92: 1124---1134.
+
+\leavevmode\vadjust pre{\hypertarget{ref-vazquez2019comparing}{}}%
+Vazquez, C.
+et al.
+2019.
+\href{https://doi.org/10.1111/2041-210X.13290}{Comparing diel activity patterns of wildlife across latitudes and seasons: Time transformations using day length}.
+- Methods in Ecology and Evolution 10: 2057--2066.
+
+\leavevmode\vadjust pre{\hypertarget{ref-venter2016sixteen}{}}%
+Venter, O.
+et al.
+2016.
+\href{https://doi.org/10.1038/ncomms12558}{Sixteen years of change in the global terrestrial human footprint and implications for biodiversity conservation}.
+- Nature communications 7: 12558.
+
+\leavevmode\vadjust pre{\hypertarget{ref-venter2018last}{}}%
+Venter, O. et al. 2018. \href{https://doi.org/10.7927/H46T0JQ4.}{Last of the wild project, version 3 (LWP-3): 2009 human footprint, 2018 release.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-wayne2017recoveries}{}}%
+Wayne, A.
+F.
+et al.
+2017.
+\href{https://doi.org/10.1093/jmammal/gyw237}{{Recoveries and cascading declines of native mammals associated with control of an introduced predator}}.
+- Journal of Mammalogy 98: 489--501.
+
+\leavevmode\vadjust pre{\hypertarget{ref-ggplot2}{}}%
+Wickham, H. 2016. \href{https://ggplot2.tidyverse.org}{{g}gplot2: Elegant graphics for data analysis}. - Springer-Verlag New York.
+
+\leavevmode\vadjust pre{\hypertarget{ref-willems2009predator}{}}%
+Willems, E. P. and Hill, R. A. 2009. \href{https://doi.org/10.1890/08-0765.1}{Predator-specific landscapes of fear and resource distribution: Effects on spatial range use}. - Ecology 90: 546--555.
+
+\leavevmode\vadjust pre{\hypertarget{ref-wirsing2021context}{}}%
+Wirsing, A. J. et al. 2021. \href{https://doi.org/10.1111/ele.13614}{The context dependence of non-consumptive predator effects}. - Ecology Letters 24: 113--129.
+
+\leavevmode\vadjust pre{\hypertarget{ref-woinarski2015ongoing}{}}%
+Woinarski, J.
+C.
+Z.
+et al.
+2015.
+\href{https://doi.org/10.1073/pnas.1417301112}{Ongoing unraveling of a continental fauna: Decline and extinction of {{A}ustralian} mammals since {European} settlement}.
+- Proceedings of the National Academy of Sciences 112: 4531--4540.
+
+\leavevmode\vadjust pre{\hypertarget{ref-woinarski2018reptiles}{}}%
+Woinarski, J. C. Z. et al. 2018. \href{https://doi.org/10.1071/WR17160}{How many reptiles are killed by cats in {{A}ustralia}?} - Wildlife Research 45: 247--266.
+
+\leavevmode\vadjust pre{\hypertarget{ref-wood2017generalized}{}}%
 Wood, S. N. 2017. Generalized additive models: An introduction with r. - CRC press.
 
+\leavevmode\vadjust pre{\hypertarget{ref-wooster2022predator}{}}%
+Wooster, E. I. et al. 2022. \href{https://doi.org/10.1111/oik.09059}{Predator protection dampens the landscape of fear}. - Oikos: e09059.
+
+\end{CSLReferences}
+
+\newpage
+\beginsupplement
+\setcounter{page}{1}
+
+\hypertarget{supporting-information}{%
+\section*{SUPPORTING INFORMATION}\label{supporting-information}}
+\addcontentsline{toc}{section}{SUPPORTING INFORMATION}
+
+\newpage
+
+\newpage
+\blandscape
+
+\begin{table}
+
+\caption{\label{tab:tab-sum}Summary of experimental camera-trap survey designs around fox control in south-west Victoria, Australia.}
+\centering
+\resizebox{\linewidth}{!}{
+\fontsize{10}{12}\selectfont
+\begin{tabular}[t]{llllrrrrrrr}
+\toprule
+Region & Source & Block & Treatment & Surveys & Year from & Year to & Sites & Min. spacing (m) & Mean spacing (m) & Max spacing (m)\\
+\midrule
+Glenelg & Glenelg Ark & Annya & non-treatment & 7 & 2013 & 2019 & 40 & 683 & 988 & 1509\\
+Glenelg & Glenelg Ark & Hotspur & non-treatment & 7 & 2013 & 2019 & 40 & 335 & 817 & 1319\\
+Glenelg & Glenelg Ark & LGNP N & non-treatment & 7 & 2013 & 2019 & 40 & 275 & 687 & 1828\\
+Glenelg & Glenelg Ark & Cobboboonee & treatment & 7 & 2013 & 2019 & 40 & 362 & 982 & 2351\\
+Glenelg & Glenelg Ark & LGNP S & treatment & 7 & 2013 & 2019 & 40 & 502 & 1120 & 2686\\
+\addlinespace
+Glenelg & Glenelg Ark & Mt Clay & treatment & 7 & 2013 & 2019 & 40 & 178 & 607 & 1068\\
+Glenelg & MWR PhD & Annya & non-treatment & 1 & 2018 & 2018 & 110 & 307 & 469 & 623\\
+Glenelg & MWR PhD & Hotspur & non-treatment & 1 & 2018 & 2018 & 99 & 242 & 428 & 594\\
+Glenelg & MWR PhD & Cobboboonee & treatment & 1 & 2018 & 2018 & 110 & 274 & 452 & 509\\
+Glenelg & MWR PhD & Mt Clay & treatment & 1 & 2018 & 2018 & 106 & 209 & 445 & 671\\
+\addlinespace
+Otways & MWR PhD & North & non-treatment & 3 & 2017 & 2019 & 104 & 238 & 434 & 714\\
+Otways & MWR PhD & South & treatment & 3 & 2017 & 2019 & 91 & 86 & 436 & 717\\
+Otways & Otway Ark & FA4 & non-treatment & 3 & 2017 & 2019 & 73 & 634 & 1560 & 4150\\
+Otways & Otway Ark & FA1 & treatment & 3 & 2016 & 2018 & 80 & 277 & 996 & 1825\\
+Otways & Otway Ark & FA2 & treatment & 3 & 2016 & 2018 & 80 & 638 & 938 & 1865\\
+\addlinespace
+Otways & Otway Ark & FA3 & treatment & 3 & 2016 & 2018 & 140 & 739 & 1491 & 3514\\
+\bottomrule
+\end{tabular}}
+\end{table}
+
+\thispagestyle{empty}
+\elandscape
+\newpage
+
+\begingroup\fontsize{10}{12}\selectfont
+
+\begin{longtable}[t]{llrrrr}
+\caption{\label{tab:diel-tab1}Summary of the number of camera-trap deployments, unique survey sites and 'independent' counts of invasive predator detections across Ecological Vegetation Class groups within the Glenelg and Otway regions, south-west Victoria, Australia.}\\
+\toprule
+Vegetation & Region & Sites & Deployments & Fox counts & Cat counts\\
+\midrule
+Dry Forest & Glenelg & 25 & 69 & 347 & 9\\
+ & Otways & 111 & 314 & 341 & 158\\
+Heathland & Glenelg & 40 & 119 & 265 & 59\\
+ & Otways & 3 & 9 & 8 & 6\\
+Heathy Woodland & Glenelg & 154 & 424 & 574 & 96\\
+\addlinespace
+ & Otways & 82 & 256 & 160 & 66\\
+Herb-rich Woodland & Glenelg & 59 & 373 & 863 & 198\\
+ & Otways & 2 & 6 & 3 & 2\\
+Lowland Forest & Glenelg & 383 & 1046 & 1900 & 290\\
+ & Otways & 52 & 163 & 190 & 35\\
+\addlinespace
+Swampy Scrub & Glenelg & 4 & 10 & 19 & 8\\
+ & Otways & 36 & 98 & 64 & 88\\
+Wet Forest & Otways & 281 & 780 & 715 & 1187\\
+\textbf{Total} & \textbf{} & \textbf{1232} & \textbf{3667} & \textbf{5449} & \textbf{2202}\\
+\bottomrule
+\end{longtable}
+\endgroup{}
+
+\newpage
+
+\begingroup\fontsize{10}{12}\selectfont
+
+\begin{longtable}[t]{llrrr}
+\caption{\label{tab:diel-tab-fits}Generalised additive model summaries for invasive predator spatiotemporal activity in south-west Victoria, Australia.}\\
+\toprule
+Species & Model & EDF & dev.expl & r.sq\\
+\midrule
+fox & 1\_spatial & 98.0605 & 0.0657 & 0.0137\\
+fox & 2\_vegetation\_type & 752.9670 & 0.3496 & 0.1155\\
+fox & 3\_habitat\_type & 757.1467 & 0.3486 & 0.1138\\
+cat & 1\_spatial & 78.4323 & 0.1110 & 0.0367\\
+cat & 2\_vegetation\_type & 566.7735 & 0.2301 & 0.0769\\
+\addlinespace
+cat & 3\_fox\_by\_habitat\_type & 585.8580 & 0.2317 & 0.0779\\
+\bottomrule
+\multicolumn{5}{l}{\rule{0pt}{1em}\textit{Note: }}\\
+\multicolumn{5}{l}{\rule{0pt}{1em}EDF - estimated degrees of freedom of all model terms.}\\
+\multicolumn{5}{l}{\rule{0pt}{1em}dev.expl - proportion of the null deviance explained by the model. }\\
+\multicolumn{5}{l}{\rule{0pt}{1em}r.sq -  adjusted r-squared value.}\\
+\end{longtable}
+\endgroup{}
+
+\newpage
+
+\begin{figure}
+
+{\centering \includegraphics[width=0.8\linewidth]{../figs/foxbaiting_effects}
+
+}
+
+\caption{Effects of 1080 poison-bait density on the spatial activity of red foxes \textit{Vulpes vulpes}, derived from each model (model 1: spatial, model 2: vegetation group / human footprint, model 3: habitat type).
+Shaded areas indicate 95\% confidence intervals.}\label{fig:fox-baits}
+\end{figure}
+
+\begin{figure}
+
+{\centering \includegraphics[width=0.8\linewidth]{../figs/fox_hfi}
+
+}
+
+\caption{Effects human footprint index on the spatial activity of red foxes \textit{Vulpes vulpes}.
+(model 2).
+The interaction effect of human foot print and fox diel activity pattern was removed from this model.
+Both the marginal and interactive effect of human footprint was removed for feral cats \textit{Felis catus}).
+Shaded areas indicate 95\% confidence intervals.}\label{fig:fox-hfi}
+\end{figure}
+
 \end{document}

### 2023-07-28 r_scripts/2_GAMs.R

@@ -1,0 +1,120 @@
+## FIT GAMS TO MODEL DIEL ACTIVITY FOR FOXES AND FERAL CATS
+# MATTHEW REES
+
+library(dplyr)
+library(mgcv)
+# for plots
+library(gratia)
+library(ggplot2)
+library(patchwork)
+library(ggpubr)
+library(viridis)
+library(forcats)
+# Set the default theme for ggplot objects 
+theme_set(theme_bw())
+theme_update(panel.grid = element_blank())
+
+
+# load data
+records <- read.csv("derived_data/predator_counts_hour.csv")
+
+
+# Transform variables -----------------------------------------------------
+records <- transform(records,
+                      hour = as.integer(hour),
+                      cat = as.integer(cat), 
+                      fox = as.integer(fox), 
+                      foxbaits = as.integer(foxbaits), 
+                      fox_pa = factor(fox_pa, ordered = FALSE), 
+                      fox_count_adj = as.numeric(fox_count_adj), 
+                      region = factor(region, ordered = FALSE), 
+                      station = factor(station, ordered = FALSE), 
+                      vegetation_group = factor(XGROUPNAME, ordered = FALSE), 
+                      habitat_type = factor(habitat_type, ordered = FALSE), 
+                      survey_duration = as.integer(survey_duration),
+                      hfi = round(hfi, digits = 2))
+
+
+
+
+# 1) Interaction between spatial and diel activity ('model 1') ---------------------
+# specify penalties for 2D Duchon spline and 1D cc spline
+m <- list(c(1,.5),rep(0,0)) 
+
+gam_fox_sp <- bam(fox ~ s(hour, bs = "cc", k = 8) + 
+                        s(x, y, bs = "ds", k = 80, m = c(1,.5)) + 
+                        ti(x, y, hour, d = c(2, 1), bs = c("ds", "cc"), k = c(80, 8), m = m) +                     
+                        s(foxbaits, region, bs = "fs", xt = list(bs = "tp"), k = 4) + 
+                        offset(log(survey_duration)), 
+                  data = records, family = nb, knots = list(hour = c(0, 23)), discrete = TRUE)
+
+gam_cat_sp <- bam(cat ~ s(hour, bs = "cc", k = 8) + 
+                        s(x, y, bs = "ds", k = 80, m = c(1,.5)) + 
+                        ti(x, y, hour, d = c(2, 1), bs = c("ds", "cc"), k = c(80, 8), m = m) +                     
[…]

### 2023-07-28 r_scripts/format_records_hour_gams.R

@@ -1,177 +1,0 @@
-# load r packages
-library(maptools)        
-library(dplyr)
-library(activity) 
-library(lubridate)
-library(reshape2)
-library(sf)
-# note: chron package is loaded further down
-
-# load function to change clock time to radial time for overlap package
-source("r_scripts/function_clock_time_to_radian.R")
-
-
-# LOAD / PREPARE DATA -----------------------------------------------------
-
-# load data
-records <- read.csv("raw_data/predator_records.csv")
-camdata <- read.csv("raw_data/camdata.csv")
-# remove stupid X's
-camdata <- subset(camdata, select = -c(X))
-records <- subset(records, select = -c(X))
-# add camdata info to records - NA for sites without pred detections?
-records <- left_join(records, camdata, by = c("station", "station_year"))
-
-
-# 1) ADD SOLAR TIMES -------------------------------------------------------
-## add average anchored solar time to species records
-# using methods from Vazquez et al. 2019 https://doi.org/10.1111/2041-210X.13290
-
-# take a matrix of the coordinates
-coords <- matrix(c(records$longitude, records$latitude), nrow = length(records$latitude))
-# add a coordinate reference system (can't be UTM)
-coords <- sp::SpatialPoints(coords, proj4string=sp::CRS("+proj=longlat +datum=WGS84")) 
-# specify detection time as a date class, as well as timezone (use Brisbane because it doesn't have daylight savings)
-records$date_time <- ymd_hms(records$date_time, tz = "Australia/Brisbane")
-records$date_start <- ymd(records$date_start, tz = "Australia/Brisbane")
-records$date_end <- ymd(records$date_end, tz = "Australia/Brisbane")
-
-
-## ADD ADJUSTED HOUR FOR EACH SPECIES DETECTION
-# calculate the time of sunrise given the location and timing of each species detection 
-sunriset_df <- as.data.frame(sunriset(coords, records$date_time, direction = "sunrise", POSIXct.out=TRUE))
-# convert to radian time 
-sunrise_radian <- ClocktimeToTimeRad(sunriset_df$time) 
-# calculate the time of sunset given the location and timing of each species detection 
-sunsett_df <- as.data.frame(sunriset(coords, records$date_time, direction = "sunset", POSIXct.out=TRUE))
-# convert to radian 
[…]

### 2023-07-28 r_scripts/models.R

@@ -1,132 +1,0 @@
-## FIT GAMS TO MODEL DIEL ACTIVITY FOR FOXES AND FERAL CATS
-# MATTHEW REES
-
-library(dplyr)
-library(mgcv)
-
-# load data
-records <- read.csv("derived_data/predator_counts_hour.csv")
-
-
-# Transform variables -----------------------------------------------------
-
-# new variable for fox counts adjusted for survey effort 
-records$fox_count_adj <- records$fox_count / records$survey_duration
-# cats more likely care about fox activity on the log scale
-records$fox_count_adj <- log(records$fox_count_adj + 1)
-
-## Transform variable class  for GAMs
-records <- transform(records,
-                     hour = as.integer(hour),
-                     cat = as.integer(cat), 
-                     fox = as.integer(fox), 
-                     foxbaits = as.integer(foxbaits), 
-                     fox_pa = factor(fox_pa, ordered = FALSE), 
-                     fox_count_adj = as.numeric(fox_count_adj), 
-                     region = factor(region, ordered = FALSE), 
-                     station = factor(station, ordered = FALSE), 
-                     vegetation_group = factor(XGROUPNAME, ordered = FALSE), 
-                     habitat_type = factor(habitat_type, ordered = FALSE), 
-                     survey_duration = as.integer(survey_duration)
-)
-
-# rename habitat type variable (for plotting)
-records$habitat_type <- recode_factor(records$habitat_type,  "dry_glenelg" = "Dry vegetation (Glenelg region)",
-                                      "dry_otway" = "Dry vegetation (Otway Ranges)",
-                                      "wet_otway" = "Wet forest (Otway Ranges)")
-
-
-
-# Changes in spatial activity across the daily cycle ----------------------
-
-# specify penalties for 2D Duchon spline and 1D cc spline
-m <- list(c(1,.5),rep(0,0)) 
-
-gam_fox_sp <- bam(fox ~ t2(x, y, hour, d = c(2, 1), bs = c("ds", "cc"), k = c(80, 8), m = m, full = TRUE) +
-                        s(foxbaits, bs = "tp", k = 4) + 
-                        s(station, bs = "re") +  
-                        offset(log(survey_duration)), 
-                   data = records, family = nb, knots = list(hour = c(0, 23)), nthreads = 3, discrete = TRUE)
-
-
[…]

### 2023-07-28 r_scripts/plot_cat_v_fox_alternative_med.R

@@ -1,55 +1,0 @@
-# PLOT FIGURE
-
-## CAT PLOT - 3D 
-df = expand.grid(hour = 0:23,
-                 habitat_type = levels(records$habitat_type),
-                 longitude = mean(records$longitude),
-                 latitude = mean(records$latitude),
-                 fox_count_adj = seq(min(records$fox_count_adj), max(records$fox_count_adj), length=150),
-                 station = "T052",
-                 survey_duration = 60)
-# predict model results into dataframe
-x = sapply(gam_cat_fox$smooth, "[[",  "label")
-df_int <- cbind(df, predict(gam_cat_fox, newdata = df, se.fit = TRUE, type = "link", exclude = c(x[4:5], "s(survey_duration)")))
-
-## CAT PLOT - 2D (same model at min / max fox)
-# min df 
-df_int_min <- df_int %>%
-  group_by(habitat_type) %>%
-  slice_min(fox_count_adj, n = 1) %>%
-  mutate(minmax = "0min")
-# median df
-df_int_med <- df_int %>% 
-  filter(abs(fox_count_adj - median(fox_count_adj)) == min(abs(fox_count_adj - median(fox_count_adj)))) %>% 
-  mutate(minmax = "1mid")
-# max df 
-df_int_max <- df_int %>%
-  group_by(habitat_type) %>%
-  slice_max(fox_count_adj, n = 1) %>%
-  mutate(minmax = "2max")
-# join df
-df_int_minmax <- bind_rows(df_int_min, df_int_med, df_int_max)
-# rename facet
-df_int_minmax$minmax <- as.factor(df_int_minmax$minmax)
-df_int_minmax$minmax <- recode_factor(df_int_minmax$minmax,  
-                                      "0min" = "Minimum",
-                                      "1mid" = "Median",
-                                      "2max" = "Maximum")
-# plot
-plot_wet_lowhigh <- ggplot(data=df_int_minmax, aes(x=hour, y=fit, group=minmax, col = minmax)) +
-  geom_line(aes(y=fit, x=hour), lwd = 1.2) +
-  facet_wrap(~minmax ~habitat_type, nrow = 3, scales = "free") +
-  ggtitle("", subtitle = "Feral cat") +
-  geom_ribbon(aes(ymin=(fit-2*se.fit), ymax=(fit+2*se.fit), fill = minmax), alpha=0.2) +
[…]

### 2023-07-28 r_scripts/plot_fox_baits_hfi.R

@@ -1,0 +1,127 @@
+# plot spatial effects of human footprint and fox control 
+
+
+# HUMAN FOOTPRINT INDEX PLOT  ---------------------------------------------
+# plot human footprint index  (no effect on diel activity - plot marginal effect only)
+fox_hfi <- smooth_estimates(gam_fox_veg, overall_uncertainty = TRUE, "s(hfi)") %>%
+  mutate(lower = est + (1.96 * se),
+         upper = est - (1.96 * se)) %>% 
+  ggplot(aes(hfi, est)) + 
+  geom_line() + 
+  geom_ribbon(aes(ymin=lower, ymax=upper), alpha=0.2)  + 
+  labs(x = "Human footprint index", y = "Partial effect")
+
+# plot together
+png("figs/fox_hfi.png", width = 6, height = 4, res = 600, units = "in")
+fox_hfi
+dev.off()
+
+
+# ESTIMATES OF FOX SUPPRESSION ----------------------------------------------------------
+# MODEL 1 
+x = sapply(gam_fox_sp$smooth, "[[",  "label")
+df <- expand.grid(hour = 0,
+                  x = mean(records$x),
+                  y = mean(records$y),
+                  foxbaits = min(records$foxbaits):max(records$foxbaits), 
+                  hfi = mean(records$hfi),
+                  region = c("glenelg", "otways"),
+                  survey_duration = 60)
+newdf_1 <- cbind(df, predict.gam(gam_fox_sp, newdata = df, se.fit = TRUE, type = "link", newdata.guaranteed = TRUE, exclude = c(x[c(1,2,3)], "s(survey_duration)"))) %>%
+  mutate(model = "1",
+         fit_resp = exp(fit), 
+         lower = exp(fit - (1.96 * se.fit)),
+         upper = exp(fit + (1.96 * se.fit))) %>%
+  filter(foxbaits == 0 | foxbaits == 19) %>%
+  select(model, region, fit_resp, lower, upper) %>% 
+  mutate(across(c(fit_resp, lower, upper), round, 3))
+
+# MODEL 2
+x = sapply(gam_fox_veg$smooth, "[[",  "label")
+df <- expand.grid(hour = 0,
+                     vegetation_group = "Dry Forest",
+                     foxbaits = min(records$foxbaits):max(records$foxbaits), 
+                     hfi = mean(records$hfi),
[…]

### 2023-07-28 r_scripts/plot_predator_spatiotemporal.R

@@ -1,377 +1,0 @@
-# PLOT SPATIO-TEMPORAL MODELS
-# 1) DIEL ACTIVITY STRENGTH ACROSS SPACE
-# 2) ANIMATED GIF FOR SPATIOTEMPORAL ACTIVITY
-
-# additional packages
-library(gganimate)
-library(sp)
-library(sf)
-library(viridis)
-
-# split records
-records_g <- filter(records, region == "glenelg")
-records_o <- filter(records, region == "otways")
-
-# cam locations 
-camdata <- records %>%
-  distinct(., station, .keep_all = TRUE) %>%
-  select(region, station, x, y) 
-# add new variable to match variable we will facet by - region x species
-camdata_fox <- camdata
-camdata_fox$species <- "Red fox"
-camdata_cat <- camdata
-camdata_cat$species <- "Feral cat"
-camdata <- rbind(camdata_fox, camdata_cat)
-camdata$facet <- paste0(camdata$species, "_", camdata$region)
-unique(camdata$facet)
-
-# MAKE SPATIAL DATAFRAMES TO PREDICT INTO ----------------------------------------------
-
-## GLENELG:
-# take just the cams
-records_g_cams <- distinct(records_g, station, .keep_all = TRUE)
-# change it to sf class
-records_g_cams <- st_as_sf(records_g_cams, coords = c("x", "y"), crs = 32754) 
-# make a 4km buffer around each camera
-glenelg_cams_buffer = st_buffer(records_g_cams, 4000)
-# dissolve the buffer
-glenelg_cams_buffer = st_union(glenelg_cams_buffer)
-# convert back to dataframe
-glenelg_buffer_df <- fortify(as_Spatial(glenelg_cams_buffer))%>%
-  transmute(x = long, y = lat, order = order, group = group)
-# split by each group (seperate polygon for each grid)
-buffer_df1 <- glenelg_buffer_df[which(glenelg_buffer_df$group == "ID1.1"),]
-buffer_df2 <- glenelg_buffer_df[which(glenelg_buffer_df$group == "ID1.2"),]
-buffer_df3 <- glenelg_buffer_df[which(glenelg_buffer_df$group == "ID1.3"),]
-buffer_df4 <- glenelg_buffer_df[which(glenelg_buffer_df$group == "ID1.4"),]
-# make a dataframe from this to predict into
-data_g_plot = expand.grid(
-  x = seq(min(glenelg_buffer_df$x), 
[…]

### 2023-07-28 r_scripts/plot_predator_vegetation_type.R

@@ -1,44 +1,0 @@
-
-
-# function to run and plot GAMs for each species
-gam_plot <- function(gam_model, data){
-  ## Plot global smooth (using the gratia package)
-  plot_global <- draw(evaluate_smooth(gam_model, overall_uncertainty = TRUE, "s(hour)"), xlab = "Hour",  title = "", subtitle = "i.   Average") + 
-    geom_vline(xintercept = c(6.16,18.34), colour = "black", size = 0.6, linetype="dotted") 
-  ## Plot the difference smooths using the gratia package
-  plot_difference <- draw(evaluate_smooth(gam_model, "s(hour,vegetation_group)"), xlab = "Hour",  title = "", subtitle = "ii.   Vegetation-specific deviations from the average") + 
-    geom_vline(xintercept = c(6.16,18.34), colour = "black", size = 0.6, linetype="dotted") 
-  ## Plot the actual hourly activity for each vegetation type   
-  # make a new dataframe to predict into
-  df <- expand.grid(hour = 0:23,
-                    vegetation_group = levels(data$vegetation_group),
-                    foxbaits = 0, 
-                    region = c("glenelg", "otways"),
-                    survey_duration = 60)
-  # get smooths to exlude
-  x = sapply(gam_model$smooth, "[[",  "label")
-  # predict model estimates (and uncertainty) into this dataframe, excluding the impact of variables other than hour:
-  df <- cbind(df, predict.gam(gam_model, newdata = df, se.fit = TRUE, type = "link", newdata.guaranteed = TRUE, exclude = c(x[3:7])))
-  plot_veg <- ggplot(data=df, aes(x=hour, y=fit, group=vegetation_group)) +
-    # ylim(-13, -2) +
-    geom_line(aes(y=fit, x=hour), lwd = 0.7) +
-    facet_wrap(~vegetation_group, nrow =1) +
-    geom_ribbon(aes(ymin=(fit-2*se.fit), ymax=(fit+2*se.fit)), alpha=0.2) +
-    geom_vline(xintercept = c(6.16,18.34), colour = "black", size = 0.6, linetype="dotted") + 
-    labs(subtitle = "iii.   Predicted activity", x = "Hour", y = "log(count)") 
-  ## Assemble plots
-  patch <- (plot_global | plot_difference) / plot_veg 
-  patch
-}
-
-# save
-png("figs/fox_veg.png", width = 9, height = 5, res = 600, units = "in")
[…]

### 2023-07-28 r_scripts/plot_spatial_only_interaction.R

@@ -1,406 +1,0 @@
-# PLOT SEPERATED MARGINAL AND INTERACTION EFFECTS FOR SPATIAL MODELS (1)
-
-
-# RUN MODEL WITH TI TENSOR PRODUCT ----------------------
-# so we can separate marginal effects from interaction
-
-# specify penalties for 2D Duchon spline and 1D cc spline
-m <- list(c(1,.5),rep(0,0)) 
-
-gam_fox_sp <- bam(fox ~ s(hour, bs = "cc", k = 8) +
-                        s(x, y, bs = "ds", k = 80, m = c(1,.5)) +
-                        ti(x, y, hour, d = c(2, 1), bs = c("ds", "cc"), k = c(80, 8), m = m) +
-                        s(foxbaits, bs = "tp", k = 4) + 
-                        s(station, bs = "re") +  
-                        offset(log(survey_duration)), 
-                  data = records, family = nb, knots = list(hour = c(0, 23)), nthreads = 3, discrete = TRUE)
-
-
-gam_cat_sp <- bam(cat ~ s(hour, bs = "cc", k = 8) +
-                        s(x, y, bs = "ds", k = 80, m = c(1,.5)) +
-                        ti(x, y, hour, d = c(2, 1), bs = c("ds", "cc"), k = c(80, 8), m = m) +
-                        s(station, bs = "re") +  
-                        offset(log(survey_duration)), 
-                  data = records, family = nb, knots = list(hour = c(0, 23)), nthreads = 3, discrete = TRUE)
-
-
-
-
-
-# PLOT AVERAGE HOUR -------------------------------------------------------
-
-# new df 
-df <- expand.grid(x = 625380,
-                  y = 5760730,
-                  survey_duration = 60,
-                  foxbaits = 0, 
-                  hour = 0:23,
-                  station = "A001")
-
-
-# predict fox
-x = sapply(gam_fox_sp$smooth, "[[",  "label")
-df_hr_fox <- cbind(df, predict.gam(gam_fox_sp, newdata = df, se.fit = TRUE, type = "link", exclude = c(x[2:4], x[5])))
-df_hr_fox$species <- "fox"
-
-# predict cat
-x = sapply(gam_cat_sp$smooth, "[[",  "label")
-df_hr_cat <- cbind(df, predict.gam(gam_cat_sp, newdata = df, se.fit = TRUE, type = "link", exclude = c(x[2:4])))
-df_hr_cat$species <- "cat"
[…]

### 2023-07-28 r_scripts/test.R

@@ -1,253 +1,0 @@
-# PLOT SPATIO-TEMPORAL MODELS
-# 1) DIEL ACTIVITY STRENGTH ACROSS SPACE
-# 2) ANIMATED GIF FOR SPATIOTEMPORAL ACTIVITY
-
-
-gam_fox_sp <- bam(fox ~ t2(x, y, hour, d = c(2, 1), bs = c("tp", "cc"), k = c(80, 8), full = TRUE) +
-                    s(foxbaits, bs = "tp", k = 4) + 
-                    s(station, bs = "re") +  
-                    offset(log(survey_duration)), 
-                  data = records, family = nb, knots = list(hour = c(0, 23)), nthreads = 3, discrete = TRUE)
-
-gam_cat_sp <- bam(cat ~ t2(x, y, hour, d = c(2, 1), bs = c("tp", "cc"), k = c(80, 8),full = TRUE) +
-                    s(station, bs = "re") +  
-                    offset(log(survey_duration)), 
-                  data = records, family = nb, knots = list(hour = c(0, 23)), nthreads = 3, discrete = TRUE)
-
-
-
-
-gam_fox_sp <- bam(fox ~ ti(x, y, bs = "tp", k = 80) + 
-                        ti(hour, bs = "cc", k = 8) +
-                        ti(x, y, hour, d = c(2, 1), bs = c("ds", "tp"), k = c(80, 8)) +
-                        s(station, bs = "re") +  
-                        offset(log(survey_duration)), 
-                  data = records, family = nb, knots = list(hour = c(0, 23)), nthreads = 3, discrete = FALSE)
-
-
-
-# additional packages
-library(sp)
-library(sf)
-library(viridis)
-
-# split records
-records_g <- filter(records, region == "glenelg")
-records_o <- filter(records, region == "otways")
-
-# cam locations 
-camdata <- records %>%
-  distinct(., station, .keep_all = TRUE) %>%
-  select(region, station, x, y) 
-# add new variable to match variable we will facet by - region x species
-camdata_fox <- camdata
-camdata_fox$species <- "Red fox"
-camdata_cat <- camdata
-camdata_cat$species <- "Feral cat"
-camdata <- rbind(camdata_fox, camdata_cat)
-camdata$facet <- paste0(camdata$species, "_", camdata$region)
-unique(camdata$facet)
-
[…]

### 2023-11-01 fit_gams.R
oikos accepted clean

@@ -1,0 +1,150 @@
+## FIT GAMS TO MODEL DIEL ACTIVITY FOR FOXES AND FERAL CATS
+# CODE FOR OIKOS PAPER 'DYNAMIC SHIFTS IN PREDATOR DIEL ACTIVITY PATTERNS ACROSS LANDSCAPES AND THREAT-LEVELS'
+# MATTHEW REES
+
+
+# SET-UP ------------------------------------------------------------------
+
+# load r-packages (all except for mgcv and dplyr are for plotting)
+library(dplyr)
+library(mgcv)
+library(gratia)
+library(ggplot2)
+library(patchwork)
+library(ggpubr)
+library(sp)
+library(sf)
+library(viridis)
+library(forcats)
+
+# load plotting functions
+source("plotting_functions/function_plot_spatiotemporal_interaction.R")
+source("plotting_functions/function_plot_vegetation_type.R")
+source("plotting_functions/function_plot_fox_human_footprint.R")
+source("plotting_functions/function_plot_cat_v_fox.R")
+source("plotting_functions/function_plot_fox_foxbaits.R")
+
+# load data: response variable = fox / cat detection counts at a camera-trap site ('station') for each hour of the day (0-23)
+records <- read.csv("data_predator_counts_hour.csv")
+
+# set the default theme for ggplot objects 
+theme_set(theme_bw())
+theme_update(panel.grid = element_blank())
+
+
+# Transform variables -----------------------------------------------------
+records <- transform(records,
+                     region = factor(region, ordered = FALSE), 
+                     station = factor(station, ordered = FALSE), 
+                     x = as.numeric(x),
+                     y = as.numeric(y),
+                     hour = as.integer(hour),
+                     cat = as.integer(cat), 
+                     fox = as.integer(fox), 
+                     fox_count_adj = as.numeric(fox_count_adj), 
+                     foxbaits = as.integer(foxbaits), 
+                     vegetation_group = factor(vegetation_group, ordered = FALSE), 
+                     habitat_type = factor(habitat_type, ordered = FALSE), 
+                     survey_duration = as.integer(survey_duration),
+                     hfi = round(hfi, digits = 2))
+
+
+
+
+# 1) Interaction between spatial and diel activity ('model 1') ---------------------
+
[…]

### 2023-11-01 old_code/r_scripts/1_reformat_data_by_hour.R

@@ -1,0 +1,192 @@
+# This script reformats data from a row per detection, to a row per hour at a site with counts of fox and cat detections
+# Also converts detection times to solar time and adds human footprint index variable 
+
+
+# load r packages
+library(maptools)        
+library(dplyr)
+library(activity) 
+library(lubridate)
+library(reshape2)
+library(sf)
+library(hms)
+library(terra)
+
+# set terra tempdir (for cleanup) + raster progress bar
+terraOptions(tempdir ="temp/", progress = 1, memfrac = 0.8)
+
+# load function to change clock time to radial time for overlap package
+source("r_scripts/function_clock_time_to_radian.R")
+
+
+
+# LOAD / PREPARE DATA -----------------------------------------------------
+# load human footprint index raster (1km)
+hfi <- terra::rast("raw_data/wildareas-v3-2009-human-footprint-geotiff/wildareas-v3-2009-human-footprint.tif")
+
+# records of species detections (1 row per detection)
+records <- read.csv("raw_data/predator_records.csv")
+# camera-trap deplopyment / site information 
+camdata <- read.csv("raw_data/camdata.csv")
+# stitch together
+records <- left_join(records, camdata, by = c("station", "station_year"))
+
+
+
+# 1) ADD SOLAR TIMES -------------------------------------------------------
+## add average anchored solar time to species records
+# using methods from Vazquez et al. 2019 https://doi.org/10.1111/2041-210X.13290
+
+# take a matrix of the coordinates
+coords <- matrix(c(records$longitude, records$latitude), nrow = length(records$latitude))
+# add a coordinate reference system (can't be UTM)
+coords <- sp::SpatialPoints(coords, proj4string=sp::CRS("+proj=longlat +datum=WGS84")) 
+# specify detection time as a date class, as well as timezone (use Brisbane because it doesn't have daylight savings)
+records$date_time <- ymd_hms(records$date_time, tz = "Australia/Brisbane")
+records$date_start <- ymd(records$date_start, tz = "Australia/Brisbane")
+records$date_end <- ymd(records$date_end, tz = "Australia/Brisbane")
+
+
+## ADD ADJUSTED HOUR FOR EACH SPECIES DETECTION
+# calculate the time of sunrise given the location and timing of each species detection 
[…]

### 2023-11-01 old_code/r_scripts/2_GAMs.R

@@ -1,0 +1,120 @@
+## FIT GAMS TO MODEL DIEL ACTIVITY FOR FOXES AND FERAL CATS
+# MATTHEW REES
+
+library(dplyr)
+library(mgcv)
+# for plots
+library(gratia)
+library(ggplot2)
+library(patchwork)
+library(ggpubr)
+library(viridis)
+library(forcats)
+# Set the default theme for ggplot objects 
+theme_set(theme_bw())
+theme_update(panel.grid = element_blank())
+
+
+# load data
+records <- read.csv("derived_data/predator_counts_hour.csv")
+
+
+# Transform variables -----------------------------------------------------
+records <- transform(records,
+                      hour = as.integer(hour),
+                      cat = as.integer(cat), 
+                      fox = as.integer(fox), 
+                      foxbaits = as.integer(foxbaits), 
+                      fox_pa = factor(fox_pa, ordered = FALSE), 
+                      fox_count_adj = as.numeric(fox_count_adj), 
+                      region = factor(region, ordered = FALSE), 
+                      station = factor(station, ordered = FALSE), 
+                      vegetation_group = factor(XGROUPNAME, ordered = FALSE), 
+                      habitat_type = factor(habitat_type, ordered = FALSE), 
+                      survey_duration = as.integer(survey_duration),
+                      hfi = round(hfi, digits = 2))
+
+
+
+
+# 1) Interaction between spatial and diel activity ('model 1') ---------------------
+# specify penalties for 2D Duchon spline and 1D cc spline
+m <- list(c(1,.5),rep(0,0)) 
+
+gam_fox_sp <- bam(fox ~ s(hour, bs = "cc", k = 8) + 
+                        s(x, y, bs = "ds", k = 80, m = c(1,.5)) + 
+                        ti(x, y, hour, d = c(2, 1), bs = c("ds", "cc"), k = c(80, 8), m = m) +                     
+                        s(foxbaits, region, bs = "fs", xt = list(bs = "tp"), k = 4) + 
+                        offset(log(survey_duration)), 
+                  data = records, family = nb, knots = list(hour = c(0, 23)), discrete = TRUE)
+
+gam_cat_sp <- bam(cat ~ s(hour, bs = "cc", k = 8) + 
+                        s(x, y, bs = "ds", k = 80, m = c(1,.5)) + 
+                        ti(x, y, hour, d = c(2, 1), bs = c("ds", "cc"), k = c(80, 8), m = m) +                     
[…]

### 2023-11-01 old_code/r_scripts/plot_cat_v_fox.R

@@ -1,0 +1,79 @@
+# PLOT FIGURE
+
+## CAT PLOT - 3D 
+df = expand.grid(hour = 0:23,
+                 habitat_type = levels(records$habitat_type),
+                 longitude = mean(records$longitude),
+                 latitude = mean(records$latitude),
+                 fox_count_adj = seq(min(records$fox_count_adj), max(records$fox_count_adj), length=150),
+                 station = "T052",
+                 survey_duration = 60)
+# predict model results into dataframe
+x = sapply(gam_cat_fox$smooth, "[[",  "label")
+df_int <- cbind(df, predict(gam_cat_fox, newdata = df, se.fit = TRUE, type = "link", exclude = c(x[4:5], "s(survey_duration)")))
+# plot
+plot_range <- ggplot(aes(hour, fox_count_adj, fill = fit), data = df_int) +
+  geom_tile() +
+  facet_wrap(~habitat_type, nrow =1) +
+  geom_contour(aes(z = fit, fill = NULL), colour = "grey40", bins = 10) +
+  scale_fill_viridis("Cat activity", option = "viridis", breaks = c(-5, -3), labels = c("Low", "High")) + 
+  ggtitle("", subtitle = "Feral cat") +
+  geom_vline(xintercept = c(6.16,18.34), colour = "black", size = 0.6, linetype="dotted") + 
+  scale_y_continuous(breaks=c(min(records$fox_count_adj), max(records$fox_count_adj)), labels = c("Low", "High")) +
+  ylab("Fox activity") + 
+  xlab("Hour") 
+
+## CAT PLOT - 2D (same model at min / max fox)
+# min df 
+df_int_min <- df_int %>%
+  group_by(habitat_type) %>%
+  slice_min(fox_count_adj, n = 1) %>%
+  mutate(minmax = "0min")
+# max df 
+df_int_max <- df_int %>%
+  group_by(habitat_type) %>%
+  slice_max(fox_count_adj, n = 1) %>%
+  mutate(minmax = "1max")
+# join df
+df_int_minmax <- rbind(df_int_min, df_int_max)
+# plot
+plot_wet_lowhigh <- ggplot(data=df_int_minmax, aes(x=hour, y=fit, group=minmax, col = minmax)) +
+  geom_line(aes(y=fit, x=hour), lwd = 1.2) +
+  facet_wrap(~habitat_type, nrow =1) +
[…]

### 2023-11-01 old_code/r_scripts/plot_fox_baits_hfi.R

@@ -1,0 +1,104 @@
+# plot spatial effects of human footprint and fox control 
+
+
+# HUMAN FOOTPRINT INDEX PLOT  ---------------------------------------------
+# plot human footprint index  (no effect on diel activity - plot marginal effect only)
+fox_hfi <- smooth_estimates(gam_fox_veg, overall_uncertainty = TRUE, "s(hfi)") %>%
+  mutate(lower = est + (1.96 * se),
+         upper = est - (1.96 * se)) %>% 
+  ggplot(aes(hfi, est)) + 
+  geom_line() + 
+  geom_ribbon(aes(ymin=lower, ymax=upper), alpha=0.2)  + 
+  labs(x = "Human footprint index", y = "Partial effect")
+
+# plot together
+png("figs/fox_hfi.png", width = 6, height = 4, res = 600, units = "in")
+fox_hfi
+dev.off()
+
+
+# ESTIMATES OF FOX SUPPRESSION ----------------------------------------------------------
+# MODEL 1 
+x = sapply(gam_fox_sp$smooth, "[[",  "label")
+df <- expand.grid(hour = 0,
+                  x = mean(records$x),
+                  y = mean(records$y),
+                  foxbaits = min(records$foxbaits):max(records$foxbaits), 
+                  hfi = mean(records$hfi),
+                  region = c("glenelg", "otways"),
+                  survey_duration = 60)
+newdf_1 <- cbind(df, predict.gam(gam_fox_sp, newdata = df, se.fit = TRUE, type = "link", newdata.guaranteed = TRUE, exclude = c(x[c(1,2,3)], "s(survey_duration)"))) %>%
+  mutate(model = "1",
+         fit_resp = exp(fit), 
+         lower = exp(fit - (1.96 * se.fit)),
+         upper = exp(fit + (1.96 * se.fit))) %>%
+ # filter(foxbaits == 0 | foxbaits == 19) %>%
+  select(model, region, foxbaits, fit_resp, lower, upper) %>% 
+  mutate(across(c(fit_resp, lower, upper), round, 3))
+
+# MODEL 2
+x = sapply(gam_fox_veg$smooth, "[[",  "label")
+df <- expand.grid(hour = 0,
+                     vegetation_group = "Dry Forest",
+                     foxbaits = min(records$foxbaits):max(records$foxbaits), 
[…]

### 2023-11-01 old_code/r_scripts/plot_spatiotemporal_interaction.R

@@ -1,0 +1,255 @@
+# PLOT SPATIO-TEMPORAL MODELS
+# 1) DIEL ACTIVITY STRENGTH ACROSS SPACE
+# 2) ANIMATED GIF FOR SPATIOTEMPORAL ACTIVITY
+
+# additional packages
+library(sp)
+library(sf)
+library(viridis)
+
+
+# split records
+records_g <- filter(records, region == "glenelg")
+records_o <- filter(records, region == "otways")
+
+# cam locations 
+camdata <- records %>%
+  distinct(., station, .keep_all = TRUE) %>%
+  select(region, station, x, y) 
+# add new variable to match variable we will facet by - region x species
+camdata_fox <- camdata
+camdata_fox$species <- "Red fox"
+camdata_cat <- camdata
+camdata_cat$species <- "Feral cat"
+camdata <- rbind(camdata_fox, camdata_cat)
+camdata$facet <- paste0(camdata$species, "_", camdata$region)
+
+
+# MAKE SPATIAL DATAFRAMES TO PREDICT INTO ----------------------------------------------
+
+## GLENELG:
+# take just the cams
+records_g_cams <- distinct(records_g, station, .keep_all = TRUE)
+# change it to sf class
+records_g_cams <- st_as_sf(records_g_cams, coords = c("x", "y"), crs = 3111) 
+# make a 4km buffer around each camera
+glenelg_cams_buffer = st_buffer(records_g_cams, 4000)
+# dissolve the buffer
+glenelg_cams_buffer = st_union(glenelg_cams_buffer)
+# but bring in the edges - don't want to predict outside of surveyed area
+glenelg_cams_buffer <- st_buffer(glenelg_cams_buffer, dist = -3500, joinStyle  = "ROUND")
+# convert back to dataframe
+glenelg_buffer_df <- fortify(as_Spatial(glenelg_cams_buffer))%>%
+  transmute(x = long, y = lat, order = order, group = group)
+# split by each group (seperate polygon for each grid)
+buffer_df1 <- glenelg_buffer_df[which(glenelg_buffer_df$group == "ID1.1"),]
+buffer_df2 <- glenelg_buffer_df[which(glenelg_buffer_df$group == "ID1.2"),]
+buffer_df3 <- glenelg_buffer_df[which(glenelg_buffer_df$group == "ID1.3"),]
+buffer_df4 <- glenelg_buffer_df[which(glenelg_buffer_df$group == "ID1.4"),]
+# make this a list for x and y cols
[…]

### 2023-11-01 old_code/r_scripts/plot_vegetation_type.R

@@ -1,0 +1,47 @@
+
+
+# function to run and plot GAMs for each species
+gam_plot <- function(gam_model, data){
+  ## Plot global smooth (using the gratia package)
+  plot_global <- draw(evaluate_smooth(gam_model, overall_uncertainty = TRUE, "s(hour)"), xlab = "Hour",  title = "", subtitle = "i.   Average") + 
+    geom_vline(xintercept = c(6.16,18.34), colour = "black", size = 0.6, linetype="dotted") 
+  ## Plot the difference smooths using the gratia package
+  plot_difference <- draw(evaluate_smooth(gam_model, "s(hour,vegetation_group)"), xlab = "Hour",  title = "", subtitle = "ii.   Vegetation-specific deviations from the average") + 
+    geom_vline(xintercept = c(6.16,18.34), colour = "black", size = 0.6, linetype="dotted") 
+  ## Plot the actual hourly activity for each vegetation type   
+  # make a new dataframe to predict into
+  df <- expand.grid(hour = 0:23,
+                    vegetation_group = levels(data$vegetation_group),
+                    foxbaits = 0, 
+                    hfi = mean(data$hfi),
+                    region = c("glenelg", "otways"),
+                    survey_duration = 60)
+  # get smooths to exlude
+  x = sapply(gam_model$smooth, "[[",  "label")
+  # predict model estimates (and uncertainty) into this dataframe, excluding the impact of variables other than hour:
+  df <- cbind(df, predict.gam(gam_model, newdata = df, se.fit = TRUE, type = "link", newdata.guaranteed = TRUE, exclude = c(x[3:7])))
+  plot_veg <- ggplot(data=df, aes(x=hour, y=fit, group=vegetation_group)) +
+    # ylim(-13, -2) +
+    geom_line(aes(y=fit, x=hour), lwd = 0.7) +
+    facet_wrap(~vegetation_group, nrow =1) +
+    geom_ribbon(aes(ymin=(fit-2*se.fit), ymax=(fit+2*se.fit)), alpha=0.2) +
+    geom_vline(xintercept = c(6.16,18.34), colour = "black", size = 0.6, linetype="dotted") + 
+    labs(subtitle = "iii.   Predicted activity in each vegetation type", x = "Hour", y = "log(count)") 
+  ## Assemble plots
+  patch <- (plot_global | plot_difference) / plot_veg 
+  patch
+}
+
+# save
[…]

### 2023-11-01 plotting_functions/function_plot_cat_v_fox.R

@@ -1,0 +1,76 @@
+plot_model_3 <- function(){
+  
+  ## CAT PLOT 3D  
+  # new dataframe to predict into
+  df = expand.grid(hour = 0:23,
+                   habitat_type = levels(records$habitat_type),
+                  # longitude = mean(records$x),
+                  # latitude = mean(records$y),
+                   fox_count_adj = seq(min(records$fox_count_adj), max(records$fox_count_adj), length=150),
+                   station = "T052",
+                   survey_duration = 60)
+  # predict model results into dataframe
+  df_int <- cbind(df, predict(gam_cat_fox, newdata = df, se.fit = TRUE, type = "link", exclude = c("s(station)", "s(survey_duration)")))
+  # plot
+  plot_range <- ggplot(aes(hour, fox_count_adj, fill = fit), data = df_int) +
+    geom_tile() +
+    facet_wrap(~habitat_type, nrow =1) +
+    geom_contour(aes(z = fit, fill = NULL), colour = "grey40", bins = 10) +
+    scale_fill_viridis("Cat activity", option = "viridis", breaks = c(-5, -3), labels = c("Low", "High")) + 
+    ggtitle("", subtitle = "Feral cat") +
+    geom_vline(xintercept = c(6.16,18.34), colour = "black", size = 0.6, linetype="dotted") + 
+    scale_y_continuous(breaks=c(min(records$fox_count_adj), max(records$fox_count_adj)), labels = c("Low", "High")) +
+    ylab("Fox activity") + 
+    xlab("Hour") 
+  
+  ## CAT PLOT - 2D (same model at min / max fox)
+  # min df 
+  df_int_min <- df_int %>%
+    group_by(habitat_type) %>%
+    slice_min(fox_count_adj, n = 1) %>%
+    mutate(minmax = "0min")
+  # max df 
+  df_int_max <- df_int %>%
+    group_by(habitat_type) %>%
+    slice_max(fox_count_adj, n = 1) %>%
+    mutate(minmax = "1max")
+  # join df
+  df_int_minmax <- rbind(df_int_min, df_int_max)
+  # plot
+  plot_wet_lowhigh <- ggplot(data=df_int_minmax, aes(x=hour, y=fit, group=minmax, col = minmax)) +
+    geom_line(aes(y=fit, x=hour), lwd = 1.2) +
+    facet_wrap(~habitat_type, nrow =1) +
[…]

### 2023-11-01 plotting_functions/function_plot_fox_foxbaits.R

@@ -1,0 +1,86 @@
+plot_foxbaits <- function(){
+
+  # ESTIMATES OF FOX SUPPRESSION ----------------------------------------------------------
+  # MODEL 1 
+  x = sapply(gam_fox_sp$smooth, "[[",  "label")
+  df <- expand.grid(hour = 0,
+                    x = mean(records$x),
+                    y = mean(records$y),
+                    foxbaits = min(records$foxbaits):max(records$foxbaits), 
+                    hfi = mean(records$hfi),
+                    region = c("glenelg", "otways"),
+                    survey_duration = 60)
+  newdf_1 <- cbind(df, predict.gam(gam_fox_sp, newdata = df, se.fit = TRUE, type = "link", newdata.guaranteed = TRUE, exclude = c(x[c(1,2,3)], "s(survey_duration)"))) %>%
+    mutate(model = "1",
+           fit_resp = exp(fit), 
+           lower = exp(fit - (1.96 * se.fit)),
+           upper = exp(fit + (1.96 * se.fit))) %>%
+   # filter(foxbaits == 0 | foxbaits == 19) %>%
+    select(model, region, foxbaits, fit_resp, lower, upper) %>% 
+    mutate(across(c(fit_resp, lower, upper), round, 3))
+  
+  # MODEL 2
+  x = sapply(gam_fox_veg$smooth, "[[",  "label")
+  df <- expand.grid(hour = 0,
+                       vegetation_group = "Dry Forest",
+                       foxbaits = min(records$foxbaits):max(records$foxbaits), 
+                       hfi = mean(records$hfi),
+                       region = c("glenelg", "otways"),
+                       survey_duration = 60)
+  
+  newdf_2 <- cbind(df, predict.gam(gam_fox_veg, newdata = df, se.fit = TRUE, type = "link", newdata.guaranteed = TRUE, exclude = c(x[c(1,2,4,5,6, 7)], "s(survey_duration)"))) %>%
+    mutate(model = "2",
+           fit_resp = exp(fit), 
+           lower = exp(fit - (1.96 * se.fit)),
+           upper = exp(fit + (1.96 * se.fit))) %>%
+    #filter(foxbaits == 0 | foxbaits == 19) %>%
+    select(model, region, foxbaits, fit_resp, lower, upper) %>% 
+    mutate(across(c(fit_resp, lower, upper), round, 3))
+           
[…]

### 2023-11-01 plotting_functions/function_plot_fox_human_footprint.R

@@ -1,0 +1,17 @@
+plot_fox_hfi <- function(){
+
+  # plot human footprint index  (no effect on diel activity - plot marginal effect only)
+  plot <- smooth_estimates(gam_fox_veg, overall_uncertainty = TRUE, "s(hfi)") %>%
+    mutate(lower = est + (1.96 * se),
+           upper = est - (1.96 * se)) %>% 
+    ggplot(aes(hfi, est)) + 
+    geom_line() + 
+    geom_ribbon(aes(ymin=lower, ymax=upper), alpha=0.2)  + 
+    labs(x = "Human footprint index", y = "Partial effect")
+  
+  # plot together
+  png("figs/fox_hfi.png", width = 6, height = 4, res = 600, units = "in")
+  print(plot)
+  dev.off()
+  
+}

### 2023-11-01 plotting_functions/function_plot_spatiotemporal_interaction.R

@@ -1,0 +1,248 @@
+plot_model_1 <- function(){
+  
+  # split records
+  records_g <- filter(records, region == "glenelg")
+  records_o <- filter(records, region == "otways")
+  
+  # cam locations 
+  camdata <- records %>%
+    distinct(., station, .keep_all = TRUE) %>%
+    select(region, station, x, y) 
+  # add new variable to match variable we will facet by - region x species
+  camdata_fox <- camdata
+  camdata_fox$species <- "Red fox"
+  camdata_cat <- camdata
+  camdata_cat$species <- "Feral cat"
+  camdata <- rbind(camdata_fox, camdata_cat)
+  camdata$facet <- paste0(camdata$species, "_", camdata$region)
+  
+  
+  # MAKE SPATIAL DATAFRAMES TO PREDICT INTO ----------------------------------------------
+  
+  ## GLENELG:
+  # take just the cams
+  records_g_cams <- distinct(records_g, station, .keep_all = TRUE)
+  # change it to sf class
+  records_g_cams <- st_as_sf(records_g_cams, coords = c("x", "y"), crs = 3111) 
+  # make a 4km buffer around each camera
+  glenelg_cams_buffer = st_buffer(records_g_cams, 4000)
+  # dissolve the buffer
+  glenelg_cams_buffer = st_union(glenelg_cams_buffer)
+  # but bring in the edges - don't want to predict outside of surveyed area
+  glenelg_cams_buffer <- st_buffer(glenelg_cams_buffer, dist = -3500, joinStyle  = "ROUND")
+  # convert back to dataframe
+  glenelg_buffer_df <- fortify(as_Spatial(glenelg_cams_buffer))%>%
+    transmute(x = long, y = lat, order = order, group = group)
+  # split by each group (seperate polygon for each grid)
+  buffer_df1 <- glenelg_buffer_df[which(glenelg_buffer_df$group == "ID1.1"),]
+  buffer_df2 <- glenelg_buffer_df[which(glenelg_buffer_df$group == "ID1.2"),]
+  buffer_df3 <- glenelg_buffer_df[which(glenelg_buffer_df$group == "ID1.3"),]
+  buffer_df4 <- glenelg_buffer_df[which(glenelg_buffer_df$group == "ID1.4"),]
+  # make this a list for x and y cols
+  buffer_df1 <- list(buffer_df1$x, buffer_df1$y)
+  buffer_df2 <- list(buffer_df2$x, buffer_df2$y)
[…]

### 2023-11-01 plotting_functions/function_plot_vegetation_type.R

@@ -1,0 +1,33 @@
+plot_model_2 <- function(gam_model, data){
+  
+  ## Plot global smooth (using the gratia package)
+  plot_global <- draw(smooth_estimates(gam_model, overall_uncertainty = TRUE, "s(hour)"), xlab = "Hour",  title = "", subtitle = "i.   Average") + 
+    geom_vline(xintercept = c(6.16,18.34), colour = "black", linewidth = 0.6, linetype="dotted") 
+  
+  ## Plot the difference smooths using the gratia package
+  plot_difference <- draw(smooth_estimates(gam_model, "s(hour,vegetation_group)"), xlab = "Hour",  title = "", subtitle = "ii.   Vegetation-specific deviations from the average") + 
+    geom_vline(xintercept = c(6.16,18.34), colour = "black", linewidth = 0.6, linetype="dotted") 
+  
+  ## Plot the actual hourly activity for each vegetation type   
+  # make a new dataframe to predict into
+  df <- expand.grid(hour = 0:23,
+                    vegetation_group = levels(data$vegetation_group),
+                    foxbaits = 0, 
+                    hfi = mean(data$hfi),
+                    region = c("glenelg", "otways"),
+                    survey_duration = 60)
+  # get smooths to exlude
+  x = sapply(gam_model$smooth, "[[",  "label")
+  # predict model estimates (and uncertainty) into this dataframe, excluding the impact of variables other than hour:
+  df <- cbind(df, predict.gam(gam_model, newdata = df, se.fit = TRUE, type = "link", newdata.guaranteed = TRUE, exclude = c(x[3:7])))
+  plot_veg <- ggplot(data=df, aes(x=hour, y=fit, group=vegetation_group)) +
+    # ylim(-13, -2) +
+    geom_line(aes(y=fit, x=hour), lwd = 0.7) +
+    facet_wrap(~vegetation_group, nrow =1) +
+    geom_ribbon(aes(ymin=(fit-2*se.fit), ymax=(fit+2*se.fit)), alpha=0.2) +
+    geom_vline(xintercept = c(6.16,18.34), colour = "black", linewidth = 0.6, linetype="dotted") + 
+    labs(subtitle = "iii.   Predicted activity in each vegetation type", x = "Hour", y = "log(count)") 
+  
+  ## Assemble plots
+  print((plot_global | plot_difference) / plot_veg)
+}

### 2023-11-01 r_scripts/1_reformat_data_by_hour.R

@@ -1,192 +1,0 @@
-# This script reformats data from a row per detection, to a row per hour at a site with counts of fox and cat detections
-# Also converts detection times to solar time and adds human footprint index variable 
-
-
-# load r packages
-library(maptools)        
-library(dplyr)
-library(activity) 
-library(lubridate)
-library(reshape2)
-library(sf)
-library(hms)
-library(terra)
-
-# set terra tempdir (for cleanup) + raster progress bar
-terraOptions(tempdir ="temp/", progress = 1, memfrac = 0.8)
-
-# load function to change clock time to radial time for overlap package
-source("r_scripts/function_clock_time_to_radian.R")
-
-
-
-# LOAD / PREPARE DATA -----------------------------------------------------
-# load human footprint index raster (1km)
-hfi <- terra::rast("raw_data/wildareas-v3-2009-human-footprint-geotiff/wildareas-v3-2009-human-footprint.tif")
-
-# records of species detections (1 row per detection)
-records <- read.csv("raw_data/predator_records.csv")
-# camera-trap deplopyment / site information 
-camdata <- read.csv("raw_data/camdata.csv")
-# stitch together
-records <- left_join(records, camdata, by = c("station", "station_year"))
-
-
-
-# 1) ADD SOLAR TIMES -------------------------------------------------------
-## add average anchored solar time to species records
-# using methods from Vazquez et al. 2019 https://doi.org/10.1111/2041-210X.13290
-
-# take a matrix of the coordinates
-coords <- matrix(c(records$longitude, records$latitude), nrow = length(records$latitude))
-# add a coordinate reference system (can't be UTM)
-coords <- sp::SpatialPoints(coords, proj4string=sp::CRS("+proj=longlat +datum=WGS84")) 
-# specify detection time as a date class, as well as timezone (use Brisbane because it doesn't have daylight savings)
-records$date_time <- ymd_hms(records$date_time, tz = "Australia/Brisbane")
-records$date_start <- ymd(records$date_start, tz = "Australia/Brisbane")
-records$date_end <- ymd(records$date_end, tz = "Australia/Brisbane")
-
-
-## ADD ADJUSTED HOUR FOR EACH SPECIES DETECTION
-# calculate the time of sunrise given the location and timing of each species detection 
[…]

### 2023-11-01 r_scripts/2_GAMs.R

@@ -1,120 +1,0 @@
-## FIT GAMS TO MODEL DIEL ACTIVITY FOR FOXES AND FERAL CATS
-# MATTHEW REES
-
-library(dplyr)
-library(mgcv)
-# for plots
-library(gratia)
-library(ggplot2)
-library(patchwork)
-library(ggpubr)
-library(viridis)
-library(forcats)
-# Set the default theme for ggplot objects 
-theme_set(theme_bw())
-theme_update(panel.grid = element_blank())
-
-
-# load data
-records <- read.csv("derived_data/predator_counts_hour.csv")
-
-
-# Transform variables -----------------------------------------------------
-records <- transform(records,
-                      hour = as.integer(hour),
-                      cat = as.integer(cat), 
-                      fox = as.integer(fox), 
-                      foxbaits = as.integer(foxbaits), 
-                      fox_pa = factor(fox_pa, ordered = FALSE), 
-                      fox_count_adj = as.numeric(fox_count_adj), 
-                      region = factor(region, ordered = FALSE), 
-                      station = factor(station, ordered = FALSE), 
-                      vegetation_group = factor(XGROUPNAME, ordered = FALSE), 
-                      habitat_type = factor(habitat_type, ordered = FALSE), 
-                      survey_duration = as.integer(survey_duration),
-                      hfi = round(hfi, digits = 2))
-
-
-
-
-# 1) Interaction between spatial and diel activity ('model 1') ---------------------
-# specify penalties for 2D Duchon spline and 1D cc spline
-m <- list(c(1,.5),rep(0,0)) 
-
-gam_fox_sp <- bam(fox ~ s(hour, bs = "cc", k = 8) + 
-                        s(x, y, bs = "ds", k = 80, m = c(1,.5)) + 
-                        ti(x, y, hour, d = c(2, 1), bs = c("ds", "cc"), k = c(80, 8), m = m) +                     
-                        s(foxbaits, region, bs = "fs", xt = list(bs = "tp"), k = 4) + 
-                        offset(log(survey_duration)), 
-                  data = records, family = nb, knots = list(hour = c(0, 23)), discrete = TRUE)
-
-gam_cat_sp <- bam(cat ~ s(hour, bs = "cc", k = 8) + 
-                        s(x, y, bs = "ds", k = 80, m = c(1,.5)) + 
-                        ti(x, y, hour, d = c(2, 1), bs = c("ds", "cc"), k = c(80, 8), m = m) +                     
[…]

### 2023-11-01 r_scripts/function_clock_time_to_radian.R

@@ -1,10 +1,0 @@
-# written by overlap package maker
-ClocktimeToTimeRad <- function(Clocktime,
-                               timeformat = "%Y-%m-%d %H:%M:%S")
-{
-  DateTime2 <- strptime(as.character(Clocktime), format = timeformat, tz = "UTC")
-  Time2     <- format(DateTime2, format = "%H:%M:%S", usetz = FALSE)
-  Time.rad  <- (as.numeric(as.POSIXct(strptime(Time2, format = "%H:%M:%S", tz = "UTC"))) -
-                  as.numeric(as.POSIXct(strptime("0", format = "%S", tz = "UTC")))) / 3600 * (pi/12)
-  return(Time.rad)
-}

### 2023-11-01 r_scripts/plot_cat_v_fox.R

@@ -1,79 +1,0 @@
-# PLOT FIGURE
-
-## CAT PLOT - 3D 
-df = expand.grid(hour = 0:23,
-                 habitat_type = levels(records$habitat_type),
-                 longitude = mean(records$longitude),
-                 latitude = mean(records$latitude),
-                 fox_count_adj = seq(min(records$fox_count_adj), max(records$fox_count_adj), length=150),
-                 station = "T052",
-                 survey_duration = 60)
-# predict model results into dataframe
-x = sapply(gam_cat_fox$smooth, "[[",  "label")
-df_int <- cbind(df, predict(gam_cat_fox, newdata = df, se.fit = TRUE, type = "link", exclude = c(x[4:5], "s(survey_duration)")))
-# plot
-plot_range <- ggplot(aes(hour, fox_count_adj, fill = fit), data = df_int) +
-  geom_tile() +
-  facet_wrap(~habitat_type, nrow =1) +
-  geom_contour(aes(z = fit, fill = NULL), colour = "grey40", bins = 10) +
-  scale_fill_viridis("Cat activity", option = "viridis", breaks = c(-5, -3), labels = c("Low", "High")) + 
-  ggtitle("", subtitle = "Feral cat") +
-  geom_vline(xintercept = c(6.16,18.34), colour = "black", size = 0.6, linetype="dotted") + 
-  scale_y_continuous(breaks=c(min(records$fox_count_adj), max(records$fox_count_adj)), labels = c("Low", "High")) +
-  ylab("Fox activity") + 
-  xlab("Hour") 
-
-## CAT PLOT - 2D (same model at min / max fox)
-# min df 
-df_int_min <- df_int %>%
-  group_by(habitat_type) %>%
-  slice_min(fox_count_adj, n = 1) %>%
-  mutate(minmax = "0min")
-# max df 
-df_int_max <- df_int %>%
-  group_by(habitat_type) %>%
-  slice_max(fox_count_adj, n = 1) %>%
-  mutate(minmax = "1max")
-# join df
-df_int_minmax <- rbind(df_int_min, df_int_max)
-# plot
-plot_wet_lowhigh <- ggplot(data=df_int_minmax, aes(x=hour, y=fit, group=minmax, col = minmax)) +
-  geom_line(aes(y=fit, x=hour), lwd = 1.2) +
-  facet_wrap(~habitat_type, nrow =1) +
[…]

### 2023-11-01 r_scripts/plot_fox_baits_hfi.R

@@ -1,127 +1,0 @@
-# plot spatial effects of human footprint and fox control 
-
-
-# HUMAN FOOTPRINT INDEX PLOT  ---------------------------------------------
-# plot human footprint index  (no effect on diel activity - plot marginal effect only)
-fox_hfi <- smooth_estimates(gam_fox_veg, overall_uncertainty = TRUE, "s(hfi)") %>%
-  mutate(lower = est + (1.96 * se),
-         upper = est - (1.96 * se)) %>% 
-  ggplot(aes(hfi, est)) + 
-  geom_line() + 
-  geom_ribbon(aes(ymin=lower, ymax=upper), alpha=0.2)  + 
-  labs(x = "Human footprint index", y = "Partial effect")
-
-# plot together
-png("figs/fox_hfi.png", width = 6, height = 4, res = 600, units = "in")
-fox_hfi
-dev.off()
-
-
-# ESTIMATES OF FOX SUPPRESSION ----------------------------------------------------------
-# MODEL 1 
-x = sapply(gam_fox_sp$smooth, "[[",  "label")
-df <- expand.grid(hour = 0,
-                  x = mean(records$x),
-                  y = mean(records$y),
-                  foxbaits = min(records$foxbaits):max(records$foxbaits), 
-                  hfi = mean(records$hfi),
-                  region = c("glenelg", "otways"),
-                  survey_duration = 60)
-newdf_1 <- cbind(df, predict.gam(gam_fox_sp, newdata = df, se.fit = TRUE, type = "link", newdata.guaranteed = TRUE, exclude = c(x[c(1,2,3)], "s(survey_duration)"))) %>%
-  mutate(model = "1",
-         fit_resp = exp(fit), 
-         lower = exp(fit - (1.96 * se.fit)),
-         upper = exp(fit + (1.96 * se.fit))) %>%
-  filter(foxbaits == 0 | foxbaits == 19) %>%
-  select(model, region, fit_resp, lower, upper) %>% 
-  mutate(across(c(fit_resp, lower, upper), round, 3))
-
-# MODEL 2
-x = sapply(gam_fox_veg$smooth, "[[",  "label")
-df <- expand.grid(hour = 0,
-                     vegetation_group = "Dry Forest",
-                     foxbaits = min(records$foxbaits):max(records$foxbaits), 
-                     hfi = mean(records$hfi),
[…]

### 2023-11-01 r_scripts/plot_spatiotemporal_interaction.R

@@ -1,255 +1,0 @@
-# PLOT SPATIO-TEMPORAL MODELS
-# 1) DIEL ACTIVITY STRENGTH ACROSS SPACE
-# 2) ANIMATED GIF FOR SPATIOTEMPORAL ACTIVITY
-
-# additional packages
-library(sp)
-library(sf)
-library(viridis)
-
-
-# split records
-records_g <- filter(records, region == "glenelg")
-records_o <- filter(records, region == "otways")
-
-# cam locations 
-camdata <- records %>%
-  distinct(., station, .keep_all = TRUE) %>%
-  select(region, station, x, y) 
-# add new variable to match variable we will facet by - region x species
-camdata_fox <- camdata
-camdata_fox$species <- "Red fox"
-camdata_cat <- camdata
-camdata_cat$species <- "Feral cat"
-camdata <- rbind(camdata_fox, camdata_cat)
-camdata$facet <- paste0(camdata$species, "_", camdata$region)
-
-
-# MAKE SPATIAL DATAFRAMES TO PREDICT INTO ----------------------------------------------
-
-## GLENELG:
-# take just the cams
-records_g_cams <- distinct(records_g, station, .keep_all = TRUE)
-# change it to sf class
-records_g_cams <- st_as_sf(records_g_cams, coords = c("x", "y"), crs = 3111) 
-# make a 4km buffer around each camera
-glenelg_cams_buffer = st_buffer(records_g_cams, 4000)
-# dissolve the buffer
-glenelg_cams_buffer = st_union(glenelg_cams_buffer)
-# but bring in the edges - don't want to predict outside of surveyed area
-glenelg_cams_buffer <- st_buffer(glenelg_cams_buffer, dist = -3500, joinStyle  = "ROUND")
-# convert back to dataframe
-glenelg_buffer_df <- fortify(as_Spatial(glenelg_cams_buffer))%>%
-  transmute(x = long, y = lat, order = order, group = group)
-# split by each group (seperate polygon for each grid)
-buffer_df1 <- glenelg_buffer_df[which(glenelg_buffer_df$group == "ID1.1"),]
-buffer_df2 <- glenelg_buffer_df[which(glenelg_buffer_df$group == "ID1.2"),]
-buffer_df3 <- glenelg_buffer_df[which(glenelg_buffer_df$group == "ID1.3"),]
-buffer_df4 <- glenelg_buffer_df[which(glenelg_buffer_df$group == "ID1.4"),]
-# make this a list for x and y cols
[…]

### 2023-11-01 r_scripts/plot_vegetation_type.R

@@ -1,47 +1,0 @@
-
-
-# function to run and plot GAMs for each species
-gam_plot <- function(gam_model, data){
-  ## Plot global smooth (using the gratia package)
-  plot_global <- draw(evaluate_smooth(gam_model, overall_uncertainty = TRUE, "s(hour)"), xlab = "Hour",  title = "", subtitle = "i.   Average") + 
-    geom_vline(xintercept = c(6.16,18.34), colour = "black", size = 0.6, linetype="dotted") 
-  ## Plot the difference smooths using the gratia package
-  plot_difference <- draw(evaluate_smooth(gam_model, "s(hour,vegetation_group)"), xlab = "Hour",  title = "", subtitle = "ii.   Vegetation-specific deviations from the average") + 
-    geom_vline(xintercept = c(6.16,18.34), colour = "black", size = 0.6, linetype="dotted") 
-  ## Plot the actual hourly activity for each vegetation type   
-  # make a new dataframe to predict into
-  df <- expand.grid(hour = 0:23,
-                    vegetation_group = levels(data$vegetation_group),
-                    foxbaits = 0, 
-                    hfi = mean(data$hfi),
-                    region = c("glenelg", "otways"),
-                    survey_duration = 60)
-  # get smooths to exlude
-  x = sapply(gam_model$smooth, "[[",  "label")
-  # predict model estimates (and uncertainty) into this dataframe, excluding the impact of variables other than hour:
-  df <- cbind(df, predict.gam(gam_model, newdata = df, se.fit = TRUE, type = "link", newdata.guaranteed = TRUE, exclude = c(x[3:7])))
-  plot_veg <- ggplot(data=df, aes(x=hour, y=fit, group=vegetation_group)) +
-    # ylim(-13, -2) +
-    geom_line(aes(y=fit, x=hour), lwd = 0.7) +
-    facet_wrap(~vegetation_group, nrow =1) +
-    geom_ribbon(aes(ymin=(fit-2*se.fit), ymax=(fit+2*se.fit)), alpha=0.2) +
-    geom_vline(xintercept = c(6.16,18.34), colour = "black", size = 0.6, linetype="dotted") + 
-    labs(subtitle = "iii.   Predicted activity in each vegetation type", x = "Hour", y = "log(count)") 
-  ## Assemble plots
-  patch <- (plot_global | plot_difference) / plot_veg 
-  patch
-}
-
-# save
[…]

### 2023-11-01 old_code/manuscript/manuscript.Rmd
store manuscript rmd files

--- 2023-07-28 manuscript/manuscript.Rmd
@@
 ---
-title: Spatial variation in predator diel activity patterns
+title: Dynamic shifts in predator diel activity patterns across landscapes and threat levels
 author:
@@
   - code: UOM
-    address: Quantitative and Applied Ecology Group, School of Ecosystem and Forest Sciences, The University of Melbourne, Parkville, VIC, 3010, Australia
+    address: Quantitative and Applied Ecology Group, School of Agriculture, Food and Ecosystem Sciences, The University of Melbourne, Parkville, VIC, 3010, Australia
   - code: CSIRO
@@
 abstract: |
-  *Open research:* Data and code will be deposited on Dryad upon acceptance and can currently be found at this link: https://github.com/matt-w-rees/spatiotemporal-gams-invasive-predators.
 keywords: 
@@
   cache     = FALSE,
-  fig.path  = "../figs/",
+  fig.path  = "",
   fig.align = "center",
@@
 
-# Spatial variation in predator diel activity patterns {.unnumbered}
-
 $~$
@@
 Here, we tested whether an invasive dominant predator (red fox *Vulpes vulpes*) alters the spatiotemporal activity of an invasive subordinate predator (feral cat *Felis catus*).
-We collated records of both species from 3,667 camera-traps deployed experimentally across two regions of south-eastern Australia with simplified predator guilds; foxes were poison-baited in some landscapes within each region.
+We collated records of both species from 3,667 camera-traps deployed experimentally across two regions of south-eastern Australia with simplified predator guilds.
+Foxes were poison-baited in some landscapes within each region.
 We used generalised additive models to quantify changes in predator spatiotemporal activity across geographic space, vegetation types, human footprint and (artificially manipulated) gradients of dominant predator activity.
@@
 However, our study demonstrates that diel activity patterns also vary across space, likely mediated by both landscape-context and fear.
-dominant predator avoidance appears to be dynamic---a key nuance which is overlooked when simply comparing the average activity overlap between two species.
+Dominant predator avoidance in time appears to be spatially dynamic---a key nuance which is overlooked when simply comparing the average activity overlap between two species.
 
@@
 # INTRODUCTION
-
 Top predators can drive ecosystem structure and dynamics by suppressing populations of rival predators and prey species [@estes2011trophic; @prugh2009rise].
-Dominant predators can suppress subordinate species directly through antagonistic interactions and predation, as well as indirectly through fear [@creel2008relationships; @allen2022fear].
+Dominant predators can suppress subordinate species directly through antagonistic interactions and predation, as well as indirectly through fear [@creel2008relationships].
 Fear-induced behavioural suppression can be as detrimental to subordinate species as predation itself, by limiting resource acquisition and breeding success [@schmitz2004trophic; @preisser2005scared; @brown1999ecology].
-Understanding how top predators&mdash;including humans&mdash;constrain the behaviour of subordinate species is therefore important to accurately predict consequences of predator management, such as reintroductions or lethal control [@suraci2016fear; @gaynor2021applied].
+Understanding how top predators&mdash;including humans&mdash;constrain the behaviour of subordinate species is therefore important to accurately predict consequences of predator management, such as reintroductions or lethal control [@gaynor2021applied].
 
@@
 However, subordinate species may not consistently employ avoidance behaviours because perceived predation risk is spatiotemporally variable, and antipredator behaviours typically involve a trade-off against resource acquisition, such as limiting or relegating activity to suboptimal places or times [@lima1999temporal; @lima1990behavioral].
-Therefore, optimal predator avoidance strategies likely vary across heterogeneous landscapes where resource availability (e.g., shelter, food) and perceived predation risks differ [@kauffman2007landscape; @willems2009predator; @wirsing2021context; @davies2021spatial].
+Therefore, optimal predator avoidance strategies likely vary across heterogeneous landscapes where resource availability (e.g., shelter, food) and perceived predation risks differ [@kauffman2007landscape; @willems2009predator; @wirsing2021context; @davies2021spatial; @gaynor2018influence].
 For example, temporal predator avoidance may be preferable over spatial avoidance if food is constantly available throughout the day, and vice versa.
 These theories are unified under the ecology of fear concept [@brown1999ecology], which has gained increasing attention in recent times [@gaynor2019landscapes].
-However, there remains a lack of robust, empirical evidence documenting dynamic changes in antipredator behaviour.
+However, there remains a lack of robust, empirical evidence documenting dynamic changes in antipredator behaviour [@gilbert2022daily].
 
@@
 For example, by fitting separate models for spatial and temporal overlap, or by repeating spatial analyses (e.g., resource selection functions) at different time periods such night and day [e.g., @basille2015plastic; @kohl2019prey; @smith2019integrating; @wooster2022predator].
-Such ad-hoc approaches make it difficult to assess the relative importance, as well as dynamic changes in spatial and temporal processes [@suraci2022beyond].
+Such approaches make it difficult to assess the relative importance, as well as dynamic changes in spatial and temporal processes.
 
 There are many approaches to modelling intraguild spatiotemporal interactions, such as time-to-encounter models and coefficients of activity overlap [@frey2017investigating].
-Time-to-encounter approaches consider fine-scale avoidance or attraction by using time intervals between species detections as the response variable, although do not estimate diel activity patterns [@harmsen2009spatial; @niedballa2019assessing; @fancourt2019introduced].
+Time-to-encounter approaches consider fine-scale avoidance or attraction by using time intervals between species detections as the response variable, but do not estimate diel activity patterns [@harmsen2009spatial; @niedballa2019assessing; @fancourt2019introduced].
 Multispecies models have been recently developed to assess overlap in spatial and diel activity between two species [@azzou2021sparse], as well as different diel activity patterns based on the presence or absence of another species at a site [@kellner2022two].
@@
 Generalised Additive Models (hereafter 'GAMs') are increasingly being used to estimate animal diel activity patterns and offer a flexible framework to jointly consider spatial activity.
-GAMs are useful to model diel activity patterns as they can implement cyclical splines&mdash;joining the end with the start of the day&mdash;as well as complexity penalties to reduce overfitting.
-If complex nonlinear (i.e., wiggly) effects are not statistically supported, smoothing penalties in GAMs 'shrink' them to a linear (i.e., no wiggliness) effect, or even to a horizontal line - removing the entire effect of the predictor variable and therefore  the need for additional tests of statistical significance (such as p-values and information criteria).
-GAMS can also capture nonlinear interactions between multiple variables with different units (i.e., 'tensor products'), and offer the ability to share information across categorical variables through hierarchical specifications [@wood2017generalized; @pedersen2019hierarchical].
-However, we are only aware of one study which allowed animal diel activity to interact with predation risk as a continuous variable in a GAM [although without considering overall activity, @cunningham2019temporal].
+GAMs are useful for modelling diel activity patterns as they can implement cyclical splines, which join the end of the one day with the start of the following day.
+GAMs can also include complexity penalties to reduce overfitting; these penalties can reduce complex nonlinear (i.e., wiggly) smooths to linear effects or even remove the effect from the model if it is not statistically supported, avoiding the need for additional tests of statistical significance (such as p-values and information criteria).
+GAMS can also capture nonlinear interactions between two variables with different units (i.e., 'tensor products'), and offer the ability to share information across categorical variables through hierarchical specifications [@wood2017generalized; @pedersen2019hierarchical].
+However, we are only aware of one other study that has used a GAM to explore how animal diel activity interacts with predation risk as a continuous variable: @cunningham2019temporal used this approach to document shifts in diel activity patterns of mammal communities following decline in a top predator (although without also considering spatial animal activity).
 
-The red fox *Vulpes vulpes* (hereafter 'fox') and feral cat *Felis catus* (hereafter 'cat') have devastating impacts on native prey throughout their introduced range, implicated in the extinction of \~10 and 63 species, respectively [@doherty2016invasive].
+Invasive predators are often lethally controlled to protect native competitor and prey species; however, the conservation benefits of invasive predator control may be dampened if this benefits other invasive species [@takimoto2022simple; @doherty2017stop].
+The red fox *Vulpes vulpes* (hereafter 'fox') and feral cat *Felis catus* (hereafter 'cat') have had devastating impacts on native prey throughout their introduced range, and are implicated in the extinction of \~10 and 63 species worldwide, respectively [@doherty2016invasive].
 The impacts of these invasive predators are particularly extreme on the Australian continent [@woinarski2015ongoing].
 Cats are more difficult to manage, and so introduced predator control programs in Australia often target only foxes [particularly through poison-baiting, @reddiex2007control].
-As foxes and cats compete for many of the same resources [@fleming2022distinctive], there is concern that lethal fox control could cause mesopredator release [@soule1988reconstructed] of cats [@glen2005complex; @doherty2017stop; @wayne2017recoveries].
-However, there is limited evidence that spatial cat activity increases in response to fox control [@hunter2018not], although there are signs of changes in behaviour and population density [@molsher2017mesopredator; @rees2023mesopredator].
-Other studies have investigated potential spatial and temporal interactions between these invasive predators [e.g., @roshier2021space], but not in response to fox control, or in a joint spatiotemporal framework that allows flexibility in cat avoidance behaviours with respect to differences in fox activity.
-Conservation managers are therefore uncertain as to whether fox control may increase the impacts of feral cats on native prey, through increases in cat density or behavioural changes which improve hunting success.
+As foxes and cats compete for many of the same resources [@fleming2022distinctive], there is concern that lethal fox control could cause mesopredator release [@soule1988reconstructed] of cats [@glen2005complex; @wayne2017recoveries].
+While there is evidence that foxes may suppress cat density and alter movement behaviour [@molsher2017mesopredator; @rees2023mesopredator], potential changes in cat diel activity have been relatively rarely studied [although see @roshier2021space].
+Changes in predator diel activity patterns may lead to changes in population-level fitness, as well as hunting success and primary prey species [@allen2022fear; @suraci2016fear].
 
@@
 We compiled data from multiple smaller-scale camera-trap studies, each designed to experimentally assess mammal responses to fox control.
-Overall, we collated 5,449 and 2,202 independent detections of foxes and cats, respectively (separated by at least 30 minutes) from 172,052 camera-trap nights (Table \@ref(tab:diel-tab1)).
+Overall, we collated 5,449 and 2,202 independent detections of foxes and cats, respectively (separated by at least 30 minutes) from 172,052 camera-trap nights (Supporting Information Table \@ref(tab:diel-tab1)).
 
@@
 Different road densities therefore result in variable densities of poison-baits.
-Foxes readily dig these poison-baits often resulting in population suppression, whereas it is generally accepted that cats are unlikely to consume buried fox-baits and there is no evidence buried 1080 fox-baits suppresses cat populations [@algar2004feral; @moseby2011use; @risbey1997control; @hunter2018not].
+Foxes readily dig these poison-baits, whereas it is generally accepted that cats are unlikely to consume buried fox-baits and there is no evidence that buried 1080 fox-baits suppress cat populations [@algar2004feral; @moseby2011use; @risbey1997control; @hunter2018not].
 Managers also frequently implement prescribed fire across both regions, primarily to reduce fuel loads to prevent large wildfires.
@@
 We also include data from an additional before-after control-impact surveys (one 'before' baiting survey and two 'after' baiting surveys) in the western section of the Otway Ranges, conducted annually 2017 - 2019 [@rees2023data].
-This added a further 195 sites and 524 camera-trap deployments (Table \@ref(tab:tab-sum)).
+This added a further 195 sites and 524 camera-trap deployments (Supporting Information Table \@ref(tab:tab-sum)).
 
@@
 Camera-traps were active for an average of 47 days (maximum 93 days), totalling 172,052 trap-nights.
-Camera-trap spacing averaged 867 m for Glenelg Ark, 1246 m for Otway Ark, and 444 m for the M.W.R PhD surveys (Table \@ref(tab:tab-sum)).
+Camera-trap spacing averaged 867 m for Glenelg Ark, 1246 m for Otway Ark, and 444 m for the M.W.R PhD surveys (Supporting Information Table \@ref(tab:tab-sum)).
 
@@
 We used the negative binomial family, as overdispersion, but not zero-inflation, was detected with a poisson distribution using the 'DHARMa' R-package [@DHARMa].
-We specified the natural log of the number of survey days as a model offset to account for differences in camera-trap survey duration, and a random intercept for each site to account for repeat sampling for models 2 and 3 (as pseudoreplication from repeat sampling in model 1 was explicitly accounted for using the spatial term).
-For fox models, we also included a smooth effect of poison-bait density [number of baits within a 2.3 km radius around each camera-trap - the average maximum distance adult foxes travel from their home-range centre in these regions, @hradsky2017human] with separate responses per region to account for the effect of fox control.
-These specifications were included in each model we fitted; models differed in their specification of the cyclical hour smooth to provide inference on variations of predator diel activity across the three questions of interest - this is detailed in the sections below.
+We specified the natural log of the number of survey days as a model offset to account for differences in camera-trap survey duration, and a random intercept for each site to account for repeat sampling for models 2 and 3 (pseudoreplication from repeat sampling in model 1 was explicitly accounted for using the spatial term).
+For all fox models, we also included a smooth effect of poison-bait density [the number of baits within a 2.3 km radius around each camera-trap - the average maximum distance adult foxes travel from their home-range centre in these regions, @hradsky2017human] with separate responses in each region to account for the effect of fox control effort.
+Poison-bait density ranged from 0 - 19 baits per 2.3km radius in both the Glenelg and Otway region.
+Models differed in their specification of the cyclical hour smooth to provide inference on variations of predator diel activity across the three questions of interest - this is detailed in the sections below.
 We used the 'ggplot2' [@ggplot2] and 'gratia' [@gratia] R-packages to plot models.
@@
 In this paper, we use the term 'spatial activity' to refer to the number of 'independent' predator detections at a site (offset to account for survey duration; analogous to an activity or abundance index), and 'diel activity pattern' to refer to fluctuations in relative activity throughout the 24-hour daily cycle.
-The first two models we fit explored how predator spatiotemporal activity varied across these regions, informing the best structure for our third model which tested the activity of foxes limits the spatiotemporal activity of cats (detailed in the sections below).
+The first two models explore how predator spatiotemporal activity varies across these regions.
+We used these to inform the best structure for our third model which tested whether the activity of foxes limits the spatiotemporal activity of cats (detailed in the sections below).
 
@@
 ### How does vegetation type and human footprint affect predator activity patterns? (model 2)
-The spatial term in model 1 estimated differences in predator activity across geographic space; we hypothesised that this was partly due to differences in vegetation type, based both on the observed spatial patterns and because vegetation type is a major driver of understorey habitat structure and prey occurrence in these regions [@swan2015predicting; @hradsky2017bayesian].
+The spatial term in model 1 estimated differences in predator activity across geographic space; we hypothesised that this would be partly due to differences in vegetation type, based both on the observed spatial patterns and because vegetation type is a major driver of understorey habitat structure and prey occurrence in these regions [@swan2015predicting; @hradsky2017bayesian].
+We further hypothesised that predators would become relatively more nocturnal with increasing human activity, reflecting temporal avoidance of humans [@gaynor2018influence].
+
 To test whether the diel activity pattern of each predator varied among vegetation types, we identified the Ecological Vegetation Class group [hereafter 'vegetation type' - standard units for vegetation classification in Victoria, @delwp2020bioregions] for each unique camera-trap site, totalling eight vegetation types.
@@
 We then estimated predator activity across vegetation types using a hierarchical model specification: a global smoother for hour (i.e., average response) and group-level smoothers with shared wiggliness for the seven vegetation types ['model GS' detailed in @pedersen2019hierarchical].
-We also fit a tensor product interaction term for hour and the human footprint index [@venter2016sixteen; @venter2018last]; as human footprint is known to impact animal activity patterns [@gaynor2018influence; @van2023influence]
+
+To influence of human activity on predator activity patterns, we fit a tensor product interaction term for hour and the human footprint index [@venter2018last].
+The human footprint index maps human pressure on environments by considering variables such as human population density, infrastructure and agriculture [@venter2016sixteen].
 
 ### Do feral cats avoid foxes in space or time? (model 3)
-Fox diel activity across vegetation types showed strong similarity between all vegetation types except wet forests.
+Model 2 showed that fox diel activity patterns was mostly uniform among all vegetation types except wet forests.
 To examine whether cats avoid foxes in space or time, we therefore modelled fox-induced changes in feral cat diel activity separately for wet forest and dry vegetation types.
-We further split dry vegetation types by region for further replication.
+We examined responses in dry vegetation types by region for further replication.
 We refer to the resulting variable as 'habitat type', which had three levels: (i) wet forests and rainforests in the western Otway Ranges ('wet_otways'), (ii) dry vegetation types in the Otway Ranges ('dry_otways') and (iii) dry vegetation types in the Glenelg region ('dry_glenelg').
-We further hypothesised that cats would avoid foxes in time by becoming more diurnal in dry vegetation types where foxes were mostly nocturnal, but not in wet forests where fox activity showed little variation across the daily cycle.
+We hypothesised that cats would avoid foxes in time by becoming more diurnal in dry vegetation types where foxes were mostly nocturnal, but not in wet forests where foxes were more consistently active throughout the 24-hour period.
 
 To investigate changes in feral cat diel activity across the range of observed fox activity, we first quantified fox activity for each camera trap deployment as the total number of fox detections for the deployment divided by the number of survey days, to adjust for differences in survey duration (hereafter 'adjusted fox counts').
-We modelled an interaction between hour and adjusted fox counts (as a thin plate regression spline with shrinkage - so complexity penalties could entirely remove this effect from the model if not supported), allowing cats to have nonlinear responses to both hour and adjusted fox counts.
+We modelled an interaction between hour and adjusted fox counts (as a thin plate regression spline with shrinkage - complexity penalties could entirely remove this effect from the model if not supported), allowing cats to have nonlinear responses to both hour and adjusted fox counts.
 We fit separate tensor product interactions for each habitat type (using a 'by-variable' term).
@@
 ```{r diel-map, fig.cap = '(ref:fig1cap)'}
-knitr::include_graphics("../figs/map_cams.png")
+knitr::include_graphics("map_cams.png")
 ```
@@
 ## Lethal suppression of foxes  
-In each model, there was a strong effect of poison fox-baiting on fox spatial activity, particularly in the Glenelg region (Supporting Information Fig.
-\@ref(fig:fox-baits)b).
-Predicted fox counts on the response scale were reduced by a factor of 12, 24, and 28 in the Glenelg region, as well as a factor of 13, 13, and 8 in the Otway Ranges, for models 1, 2, and 3, respectively.
-For example, model 2 predicted fox counts in the Glenelg region to range from 0.097 (95% CI: 0.077 - 0.122) in sites with no fox-baiting, to 0.004 (95% CI: 0.002 - 0.012) at sites with the maximum poison-bait density; in the Otways ranging from 0.041 (95% CI: 0.033 - 0.05) - 0.008 (95% CI: 0.003 - 0.020) across the gradient of poison-bait density.
+In all fox models, the density of poison-baits had a strong negative association with fox spatial activity (Supporting Information Fig.
+\@ref(fig:fox-baits)).
+In the absence of baiting, fox spatial activity was substantially higher in the Glenelg region than the Otway Ranges.
+Baiting had a stronger effect on fox spatial activity in the Glenelg region, likely reflecting differences in baiting intensity.
+Across the observed range of bait densities, fox counts on the response scale were predicted to be reduced by 91% - 96% in the Glenelg region, and 87% - 93% in the Otway Ranges across the three models.
+For example, model 2 predicted fox counts in the Glenelg region to decline from 0.097 (95% CI: 0.077 - 0.122) in sites without fox-baiting to 0.004 (95% CI: 0.002 - 0.012) at sites with the maximum poison-bait density, while fox counts in the Otways declined from 0.041 (95% CI: 0.033 - 0.05) at unbaited sites to 0.008 (95% CI: 0.003 - 0.020) at the maximum bait density.
 
 ## How does predator diel activity patterns vary across geographic space? (model 1)
-Predator activity varied considerably across space and throughout the 24-hour daily cycle, and there was some variation in the predator diel activity patterns across space.
-On average, both predators showed similar diel activity patterns, with peaks in activity around sunrise and sunset (i.e., crepuscular), although were more likely to be active during the night than the day (Fig.
+Predator activity varied considerably across space and throughout the 24-hour daily cycle.
+Both predators had similar average diel activity patterns: on average, their activity peaked around sunrise and sunset (i.e., crepuscular), they tended to be more active during the early evening than at sunrise, and their activity was lowest during the middle of the day (Fig.
 \@ref(fig:diel-st-int)i).
-On average, the main difference between the species was that fox activity peaked just after sunset and they were less likely to be active during the day than cats.
-Cats also tended to be more active at sunset relative to sunrise (Fig.
+The main differences in average activity between the species were that fox activity peaked a couple of hours after sunset, whereas cat activity peaked at sunset, and foxes were less likely to be active during the day than cats (Fig.
 \@ref(fig:diel-st-int)i).
 
-The interaction between hourly and spatial smooths revealed deviations from marginal diel activity pattern in space, particularly across the two regions, and within the Otway Ranges (Fig.
+In addition, there was considerable spatial variation in the diel activity of each predator (away from these average patterns).
+The interaction between hourly and spatial smooths revealed deviations from marginal diel activity pattern in space across and within (particularly within the Otway Ranges) the two regions (Fig.
 \@ref(fig:diel-st-int)).
-For example, at midday, foxes were less active in the Glenelg region and more active across much of the Otway Ranges (except for the north-east) than the average diel activity pattern.
-Relative to foxes, cats had slightly more complex (i.e., wiggly) deviations in space from their average diel activity pattern.
-Most notably, cats were less active during sunset and more active at midnight in the Otway Ranges relative to the Glenelg region (Fig.
-\@ref(fig:diel-st-int)).
+For example, foxes were less diurnal in the Glenelg region and the north-east of the Otway Ranges (i.e., a negative partial effect at midday) and much more strongly diurnal across the southern Otway Ranges, relative to their average diel activity pattern (Fig.
+\@ref(fig:diel-st-int)a).
+Cats had slightly more complex (i.e., wiggly) deviations in space from their average diel activity pattern than foxes.
+Most notably, cats were more active than average at midnight and less active than average at sunset in the Otway Ranges; the opposite was true in the Glenelg region (Fig.
+\@ref(fig:diel-st-int)b).
 
 ## How does vegetation type and human footprint affect predator activity patterns? (model 2)
-Vegetation type had a strong influence on the spatial and diel activity patterns of foxes and cats, whereas human footprint only affected fox spatial activity.
-Spatial levels of fox activity were similar across all vegetation types, except wet forests where fox activity was considerably lower.
+Vegetation type had a strong influence on the spatial and diel activity patterns of foxes and cats.
+Spatial levels of fox activity were similar across all vegetation types, except wet forests where fox activity was considerably lower (Fig.
+\@ref(fig:diel-veg)a).
 Spatial cat activity was relatively more variable across vegetation types; lowest in heathy woodlands and highest in wet forests (Fig.
@@
 
-There was no evidence that human footprint affected diel activity patterns for either predator species; the interaction term between hour and human footprint index was removed from both models.
-Fox spatial activity increased linearly with the human footprint index (Supporting Information Fig.
-\@ref(fig:fox-hfi)a), although human footprint did not impact cat spatial activity (the marginal effect of human footprint was also removed from the cat model).
+In contrast, there was no evidence that the human footprint index was associated with the diel activity pattern of either predator species: the interaction term between hour and human footprint index was removed from both models.
+Further, fox spatial activity increased linearly with the human footprint index (Supporting Information Fig.
+\@ref(fig:fox-hfi)), but cat spatial activity showed no relationship with the human footprint index (the marginal effect of human footprint was also removed from the cat model).
 
@@
 \@ref(fig:diel-cat-fox)).
-Fox activity did not impact spatial cat activity in the Otway Ranges, whereas cat spatial activity slightly increased with adjusted fox counts in the Glenelg region.
+Fox activity did not impact spatial cat activity in the Otway Ranges, while cat spatial activity slightly increased with adjusted fox counts in the Glenelg region.
 
-Across all habitat types, feral cat diel activity patterns changed across gradients of fox activity (Fig.
+Across all habitat types, feral cat diel activity patterns changed across gradients of fox activity.
+In the Glenelg region and Otway dry habitat types, feral cats had a nocturnal-crepuscular diel activity pattern where fox activity was low, but were most active during the day where fox activity was high.
+In contrast, in the wet forests of the Otway Ranges, feral cats were more strongly nocturnal when fox activity was high, avoiding dawn in particular (Fig.
 \@ref(fig:diel-cat-fox)).
-In the Glenelg region and Otway dry habitat types, feral cats had a nocturnal-crepuscular diel activity pattern where fox activity was low, but were most active during the day where fox activity was high.
-In contrast, in the wet forests of the Otway Ranges, feral cats were more strongly nocturnal when fox activity was high, avoiding dawn in particular.
 
@@
 
-```{r diel-st-int, out.width = "70%", fig.cap = "Marginal diel activity (i) and space-time interaction effect in the Glenelg region (ii) and Otway Ranges (iii) at midnight (0), sunrise (6), midday (12) and sunset (18) times for introduced red foxes \\textit{Vulpes vulpes} and feral cats \\textit{Felis catus}.
+```{r diel-st-int, out.width = "70%", fig.cap = "Marginal diel activity (i.e., average; i) and space-time interaction effect in the Glenelg region (ii) and Otway Ranges (iii) at midnight (0), sunrise (6), midday (12) and sunset (18) times for introduced red foxes \\textit{Vulpes vulpes} and feral cats \\textit{Felis catus}.
+Negative partial effects in ii and iii (blue shades) indicate that the predator was less active in these areas than the estimated marginal diel activity pattern (i), while red shades indicate higher activity than the marginal diel activity pattern.
 These effects were estimated by a Generalised Additive Model with a tensor product interaction between spatial and hourly smooths, fit to camera-trap data ('model 1').
 Dotted, vertical lines represent average sunrise and sunset times, shaded areas indicate 95\\% confidence intervals (i); black dots depict camera-trap sites (ii; iii)."}
-knitr::include_graphics("../figs/spatial_interaction.png")
+knitr::include_graphics("spatial_interaction.png")
 
@@
 These effects were estimated by a hierarchical Generalised Additive Model with a global smoother for hour (i.e., average response) and group-level smoothers with shared wiggliness for vegetation type (EVC groups), fit to camera-trap data ('model 2').
+Panels i and ii in Fig.
+3 show the partial effects of each smooth as implemented in the GAM (which can interpreted as log odds ratios; conditional on the other estimated model terms), while Panel iii shows the sum of the partial effects in Panels i and ii, including the intercept.
 Dotted, vertical lines represent average sunrise and sunset times.
@@
 The overall level of activity was relatively consistent across EVC groups for foxes (a – iii), whereas it differed substantially for feral cats (b - iii)."}
-knitr::include_graphics("../figs/predator_veg_leg.png")
+knitr::include_graphics("predator_veg_leg.png")
 ```
@@
 In the Otway Ranges, feral cat activity also peaked during the day where fox activity was high in dry vegetation types (b), but was more nocturnal where fox activity was high in the rainforests and wet forests (c)."}
-knitr::include_graphics("../figs/cat_fox_count.png")
+knitr::include_graphics("cat_fox_count.png")
 ```
@@
 A key question in ecological theory is whether animals are evolutionarily hardwired to occupy particular temporal niches, or have circadian rhythms that are responsive to changing environmental conditions and interactions with other species [@daan1981adaptive; @schoener1974compression; @lima1990behavioral].
-Here we demonstrate that diel activity patterns are not fixed, but vary across space based on landscape context and fear.
+Here we demonstrate that predator diel activity patterns are not fixed, but vary across space based on landscape context and fear.
 In our study, sympatric invasive predators had similar diel activity patterns when averaged across broad regions [i.e., high circular overlap, as did @roshier2021space], but behaviours varied considerably within landscapes.
-Foxes were nocturnal in dry habitat types, but showed little variation in activity through the daily cycle in wet forests (Fig.
-\@ref(fig:diel-veg)a).
-In contrast, cats were mostly nocturnal in wet forests but crepuscular in dry vegetation types (Fig.
-\@ref(fig:diel-veg)b).
 Within broad habitat types, cats altered their diel activity patterns at sites with higher fox activity towards less-risky times of day (i.e., times when fox activity was lower), however we found no signs of foxes suppressing cat spatial activity (Fig.
 \@ref(fig:diel-cat-fox)).
-Control programs that reduce invasive fox activity may therefore shift cat diel activity patterns, which in turn may alter cat impacts on native prey species.
+Lethal control programs that reduce invasive fox activity may therefore shift cat diel activity patterns, which in turn may alter cat impacts on native prey species.
 Quantifying changes in diel activity patterns provides important context for understanding species interactions, which is key for effective ecosystem management [@gaynor2021applied].
@@
 Shifting diel activity patterns may facilitate spatial coexistence of dominant and subordinate species [@carothers1984time].
-For cats, altering diel activity patterns to less-preferred times of day may be worthwhile to persist in high-quality habitat.
-There have been an increasing number of studies which document slight changes in diel activity peaks of subordinate species in response to dominant predators, but few studies have demonstrated predator-induced reversals in diel activity [@kronfeld2003partitioning].
-Notably, ship rats *Rattus norvegicus* were also found to switch from nocturnal to diurnal behaviour in response to fox activity [@fenn1995use] and a similar nocturnal-diurnal shift was observed in American mink *Neovison vison* following the recolonisation of native predators [@harrington2009impact].
-Our study is particularly unique in that we replicated this behaviour across two separate regions with experimental manipulation, although there was considerable uncertainty in cat diel activity patterns where fox activity was highest (because there were fewer sites with high fox activity relative to low).
+For cats, radically altering diel activity patterns to less-preferred times of day may be worthwhile to persist in high-quality habitat.
+An increasing number of studies have documented changes in the magnitude of diel activity peaks of subordinate species in response to dominant predators [e.g., @cunningham2019temporal; @crawford2021predation], but few studies have demonstrated predator-induced reversals in diel activity patterns [@kronfeld2003partitioning], as we observed among cats in the dry habitat types of our study landscapes.
+Notably, ship rats *Rattus norvegicus* were also found to switch from nocturnal to diurnal behaviour in response to fox activity [@fenn1995use], as did American mink *Neovison vison* following the recolonisation of native predators [@harrington2009impact].
+Our study is unique in that we documented this behaviour across two separate replicate regions with experimental manipulation, although there was considerable uncertainty in cat diel activity patterns where fox activity was highest (because there were fewer sites with high fox activity than low fox activity).
 
-For cats in our study, a switch to diurnal behaviour where fox activity was high in the dry vegetation types may have been facilitated by the higher abundance of reptiles in these habitat types relative to wet forests [which are mostly diurnal, @woinarski2018reptiles].
-At wet forest sites with high fox activity, cats concentrated their activity away from sunrise and sunset towards midnight, despite a diurnal shift appearing to similarly reduce the risk of a fox encounter.
-In this situation, we expect becoming more nocturnal to be favourable over a diurnal shift because this is when small mammals are active and cats would be least visible to foxes [cats here mostly had black or grey coats, @rees2023mesopredator].
-Spatial fox activity was twice as low in the wet forests relative to dry habitat types, and so cats were likely under less pressure to radically alter their diel activity patterns.
+For cats in the dry vegetation types, a switch to diurnal behaviour where fox activity was high  may have been facilitated by the relatively high abundance of reptiles in this habitat type, compared to wet forests [reptiles are mostly diurnal, @woinarski2018reptiles].
+In contrast, at wet forest sites with high fox activity, cats concentrated their activity away from sunrise and sunset towards midnight.
+An diurnal shift in wet forests would potentially result in a similar reduction in the risk of fox encounter, but increasing activity at night instead is likely be more favourable, because small mammals are most active at night and cats would be least visible to foxes [cats here mostly had black or grey coats, @rees2023mesopredator].
+Further, spatial fox activity was twice as low in the wet forests relative to dry habitat types, and so cats were likely under less pressure to radically alter their diel activity patterns here.
 Understanding how these potential avoidance behaviours impacts native prey is a key research priority to improve invasive predator management.
+Mesopredators need to be highly adaptable to minimise top predator encounters while simultaneously maximising prey encounters; flexible diel activity patterns could be key strategy to achieve this trade-off across broad landscapes.
 
-Cats appeared to avoid foxes in time, but we saw no evidence of spatial avoidance (i.e.
-no fine-scale negative association between fox and cat spatial activity.
-In fact, there was some evidence that cat and fox spatial activity was positively correlated in the Glenelg region, suggesting temporal partitioning may facilitate predator co-abundance in quality habitat.
-Nonetheless, cats may have still avoided foxes spatially, but transiently on the scale of hours to days, or at finer spatial scales than most camera-trap studies can detect [and which telemetry studies can highlight, e.g., @kohl2019prey; @smith2019integrating].
-Short-term spatial avoidance is quite plausible given foxes mark territories using scats and odours, which cats could tangibly associate with high risk shortly after.
+Cats appeared to avoid foxes in time, but we saw no evidence of spatial avoidance (i.e., no fine-scale negative association between fox and cat spatial activity).
+In fact, there was some evidence that cat and fox spatial activity was positively correlated in the Glenelg region, suggesting temporal partitioning may facilitate high predator co-abundance in quality habitat.
+Nonetheless, cats may have still avoided foxes spatially but transiently, on the scale of hours to days, or at finer spatial scales than most camera-trap studies can detect [and which telemetry studies can highlight, e.g., @kohl2019prey; @smith2019integrating].
+Short-term spatial avoidance is plausible given foxes mark territories using scats and odours, which cats could associate with short-term increases in risk.
 Subordinate predators may prefer temporal avoidance of dominant predators (changing diel patterns or transient site avoidance) over longer-term site avoidance when both species have similar resource requirements and share many prey species.
 
-Many species increase nightly activity to avoid humans [@gaynor2018influence; @van2023influence], however we effect of human footprint on either predator species on our study.
-This is somewhat unsurprising given there was little variation in human activity among our sites; we studied predators within rugged conservation reserves largely free from human pressures (our camera-trap sites encompassed only the lower 32% of the total human footprint index range).
-On the other hand, foxes within our study region often select for human-modified features more during night than day [@hradsky2017human].
-Subordinate predators may still prefer to avoid dominant (animal) predators over humans, even when this decision leads to higher mortality rates [@prugh2023fear].
+Many species increase nightly activity to avoid humans [@gaynor2018influence; @van2023influence], however we found no effect of human footprint on the diel activity of either predator.
+A telemetry study has shown that that foxes in our study region often select for human-modified habitat features, such as roads and urban areas, particularly during the night [@hradsky2017human].
+The mismatch between the diel effect observed in the telemetry study and our results could reflect the relatively low and consistent amount of human activity among our sites; our camera-trap sites were all located within conservation reserves and encompassed only the lower 32% of the total human footprint index range.
+On the other hand, we did detect a strong, positive effect of human footprint on spatial fox activity [similar to @hradsky2017human].
+We found no evidence human activity impacted cats&mdash;subordinate predators may prefer to avoid dominant (animal) predators over humans, even when this decision leads to higher mortality rates [@prugh2023fear].
 
-A distinction of our study from others is that we modelled potential avoidance behaviours in a simple predator guild, where dominant predator activity was artificially manipulated (suppressed by up to 21-fold).
-The simple predator guild removes bias from unmodelled impacts of other predators in the system [@levi2012wolves].
-Artificial manipulation of the dominant predator also reduces potential bias from differences in niche preferences between the predator species.
+One strength of our study is that we modelled potential avoidance behaviours in a simple predator guild, where dominant predator activity was artificially manipulated through lethal control.
+The simple predator guild reduced bias from unmodelled impacts of other predators in the system [@levi2012wolves].
+Artificial suppression of the dominant predator allowed us to better infer the effects of species interactions, separate from  underlying differences in niche preferences of the two predator species.
 We also included replication across different habitat types.
-However, because our study did not consider associations with prey species, we cannot distinguish whether changes in cat diel activity patterns were the result of direct fox avoidance or indirect associations with shared prey.
+However, because our study did not consider associations with prey species, we could not distinguish whether changes in cat diel activity patterns were the result of direct fox avoidance or indirect associations with shared prey.
 For example, low fox activity may promote the availability of a preferred shared prey species with a diel pattern which differs from those of cats on average, and as a result, cats might shift diel activity patterns at sites with low fox activity to more closely match those of the more abundant prey species.
@@
 
-Flexible antipredator behaviours make evolutionary sense, but have been rarely demonstrated in terms of spatiotemporal predator avoidance, because this often requires manipulative experiments or at least more complicated models [although, see @relyea2003predators; @brown2013phenotypically; @cunningham2019temporal].
-Our study demonstrates that GAMs offer a powerful tool for modelling continuous shifts in animal activity across both space and time, capable of capturing complex interactions and sharing information across categorical variables.
-We are only aware of one study which allowed animal diel activity to interact with predation risk as a continuous variable in a GAM [although without allowing for nonlinear shifts via a tensor product or jointly considering spatial activity, @cunningham2019temporal].
-Further, complexity penalties are particularly beneficial in assessing spatiotemporal responses, because they are less prone to spurious estimates from small sample sizes or noisy data, and allow multiple hypotheses to be tested within one model without the need for ad-hoc significance testing (e.g., running a separate model for each scenario and comparing with Information Criteria).
-For example, our model specification allowed cats to respond to fox activity with (a) nonlinear shifts in both spatial and diel activity (as was observed in the Glenelg region), (b) nonlinear shifts in spatial but not diel activity, (c) nonlinear shifts in diel but not spatial activity (as was observed in both habitat types of the Otway Ranges), as well as their respective linear counterparts and also different linear-nonlinear combinations of the first scenario ('model 3').
-**ADD CLOSER**
+Adaptable antipredator behaviours make evolutionary sense but have been rarely demonstrated in terms of spatiotemporal predator avoidance, because this often requires manipulative experiments or at least more complicated statistical models [although see @relyea2003predators; @brown2013phenotypically; @cunningham2019temporal].
+Our study demonstrates that GAMs offer a powerful tool for modelling continuous shifts in animal activity across both space and time, as they are capable of capturing complex interactions and sharing information across categorical variables.
+Furthermore, complexity penalties in GAMs are particularly beneficial for assessing spatiotemporal responses, reducing spurious estimates due to small sample sizes or noisy data.
+Complexity penalties also allow multiple hypotheses to be tested within one model without post-hoc significance testing.
+For example, our model specification for 'model 3' allowed us to simultaneously test whether cats respond to differences in fox activity with non-linear or linear shifts in their spatial and/or diel activity patterns.
+Importantly, GAMs implemented in the 'mgcv' package have a relatively low entry barrier for ecologists familiar with generalised linear models, allowing ready access to complex and flexible spatiotemporal models [@wood2017generalized].
 
 The alternative approach of simply comparing average diel activity overlap between two species [@ridout2009estimating] would have been misleading for two reasons.
-Firstly, predator diel activity patterns varied 'naturally' across heterogeneous landscapes (requiring avoidance to be tested in wet forests and dry vegetation types separately, Fig.
+Firstly, 'baseline' predator diel activity patterns varied 'naturally' across our heterogeneous landscapes (requiring avoidance to be tested in wet forests and dry vegetation types separately, Fig.
 \@ref(fig:diel-veg)).
-Secondly, dominant predator temporal avoidance strategies were not consistently employed, but depended on spatial dominant predator activity and habitat type (Fig.
+Secondly, temporal avoidance strategies were not consistent, but depended on the level of spatial activity of the dominant predator and the habitat type (Fig.
 \@ref(fig:diel-cat-fox)).
-Despite their underlying statistical complexity, GAMs in the 'mgcv' R-package are straightforward to fit.
 Our GAM framework for modelling spatiotemporal activity can be used on any species with time-stamped detections, including datasets with categorical or continuous covariates and hierarchical groupings.
@@
 Our results demonstrate the importance of (a) considering diel activity in regards to species interactions, (b) modelling changes in animal behaviour rather than overlap with other species, and (c) testing avoidance behaviours within a joint spatiotemporal framework.
-Our study adds to the growing body of evidence that dominant predators may cause fear which is powerful enough to reverse the diel activity patterns of subordinate species [@kronfeld2003partitioning], and crucially, demonstrates that these effects are habitat-dependent.
+Our study adds to the growing body of evidence that dominant predators may cause fear that is powerful enough to reverse the diel activity patterns of subordinate species [@kronfeld2003partitioning], and crucially demonstrates that these effects are habitat-dependent.
 
 \newpage
+
+# ACKNOWLEDGEMENTS {.unnumbered}
+
+We acknowledge and pay respects to the Gunditjmara and Eastern Maar peoples whose traditional
+lands this study took place. We thank Nick Golding who helped inspire the approach of this paper and provided statistical
+advice. We are grateful to Alan Robley, the Department of Environment, Land, Water, and Planning
+(DELWP) Heywood team and Parks Victoria West Coast District Team for sharing camera-trapping datasets.
+The M.W.R PhD surveys were conducted with the assistance of many fieldwork volunteers, under the
+University of Melbourne Animal Ethics Committee approval 1714119.5 and Victorian Government DELWP
+Research Permit 10008273. This study was generously supported by the Conservation Ecology Centre, the
+Victorian Government DELWP, Parks Victoria, Holsworth Wildlife Research Endowment – Equity Trustees
+Charitable Foundation and Ecological Society of Australia, Australian Research Council Linkage Project
+LP170101134, and the Australian Government’s National Environmental Science Program through the
+Threatened Species Recovery Hub. M.W.R also received funding from an Australian Government Research
+Training Program Scholarship and a Victorian Government DELWP top-up scholarship.
+
+# DATA AVAILABILITY {.unnumbered}
+
+Data and code available from the Dryad Digital Repository https://doi.org/10.5061/dryad.msbcc2g4p [@rees2023data].
 
@@
 \newpage
-
-```{=tex}
-\newpage
-\blandscape
-```
-
-```{r tab-sum}
-
-cam_dist <- read.csv("../raw_data/survey_summary_distances.csv")
-
-cam_dist <- select(cam_dist, !c(X, sess))
-
-#cam_dist2 <- cam_dist %>%
-#  group_by(data_source) %>%
-#  mutate(mean = mean(mean_mindist))
-
-cam_dist <- arrange(cam_dist, region, data_source, treatment)
-
-# make table
-kbl(cam_dist, booktabs = T, longtable = F, digits = 0,
-  caption = "Summary of experimental camera-trap survey designs around fox control in south-west Victoria, Australia.",
-   col.names = c("Region", "Source", "Block", "Treatment", "Surveys", "Year from", "Year to", "Sites", "Min. spacing (m)", "Mean spacing (m)", "Max spacing (m)")) %>% 
-  kable_styling(font_size = 10, latex_options = "scale_down") 
-```
-
-```{=tex}
-\thispagestyle{empty}
-\elandscape
-\newpage
-```
 
@@
 
+```{=tex}
+\newpage
+\blandscape
+```
+
+```{r tab-sum}
+
+cam_dist <- read.csv("../raw_data/survey_summary_distances.csv")
+
+cam_dist <- select(cam_dist, !c(X, sess))
+
+#cam_dist2 <- cam_dist %>%
+#  group_by(data_source) %>%
+#  mutate(mean = mean(mean_mindist))
+
+cam_dist <- arrange(cam_dist, region, data_source, treatment)
+
+# make table
+kbl(cam_dist, booktabs = T, longtable = F, digits = 0,
+  caption = "Summary of experimental camera-trap survey designs around fox control in south-west Victoria, Australia.",
+   col.names = c("Region", "Source", "Block", "Treatment", "Surveys", "Year from", "Year to", "Sites", "Min. spacing (m)", "Mean spacing (m)", "Max spacing (m)")) %>% 
+  kable_styling(font_size = 10, latex_options = "scale_down") 
+```
+
+```{=tex}
+\thispagestyle{empty}
+\elandscape
+\newpage
+```
+
 ```{r diel-tab-fits}
@@
 Shaded areas indicate 95\\% confidence intervals."}
-knitr::include_graphics("../figs/foxbaiting_effects.png")
+knitr::include_graphics("foxbaiting_effects.png")
 ```
+
+\newpage
 
@@
 Shaded areas indicate 95\\% confidence intervals."}
-knitr::include_graphics("../figs/fox_hfi.png")
+knitr::include_graphics("fox_hfi.png")
 ```
